### 2019-07-19 paper/abstract.tex
adding paper

% The distribution of rotation periods of K and M stars, measured from light
% curves obtained from the \kepler\ spacecraft, has a sharp mass-dependent gap
% at around 10-20 days.
% This gap traces a line of constant age and constant Rossby number in the
% rotation period-effective temperature plane, indicating that the cause could
% be related to a discontinuity in either the local star formation history, or
% the magnetic braking evolution of stars.
% A third explanation for the rotation period gap is measurement error caused by
% confounders such as binary companions or aliasing.
% For example, the lower rotation sequence could be a reflection of the upper
% sequence, caused by incorrect measurements at half the true period.
% In this paper, we rule out the possibility that this gap could be caused by
% incorrect period measurements or binary companions, by showing that the
% rapidly rotating stars are dynamically young.

% We also find that nai/"vely applying gyrochronology relations to stars hotter
% than around 5000 K can result in inaccurate ages.

% Finally, we show that the changing shape of the rotation-temperature relations
% causes the `M dwarf dip', providing further evidence that the gyrochronology
% relations are not separable functions of color and age.

The rotational evolution of FGKM dwarfs is still mostly unconstrained after
around 2-3 billion years.
This limit is set by the oldest open clusters whose members have precise light
curves, and therefore photometric rotation periods.
However, recent evidence suggests that the relationship between rotation
period and color/temperature/mass does not remain constant over large
timescales, as has long been assumed.
Old clusters seem to have a flatter relationship between rotation period and
color than young clusters.
In addition, old field stars with asteroseismic ages, observed by the \kepler\
spacecraft, appear to be rotating more rapidly than expected for their ages
and masses.
In this work we use the velocity dispersion of groups of stars as a proxy for
age in order to demonstrate that the relationship between rotation period and
effective temperature not only changes over time but even appears to invert.
At young ages G stars rotate more rapidly than K and M dwarfs, but at old ages
they appear to rotate more rapidly.
In addition, we show that stars rotating just below the rotation period gap
are dynamically young, we ruling out the possibility that the rotation period
gap is caused by incorrect period measurements or binary companions.

### 2019-07-19 paper/intro.tex
adding intro.tex

% Word limit: 500
\section{Introduction}

\subsection{Gyrochronology}
The rotation periods of FGKM dwarfs increase with the square-root of time, and
this phenomenon was first observed several decades ago \citep{skumanich1972}.
This characteristic of main-sequence stars allows them to be dated via their
rotation periods, a practice known as gyrochronology.
This is convenient, since the ages of main-sequence stars are extremely
difficult, and at times impossible, to measure via the traditional age-dating
method of isochrone placement.
It has also been known for a long time that stars of the same age but
different masses have different rotation periods \racomment{(citation)}.
However, an underlying assumption behind many of the empirical gyrochronology
relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015, angus2019} is that the relationships between rotation period and
color, and rotation period and age are {\it separable}.
In other words, the period-color relation is the same at all ages and the
period-age relation is the same at all colors.
It was recently shown that old stars appear to be too rapidly rotating for
their age, \citep{angus2015, vansaders2016, vansaders2018} and that a
mass-dependent modification to the classical \citet{skumanich1972} spin-down
law was needed in order to fit the data \citep{vansaders2016, vansaders2018}.
In addition, an even more recent analysis of middle-aged open clusters provide
further hints that the spin-down rate is mass-dependent \citep{curtis2019}.
Unfortunately, without rotation periods for members of old open clusters, or
any large group of old stars with precisely measured ages, it is extremely
difficult to calibrate the rotation-color relations at old ages.
For this reason, we turn to a population-based stellar age proxy: velocity
dispersion.

\subsection{Kinematics as an age proxy}

Stars are usually born in the thin disk of the Milky Way and orbit the center
of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or $v_z$).
Young stars have relatively low vertical velocities, but gain angular momentum
in the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds are thought to play an important role.

Stellar velocities have a long history of being used as an age proxy, with
several notable examples within stellar astronomy \citep[\eg][]{faherty2009,
west2011}.

A star's vertical {\it action} is a better indication of its age than its
velocity, because it is integrated over the Milky Way's potential, and
therefore position invariant -- \eg\ a star will have the same action at
periapsis and apoapsis.
In contrast, a star's velocity will be different at periapsis and apoapsis --
so in this sense, {\it actions} are the natural quantities to use as age
proxies.
However, actions can only be calculated will full 6-dimensional position and
velocity information, yet most stars with rotation periods do not have radial
velocity measurements because they are \kepler\ targets and most \kepler\
targets are relatively faint (between $\sim$11th and $~$18th magnitudes).
For this reason, we used an alternative age proxy: velocity in the direction
of galactic latitude, \vb.
Since these stars only have proper motions, parallaxes and positions, with no
radial velocites, we could not calculate full 3D velocities or actions.
However, since the \kepler\ field is at low galactic latitude, \vb is a close
approximation to $v_z$.

% The rotation periods of middle-aged FGKM stars are mostly determined by their
% mass, age and evolutionary stage, and this is due, for the most part, to
% angular momentum loss through magnetic braking.
% Although stars are born with a range of rotation periods, a steep dependence
% of angular momentum loss, $J$ angular frequency, $\omega$, $J \propto
% \omega^3$ causes stellar rotation periods to converge after a characteristic
% timescale.
% This timescale is on the order of a few tens of millions of years for F and G
% stars, but hundreds of millions of years for K and M stars.
% Once stars exceed this age, they steadily lose angular momentum via magnetized
% stellar winds that remove angular momentum from the star at a radius that is
% determined by the magnetic field strength.
% The Rossby number, the ratio of rotation period to convective turnover
% timescale, is linked to the strength of stellar magnetic dynamos.
% Large-scale polodial and toroidal stellar magnetic fields are thought to be
% generated by the $\alpha-\Omega$ mechanism, whereby convective motion of
% plasma and differential stellar rotation cause the winding up of magnetic
% field lines, generating a strong magnetic field \racomment{(citations)}.
% In general, the ratio between a star's rotation period and its characteristic
% timescale of convective motion (the Rossby number) determines the strength of
% the magnetic field \racomment{(citation)}.
% Stars that rotate quickly and have deep convection zones, therefore long
% convective turnover times, have small Rossby numbers and strong magnetic
% fields.
% Young M dwarfs typically have the strongest magnetic fields.
% Slowly rotating stars with shallow convective zones, such as old F and G
% dwarfs, have weak magnetic fields \racomment{(citations)}.
% Since magnetic field strength depends on stellar rotation period and mass via
% the Rossby number, it follows that angular momentum loss driven by magnetized
% stellar winds also depends on Rossby number.

### 2019-07-19 paper/intro.tex
removing old text from intro.tex
adding more words about velocity dispersino

@@
 difficult to calibrate the rotation-color relations at old ages.
-For this reason, we turn to a population-based stellar age proxy: velocity
-dispersion.
+For this reason, we turned to a population-based stellar age proxy: velocity
+dispersion in order to investigate the period-temperature relations at old
+ages.
 
@@
 
-Stars are usually born in the thin disk of the Milky Way and orbit the center
-of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or $v_z$).
-Young stars have relatively low vertical velocities, but gain angular momentum
-in the vertical direction over time.
+Stars are thought to be born in the thin disk of the Milky Way, orbiting the
+center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
+$v_z$).
+Young stars have relatively small vertical velocities, but gain angular
+momentum in the vertical direction over time.
 Although the cause of orbital heating is not well understood, interactions
@@
 
-A star's vertical {\it action} is a better indication of its age than its
-velocity, because it is integrated over the Milky Way's potential, and
-therefore position invariant -- \eg\ a star will have the same action at
-periapsis and apoapsis.
+Vertical {\it actions} are better age indicators than velocities, because
+actions are calculated by integrating angular momentum over the Milky Way's
+potential, and are therefore position invariant -- \eg\ a star will have the
+same action at periapsis and apoapsis.
 In contrast, a star's velocity will be different at periapsis and apoapsis --
@@
 However, actions can only be calculated will full 6-dimensional position and
-velocity information, yet most stars with rotation periods do not have radial
-velocity measurements because they are \kepler\ targets and most \kepler\
-targets are relatively faint (between $\sim$11th and $~$18th magnitudes).
+velocity information, yet most stars with measured rotation periods do not
+have radial velocity measurements because they are \kepler\ targets and most
+\kepler\ targets are relatively faint (between $\sim$11th and $~$18th
+magnitudes).
 For this reason, we used an alternative age proxy: velocity in the direction
@@
 
-% The rotation periods of middle-aged FGKM stars are mostly determined by their
-% mass, age and evolutionary stage, and this is due, for the most part, to
-% angular momentum loss through magnetic braking.
-% Although stars are born with a range of rotation periods, a steep dependence
-% of angular momentum loss, $J$ angular frequency, $\omega$, $J \propto
-% \omega^3$ causes stellar rotation periods to converge after a characteristic
-% timescale.
-% This timescale is on the order of a few tens of millions of years for F and G
-% stars, but hundreds of millions of years for K and M stars.
-% Once stars exceed this age, they steadily lose angular momentum via magnetized
-% stellar winds that remove angular momentum from the star at a radius that is
-% determined by the magnetic field strength.
-% The Rossby number, the ratio of rotation period to convective turnover
-% timescale, is linked to the strength of stellar magnetic dynamos.
-% Large-scale polodial and toroidal stellar magnetic fields are thought to be
-% generated by the $\alpha-\Omega$ mechanism, whereby convective motion of
-% plasma and differential stellar rotation cause the winding up of magnetic
-% field lines, generating a strong magnetic field \racomment{(citations)}.
-% In general, the ratio between a star's rotation period and its characteristic
-% timescale of convective motion (the Rossby number) determines the strength of
-% the magnetic field \racomment{(citation)}.
-% Stars that rotate quickly and have deep convection zones, therefore long
-% convective turnover times, have small Rossby numbers and strong magnetic
-% fields.
-% Young M dwarfs typically have the strongest magnetic fields.
-% Slowly rotating stars with shallow convective zones, such as old F and G
-% dwarfs, have weak magnetic fields \racomment{(citations)}.
-% Since magnetic field strength depends on stellar rotation period and mass via
-% the Rossby number, it follows that angular momentum loss driven by magnetized
-% stellar winds also depends on Rossby number.
+Although the velocity of any individual star will only provide a weak age
+constraint, the velocity dispersion of a group of stars can indicate whether,
+on average, that group is old or young relative to other groups.
+In this work we compare the velocity dispersions of groups of stars to
+ascertain which groups are older and which younger and draw conclusions based
+on the implied relative ages of populations.

### 2019-07-19 paper/abstract.tex
removing old version
better

@@
-% The distribution of rotation periods of K and M stars, measured from light
-% curves obtained from the \kepler\ spacecraft, has a sharp mass-dependent gap
-% at around 10-20 days.
-% This gap traces a line of constant age and constant Rossby number in the
-% rotation period-effective temperature plane, indicating that the cause could
-% be related to a discontinuity in either the local star formation history, or
-% the magnetic braking evolution of stars.
-% A third explanation for the rotation period gap is measurement error caused by
-% confounders such as binary companions or aliasing.
-% For example, the lower rotation sequence could be a reflection of the upper
-% sequence, caused by incorrect measurements at half the true period.
-% In this paper, we rule out the possibility that this gap could be caused by
-% incorrect period measurements or binary companions, by showing that the
-% rapidly rotating stars are dynamically young.
-
-% We also find that nai/"vely applying gyrochronology relations to stars hotter
-% than around 5000 K can result in inaccurate ages.
-
-% Finally, we show that the changing shape of the rotation-temperature relations
-% causes the `M dwarf dip', providing further evidence that the gyrochronology
-% relations are not separable functions of color and age.
-
 The rotational evolution of FGKM dwarfs is still mostly unconstrained after
@@
 period and color/temperature/mass does not remain constant over large
-timescales, as has long been assumed.
-Old clusters seem to have a flatter relationship between rotation period and
-color than young clusters.
-In addition, old field stars with asteroseismic ages, observed by the \kepler\
-spacecraft, appear to be rotating more rapidly than expected for their ages
-and masses.
+timescales, contradicting the long-held assumption.
+New results show that old clusters appear to have a flatter relationship
+between rotation period and color than young clusters.
 In this work we use the velocity dispersion of groups of stars as a proxy for
 age in order to demonstrate that the relationship between rotation period and
-effective temperature not only changes over time but even appears to invert.
-At young ages G stars rotate more rapidly than K and M dwarfs, but at old ages
-they appear to rotate more rapidly.
+effective temperature not only changes over time.
+Not only this, the slope of the period-\teff\ relation even appears to invert:
+at young ages G stars rotate more rapidly than K and M dwarfs, but at old ages
+they rotate more rapidly.
 In addition, we show that stars rotating just below the rotation period gap
-are dynamically young, we ruling out the possibility that the rotation period
-gap is caused by incorrect period measurements or binary companions.
+are dynamically young, ruling out the possibility that the rotation period gap
+is caused by incorrect period measurements or binary companions.

### 2019-07-22 color_conversions.py
adding files and things to .gitignore

@@ -1,0 +1,69 @@
+import numpy as np
+
+
+def bprp_to_teff(bprp):
+    """
+    Calculate an effective temperature from a bp-rp color.
+
+    """
+
+    p = [0.01868672, -0.14779631, 0.17718147, 1.44461987, -5.50943128,
+         9.26444289]
+    return np.polyval(p, bprp)*1000
+
+
+def teff_to_bprp(teff):
+    """
+    Calculate a bp-rp color from an effective temperature.
+
+    """
+
+    p = [-2.6948047689554756e-18, 8.945100145214491e-14,
+         -1.1752321538397444e-09, 7.673913205006288e-06,
+         -0.025302316464295933, 35.17655191550194]
+    return np.polyval(p, teff)
+
+
+def bprp_to_bv(bprp):
+    """
+    Calculate a B-V color from a G_BP - G_RP color.
+
+    Poor fit at large and small bp-rp. (Especially bad for low-mass stars).
+
+    """
+
+    p = [-0.0334115504231673, 0.3754200727545447, -1.4691791496336342,
+         2.209433533403669, -0.39641699367657274, 0.14752490961352926]
+    return np.polyval(p, bprp)
+
+
+def bprp_to_bv(bv):
+    """
+    Calculate a G_BP - G_RP color from a B-V color.
+
+    Poor fit at large and small B-V. (Especially bad for low-mass stars).
+
+    """
+
+    p = [-0.9933881241365243, 2.20310834260743, 1.839433873944796,
+         -6.714250044137177, 4.501655472990212, 0.2647400296095147,
+         0.0845117268687072]
+
+    return np.polyval(p, bv)
+
+
+def teff_to_bv(teff):
+
+    p = [-1.6959955278110665e-19, 9.283100332998309e-16,
+         4.761475302858526e-11, -6.354265999828208e-07,
+         0.0024740588508326837, -1.5621606432703632]
+
+    return np.polyval(p, teff)
+
+
+def bv_to_teff(bv):
+
+    p = [3578.180565278925, -13195.543683684042, 15201.484715176379,
+         -3669.547034686415, -6212.974928561518, 9105.456630795197]
+
+    return np.polyval(p, bv)

### 2019-07-22 dispersion.py

@@ -1,0 +1,150 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from tqdm import trange
+
+
+def MC_dispersion(x, y, xerr, yerr, bins, nsamps):
+    """
+    Calculate the dispersion in a set of bins, with Monte Carlo uncertainties.
+
+    Args:
+        x (array): The x-values.
+        y (array): The y-values.
+        xerr (array): The x-value uncertainties.
+        yerr (array): The y-value uncertainties.
+        bins (array): The bin edges. length = number of bins + 1
+        nsamps (int): Number of Monte Carlo samples.
+
+    Returns:
+        dispersion (array): The dispersion array. length = number of bins.
+        dispersion_err (array): The uncertainty on the dispersion.
+            length = number of bins.
+
+    """
+    nbins = len(bins) - 1
+
+    # Sample from Gaussians
+    xsamps = np.zeros((len(x), nsamps))
+    ysamps = np.zeros((len(y), nsamps))
+    dsamps = np.zeros((nbins, nsamps))
+
+    for i in trange(nsamps):
+        xsamps[:, i] = np.random.randn(len(x))*xerr + x
+        ysamps[:, i] = np.random.randn(len(y))*yerr + y
+        dsamps[:, i] = dispersion(xsamps[:, i], ysamps[:, i], bins)
+
+    return np.mean(dsamps, axis=1), np.std(dsamps, axis=1)
+
+
+def dispersion(x, y, bins):
+    """
+    Calculate the dispersion in a set of bins.
+
+    Args:
+        x (array): The x-values.
+        y (array): The y-values.
+        bins (array): The bin edges. length = number of bins + 1
+
+    Returns:
+        dispersion (array): The dispersion array. length = number of bins.
+
+    """
+
+    d = np.zeros(len(bins)-1)
+    for i in range(len(bins)-1):
+        m = (bins[i] < x) * (x < bins[i+1])
+        d[i] = np.std(y[m])
+    return d
+
+
+def running_dispersion(x, y, bsize, mad=False):
+
+    assert x[0] == np.sort(x)[0], "arrays must be sorted on x."
+
+    # Calculate running std of points in bin of bsize
+    d, newx = [], []
+    for i in range(len(x)):
+        if i+bsize < len(x):
+            if not mad:
[…]

### 2019-07-22 paper/method.tex
adding some words to the method

% 270 words
% Limit: 500 words
\section{Method}

\subsection{Calculating velocites}

We crossmatched the \mct\ catalog of stellar rotation periods, measured from
\kepler\ light curves with the \gaia\ DR2 catalog.
We added 2MASS K-band photometry from the Kepler Input Catalog
\citep{brown2011} and SDSS g-band photometry \racomment{(citations)}.

\begin{figure}
  \caption{
A color magnitude diagram in SDSS g and 2MASS K bands showing the
\citet{mcquillan2014} sample, colored by their rotation periods.
We excluded photometric binaries and subgiants from our analysis by removing
stars above the two dashed lines.
The rotation periods of binaries and subgiants do not follow a Skumanich-like
braking law and were removed from our sample.
% since we rely on a simple scaling
% between rotation period and age to make the argument that the rotation period
% is not caused by incorrect period measurements.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts}
\label{fig:CMD_cuts}
\end{figure}

In order to explore the age of this stellar population from a rotation-period
standpoint, it was first necessary to remove visual binaries and subgiants
from the sample.
The rotation-period evolution of these two types of stars is generally
different to that of single stars which more usually follow a Skumanich-like
spin-down law.
Tidal and magnetic interactions between the two components of a binary system
can influence the rotation periods of both stars, and the expanding envelopes
of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries from the sample by fitting a 6th-order polynomial
to the main sequence and raising it by subtracting 0.15 magnitudes to
approximate the line between the single-star and binary main sequences.
This polynomial is shown in figure \ref{fig:CMD_cuts}.
\racomment{Why did we do this in g-K?}
In order to remove subgiants, we removed stars brighter than 6th magnitude in
the g-band from the sample.

We used the {\tt Pyia} {\it Python} package to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the covariance between
these properties.
We calculated \vb\ for each star in the \mct\ catalog that survived the visual
binary and subgiant cuts.

% \begin{figure}
%   \caption{
% the rotation periods of stars in the \citet{mcquillan2014} sample vs.
% effective temperature.
% black points are dwarf stars identified as non-photometric binaries, hotter
% than 4800 k.
% blue circle points are non-photometric binary dwarfs, cooler than 4800 k, with
% a rotation period and \gaia\ color indicating they are older than 1.1 gyr.
% orange squares are stars that with rotation periods that fall just below the
% gap: they have rotation-ages between 0.5 and 1.1 gyrs.
% green triangles are stars with rotation periods faster than the main envelop
% of stars.
% these are probably binaries whose rotation periods are synchronized to their
% orbits and have been spun-up via tidal interactions.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{period_teff}
% \label{fig:period_teff}
% \end{figure}

### 2019-07-22 paper/results.tex
adding old results.tex

% word limit: 500
\begin{itemize}
\item{Showing that the gap stars are young.}
\item{Is there a lack of old M dwarfs is this sample?}
\end{itemize}

\begin{figure}
  \caption{
Rotation period vs. velocity in the galactic latitute direction.
}
  \centering
    \includegraphics[width=1\textwidth]{rotation_vb_dispersion}
\label{fig:rotation_vb_dispersion}
\end{figure}

\begin{figure}
  \caption{
Age vs. velocity in the galactic latitute direction.
The velocity of stars as a function of their age.
}
  \centering
    \includegraphics[width=1\textwidth]{age_vb_dispersion}
\label{fig:age_vb_dispersion}
\end{figure}
figure \ref{age_vb_dispersion} shows the velocity of stars in the \bvector\
direction, plotted against their gyrochronal ages.
These ages were calculated using equation 1 of \citet{stardate_paper},
implemented in the \sd\ \python\ package \citep{stardate}.
The two vertical lines show the approximate locations of the synchronized
binary upper limit and the rotation gap.
Stars to the left of the rotation gap (age $\lesssim$ 1 Gyr) and to the right
of the synchronized binary regime (age $\gtrsim$ 0.5 Gyr) have lower $V_b$
dispersion than stars to the right of the rotation gap (age $gtrsim$ 1 Gyr),
indicating that they are kinematically young.
This increase in velocity dispersion can be seen in both the individually
plotted stars, and in the bins of standard deviation shown as the solid black
line, which increase with age.
This figure clearly indicates that stars with rotation periods that indicate
they are young (but are greater than 7 days), are indeed young.
This rules out the possibility that the rotation period gap is caused by
either incorrect rotation periods, or binarity.

This figure also shows a lack of M dwarfs at old ages caused by the `M dwarf
dip' \citep{vansaders2019, mcquillan2014}.
This is a feature of the \citet{mcquillan2014} sample where there appears to
be a lack of slowly rotating stars at temperatures cooler than 4000 K.
Like the rotation period gap, the origin of the M dwarf dip could either be
due to a lack of stars with long rotation periods, or due to a suppression in
the amplitudes of variability for these stars.
However, in general the completeness of the \citet{mcquillan2014} catalog
increases as a function of decreasing temperature, and is around 80\% for M
dwarfs \citep{mcquillan2013, mcquillan2014, simonian2019}.
This suggests that the age-rotation relation has a different shape at
different ages, as indicated by the 1 Gyr NGC 6811 cluster that has a
different rotation period-age relation to those of younger clusters
\citep{curtis2019}.

### 2019-07-23 paper/results.tex
adding figure description

@@
 % word limit: 500
-\begin{itemize}
-\item{Showing that the gap stars are young.}
-\item{Is there a lack of old M dwarfs is this sample?}
-\end{itemize}
+\section{Results}
 
+To explore the shape of the rotation period-\teff\ relation, we selected
+groups of stars within different age ranges, and calculated the velocity
+dispersion, $\sigma_{v{\bf b}}$ as a function of effective temperature for
+each group.
+We chose to use effective temperatures and not colors in this analysis as it
+is the linear quantity and therefore easier to divide into bins of roughly
+equal numbers of stars.
+Ages were calculated with the \citet{angus2019} gyrochronology relation, which
+is based on Praesepe and the Sun and calibrated in \gaia\ \gcolor\ color.
+This relation was calibrated by fitting a 5th-order polynomial to the relation
+between rotation period and \gaia\ \gcolor\ color for around 800 members of
+the Praesepe cluster, and a straight line in age to Praesepe and the Sun.
+Although only calibrated using Praesepe and the Sun, this gyrochronology
+relation was tested on NGC 6819, a 2.5 Gyr cluster, and predicted accurate
+ages for members of that cluster.
+In addition, the period-color relation of Praesepe is nearly identical to the
+period-color relation of the Hyades, a cluster of around the same age: $\sim$
+650 years \citep{douglas2019}.
+However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
+Gyr cluster \citep{curtis2019}.
+The G dwarfs in this cluster rotate at the same rate as the K dwarfs.
+
+The \citet{angus2019} gyrochronology relation assumes that the rotation
+period-color relation of Praesepe is applicable to stars of all ages: the same
+polynomial relation fit to Praesepe is used to describe the period-color
+relation at all ages.
+If the rotation-color relation of the \citet{angus2019} gyrochronology
+relation were a perfect model for the rotational evolution of stars, then
+groups of stars selected to be similar ages using this relation should fall on
+an isochrone on both a CMD, and in period-\teff\ space.
+Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
+metallicity and extinction blur out the main sequence enough that differences
+between stars in different age and rotation bins are not easily discernable on
+the CMD, although there is still a general age and rotation period gradient on
+the CMD, as seen in figures \ref{fig:CMD_cuts} and \ref{fig:age_gradient}.
 \begin{figure}
   \caption{
-Rotation period vs. velocity in the galactic latitute direction.
+Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
+    binaries removed.
+    Points are colored by their gyrochronal age, according to the
+    \citet{angus2019} gyrochronology relation.
+    A general age gradient is visible across the main sequence.
 }
   \centering
-    \includegraphics[width=1\textwidth]{rotation_vb_dispersion}
-\label{fig:rotation_vb_dispersion}
+    \includegraphics[width=1\textwidth]{age_gradient}
+\label{fig:age_gradient}
+\end{figure}
+In contrast, isochrones and stellar evolution tracks are highly dependent on
+choices made about input physics and assumptions and have often been
+calibrated using different types of stars.
+As a result, different sets of models can have very different shapes on the
+CMD, particularly at low masses.
+Instead of relying on CMD position to age-date groups of stars, we opted to
+explore age trends via kinematics.
+Kinematic age-dating has the advantage of being relatively model independent,
+or at least, having a very simple model: that velocity dispersion increases
+over time.
+This means that it is relatively easy to rank groups of stars by age: older
+groups have a larger velocity dispersion.
+It is less easy to rank-order groups of stars by age on the CMD because
+stellar evolution is highly mass and metallicity dependent, and because
+different stellar evolution models predict different ages for the same CMD
+position.
+This does not mean that it is not possible to calibrate gyrochronology using
+isochrones, however we leave this exercise for a future study.
+
+\subsection{Selecting groups of stars to examine the gyrochronology relations}
+
+In order to examine the rotation period-temperature/color relation, we
+selected stars in the \mct\ sample with the same gyrochronal age, calculated
+using the \citet{angus2019} relations.
+The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample (minus
+visual binaries and subgiants) in grey, with groups selected by age shown in
+color.
+The bottom panel shows the velocity dispersion of each age group, as a
+function of effective temperature.
+If the stars in each group have the same age across the temperature range,
+their velocity dispersion should be constant.
+Although the youngest age groups have a relatively constant velocity
+dispersion across temperatures, the oldest age groups do not.
+This indicates that the shape of the period-color relation does not remain
+constant over time -- it flattens out.
+In the oldest age bin, the cool stars have a larger velocity dispersion than
+the hot stars, suggesting that the cool stars are older than the hot stars
+(\ie\ these stars do not actually belong in the same age bin).
+Either the cool stars are being grouped into an age bin that is too young, or
+the hot stars are grouped into an age bin that is too old, but either way, the
+isochrones used to make the age selection are too steeply sloped.
+\begin{figure}
+  \caption{
+Top: rotation period vs effective temperature for stars in the \mct\
+    catalog.
+    The full catalog, with subgiants and visual binaries removed is shown in
+    grey, and stars selected to be in different age groups are overlayed in
+    color.
+    These age groups were selected using the \citet{angus2019} gyrochronology
+    relation.
+The legend in the center of the figure lists the age range of each group.
+    Bottom: velocity dispersion vs effective temperature for each age group.
+The color of the line corresponds to the color of the group shown in the top
+    panel.
+If the gyrochronal model were correct at all ages and the stars in each group
+    were the same age across temperatures, the velocity dispersion would be
+    constant as a function of \teff.
+However, the velocity dispersions of the oldest age groups depend on \teff,
+    indicating that the \citet{angus2019} gyrochronology model either
+    overpredicts the ages of old M dwarfs or underpredicts the ages of old G
+    and K dwarfs.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{age_cut}
+\label{fig:age_cut}
 \end{figure}
 
+We applied the same analysis to groups of stars selected using simple rotation
+period cuts.
+The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with
+selected groups of stars shown in different colors.
+The bottom panel shows the velocity dispersion as a function of effective
+temperature for each group.
 \begin{figure}
   \caption{
-Age vs. velocity in the galactic latitute direction.
-The velocity of stars as a function of their age.
+This figure is similar to figure \ref{fig:age_cut}, with stars divided into
+    period groups rather than age groups.
+The velocity dispersion is more constant across effective temperatures for the
+    most slowly rotating stars, compared to the stars selected with the
+    \citet{angus2018} gyrochronology model.
+    This indicates that the gyrochronology models flatten out, and possibly
+    even invert, at old ages.
 }
   \centering
-    \includegraphics[width=1\textwidth]{age_vb_dispersion}
-\label{fig:age_vb_dispersion}
+    \includegraphics[width=1\textwidth]{period_cut}
+\label{fig:period_cut}
 \end{figure}
-figure \ref{age_vb_dispersion} shows the velocity of stars in the \bvector\
-direction, plotted against their gyrochronal ages.
-These ages were calculated using equation 1 of \citet{stardate_paper},
-implemented in the \sd\ \python\ package \citep{stardate}.
-The two vertical lines show the approximate locations of the synchronized
-binary upper limit and the rotation gap.
-Stars to the left of the rotation gap (age $\lesssim$ 1 Gyr) and to the right
-of the synchronized binary regime (age $\gtrsim$ 0.5 Gyr) have lower $V_b$
-dispersion than stars to the right of the rotation gap (age $gtrsim$ 1 Gyr),
-indicating that they are kinematically young.
-This increase in velocity dispersion can be seen in both the individually
-plotted stars, and in the bins of standard deviation shown as the solid black
-line, which increase with age.
-This figure clearly indicates that stars with rotation periods that indicate
-they are young (but are greater than 7 days), are indeed young.
-This rules out the possibility that the rotation period gap is caused by
-either incorrect rotation periods, or binarity.
 
-This figure also shows a lack of M dwarfs at old ages caused by the `M dwarf
-dip' \citep{vansaders2019, mcquillan2014}.
-This is a feature of the \citet{mcquillan2014} sample where there appears to
-be a lack of slowly rotating stars at temperatures cooler than 4000 K.
-Like the rotation period gap, the origin of the M dwarf dip could either be
-due to a lack of stars with long rotation periods, or due to a suppression in
-the amplitudes of variability for these stars.
-However, in general the completeness of the \citet{mcquillan2014} catalog
-increases as a function of decreasing temperature, and is around 80\% for M
-dwarfs \citep{mcquillan2013, mcquillan2014, simonian2019}.
-This suggests that the age-rotation relation has a different shape at
-different ages, as indicated by the 1 Gyr NGC 6811 cluster that has a
-different rotation period-age relation to those of younger clusters
-\citep{curtis2019}.
+% \begin{figure}
+%   \caption{
+% Rotation period vs. velocity in the galactic latitute direction.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{rotation_vb_dispersion}
+% \label{fig:rotation_vb_dispersion}
+% \end{figure}
+
+% \begin{figure}
+%   \caption{
+% Age vs. velocity in the galactic latitute direction.
+% The velocity of stars as a function of their age.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{age_vb_dispersion}
+% \label{fig:age_vb_dispersion}
+% \end{figure}
+
+% figure \ref{age_vb_dispersion} shows the velocity of stars in the \bvector\
+% direction, plotted against their gyrochronal ages.
+% These ages were calculated using equation 1 of \citet{stardate_paper},
+% implemented in the \sd\ \python\ package \citep{stardate}.
+% The two vertical lines show the approximate locations of the synchronized
+% binary upper limit and the rotation gap.
+% Stars to the left of the rotation gap (age $\lesssim$ 1 Gyr) and to the right
+% of the synchronized binary regime (age $\gtrsim$ 0.5 Gyr) have lower $V_b$
+% dispersion than stars to the right of the rotation gap (age $gtrsim$ 1 Gyr),
+% indicating that they are kinematically young.
+% This increase in velocity dispersion can be seen in both the individually
+% plotted stars, and in the bins of standard deviation shown as the solid black
+% line, which increase with age.
+% This figure clearly indicates that stars with rotation periods that indicate
+% they are young (but are greater than 7 days), are indeed young.
+% This rules out the possibility that the rotation period gap is caused by
+% either incorrect rotation periods, or binarity.
+
+% This figure also shows a lack of M dwarfs at old ages caused by the `M dwarf
+% dip' \citep{vansaders2019, mcquillan2014}.
+% This is a feature of the \citet{mcquillan2014} sample where there appears to
+% be a lack of slowly rotating stars at temperatures cooler than 4000 K.
+% Like the rotation period gap, the origin of the M dwarf dip could either be
+% due to a lack of stars with long rotation periods, or due to a suppression in
+% the amplitudes of variability for these stars.
+% However, in general the completeness of the \citet{mcquillan2014} catalog
+% increases as a function of decreasing temperature, and is around 80\% for M
+% dwarfs \citep{mcquillan2013, mcquillan2014, simonian2019}.
+% This suggests that the age-rotation relation has a different shape at
+% different ages, as indicated by the 1 Gyr NGC 6811 cluster that has a
+% different rotation period-age relation to those of younger clusters
+% \citep{curtis2019}.

### 2019-07-23 paper/intro.tex
small changes

@@
 law was needed in order to fit the data \citep{vansaders2016, vansaders2018}.
-In addition, an even more recent analysis of middle-aged open clusters provide
-further hints that the spin-down rate is mass-dependent \citep{curtis2019}.
+In addition, an even more recent analysis of middle-aged open clusters
+provides further hints that the spin-down rate is mass-dependent
+\citep{curtis2019}.
+Rotation period measurements of the 1.1 Gyr open cluster NGC 6811 reveal a
+flattened relationship between rotation period and color, where the G dwarfs
+are rotating at the same rate as the K dwarfs.
+
 Unfortunately, without rotation periods for members of old open clusters, or

### 2019-07-23 paper/method.tex
small changes

@@
 braking law and were removed from our sample.
+The rotation period gradient across the main sequence is visible by eye in
+    this figure: young, rapidly rotating stars are located below the old,
+    slowly rotating stars.
 % since we rely on a simple scaling
@@
 of subgiants drive rapid spin-down through conservation of angular momentum.
-We removed visual binaries from the sample by fitting a 6th-order polynomial
-to the main sequence and raising it by subtracting 0.15 magnitudes to
-approximate the line between the single-star and binary main sequences.
+We removed visual binaries and subgiants from the sample by applying cuts to
+the color-magnitude diagram (CMD).
+We fit a 6th-order polynomial to the main sequence and raised it by
+subtracting 0.15 magnitudes to approximate the line between the single-star
+and binary main sequences.
 This polynomial is shown in figure \ref{fig:CMD_cuts}.
+We removed all stars above this line from the sample.
 \racomment{Why did we do this in g-K?}

### 2019-07-23 paper/discussion.tex
adding refs and discussion

\section{Discussion}

What if the M dwarfs just have larger velocity dispersion because they're less
massive?

### 2019-07-23 paper/acknowledgements.tex
adding acknowledgements

This work was partly developed at the KITP conference 'Better stars, better
planets'.

Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
\Kepler\ mission is provided by the NASA Science Mission directorate.

This work made use of the gaia-kepler.fun crossmatch database created by Megan
Bedell

This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.

### 2019-07-23 tools.py
adding tools.py

@@ -1,0 +1,316 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+from tqdm import tqdm, trange
+import emcee
+
+import read_mist_models
+from isochrones.mist import MIST_Isochrone
+from isochrones import StarModel
+
+from stardate.lhf import gyro_model_rossby, gyro_model, sigma, calc_rossby_number
+
+bands = ["B", "V", "J", "H", "K", "BP", "RP", "G"]
+mist = MIST_Isochrone(bands)
+mist.initialize()
+
+def color_err(c):
+    c_err = np.zeros(len(c))
+    bright = c < 13
+    medium = (13 < c) * (c < 17)
+    faint = 17 <= c
+    c_err[bright] = np.ones(len(c_err[bright]))*.002
+    c_err[medium] = np.ones(len(c_err[medium]))*.01
+    c_err[faint] = np.ones(len(c_err[faint]))*.2
+    return c_err
+
+
+def photometric_noise(G, bp, rp):
+    """
+    Calculates the noise on Gaia G, bp, rp and parallax.
+
+    Args:
+        G (array): Gaia G band magnitude.
+        bp (array): Gaia G_BP band magnitude.
+        rp (array): Gaia G_RP band magnitude.
+
+    Returns:
+        G (array): Uncertainty on Gaia G band magnitude.
+        bp (array): Uncertainty on Gaia G_BP band magnitude.
+        rp (array): Uncertainty on Gaia G_RP band magnitude.
+        parallax_err (array): Photometry-based uncertainty on Gaia parallax.
+
+    """
+    G_err = np.zeros(len(G))
+    bright = G < 13
+    medium = (13 < G) * (G < 17)
+    faint = 17 <= G
+    G_err[bright] = np.ones(len(G_err[bright]))*.0003
+    G_err[medium] = np.ones(len(G_err[medium]))*.002
+    G_err[faint] = np.ones(len(G_err[faint]))*.01
+
+    bp_err = color_err(bp)
+    rp_err = color_err(rp)
+
+    parallax_err = np.zeros(len(G))
+    bright = G < 15
+    medium = (15 < G) * (G < 17)
+    faint = (17 < G) * (G < 20)
+    ultra_faint = 21 <= G
+    parallax_err[bright] = .03  # milliarcseconds
+    parallax_err[medium] = .1
+    parallax_err[faint] = .7
+    parallax_err[ultra_faint] = 2
+
[…]

### 2019-07-31 paper/abstract.tex
updating draft and adding figures

@@
-The rotational evolution of FGKM dwarfs is still mostly unconstrained after
+% Word limit: 250
+
+Due to a lack of precise ages for old stars with measured rotation periods,
+the rotational evolution of FGKM dwarfs is still mostly unconstrained after
 around 2-3 billion years.
-This limit is set by the oldest open clusters whose members have precise light
-curves, and therefore photometric rotation periods.
-However, recent evidence suggests that the relationship between rotation
-period and color/temperature/mass does not remain constant over large
-timescales, contradicting the long-held assumption.
-New results show that old clusters appear to have a flatter relationship
-between rotation period and color than young clusters.
-In this work we use the velocity dispersion of groups of stars as a proxy for
-age in order to demonstrate that the relationship between rotation period and
-effective temperature not only changes over time.
-Not only this, the slope of the period-\teff\ relation even appears to invert:
-at young ages G stars rotate more rapidly than K and M dwarfs, but at old ages
-they rotate more rapidly.
-In addition, we show that stars rotating just below the rotation period gap
-are dynamically young, ruling out the possibility that the rotation period gap
-is caused by incorrect period measurements or binary companions.
+% This limit is set by the oldest open clusters whose members have precise light
+% curves, and therefore photometric rotation periods.
+Recent evidence suggests that the relationship between rotation period and
+color, temperature, and mass does not remain constant over large timescales,
+contradicting the long-held assumption that underlies many gyrochronology
+relations.
+% For example, old clusters appear to have a flatter relationship between
+% rotation period and color than young clusters.
+In this work we use the kinematic ages of star populations to reveal the
+evolving relationship between rotation period and effective temperature.
+Not only does the period-color relation change over time, its slope appears to
+{\it invert}: at young ages early K stars rotate more rapidly than late K
+stars, but at old ages they rotate more rapidly.
+% Although the gyrochronology relations fail to correctly predict the ages of
+% old cool stars, we show that gyrochronology accurately predicts the ages of
+% early K dwarfs, and young K dwarfs of all temperatures and those ages are
+% consistent with ages predicted from age-velocity dispersion relations.
+We find that the ages of field stars with measured rotation periods do not
+exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
+braking efficiency.
+% It was previously unknown whether the upper-limit of the stellar rotation
+% period distribution was created by a ceasura of magnetic braking (\ie old
+% stars do not rotate more slowly after a critical age) or an inability to
+% measure rotation periods for old stars.
+% We show that the lack of long rotation periods is caused by enhanced
+% measurement difficulty.
+We also show that stars rotating just below the rotation period gap are
+dynamically young, ruling out the possibility that the rotation period gap is
+caused by incorrect period measurements or binary companions.
+% Finally, we suggest that this gap may mark a critical point in the rotation
+% period evolution of K dwarfs.
+The main result of this paper could be produced by either an evolving
+period-color relationship {\it or} mass-dependent dynamical heating.
+Although we provide some evidence to suggest that the former effect is
+dominant, we cannot be sure and therefore caution that the results presented
+here are subject to this caveat.

### 2019-07-31 paper/conclusion.tex
updating draft and adding figures

\section{Conclusion}

We examined the rotation period-color component of the \citep{angus2019}
gyrochronology relation and found that groups of stars, selected to be the
same age using this relation, do {\it not} have the same velocity dispersion
across all temperatures.
This indicates that {\it either} the \citep{angus2019} period-color relation
is wrong, {\it or} that mass-dependent dynamical heating is affecting this
data set.
We found that the exponent of the AVR relation, calibrated using F and G stars
in the Geneva-Copenhagen Survey, with ages estimated via isochrone-fitting, is
greater than the exponent we find for the AVR relation fit to the most massive
K dwarfs in our sample.
Since mass-dependent dynamical heating should lead to a larger AVR exponent
for lower-mass stars, this suggests that mass-dependent heating is not
strongly affecting our sample.
However, we used velocity in galactic latitude, \vb, not \vz, to calculate our
AVRs, and while \vb\ is a close approximation to \vz\ at low galactic
latitudes, it is not identical.
Indeed, the \vz\ AVR should be larger than the \vb\ AVR, since orbital
scattering is expected to have a strongest effect in the vertical direction.
For the 290 stars in our sample with \gaia\ radial velocity measurements we
measured the \vz\ AVR exponent to be larger than the \vb\ AVR exponent.

For now, we cannot draw a strong conclusion regarding the origin of the
observed increased velocity dispersion for stars of lower masses since either
an incorrectly calibrated gyrochronology relation, or mass-dependent dynamical
heating could be the cause.

### 2019-07-31 paper/intro.tex
updating draft and adding figures

--- 2019-07-19 paper/intro.tex
@@
 \subsection{Gyrochronology}
-The rotation periods of FGKM dwarfs increase with the square-root of time, and
-this phenomenon was first observed several decades ago \citep{skumanich1972}.
+It has long been known that the rotation periods of FGKM dwarfs increase over
+time \citep{skumanich1972}.
 This characteristic of main-sequence stars allows them to be dated via their
-rotation periods, a practice known as gyrochronology.
-This is convenient, since the ages of main-sequence stars are extremely
-difficult, and at times impossible, to measure via the traditional age-dating
-method of isochrone placement.
-It has also been known for a long time that stars of the same age but
-different masses have different rotation periods \racomment{(citation)}.
+rotation periods in a practice known as gyrochronology, which is convenient
+since the ages of main-sequence stars are extremely difficult to measure via
+the traditional age-dating method of isochrone placement.
+It is also well established that stars of the same age but different {\it
+masses} have different rotation periods \racomment{(citation)}, assumed to be
+caused by the deeper convective zones, and therefore stronger magnetic dynamos
+(and more efficient magnetic braking) in lower-mass stars
+\footnote{Color is the directly observable quantity to which empirical
+gyrochronology relation are usually calibrated to, rather than mass or
+effective temperature.}.
 However, an underlying assumption behind many of the empirical gyrochronology
@@
 period-age relation is the same at all colors.
-It was recently shown that old stars appear to be too rapidly rotating for
-their age, \citep{angus2015, vansaders2016, vansaders2018} and that a
-mass-dependent modification to the classical \citet{skumanich1972} spin-down
-law was needed in order to fit the data \citep{vansaders2016, vansaders2018}.
-In addition, an even more recent analysis of middle-aged open clusters provide
-further hints that the spin-down rate is mass-dependent \citep{curtis2019}.
-Unfortunately, without rotation periods for members of old open clusters, or
-any large group of old stars with precisely measured ages, it is extremely
-difficult to calibrate the rotation-color relations at old ages.
-For this reason, we turned to a population-based stellar age proxy: velocity
-dispersion in order to investigate the period-temperature relations at old
+It was recently shown that old stars rotate more rapidly than a simple,
+separable gyrochronology relation would predict \citep{angus2015,
+vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
+modification to the classical \citet{skumanich1972} spin-down law is required
+to reproduce the data \citep{vansaders2016, vansaders2018}.
+An even more recent analysis of middle-aged open clusters provides further
+indication that the exponent of the period-age relation is mass-dependent.
+New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
+revealed a flattened relationship between rotation period and color
+\citep{curtis2019}.
+In this cluster, the G dwarfs rotate at the same rate as the K dwarfs.
+
+Unfortunately, the open clusters with rotation period measurements are mostly
+young -- currently the oldest is 2.5 Gyr (NGC 6819).
+Without rotation periods for precisely dated old stars, it is extremely
+difficult to calibrate the relationship between rotation period and color at
+old ages.
+For this reason, we turned to a population-based stellar age indicator,
+velocity dispersion, to investigate the period-temperature relations at old
 ages.
+% Since a separable period-age and period-color relation provided an adequate
+% fit for the young stars, it was naturally assumed to apply to old stars too.
 
@@
 $v_z$).
-Young stars have relatively small vertical velocities, but gain angular
-momentum in the vertical direction over time.
+Young stars have relatively small vertical velocities, but gain momentum in
+the vertical direction over time.
 Although the cause of orbital heating is not well understood, interactions
-with giant molecular clouds are thought to play an important role.
-
-Stellar velocities have a long history of being used as an age proxy, with
-several notable examples within stellar astronomy \citep[\eg][]{faherty2009,
-west2011}.
-
-Vertical {\it actions} are better age indicators than velocities, because
-actions are calculated by integrating angular momentum over the Milky Way's
-potential, and are therefore position invariant -- \eg\ a star will have the
-same action at periapsis and apoapsis.
-In contrast, a star's velocity will be different at periapsis and apoapsis --
-so in this sense, {\it actions} are the natural quantities to use as age
-proxies.
-However, actions can only be calculated will full 6-dimensional position and
-velocity information, yet most stars with measured rotation periods do not
-have radial velocity measurements because they are \kepler\ targets and most
-\kepler\ targets are relatively faint (between $\sim$11th and $~$18th
-magnitudes).
-For this reason, we used an alternative age proxy: velocity in the direction
-of galactic latitude, \vb.
-Since these stars only have proper motions, parallaxes and positions, with no
-radial velocites, we could not calculate full 3D velocities or actions.
-However, since the \kepler\ field is at low galactic latitude, \vb is a close
-approximation to $v_z$.
-
+with giant molecular clouds are thought to play an important role
+\racomment{(citations)}.
 Although the velocity of any individual star will only provide a weak age
@@
 on the implied relative ages of populations.
+% Stellar velocities have a history of being used as an age proxy, with several
+% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
+% west2011}.
+Since the age-velocity dispersion relations (AVRs) are themselves still
+actively being calibrated, it is difficult to directly compare gyrochronal
+ages with kinematic ones.
+However, regardless of the exact relation between velocity dispersion and
+stellar age, it is expected to be monotonic relationship, therefore velocity
+dispersion can be used effectively to {\it rank} groups of stars by age.
+\racomment{Say some words here to prime the reader for the upcoming caveats.}
+
+% Vertical {\it actions} are better age indicators than velocities, because
+% actions are calculated by integrating angular momentum over the Milky Way's
+% potential, and are therefore position invariant -- \eg\ a star will have the
+% same action at periapsis and apoapsis.
+% In contrast, orbital velocities are different at periapsis and apoapsis -- so
+% in this sense, {\it actions} are the natural quantities to use as age proxies.
+Vertical action is a better age indicator than vertical velocity, however both
+vertical action and vertical velocity (\vz/W) can only be calculated with full
+6-dimensional position and velocity information, yet most stars with measured
+rotation periods do not have radial velocity measurements because they are
+\kepler\ targets which are relatively faint (between $\sim$11th and $~$18th
+magnitudes).
+For this reason, we used an alternative age proxy: velocity in the direction
+of galactic latitude, \vb.
+The \kepler\ field is at low galactic latitude (b=5-20\degrees), so \vb is a
+close approximation to \vz.
+% Since the stars in our sample only have proper motions, parallaxes and
+% positions, with no radial velocites, we could not calculate full 3D velocities
+% or actions.
+
+% \subsection{The degeneracy between gyrochronology and mass-dependent heating}
+
+In this paper we tested the \citet{angus2019} gyrochronology relation which
+was calibrated using the period-color relation of Praesepe and the period-age
+relation of Praesepe and the Sun and is calibrated in \gaia\ \gcolor\ color.
+% This relation was calibrated by fitting a 5th-order polynomial to the relation
+% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
+% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
+% and the Sun.
+Although only calibrated using Praesepe and the Sun, this gyrochronology
+relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
+accurate ages
+In addition, the period-color relation of Praesepe is nearly identical to the
+period-color relation of the Hyades, a cluster of around the same age: $\sim$
+650 years \citep{douglas2019}.
+However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
+Gyr cluster \citep{curtis2019}.
+The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
+appears as though the K stars have `stalled' -- their spin-down has been
+halted.
+The \citet{angus2019} gyrochronology relation assumes that the rotation
+period-color relation of Praesepe is applicable to stars of all ages: the same
+polynomial relation fit to Praesepe is used to describe the period-color
+relation for all stars.
+However, the NGC 6811 cluster suggests that this assumption does not hold past
+around 1 Gyr.
+If the period-color relation of the \citet{angus2019} gyrochronology model
+{\it were} a perfect model for the rotational evolution of stars, then groups
+of stars selected to be similar ages using this relation should fall on an
+isochrone on both a CMD, and in period-\teff\ space.
+Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
+variations in metallicity and extinction, blur out the main sequence enough
+that differences between stars in different age and rotation bins are not
+easily discernable on the CMD, although there is still a general age and
+rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
+\ref{fig:age_gradient}.
+For this reason, we chose to use {\it kinematic} `isochrones', instead of
+magnitude and color-based isochrones: although kinematics as an age indicator
+is not necessarily as well calibrated to an absolute age scale as CMD
+position, it can be very sensitive to {\it differences} in the ages of stellar
+populations.
+\begin{figure}
+  \caption{
+Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
+    binaries removed.
+    Points are colored by their gyrochronal age, according to the
+    \citet{angus2019} gyrochronology relation.
+    A general age gradient is visible across the main sequence.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{age_gradient}
+\label{fig:age_gradient}
+\end{figure}
+Isochrones and stellar evolution tracks are highly dependent on choices made
+about input physics and assumptions and have often been calibrated using
+different types of stars.
+As a result, different sets of models can have very different shapes on the
+CMD, particularly at low masses.
+Instead of relying on CMD position to age-date groups of stars, we opted to
+explore age trends via kinematics.
+Kinematic age-dating has the advantage of being relatively model independent,
+or at least, having a very simple model: that velocity dispersion increases
+over time.
+This means that it is relatively easy to rank groups of stars by age: older
+groups have a larger velocity dispersion.
+It is less easy to rank-order groups of stars by age on the CMD because
+stellar evolution is highly mass and metallicity dependent, and because
+different stellar evolution models predict different ages for the same CMD
+position.
+This does not mean that it is not possible to calibrate gyrochronology using
+isochrones, however it is not the focus of this study.

### 2019-07-31 paper/method.tex
updating draft and adding figures

--- 2019-07-22 paper/method.tex
@@
 
-\subsection{Calculating velocites}
+\subsection{The data}
 
 We crossmatched the \mct\ catalog of stellar rotation periods, measured from
-\kepler\ light curves with the \gaia\ DR2 catalog.
-We added 2MASS K-band photometry from the Kepler Input Catalog
-\citep{brown2011} and SDSS g-band photometry \racomment{(citations)}.
+\kepler\ light curves, with the \gaia\ DR2 catalog.
+Reddening and extinction from dust was calculated for each star using the
+Bayestar dust map implemented in the dustmaps {\it Python} package
+\citep{green2018}.
+We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
+using an 8th-order polynomial relation calibrated using .... stars
+\racomment{ask Jason for details}.
+\begin{equation}
+    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
+    301C^6 - 26C^7,
+% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
+%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
+%           301.13205924630165, -25.923997443169355]
+\end{equation}
+where C is \gaia\ \gcolor.
 
@@
   \caption{
-A color magnitude diagram in SDSS g and 2MASS K bands showing the
-\citet{mcquillan2014} sample, colored by their rotation periods.
+A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
+    extinction-corrected magnitudes, colored by their rotation periods.
 We excluded photometric binaries and subgiants from our analysis by removing
 stars above the two dashed lines.
-The rotation periods of binaries and subgiants do not follow a Skumanich-like
-braking law and were removed from our sample.
+% The rotation periods of binaries and subgiants do not follow a Skumanich-like
+% braking law.
+% The rotation period gradient across the main sequence is visible by eye in
+%     this figure: young, rapidly rotating stars are located below the old,
+%     slowly rotating stars.
 % since we rely on a simple scaling
@@
 \end{figure}
-
-In order to explore the age of this stellar population from a rotation-period
+To explore the age of this stellar population from a rotation-period
 standpoint, it was first necessary to remove visual binaries and subgiants
@@
 of subgiants drive rapid spin-down through conservation of angular momentum.
-We removed visual binaries from the sample by fitting a 6th-order polynomial
-to the main sequence and raising it by subtracting 0.15 magnitudes to
-approximate the line between the single-star and binary main sequences.
-This polynomial is shown in figure \ref{fig:CMD_cuts}.
-\racomment{Why did we do this in g-K?}
-In order to remove subgiants, we removed stars brighter than 6th magnitude in
-the g-band from the sample.
+We removed visual binaries and subgiants from the sample by applying cuts to
+the color-magnitude diagram (CMD), shown in figure \ref{fig:CMD_cuts}.
+We fit a 6th-order polynomial to the main sequence and raised it by 0.22 dex,
+to approximate the division between single stars and visual binaries.
+We removed all stars above this line from the sample.
+We removed subgiants by eliminating stars brighter than 6th magnitude in
+\gaia\ G-band from the sample.
 
+We removed stars with negative parallaxes and parallax signal-to-noise ratios
+below 10 from the sample.
+We also removed a small number of stars fainter than 16th magnitude from the
+sample.
 We used the {\tt Pyia} {\it Python} package to calculate stellar velocities.
@@
 It therefore not only incorporates uncertainties on the \gaia\ positions
-parallaxes and proper motions, it also accounts for the covariance between
-these properties.
-We calculated \vb\ for each star in the \mct\ catalog that survived the visual
-binary and subgiant cuts.
+parallaxes and proper motions, it also accounts for the {\it covariance}
+between these properties.
+We removed stars with absolute \vb\ uncertainties greater than 1 \kms\ from
+the sample.
 

### 2019-07-31 paper/results.tex
updating draft and adding figures

@@
 
-To explore the shape of the rotation period-\teff\ relation, we selected
-groups of stars within different age ranges, and calculated the velocity
-dispersion, $\sigma_{v{\bf b}}$ as a function of effective temperature for
-each group.
+\subsection{Selecting groups of stars to examine the gyrochronology relations}
+
+To explore the relationship between rotation period, \teff\ and age, we
+selected groups of stars within different age ranges \citep[where age was
+calculated using the][gyrochronology relation]{angus2019}, and calculated the
+velocity dispersion, the standard deviation of velocities, $\sigma_{v{\bf b}}$
+as a function of effective temperature for each age group.
+Ages were calculated using dereddened \gaia\ \gcolor\, however we throughout
+this paper we show rotation periods as a function of effective temperature.
 We chose to use effective temperatures and not colors in this analysis as it
@@
 equal numbers of stars.
-Ages were calculated with the \citet{angus2019} gyrochronology relation, which
-is based on Praesepe and the Sun and calibrated in \gaia\ \gcolor\ color.
-This relation was calibrated by fitting a 5th-order polynomial to the relation
-between rotation period and \gaia\ \gcolor\ color for around 800 members of
-the Praesepe cluster, and a straight line in age to Praesepe and the Sun.
-Although only calibrated using Praesepe and the Sun, this gyrochronology
-relation was tested on NGC 6819, a 2.5 Gyr cluster, and predicted accurate
-ages for members of that cluster.
-In addition, the period-color relation of Praesepe is nearly identical to the
-period-color relation of the Hyades, a cluster of around the same age: $\sim$
-650 years \citep{douglas2019}.
-However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
-Gyr cluster \citep{curtis2019}.
-The G dwarfs in this cluster rotate at the same rate as the K dwarfs.
+Since the processes that produce dynamical heating in the galactic disk are
+expected to generate Gaussian-distributed velocities, we performed
+sigma-clipping on stars in each age and temperature bin, to remove
+non-Gaussian outliers.
 
-The \citet{angus2019} gyrochronology relation assumes that the rotation
-period-color relation of Praesepe is applicable to stars of all ages: the same
-polynomial relation fit to Praesepe is used to describe the period-color
-relation at all ages.
-If the rotation-color relation of the \citet{angus2019} gyrochronology
-relation were a perfect model for the rotational evolution of stars, then
-groups of stars selected to be similar ages using this relation should fall on
-an isochrone on both a CMD, and in period-\teff\ space.
-Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
-metallicity and extinction blur out the main sequence enough that differences
-between stars in different age and rotation bins are not easily discernable on
-the CMD, although there is still a general age and rotation period gradient on
-the CMD, as seen in figures \ref{fig:CMD_cuts} and \ref{fig:age_gradient}.
-\begin{figure}
-  \caption{
-Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
-    binaries removed.
-    Points are colored by their gyrochronal age, according to the
-    \citet{angus2019} gyrochronology relation.
-    A general age gradient is visible across the main sequence.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{age_gradient}
-\label{fig:age_gradient}
-\end{figure}
-In contrast, isochrones and stellar evolution tracks are highly dependent on
-choices made about input physics and assumptions and have often been
-calibrated using different types of stars.
-As a result, different sets of models can have very different shapes on the
-CMD, particularly at low masses.
-Instead of relying on CMD position to age-date groups of stars, we opted to
-explore age trends via kinematics.
-Kinematic age-dating has the advantage of being relatively model independent,
-or at least, having a very simple model: that velocity dispersion increases
-over time.
-This means that it is relatively easy to rank groups of stars by age: older
-groups have a larger velocity dispersion.
-It is less easy to rank-order groups of stars by age on the CMD because
-stellar evolution is highly mass and metallicity dependent, and because
-different stellar evolution models predict different ages for the same CMD
-position.
-This does not mean that it is not possible to calibrate gyrochronology using
-isochrones, however we leave this exercise for a future study.
+% The \citet{angus2019} gyrochronology relation was calibrated using the
+% period-color relation of Praesepe and the period-age relation of Praesepe and
+% the Sun and is calibrated in \gaia\ \gcolor\ color.
+% This relation was calibrated by fitting a 5th-order polynomial to the relation
+% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
+% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
+% and the Sun.
+% Although only calibrated using Praesepe and the Sun, this gyrochronology
+% relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
+% accurate ages.
+% In addition, the period-color relation of Praesepe is nearly identical to the
+% period-color relation of the Hyades, a cluster of around the same age: $\sim$
+% 650 years \citep{douglas2019}.
+% However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
+% Gyr cluster \citep{curtis2019}.
+% The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
+% appears as though the K stars have `stalled' -- their spin-down has been
+% halted.
+% The \citet{angus2019} gyrochronology relation assumes that the rotation
+% period-color relation of Praesepe is applicable to stars of all ages: the same
+% polynomial relation fit to Praesepe is used to describe the period-color
+% relation for all stars.
+% However, the NGC 6811 cluster suggests that this assumption does not hold past
+% around 1 Gyr.
+% If the period-color relation of the \citet{angus2019} gyrochronology model
+% {\it were} a perfect model for the rotational evolution of stars, then groups
+% of stars selected to be similar ages using this relation should fall on an
+% isochrone on both a CMD, and in period-\teff\ space.
+% Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
+% variations in metallicity and extinction, blur out the main sequence enough
+% that differences between stars in different age and rotation bins are not
+% easily discernable on the CMD, although there is still a general age and
+% rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
+% \ref{fig:age_gradient}.
+% For this reason, we chose to use {\it kinematic} `isochrones', instead of
+% magnitude and color-based isochrones: although kinematics as an age indicator
+% is not necessarily as well calibrated to an absolute age scale as CMD
+% position, it can be very sensitive to {\it differences} in the ages of stellar
+% populations.
+% \begin{figure}
+%   \caption{
+% Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
+%     binaries removed.
+%     Points are colored by their gyrochronal age, according to the
+%     \citet{angus2019} gyrochronology relation.
+%     A general age gradient is visible across the main sequence.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{age_gradient}
+% \label{fig:age_gradient}
+% \end{figure}
+% Isochrones and stellar evolution tracks are highly dependent on choices made
+% about input physics and assumptions and have often been calibrated using
+% different types of stars.
+% As a result, different sets of models can have very different shapes on the
+% CMD, particularly at low masses.
+% Instead of relying on CMD position to age-date groups of stars, we opted to
+% explore age trends via kinematics.
+% Kinematic age-dating has the advantage of being relatively model independent,
+% or at least, having a very simple model: that velocity dispersion increases
+% over time.
+% This means that it is relatively easy to rank groups of stars by age: older
+% groups have a larger velocity dispersion.
+% It is less easy to rank-order groups of stars by age on the CMD because
+% stellar evolution is highly mass and metallicity dependent, and because
+% different stellar evolution models predict different ages for the same CMD
+% position.
+% This does not mean that it is not possible to calibrate gyrochronology using
+% isochrones, however it is not the focus of this study.
 
-\subsection{Selecting groups of stars to examine the gyrochronology relations}
+% To examine the rotation period-temperature/color relation, we
+% selected stars in the \mct\ sample with the same gyrochronal age, calculated
+% using the \citet{angus2019} relations.
+% Ages were calculated using dereddened \gaia\ \gcolor\, however we throughout
+% this paper we show rotation periods as a function of effective temperature.
 
-In order to examine the rotation period-temperature/color relation, we
-selected stars in the \mct\ sample with the same gyrochronal age, calculated
-using the \citet{angus2019} relations.
-The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample (minus
-visual binaries and subgiants) in grey, with groups selected by age shown in
-color.
-The bottom panel shows the velocity dispersion of each age group, as a
-function of effective temperature.
-If the stars in each group have the same age across the temperature range,
-their velocity dispersion should be constant.
-Although the youngest age groups have a relatively constant velocity
-dispersion across temperatures, the oldest age groups do not.
-This indicates that the shape of the period-color relation does not remain
-constant over time -- it flattens out.
-In the oldest age bin, the cool stars have a larger velocity dispersion than
-the hot stars, suggesting that the cool stars are older than the hot stars
-(\ie\ these stars do not actually belong in the same age bin).
-Either the cool stars are being grouped into an age bin that is too young, or
-the hot stars are grouped into an age bin that is too old, but either way, the
-isochrones used to make the age selection are too steeply sloped.
 \begin{figure}
@@
 \end{figure}
+The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample (minus
+visual binaries and subgiants) in grey, with groups selected by age shown in
+color.
+The color of the points corresponds to the age ranges specified in the legend,
+which also apply to the lines in the lower panel.
+The bottom panel shows the velocity dispersion of each age group, as a
+function of effective temperature.
+We only included stars within a temperature range of 5000 - 3500 K in our
+analysis, as hotter stars are more likely to have stopped magnetic braking
+\citep{vansaders2016}, which could bias the results.
+M dwarfs were not included in our analysis because such faint stars cannot be
+observed at large heights above the plane (because of the low galactic
+latitude of the \kepler\ field, stars at high-Z are more distant), which
+introduces a mass-dependent velocity bias: cooler populations of stars are
+skewed towards lower velocity dispersions.
+The coolest temperature bins in figure \ref{fig:age_cut} have low velocity
+dispersions, indicating that this effect may already become important at
+temperatures lower than $\sim$ 4000 K.
 
-We applied the same analysis to groups of stars selected using simple rotation
-period cuts.
+Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
+with gyrochronal age across all temperatures.
+However, if the stars in each selected age group had the {\it same} age across
+the temperature range, their velocity dispersion would be a constant function
+of \teff.
+Although the youngest age groups have a relatively constant velocity
+dispersions across temperatures, the oldest age groups do not.
+This indicates {\it either} that the shape of the period-color relation does
+not remain constant over time, \ie\ it flattens out, {\it or} that cool stars
+experience more efficient dynamical heating than hot stars.
+
+Mass-dependent dynamical heating could occur because lower-mass stars
+experience greater velocity changes when gravitationally perturbed.
+However, mass-dependent orbital heating has not yet been unambiguously
+observed in the galactic disk because of the strong anti-correlation between
+stellar mass and stellar age.
+Less massive stars do, indeed have larger velocity dispersions, however they
+are also older on average.
+This mass-age degeneracy is highly reduced in M dwarfs because their
+main-sequence lifetimes are longer than the age of the Universe, however no
+evidence for mass-dependent heating has been detected in these low mass stars
+\citep{faherty2009}.
+
+To investigate further, we calculated the exponent of the AVR for each
+temperature bin in figure \ref{fig:age_cut}.
+If mass-dependent heating is strongly affecting our data, the exponent of the
+AVR should increase with decreasing effective temperature, \ie\ the heating
+rate should be greater for lower-mass stars.
+Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
+decreasing effective temperature, however, the AVR exponent calculated using
+more massive stars from the Geneva Copenhagen Survey (GCS)
+\citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
+Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
+in our study.
+If mass-dependent orbital heating is the main cause the rise in velocity
+dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
+AVR exponent of the more massive GCS stars should fall {\it below} the AVR
+exponent for the most massive stars in our sample.
+In other words, the dynamical heating rate should be lowest for the highest
+mass stars and highest for the lowest mass stars.
+\begin{figure}
+  \caption{
+The exponent of the AVR as a function of effective temperature.
+    The AVR exponent of \citet{holmberg2009}, calculated using a more massive
+    sample of stars, is shown as a horizontal dashed line.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{AVR_exponent}
+\label{fig:AVR_exponent}
+\end{figure}
+% nordstrom2004, jorgensen2005,
+These AVR exponents were calculated with different populations of stars,
+subject to different selection effects, so directly comparing them comes with
+some risk.
+The most important difference is that the GCS AVR is calculated using \vz,
+however our AVRs are calculated using \vb\ since most stars in our sample do
+not have radial velocities.
+The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
+latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
+close approximation to \vz.
+Only 290 out of the 6820 stars in our sample, plotted in color in figure
+\ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
+that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
+between 3500 and 5000 K.
+For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
+AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
+The similarity of these two AVR exponents suggests that directly comparing
+these two quantities {\it may} be a reasonable approach.
+For now, given the large difference between the (\vz) AVR exponent of the GCS
+sample and the (\vb) AVR exponent of the hottest stars in our sample, we
+assume that the increased velocity dispersion at cooler temperatures is mostly
+caused by incorrect age-grouping due to an incorrect period-color relation at
+old ages, and that any mass-dependent heating, while it may contribute at a
+low level to this result, is not the dominant driver.
+However, differentiating the effects of mass-dependent heating and the shape
+of the gyrochronology relations is certainly warranted in a follow-up study.
+
+% We explore the differences between these two scenarios in section
+% \ref{sec:discussion} and find that an incorrect period-color relation is more
+% likely to be the cause of this effect than mass-dependent heating.
+
+% In the oldest age group, the cool stars have a larger velocity dispersion than
+% the hot stars, suggesting that the cool stars are older than the hot stars
+% (\ie\ these stars do not actually belong in the same age group).
+% The velocity dispersion of the early Ks show remarkable agreement with
+% observations of $\sigma_W$ as a function of age from the Geneva-Copenhagen
+% survey \citep{Holmwood2009}.
+% This indicates that the ages of early K stars are correctly predicted by this
+% gyrochronolgy relation but the ages of late Ks are under-predicted.
+% Alternatively, the cause of the increased dispersion as a function of
+% temperature could be mass-dependent dynamical heating, whereby less massive
+% stars experience larger velocity kicks than more massive stars.
+% We discuss the differences between these two scenarios in section
+% \ref{sec:results}.
+
+% Figure \ref{fig:AVR_exponent} shows that AVR of stars between around 4170 K
+% and 4330 K most closely matches the GCS AVR.
+
+The Praesepe-calibrated gyrochronology relation seems to accurately predict
+the relative ages of {\it young} field stars.
+This is shown by the consistent velocity dispersion of young stars as a
+function of temperature in figure \ref{fig:age_cut}.
+To our knowledge, no gyrochronology relation has never been demonstrated to
+correctly predict ages (either relative or absolute) for such cool or such
+young field stars.
+This is not particularly remarkable however, since these young stars are a
+similar age to the Praesepe cluster, which was used to calibrate the
+period-color relation.
+Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
+the \citet{angus2019} gyrochronology relation, compared with the ages of star
+groups, predicted with the \citet{holmberg2009} AVR.
+\citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
+relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
+dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
+dispersion across all temperatures, of our sample.
+\begin{figure}
+  \caption{
+      Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
+    predicted by the \citet{angus2019} gyrochronology relation, for stars in
+    different temperature ranges.
+Points are colored by the effective temperature in the center of the \teff\
+bin.
+The dashed line shows the y=x relation.
+Predicted kinematic ages fan out over gyrochronal time, either because the AVR
+    is temperature-dependent or because the gyrochronology relations have an
+    age-dependent period-color relation.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{age_comparison}
+\label{fig:age_comparison}
+\end{figure}
+Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
+the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
+predicted by the \citet{angus2019} gyrochronology relation, for stars in
+different temperature ranges.
+The predicted kinematic ages fan out over gyrochronal time, either because the
+AVR is temperature-dependent or because the gyrochronology relations have an
+age-dependent period-color relation.
+Figure \ref{fig:age_comparion} also shows that the AVR seems to break down for
+the youngest age bin, 0.5 Gyr, which has a lower velocity dispersion than
+expected, given its age.
+Alternatively, the ages of these young stars could be under-predicted by
+gyrochronology.
+These young stars have similar rotation periods to the Praesepe cluster, which
+was used to calibrate the gyrochronology relation used in this analysis.
+The ages of these young stars should, therefore, be the most accurate of the
+entire sample.
+However, they are tied to the age of the Praesepe cluster, whose age is not
+accurately known.
+\citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
+kinematic age-prediction for these stars is around 400 Myrs.
+Other studies have assigned a much older age of 800 Myrs to Praesepe
+\citep{brandt2015}.
+If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
+calibration was inaccurate, the entire age {\it scale} would be incorrect.
+However, this is not what figure \ref{fig:age_comparion} shows -- it shows
+that the {\it relative} age of these youngest stars are under-predicted by
+kinematics or over-predicted by gyrochronology.
+
+% Either the cool stars are being grouped into an age bin that is too young, or
+% the hot stars are grouped into an age bin that is too old, but either way, the
+% isochrones used to make the age selection are too steeply sloped.
+
+We applied the same analysis described at the beginning of this section to
+groups of stars selected using rotation period groups, instead of age groups.
 The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with
-selected groups of stars shown in different colors.
-The bottom panel shows the velocity dispersion as a function of effective
-temperature for each group.
+stars in different period ranges plotted in different colors.
+The bottom panel shows the velocity dispersion of each group as a function of
+effective temperature.
 \begin{figure}
@@
 \end{figure}
+Once again, the velocity dispersion increases with rotation period overall.
+For the most rapidly rotating groups of stars, velocity dispersion decreases
+with \teff.
+This is expected given the positively sloped period-color relation of Praesepe
+and other young clusters: late K dwarfs rotate more slowly than early K dwarfs
+of the same age.
+Between 15 and 25 day rotation periods, the temperature dependence of the
+velocity dispersion starts to disappear, indicating that the period-color
+relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
+of the same age.
+At long rotation periods, the velocity dispersion still increases with
+effective temperature, although the increase is much more modest than in
+figure \ref{fig:age_cut}.
+This suggests that the period-color relation may have a negative slope at old
+ages: late K dwarfs may rotate {\it more} rapidly than early K dwarfs.
+This would be a paradigm shift for gyrochronology, since the assumption has
+always been that stellar spin-down rate is tied to magnetic field strength.
+The deeper convection zones of later type stars generate stronger magnetic
+fields which {\it should} lead to more efficient angular momentum loss.
 

### 2019-08-01 paper/abstract.tex
tidying up and refining text

--- 2019-07-19 paper/abstract.tex
@@
-The rotational evolution of FGKM dwarfs is still mostly unconstrained after
-around 2-3 billion years.
-This limit is set by the oldest open clusters whose members have precise light
-curves, and therefore photometric rotation periods.
-However, recent evidence suggests that the relationship between rotation
-period and color/temperature/mass does not remain constant over large
-timescales, contradicting the long-held assumption.
-New results show that old clusters appear to have a flatter relationship
-between rotation period and color than young clusters.
-In this work we use the velocity dispersion of groups of stars as a proxy for
-age in order to demonstrate that the relationship between rotation period and
-effective temperature not only changes over time.
-Not only this, the slope of the period-\teff\ relation even appears to invert:
-at young ages G stars rotate more rapidly than K and M dwarfs, but at old ages
-they rotate more rapidly.
-In addition, we show that stars rotating just below the rotation period gap
-are dynamically young, ruling out the possibility that the rotation period gap
-is caused by incorrect period measurements or binary companions.
+% Word limit: 250
+
+Due to a lack of precise ages and rotation periods for old main-sequence
+stars, the rotational evolution of FGKM dwarfs is mostly unconstrained
+after around 2-3 billion years.
+% This limit is set by the oldest open clusters whose members have precise light
+% curves, and therefore photometric rotation periods.
+Recent evidence suggests that the relationship between rotation period and
+color/temperature/mass does not remain constant over large timescales,
+contradicting the long-held assumption that underlies many gyrochronology
+relations.
+% For example, old clusters appear to have a flatter relationship between
+% rotation period and color than young clusters.
+In this work we use the kinematic ages of populations of K dwarfs to reveal
+the evolving relationship between rotation period and effective temperature.
+We find that groups of stars, selected to be the same age, do {\it not} have
+the same velocity dispersion at all temperatures; cooler stars have greater
+velocity dispersions than hotter stars.
+Two different scenarios could be responsible for producing this observation:
+either the rotation period-color relation changes over time, or stars of
+different masses experience different rates of dynamical heating.
+We show however, that if mass-dependent heating is responsible, the
+differential heating rate is not constant over logarithmic time intervals
+which implies that different heating processes must be acting before and after
+ages of around 1 Gyr.
+Perhaps a more likely scenario is that the relation between rotation period
+and color changes over time, undergoing a particularly sharp transition at
+around 1 Gyr.
+If we assume that mass-dependent heating does not affect the data, our results
+indicate that the slope of the period-color relation actually {\it inverts}:
+at young ages early K stars rotate more rapidly than late K stars, but beyond
+$\sim$ 2 Gyr they rotate more slowly.
+% (similar to the age of the rotation gap and the 1.1 Gyr NGC 6811
+% cluster, shown to have a different period-color relation to Praesepe).
+% We find that the ages of field stars with measured rotation periods do not
+% exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
+% braking efficiency.
+% We also show that stars rotating just below the rotation period gap are
+% dynamically young, ruling out the possibility that the rotation period gap is
+% caused by incorrect period measurements or binary companions.
+% The main result of this paper could be produced by either an evolving
+% period-color relationship {\it or} mass-dependent dynamical heating.
+% Although we provide some evidence to suggest that the former effect is
+% dominant, we cannot be sure and therefore caution that the results presented
+% here are subject to this caveat.

### 2019-08-01 paper/intro.tex
tidying up and refining text

@@
 This characteristic of main-sequence stars allows them to be dated via their
-rotation periods in a practice known as gyrochronology, which is convenient
+rotation periods in a practice known as gyrochronology, which is useful
 since the ages of main-sequence stars are extremely difficult to measure via
@@
 It is also well established that stars of the same age but different {\it
-masses} have different rotation periods \racomment{(citation)}, assumed to be
+masses} have different rotation periods \racomment{(citation)}, thought to be
 caused by the deeper convective zones, and therefore stronger magnetic dynamos
@@
 New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
-revealed a flattened relationship between rotation period and color
+reveal a flattened relationship between rotation period and color
 \citep{curtis2019}.
@@
 However, regardless of the exact relation between velocity dispersion and
-stellar age, it is expected to be monotonic relationship, therefore velocity
+stellar age, it is expected to be a monotonic relationship, therefore velocity
 dispersion can be used effectively to {\it rank} groups of stars by age.
@@
 Vertical action is a better age indicator than vertical velocity, however both
-vertical action and vertical velocity (\vz/W) can only be calculated with full
-6-dimensional position and velocity information, yet most stars with measured
-rotation periods do not have radial velocity measurements because they are
-\kepler\ targets which are relatively faint (between $\sim$11th and $~$18th
-magnitudes).
+vertical action ($J_z$) and vertical velocity (\vz/W) can only be calculated
+with full 6-dimensional position and velocity information, yet most stars with
+measured rotation periods do not have radial velocity measurements because
+they are relatively faint \kepler\ targets ($\sim$11th-18th magnitudes).
 For this reason, we used an alternative age proxy: velocity in the direction
 of galactic latitude, \vb.
-The \kepler\ field is at low galactic latitude (b=5-20\degrees), so \vb is a
-close approximation to \vz.
-% Since the stars in our sample only have proper motions, parallaxes and
-% positions, with no radial velocites, we could not calculate full 3D velocities
-% or actions.
+The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
+\vb\ is a close approximation to \vz.
 
@@
 
-In this paper we tested the \citet{angus2019} gyrochronology relation which
+In this paper we test the \citet{angus2019} gyrochronology relation, which
 was calibrated using the period-color relation of Praesepe and the period-age
-relation of Praesepe and the Sun and is calibrated in \gaia\ \gcolor\ color.
+relation of Praesepe and the Sun.
+It is calibrated to \gaia\ \gcolor\ color.
 % This relation was calibrated by fitting a 5th-order polynomial to the relation
@@
 relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
-accurate ages
+accurate ages.
 In addition, the period-color relation of Praesepe is nearly identical to the
@@
 of stars selected to be similar ages using this relation should fall on an
-isochrone on both a CMD, and in period-\teff\ space.
+isochrone on a CMD, an isochrone in period-color space, and an
+isochrone in velocity dispersion-color space.
 Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
@@
 populations.
-\begin{figure}
-  \caption{
-Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
-    binaries removed.
-    Points are colored by their gyrochronal age, according to the
-    \citet{angus2019} gyrochronology relation.
-    A general age gradient is visible across the main sequence.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{age_gradient}
-\label{fig:age_gradient}
-\end{figure}
+
 Isochrones and stellar evolution tracks are highly dependent on choices made

### 2019-08-01 paper/results.tex
tidying up and refining text

@@
 
-\subsection{Selecting groups of stars to examine the gyrochronology relations}
+\subsection{Selecting groups of coeval stars}
 
@@
 calculated using the][gyrochronology relation]{angus2019}, and calculated the
-velocity dispersion, the standard deviation of velocities, $\sigma_{v{\bf b}}$
-as a function of effective temperature for each age group.
-Ages were calculated using dereddened \gaia\ \gcolor\, however we throughout
-this paper we show rotation periods as a function of effective temperature.
-We chose to use effective temperatures and not colors in this analysis as it
-is the linear quantity and therefore easier to divide into bins of roughly
-equal numbers of stars.
+velocity dispersion: the standard deviation of velocities, \sigmavb, as a
+function of effective temperature for each age group.
 Since the processes that produce dynamical heating in the galactic disk are
@@
 non-Gaussian outliers.
-
-% The \citet{angus2019} gyrochronology relation was calibrated using the
-% period-color relation of Praesepe and the period-age relation of Praesepe and
-% the Sun and is calibrated in \gaia\ \gcolor\ color.
-% This relation was calibrated by fitting a 5th-order polynomial to the relation
-% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
-% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
-% and the Sun.
-% Although only calibrated using Praesepe and the Sun, this gyrochronology
-% relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
-% accurate ages.
-% In addition, the period-color relation of Praesepe is nearly identical to the
-% period-color relation of the Hyades, a cluster of around the same age: $\sim$
-% 650 years \citep{douglas2019}.
-% However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
-% Gyr cluster \citep{curtis2019}.
-% The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
-% appears as though the K stars have `stalled' -- their spin-down has been
-% halted.
-% The \citet{angus2019} gyrochronology relation assumes that the rotation
-% period-color relation of Praesepe is applicable to stars of all ages: the same
-% polynomial relation fit to Praesepe is used to describe the period-color
-% relation for all stars.
-% However, the NGC 6811 cluster suggests that this assumption does not hold past
-% around 1 Gyr.
-% If the period-color relation of the \citet{angus2019} gyrochronology model
-% {\it were} a perfect model for the rotational evolution of stars, then groups
-% of stars selected to be similar ages using this relation should fall on an
-% isochrone on both a CMD, and in period-\teff\ space.
-% Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
-% variations in metallicity and extinction, blur out the main sequence enough
-% that differences between stars in different age and rotation bins are not
-% easily discernable on the CMD, although there is still a general age and
-% rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
-% \ref{fig:age_gradient}.
-% For this reason, we chose to use {\it kinematic} `isochrones', instead of
-% magnitude and color-based isochrones: although kinematics as an age indicator
-% is not necessarily as well calibrated to an absolute age scale as CMD
-% position, it can be very sensitive to {\it differences} in the ages of stellar
-% populations.
-% \begin{figure}
-%   \caption{
-% Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
-%     binaries removed.
-%     Points are colored by their gyrochronal age, according to the
-%     \citet{angus2019} gyrochronology relation.
-%     A general age gradient is visible across the main sequence.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{age_gradient}
-% \label{fig:age_gradient}
-% \end{figure}
-% Isochrones and stellar evolution tracks are highly dependent on choices made
-% about input physics and assumptions and have often been calibrated using
-% different types of stars.
-% As a result, different sets of models can have very different shapes on the
-% CMD, particularly at low masses.
-% Instead of relying on CMD position to age-date groups of stars, we opted to
-% explore age trends via kinematics.
-% Kinematic age-dating has the advantage of being relatively model independent,
-% or at least, having a very simple model: that velocity dispersion increases
-% over time.
-% This means that it is relatively easy to rank groups of stars by age: older
-% groups have a larger velocity dispersion.
-% It is less easy to rank-order groups of stars by age on the CMD because
-% stellar evolution is highly mass and metallicity dependent, and because
-% different stellar evolution models predict different ages for the same CMD
-% position.
-% This does not mean that it is not possible to calibrate gyrochronology using
-% isochrones, however it is not the focus of this study.
-
-% To examine the rotation period-temperature/color relation, we
-% selected stars in the \mct\ sample with the same gyrochronal age, calculated
-% using the \citet{angus2019} relations.
-% Ages were calculated using dereddened \gaia\ \gcolor\, however we throughout
-% this paper we show rotation periods as a function of effective temperature.
+Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+throughout this paper we show rotation periods as a function of {\it effective
+temperature}.
+We chose to use effective temperatures and not colors in this analysis as it
+is the linear quantity and therefore easier to divide into bins of roughly
+equal numbers of stars.
 
@@
 \end{figure}
-The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample (minus
-visual binaries and subgiants) in grey, with groups selected by age shown in
+The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
+(excluding visual binaries and subgiants) in grey, with coeval groups shown in
 color.
@@
 which also apply to the lines in the lower panel.
-The bottom panel shows the velocity dispersion of each age group, as a
-function of effective temperature.
+The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
+as a function of effective temperature.
 We only included stars within a temperature range of 5000 - 3500 K in our
@@
 stellar mass and stellar age.
-Less massive stars do, indeed have larger velocity dispersions, however they
+Less massive stars do indeed have larger velocity dispersions, however they
 are also older on average.
@@
 
-% We explore the differences between these two scenarios in section
-% \ref{sec:discussion} and find that an incorrect period-color relation is more
-% likely to be the cause of this effect than mass-dependent heating.
-
-% In the oldest age group, the cool stars have a larger velocity dispersion than
-% the hot stars, suggesting that the cool stars are older than the hot stars
-% (\ie\ these stars do not actually belong in the same age group).
-% The velocity dispersion of the early Ks show remarkable agreement with
-% observations of $\sigma_W$ as a function of age from the Geneva-Copenhagen
-% survey \citep{Holmwood2009}.
-% This indicates that the ages of early K stars are correctly predicted by this
-% gyrochronolgy relation but the ages of late Ks are under-predicted.
-% Alternatively, the cause of the increased dispersion as a function of
-% temperature could be mass-dependent dynamical heating, whereby less massive
-% stars experience larger velocity kicks than more massive stars.
-% We discuss the differences between these two scenarios in section
-% \ref{sec:results}.
-
-% Figure \ref{fig:AVR_exponent} shows that AVR of stars between around 4170 K
-% and 4330 K most closely matches the GCS AVR.
-
 The Praesepe-calibrated gyrochronology relation seems to accurately predict
@@
 expected, given its age.
-Alternatively, the ages of these young stars could be under-predicted by
-gyrochronology.
-These young stars have similar rotation periods to the Praesepe cluster, which
-was used to calibrate the gyrochronology relation used in this analysis.
-The ages of these young stars should, therefore, be the most accurate of the
-entire sample.
-However, they are tied to the age of the Praesepe cluster, whose age is not
-accurately known.
-\citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
-kinematic age-prediction for these stars is around 400 Myrs.
-Other studies have assigned a much older age of 800 Myrs to Praesepe
-\citep{brandt2015}.
-If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
-calibration was inaccurate, the entire age {\it scale} would be incorrect.
-However, this is not what figure \ref{fig:age_comparion} shows -- it shows
-that the {\it relative} age of these youngest stars are under-predicted by
-kinematics or over-predicted by gyrochronology.
-
-% Either the cool stars are being grouped into an age bin that is too young, or
-% the hot stars are grouped into an age bin that is too old, but either way, the
-% isochrones used to make the age selection are too steeply sloped.
+% Alternatively, the ages of these young stars could be under-predicted by
+% gyrochronology.
+% These young stars have similar rotation periods to the Praesepe cluster, which
+% was used to calibrate the gyrochronology relation used in this analysis.
+% The ages of these young stars should, therefore, be the most accurate of the
+% entire sample.
+% However, they are tied to the age of the Praesepe cluster, whose age is not
+% accurately known.
+% \citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
+% kinematic age-prediction for these stars is around 400 Myrs.
+% Other studies have assigned a much older age of 800 Myrs to Praesepe
+% \citep{brandt2015}.
+% If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
+% calibration was inaccurate, the entire age {\it scale} would be incorrect.
+% However, this is not what figure \ref{fig:age_comparion} shows -- it shows
+% that the {\it relative} age of these youngest stars are under-predicted by
+% kinematics or over-predicted by gyrochronology.
 
@@
     most slowly rotating stars, compared to the stars selected with the
-    \citet{angus2018} gyrochronology model.
-    This indicates that the gyrochronology models flatten out, and possibly
-    even invert, at old ages.
+    \citet{angus2019} gyrochronology model, indicating that the gyrochronology
+    models flatten out, and possibly even invert, at old ages.
 }
@@
 ages: late K dwarfs may rotate {\it more} rapidly than early K dwarfs.
-This would be a paradigm shift for gyrochronology, since the assumption has
-always been that stellar spin-down rate is tied to magnetic field strength.
+This would be a paradigm shift for gyrochronology, stellar spin-down rate is
+thought to be directly tied to magnetic field strength.
 The deeper convection zones of later type stars generate stronger magnetic
 fields which {\it should} lead to more efficient angular momentum loss.
-
-% \begin{figure}
-%   \caption{
-% Rotation period vs. velocity in the galactic latitute direction.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{rotation_vb_dispersion}
-% \label{fig:rotation_vb_dispersion}
-% \end{figure}
-
-% \begin{figure}
-%   \caption{
-% Age vs. velocity in the galactic latitute direction.
-% The velocity of stars as a function of their age.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{age_vb_dispersion}
-% \label{fig:age_vb_dispersion}
-% \end{figure}
-
-% figure \ref{age_vb_dispersion} shows the velocity of stars in the \bvector\
-% direction, plotted against their gyrochronal ages.
-% These ages were calculated using equation 1 of \citet{stardate_paper},
-% implemented in the \sd\ \python\ package \citep{stardate}.
-% The two vertical lines show the approximate locations of the synchronized
-% binary upper limit and the rotation gap.
-% Stars to the left of the rotation gap (age $\lesssim$ 1 Gyr) and to the right
-% of the synchronized binary regime (age $\gtrsim$ 0.5 Gyr) have lower $V_b$
-% dispersion than stars to the right of the rotation gap (age $gtrsim$ 1 Gyr),
-% indicating that they are kinematically young.
-% This increase in velocity dispersion can be seen in both the individually
-% plotted stars, and in the bins of standard deviation shown as the solid black
-% line, which increase with age.
-% This figure clearly indicates that stars with rotation periods that indicate
-% they are young (but are greater than 7 days), are indeed young.
-% This rules out the possibility that the rotation period gap is caused by
-% either incorrect rotation periods, or binarity.
-
-% This figure also shows a lack of M dwarfs at old ages caused by the `M dwarf
-% dip' \citep{vansaders2019, mcquillan2014}.
-% This is a feature of the \citet{mcquillan2014} sample where there appears to
-% be a lack of slowly rotating stars at temperatures cooler than 4000 K.
-% Like the rotation period gap, the origin of the M dwarf dip could either be
-% due to a lack of stars with long rotation periods, or due to a suppression in
-% the amplitudes of variability for these stars.
-% However, in general the completeness of the \citet{mcquillan2014} catalog
-% increases as a function of decreasing temperature, and is around 80\% for M
-% dwarfs \citep{mcquillan2013, mcquillan2014, simonian2019}.
-% This suggests that the age-rotation relation has a different shape at
-% different ages, as indicated by the 1 Gyr NGC 6811 cluster that has a
-% different rotation period-age relation to those of younger clusters
-% \citep{curtis2019}.

### 2019-08-01 paper/abstract.tex
outlining three arguments

@@
 after around 2-3 billion years.
-% This limit is set by the oldest open clusters whose members have precise light
-% curves, and therefore photometric rotation periods.
-Recent evidence suggests that the relationship between rotation period and
-color/temperature/mass does not remain constant over large timescales,
-contradicting the long-held assumption that underlies many gyrochronology
-relations.
-% For example, old clusters appear to have a flatter relationship between
-% rotation period and color than young clusters.
+However, recent evidence suggests that the relationship between rotation
+period and color/temperature/mass does not remain constant over large
+timescales, contradicting a long-held assumption that underlies many
+gyrochronology relations.
 In this work we use the kinematic ages of populations of K dwarfs to reveal
@@
 different masses experience different rates of dynamical heating.
-We show however, that if mass-dependent heating is responsible, the
-differential heating rate is not constant over logarithmic time intervals
-which implies that different heating processes must be acting before and after
-ages of around 1 Gyr.
-Perhaps a more likely scenario is that the relation between rotation period
-and color changes over time, undergoing a particularly sharp transition at
-around 1 Gyr.
+% We show however, that if mass-dependent heating is responsible, the
+% differential heating rate is not constant over logarithmic time intervals
+% which implies that different heating processes must be acting before and after
+% ages of around 1 Gyr.
+% Perhaps a more likely scenario is that the relation between rotation period
+% and color changes over time, undergoing a particularly sharp transition at
+% around 1 Gyr.
 If we assume that mass-dependent heating does not affect the data, our results
@@
 $\sim$ 2 Gyr they rotate more slowly.
+
 % (similar to the age of the rotation gap and the 1.1 Gyr NGC 6811
@@
 % caused by incorrect period measurements or binary companions.
-% The main result of this paper could be produced by either an evolving
-% period-color relationship {\it or} mass-dependent dynamical heating.
-% Although we provide some evidence to suggest that the former effect is
-% dominant, we cannot be sure and therefore caution that the results presented
-% here are subject to this caveat.

### 2019-08-01 paper/conclusion.tex
outlining three arguments

@@
 
-We examined the rotation period-color component of the \citep{angus2019}
-gyrochronology relation and found that groups of stars, selected to be the
-same age using this relation, do {\it not} have the same velocity dispersion
-across all temperatures.
+We found that groups of K dwarfs, selected to be coeval using the
+\citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
+dispersion across all temperatures.
 This indicates that {\it either} the \citep{angus2019} period-color relation
@@
 data set.
-We found that the exponent of the AVR relation, calibrated using F and G stars
-in the Geneva-Copenhagen Survey, with ages estimated via isochrone-fitting, is
-greater than the exponent we find for the AVR relation fit to the most massive
-K dwarfs in our sample.
-Since mass-dependent dynamical heating should lead to a larger AVR exponent
-for lower-mass stars, this suggests that mass-dependent heating is not
-strongly affecting our sample.
-However, we used velocity in galactic latitude, \vb, not \vz, to calculate our
-AVRs, and while \vb\ is a close approximation to \vz\ at low galactic
-latitudes, it is not identical.
-Indeed, the \vz\ AVR should be larger than the \vb\ AVR, since orbital
-scattering is expected to have a strongest effect in the vertical direction.
-For the 290 stars in our sample with \gaia\ radial velocity measurements we
-measured the \vz\ AVR exponent to be larger than the \vb\ AVR exponent.
+We presented three pieces of evidence that suggest mass-dependent heating
+cannot be responsible for all of the increasing velocity dispersion.
 
-For now, we cannot draw a strong conclusion regarding the origin of the
-observed increased velocity dispersion for stars of lower masses since either
-an incorrectly calibrated gyrochronology relation, or mass-dependent dynamical
-heating could be the cause.
+\begin{itemize}
+
+\item{No strong evidence for mass-dependent heating has been found among M, L
+and T-type stars.
+These low-mass stars live longer than the current age of the Universe and
+therefore do not suffer from a mass-age degeneracy \citep{faherty2009}.}
+
+\item{Literature measurements of the (\vz) AVR exponent are slightly higher
+than the value measured from our data set, but calculated using more massive
+stars.
+Mass-dependent dynamical heating should however produce a {\it smaller} AVR
+exponent for higher-mass stars.
+We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
+subject to different selection biases than those used to calculate AVRs in the
+literature \citep{holmberg2009, aumer2009, yu2018}, so drawing direct
+comparisons between them is not a robust test.}
+
+\item{The differences in the (\vb) AVR exponent between the hottest and
+coolest K dwarfs in our sample are extremely large, spanning values between
+0.3 and 0.65, yet the difference in mass is less than a factor of two.
+Such a large spread in AVR exponents cannot be explained by mass-dependent
+heating alone.}
+
+\end{itemize}
+
+\racomment{Words about the significance of an inverted period-color relation.}

### 2019-08-01 paper/intro.tex
outlining three arguments

@@
 Although the cause of orbital heating is not well understood, interactions
-with giant molecular clouds are thought to play an important role
-\racomment{(citations)}.
+with giant molecular clouds and spiral arms are thought to play an important
+role \racomment{(citations)}.
 Although the velocity of any individual star will only provide a weak age
@@
 dispersion can be used effectively to {\it rank} groups of stars by age.
-\racomment{Say some words here to prime the reader for the upcoming caveats.}
 
@@
 vertical action ($J_z$) and vertical velocity (\vz/W) can only be calculated
-with full 6-dimensional position and velocity information, yet most stars with
-measured rotation periods do not have radial velocity measurements because
-they are relatively faint \kepler\ targets ($\sim$11th-18th magnitudes).
+with full 6-dimensional position and velocity information.
+Unfortunately, most stars with measured rotation periods do not have radial
+velocity (RV) measurements because they are relatively faint \kepler\ targets
+($\sim$11th-18th magnitudes).
 For this reason, we used an alternative age proxy: velocity in the direction
@@
 
-In this paper we test the \citet{angus2019} gyrochronology relation, which
-was calibrated using the period-color relation of Praesepe and the period-age
-relation of Praesepe and the Sun.
-It is calibrated to \gaia\ \gcolor\ color.
+In this paper, we tested the \citet{angus2019} gyrochronology relation, which
+was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
+color) and the period-age relation of Praesepe and the Sun.
 % This relation was calibrated by fitting a 5th-order polynomial to the relation
@@
 period-color relation of the Hyades, a cluster of around the same age: $\sim$
-650 years \citep{douglas2019}.
+650 years \citep{douglas2016, douglas2017, rebull2017}.
 However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1

### 2019-08-01 paper/results.tex
outlining three arguments

@@
 Less massive stars do indeed have larger velocity dispersions, however they
-are also older on average.
+are also older on average \racomment{(citations)}.
 This mass-age degeneracy is highly reduced in M dwarfs because their
@@
 
-To investigate further, we calculated the exponent of the AVR for each
+The \vb\ AVR is not directly comparable to the \vz\ AVR, so, to draw a
+comparison with results from the literature, we calculated a \vz\ AVR for
+the subset of 290 stars in our sample with \gaia\ RVs.
+We measured an AVR exponent of 0.47 $\pm$ 0.1, which falls within the range of
+values (0.45-0.53) reported from measurements of F and G stars in the Solar
+neighborhood \citep{holmberg2009, aumer2009, aumer2016}.
+It is a little lower than the values of 0.56 $\pm$ 0.14 (for low-z) and 0.51
+$\pm$ 0.15 (for high-z stars), calculated using \LAMOST\ \racomment{(LAMOST
+citation)} K giants \citep{yu2018}.
+K giants are more massive than K dwarfs \racomment{(how much more massive?)},
+so this slightly higher value contradicts the mass-dependent heating
+hypothesis.
+\racomment{Add some words about selection functions...}
+
+To investigate further, we calculated the exponent of the (\vb) AVR for each
 temperature bin in figure \ref{fig:age_cut}.
-If mass-dependent heating is strongly affecting our data, the exponent of the
-AVR should increase with decreasing effective temperature, \ie\ the heating
-rate should be greater for lower-mass stars.
-Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
-decreasing effective temperature, however, the AVR exponent calculated using
-more massive stars from the Geneva Copenhagen Survey (GCS)
-\citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
-Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
-in our study.
-If mass-dependent orbital heating is the main cause the rise in velocity
-dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
-AVR exponent of the more massive GCS stars should fall {\it below} the AVR
-exponent for the most massive stars in our sample.
-In other words, the dynamical heating rate should be lowest for the highest
-mass stars and highest for the lowest mass stars.
 \begin{figure}
@@
 \end{figure}
-% nordstrom2004, jorgensen2005,
-These AVR exponents were calculated with different populations of stars,
-subject to different selection effects, so directly comparing them comes with
-some risk.
-The most important difference is that the GCS AVR is calculated using \vz,
-however our AVRs are calculated using \vb\ since most stars in our sample do
-not have radial velocities.
-The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
-latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
-close approximation to \vz.
-Only 290 out of the 6820 stars in our sample, plotted in color in figure
-\ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
-that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
-between 3500 and 5000 K.
-For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
-AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
-The similarity of these two AVR exponents suggests that directly comparing
-these two quantities {\it may} be a reasonable approach.
-For now, given the large difference between the (\vz) AVR exponent of the GCS
-sample and the (\vb) AVR exponent of the hottest stars in our sample, we
-assume that the increased velocity dispersion at cooler temperatures is mostly
-caused by incorrect age-grouping due to an incorrect period-color relation at
-old ages, and that any mass-dependent heating, while it may contribute at a
-low level to this result, is not the dominant driver.
-However, differentiating the effects of mass-dependent heating and the shape
-of the gyrochronology relations is certainly warranted in a follow-up study.
 
-The Praesepe-calibrated gyrochronology relation seems to accurately predict
-the relative ages of {\it young} field stars.
-This is shown by the consistent velocity dispersion of young stars as a
-function of temperature in figure \ref{fig:age_cut}.
+% If mass-dependent heating is strongly affecting our data, the exponent of the
+% AVR should increase with decreasing effective temperature, \ie\ the heating
+% rate should be greater for lower-mass stars.
+% Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
+% decreasing effective temperature, however, the AVR exponent calculated using
+% more massive stars from the Geneva Copenhagen Survey (GCS)
+% \citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
+% Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
+% in our study.
+% If mass-dependent orbital heating is the main cause the rise in velocity
+% dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
+% AVR exponent of the more massive GCS stars should fall {\it below} the AVR
+% exponent for the most massive stars in our sample.
+% In other words, the dynamical heating rate should be lowest for the highest
+% mass stars and highest for the lowest mass stars.
+% % nordstrom2004, jorgensen2005,
+% These AVR exponents were calculated with different populations of stars,
+% subject to different selection effects, so directly comparing them comes with
+% some risk.
+% The most important difference is that the GCS AVR is calculated using \vz,
+% however our AVRs are calculated using \vb\ since most stars in our sample do
+% not have radial velocities.
+% The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
+% latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
+% close approximation to \vz.
+% Only 290 out of the 6820 stars in our sample, plotted in color in figure
+% \ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
+% that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
+% between 3500 and 5000 K.
+% For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
+% AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
+% The similarity of these two AVR exponents suggests that directly comparing
+% these two quantities {\it may} be a reasonable approach.
+% For now, given the large difference between the (\vz) AVR exponent of the GCS
+% sample and the (\vb) AVR exponent of the hottest stars in our sample, we
+% assume that the increased velocity dispersion at cooler temperatures is mostly
+% caused by incorrect age-grouping due to an incorrect period-color relation at
+% old ages, and that any mass-dependent heating, while it may contribute at a
+% low level to this result, is not the dominant driver.
+% However, differentiating the effects of mass-dependent heating and the shape
+% of the gyrochronology relations is certainly warranted in a follow-up study.
+% }
+
+The consistent velocity dispersion of young stars as a function of temperature
+in figure \ref{fig:age_cut} shows that the Praesepe-calibrated gyrochronology
+relation accurately predicts the relative ages of {\it young} field stars.
 To our knowledge, no gyrochronology relation has never been demonstrated to
@@
 period-color relation.
-Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
-the \citet{angus2019} gyrochronology relation, compared with the ages of star
-groups, predicted with the \citet{holmberg2009} AVR.
-\citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
-relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
-dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
-dispersion across all temperatures, of our sample.
-\begin{figure}
-  \caption{
-      Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
-    predicted by the \citet{angus2019} gyrochronology relation, for stars in
-    different temperature ranges.
-Points are colored by the effective temperature in the center of the \teff\
-bin.
-The dashed line shows the y=x relation.
-Predicted kinematic ages fan out over gyrochronal time, either because the AVR
-    is temperature-dependent or because the gyrochronology relations have an
-    age-dependent period-color relation.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{age_comparison}
-\label{fig:age_comparison}
-\end{figure}
-Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
-the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
-predicted by the \citet{angus2019} gyrochronology relation, for stars in
-different temperature ranges.
-The predicted kinematic ages fan out over gyrochronal time, either because the
-AVR is temperature-dependent or because the gyrochronology relations have an
-age-dependent period-color relation.
-Figure \ref{fig:age_comparion} also shows that the AVR seems to break down for
-the youngest age bin, 0.5 Gyr, which has a lower velocity dispersion than
-expected, given its age.
+
+% Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
+% the \citet{angus2019} gyrochronology relation, compared with the ages of star
+% groups, predicted with the \citet{holmberg2009} AVR.
+% \citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
+% relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
+% dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
+% dispersion across all temperatures, of our sample.
+% \begin{figure}
+%   \caption{
+%       Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
+%     predicted by the \citet{angus2019} gyrochronology relation, for stars in
+%     different temperature ranges.
+% Points are colored by the effective temperature in the center of the \teff\
+% bin.
+% The dashed line shows the y=x relation.
+% Predicted kinematic ages fan out over gyrochronal time, either because the AVR
+%     is temperature-dependent or because the gyrochronology relations have an
+%     age-dependent period-color relation.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{age_comparison}
+% \label{fig:age_comparison}
+% \end{figure}
+% Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
+% the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
+% predicted by the \citet{angus2019} gyrochronology relation, for stars in
+% different temperature ranges.
+% The predicted kinematic ages fan out over gyrochronal time, either because the
+% AVR is temperature-dependent or because the gyrochronology relations have an
+% age-dependent period-color relation.
+% Figure \ref{fig:age_comparion} also suggests that dynamical heating does not
+% begin until after around ... Gyr:
+
 % Alternatively, the ages of these young stars could be under-predicted by
@@
 
-We applied the same analysis described at the beginning of this section to
-groups of stars selected using rotation period groups, instead of age groups.
-The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with
-stars in different period ranges plotted in different colors.
+For now, we assume that the rise in velocity dispersions at cooler effective
+temperatures is caused by an inaccurate period-color relation rather than
+mass-dependent dynamical heating.
+This assumption allows us to estimate what the shape of the period-color
+relations {\it should} be.
+Figure \ref{fig:age_cut} indicates that the period-color relation flattens
+out, so we applied the same analysis to groups of stars with similar rotation
+periods, equivalent to a completely flat period-color relation.
+The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
+in different period ranges plotted in different colors.
 The bottom panel shows the velocity dispersion of each group as a function of
@@
 For the most rapidly rotating groups of stars, velocity dispersion decreases
-with \teff.
-This is expected given the positively sloped period-color relation of Praesepe
-and other young clusters: late K dwarfs rotate more slowly than early K dwarfs
-of the same age.
-Between 15 and 25 day rotation periods, the temperature dependence of the
+with \teff\ as expected given the positively sloped period-color relation of
+Praesepe and other young clusters: late K dwarfs rotate more slowly than early
+K dwarfs of the same age.
+Between rotation periods of 15 and 25 days, the temperature dependence of the
 velocity dispersion starts to disappear, indicating that the period-color
@@
 At long rotation periods, the velocity dispersion still increases with
-effective temperature, although the increase is much more modest than in
-figure \ref{fig:age_cut}.
-This suggests that the period-color relation may have a negative slope at old
-ages: late K dwarfs may rotate {\it more} rapidly than early K dwarfs.
+effective temperature, although the increase is much more modest compared to
+the oldest stars in figure \ref{fig:age_cut}.
+This suggests that rotation period may {\it decrease} with effective
+temperature, \ie\ late K dwarfs may rotate {\it more} rapidly than early K
+dwarfs.
 This would be a paradigm shift for gyrochronology, stellar spin-down rate is
@@
 fields which {\it should} lead to more efficient angular momentum loss.
+Unfortunately, we still cannot rule out the possibility that mass-dependent
+heating is acting on these stars which could be responsible for some, if not
+all, of the increased velocity dispersion at cooler effective temperatures.

### 2019-08-01 dispersion.py
adding err_on_sample_std

@@ -1,9 +1,10 @@
 import numpy as np
-import matplotlib.pyplot as plt
 from tqdm import trange
+import scipy.stats as sps
+import matplotlib.pyplot as plt
 
 
-def MC_dispersion(x, y, xerr, yerr, bins, nsamps):
+def MC_dispersion(x, y, xerr, yerr, bins, nsamps, method="std"):
     """
     Calculate the dispersion in a set of bins, with Monte Carlo uncertainties.
 
@@ -14,10 +15,15 @@ def MC_dispersion(x, y, xerr, yerr, bins, nsamps):
         yerr (array): The y-value uncertainties.
         bins (array): The bin edges. length = number of bins + 1
         nsamps (int): Number of Monte Carlo samples.
+        method (str): The method to use. Either "std" for standard deviation
+            or "mad" for median absolute deviation.
 
     Returns:
         dispersion (array): The dispersion array. length = number of bins.
         dispersion_err (array): The uncertainty on the dispersion.
+            length = number of bins.
+        kurtosis (array): The kurtosis array. length = number of bins.
+        kurtosis_err (array): The uncertainty on the kurtosis.
             length = number of bins.
 
     """
@@ -26,17 +32,22 @@ def MC_dispersion(x, y, xerr, yerr, bins, nsamps):
     # Sample from Gaussians
     xsamps = np.zeros((len(x), nsamps))
     ysamps = np.zeros((len(y), nsamps))
-    dsamps = np.zeros((nbins, nsamps))
+    dsamps, ksamps = [np.zeros((nbins, nsamps)) for i in range(2)]
 
     for i in trange(nsamps):
         xsamps[:, i] = np.random.randn(len(x))*xerr + x
         ysamps[:, i] = np.random.randn(len(y))*yerr + y
-        dsamps[:, i] = dispersion(xsamps[:, i], ysamps[:, i], bins)
+        d, k = dispersion(xsamps[:, i], ysamps[:, i], bins, method=method)
+        dsamps[:, i] = d
+        ksamps[:, i] = k
 
-    return np.mean(dsamps, axis=1), np.std(dsamps, axis=1)
+    dispersion_err = np.std(dsamps, axis=1)
+    kurtosis_err = np.std(ksamps, axis=1)
+    return np.mean(dsamps, axis=1), dispersion_err, np.mean(ksamps, axis=1), \
+        kurtosis_err
[…]

### 2019-08-01 test_dispersion.py
adding tests

@@ -1,0 +1,122 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from dispersion import *
+import scipy.stats as sps
+import pandas as pd
+
+
+def test_dispersion():
+    np.random.seed(42)
+    N = 10000
+    x = np.random.uniform(0, 100, N)
+    y = np.random.randn(N)*x
+    inds = np.argsort(x)
+    x, y = x[inds], y[inds]
+
+    # Running dispersion
+    newx, d = running_dispersion(x, y, 100, mad=False)
+
+    AT = np.vstack((newx, np.ones_like(newx)))
+    ATA = np.dot(AT, AT.T)
+    w = np.linalg.solve(ATA, np.dot(AT, d))
+
+    # Binned dispersion
+    bins, dbins, err, mean = binned_dispersion(x, y, 10, method="rms")
+
+    # Dispersion where you define the bins.
+    db, k = dispersion(x, y, bins, method="std")
+
+    plt.figure(figsize=(10, 6))
+    plt.plot(x, y, ".")
+    plt.plot(x, x, zorder=3, label="True dispersion")
+    plt.plot(newx, d, ".", label="running")
+    plt.plot(x, w[0]*x + w[1], label="fit to running")
+    plt.plot(np.diff(bins)*.5+bins[:-1], db, "k*", label="pre-set bins",
+             zorder=10)
+    plt.title("TESTING DISPERSION")
+    plt.errorbar(bins, dbins, yerr=err, fmt="wo", markeredgecolor="k",
+                 label="RMS")
+    plt.legend()
+    plt.savefig("dispersion_test")
+
+    assert np.isclose(w[0], 1, atol=.1)
+
+
+def test_MC():
+    np.random.seed(42)
+    N = 10000
+    x = np.random.uniform(0, 100, N)
+    y = np.random.randn(N)*x
+    xerr, yerr = 5, 10
+    xerrs = np.random.randn(N)*xerr
+    yerrs = np.random.randn(N)*yerr
+    bins = np.linspace(0, 100, 10)
+    d, d_err, k, k_err = MC_dispersion(x+xerrs, y+yerrs, xerrs, yerrs, bins,
+                                       100, method="std")
+    d2, d2_err, k2, k2_err = MC_dispersion(x+xerrs, y+yerrs, xerrs, yerrs,
[…]

### 2019-08-01 paper/conclusion.tex
AVR argument

@@
 
-We found that groups of K dwarfs, selected to be coeval using the
-\citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
+We found that old groups of K dwarfs selected to be coeval using the
+\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
 dispersion across all temperatures.
-This indicates that {\it either} the \citep{angus2019} period-color relation
-is wrong, {\it or} that mass-dependent dynamical heating is affecting this
-data set.
+This indicates that either the \citep{angus2019} period-color relation is
+wrong, or that mass-dependent dynamical heating is affecting this data set.
 We presented three pieces of evidence that suggest mass-dependent heating
@@
 
-\item{No strong evidence for mass-dependent heating has been found among M, L
+\item{
+        No strong evidence for mass-dependent heating has been found among M, L
 and T-type stars.
 These low-mass stars live longer than the current age of the Universe and
-therefore do not suffer from a mass-age degeneracy \citep{faherty2009}.}
+        therefore do not suffer from a mass-age degeneracy
+        \citep{faherty2009}.
+}
 
-\item{Literature measurements of the (\vz) AVR exponent are slightly higher
+\item{
+        Literature measurements of the (\vz) AVR exponent are slightly higher
 than the value measured from our data set, but calculated using more massive
@@
 We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
-subject to different selection biases than those used to calculate AVRs in the
-literature \citep{holmberg2009, aumer2009, yu2018}, so drawing direct
-comparisons between them is not a robust test.}
+        subject to different selection biases than those used to calculate
+        AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
+        drawing direct comparisons between them does not provide a reliable
+        test.
+}
 
-\item{The differences in the (\vb) AVR exponent between the hottest and
+\item{
+        The differences in the (\vb) AVR exponent between the hottest and
 coolest K dwarfs in our sample are extremely large, spanning values between
-0.3 and 0.65, yet the difference in mass is less than a factor of two.
+0.4 and 0.75, yet the difference in mass is less than a factor of two.
 Such a large spread in AVR exponents cannot be explained by mass-dependent
-heating alone.}
+        heating alone.
+}
 

### 2019-08-01 paper/results.tex
AVR argument

@@
   \caption{
-The exponent of the AVR as a function of effective temperature.
-    The AVR exponent of \citet{holmberg2009}, calculated using a more massive
-    sample of stars, is shown as a horizontal dashed line.
+      The exponent of the (\vb) AVR as a function of effective temperature.
 }
@@
 \end{figure}
+Figure \ref{fig:AVR_exponent} shows the significant rise the in AVR exponent
+as a function of effective temperature.
+The hottest stars in our sample (\teff\ = 4833-5000 K) have a (\vb) AVR
+exponent of 0.39 $\pm$ 0.02 and the coolest (\teff\ = 3500-3667 K) have an
+exponent of 0.74 $\pm$ 0.02.
+The masses of stars in our sample differ by less than a factor of two,
+so mass-dependent heating cannot be entirely responsible for this large spread
+in heating rates.
+\racomment{Need to look into this further.}
 

### 2019-08-02 paper/intro.tex
not spiral

@@
 Although the cause of orbital heating is not well understood, interactions
-with giant molecular clouds and spiral arms are thought to play an important
+with giant molecular clouds are thought to play an important
 role \racomment{(citations)}.

### 2019-08-03 paper/intro.tex
some more edits

--- 2019-08-01 paper/intro.tex
@@
 caused by the deeper convective zones, and therefore stronger magnetic dynamos
-(and more efficient magnetic braking) in lower-mass stars
-\footnote{Color is the directly observable quantity to which empirical
-gyrochronology relation are usually calibrated to, rather than mass or
-effective temperature.}.
-However, an underlying assumption behind many of the empirical gyrochronology
+(and more efficient magnetic braking) in lower-mass stars.
+However, an underlying assumption behind many empirical gyrochronology
 relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
 angus2015, angus2019} is that the relationships between rotation period and
-color, and rotation period and age are {\it separable}.
+photometric color\footnote{As the directly observable quantity, color is
+usually used as a mass-proxy, and empirical gyrochronology relations are
+usually calibrated in color, rather than mass or effective temperature.
+}, and rotation period and age are {\it separable}.
 In other words, the period-color relation is the same at all ages and the
 period-age relation is the same at all colors.
-It was recently shown that old stars rotate more rapidly than a simple,
+It was recently shown that old field stars rotate more rapidly than a simple,
 separable gyrochronology relation would predict \citep{angus2015,
@@
 old ages.
-For this reason, we turned to a population-based stellar age indicator,
-velocity dispersion, to investigate the period-temperature relations at old
-ages.
+For this reason, we used a population-based stellar age indicator, velocity
+dispersion, to investigate the period-color relations at old ages.
 % Since a separable period-age and period-color relation provided an adequate
@@
 color) and the period-age relation of Praesepe and the Sun.
+The large number of Praesepe members with precise rotation periods from the
+\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
+through early M, makes it the ideal cluster for calibrating the period-color
+relation of stars at 650 Myrs.
 % This relation was calibrated by fitting a 5th-order polynomial to the relation
@@
 % and the Sun.
-Although only calibrated using Praesepe and the Sun, this gyrochronology
-relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
-accurate ages.
-In addition, the period-color relation of Praesepe is nearly identical to the
-period-color relation of the Hyades, a cluster of around the same age: $\sim$
-650 years \citep{douglas2016, douglas2017, rebull2017}.
-However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
-Gyr cluster \citep{curtis2019}.
+Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
+Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
+members of NGC 6819, a 2.5 Gyr open cluster.
+However there are no rotation periods for K or M dwarfs in this cluster -- its
+coolest members with rotation periods are G dwarfs.
+The period-color relation of Praesepe is nearly identical to the period-color
+relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
+\citep{douglas2016, douglas2017, rebull2017}.
+However, Praesepe and the Hyades do {\it not} have the same period-color
+relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
 The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
@@
 relation for all stars.
-However, the NGC 6811 cluster suggests that this assumption does not hold past
-around 1 Gyr.
+However, the NGC 6811 cluster suggests that this assumption does not hold for
+K dwarfs past around 1 Gyr.
 If the period-color relation of the \citet{angus2019} gyrochronology model
@@
 of stars selected to be similar ages using this relation should fall on an
-isochrone on a CMD, an isochrone in period-color space, and an
-isochrone in velocity dispersion-color space.
+isochrone on a CMD.
+% and have the same velocity dispersion across all colors.
 Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
@@
 Isochrones and stellar evolution tracks are highly dependent on choices made
-about input physics and assumptions and have often been calibrated using
-different types of stars.
+about input physics and assumptions.
+% and have often been calibrated using
+% different types of stars.
 As a result, different sets of models can have very different shapes on the
 CMD, particularly at low masses.
+In fact, only empirical models, not physical ones, are currently able to
+reproduce the CMD positions of M dwarfs.
 Instead of relying on CMD position to age-date groups of stars, we opted to
@@
 groups have a larger velocity dispersion.
-It is less easy to rank-order groups of stars by age on the CMD because
-stellar evolution is highly mass and metallicity dependent, and because
-different stellar evolution models predict different ages for the same CMD
-position.
-This does not mean that it is not possible to calibrate gyrochronology using
-isochrones, however it is not the focus of this study.
+% It is less easy to rank-order groups of stars by age on the CMD because
+% stellar evolution is highly mass and metallicity dependent, and because
+% different stellar evolution models predict different ages for the same CMD
+% position.
+% This does not mean that it is not possible to calibrate gyrochronology using
+% isochrones, however it is not the focus of this study.

### 2019-08-03 paper/method.tex
tidying up and refining text
some more edits

@@
 A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
-    extinction-corrected magnitudes, colored by their rotation periods.
+    extinction-corrected magnitudes, colored by rotation period.
 We excluded photometric binaries and subgiants from our analysis by removing
@@
 spin-down law.
-Tidal and magnetic interactions between the two components of a binary system
-can influence the rotation periods of both stars, and the expanding envelopes
-of subgiants drive rapid spin-down through conservation of angular momentum.
+% Tidal and magnetic interactions between the two components of a binary system
+% can influence the rotation periods of both stars, and the expanding envelopes
+% of subgiants drive rapid spin-down through conservation of angular momentum.
 We removed visual binaries and subgiants from the sample by applying cuts to
@@
 We removed all stars above this line from the sample.
-We removed subgiants by eliminating stars brighter than 6th magnitude in
-\gaia\ G-band from the sample.
+We also removed subgiants by eliminating stars brighter than 6th magnitude in
+\gaia\ G-band.
+\begin{figure}
+  \caption{
+Dereddened main sequence stars with \mct\ rotation periods on the \gaia\ CMD
+    with visual binaries removed.  Points are colored by their gyrochronal
+    age, according to the
+    \citet{angus2019} gyrochronology relation.
+    A general age gradient is visible across the main sequence.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{age_gradient}
+\label{fig:age_gradient}
+\end{figure}
 
@@
 below 10 from the sample.
-We also removed a small number of stars fainter than 16th magnitude from the
-sample.
-We used the {\tt Pyia} {\it Python} package to calculate stellar velocities.
+We also removed a small number of stars fainter than 16th magnitude.
+We used the {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy}
+\citep{astropy2013, astropy2018} {\it Python} packages to calculate stellar
+velocities.
 {\tt Pyia} has built-in functionality for calculating velocity samples from

### 2019-08-03 paper/results.tex
some more edits

@@
     The full catalog, with subgiants and visual binaries removed is shown in
-    grey, and stars selected to be in different age groups are overlayed in
-    color.
+    grey, and stars selected to be in different age groups (between 3500 and
+    5000 K) are overlayed in color.
 These age groups were selected using the \citet{angus2019} gyrochronology
     relation.
-The legend in the center of the figure lists the age range of each group.
-    Bottom: velocity dispersion vs effective temperature for each age group.
+The legend in the center of the figure lists the age range, in Gyr, of each
+    group.
+Bottom: velocity dispersion vs effective temperature for each age
+    group.
 The color of the line corresponds to the color of the group shown in the top
     panel.
-If the gyrochronal model were correct at all ages and the stars in each group
+If the gyrochronal model were correct at all ages, and the stars in each group
     were the same age across temperatures, the velocity dispersion would be
     constant as a function of \teff.
-However, the velocity dispersions of the oldest age groups depend on \teff,
-    indicating that the \citet{angus2019} gyrochronology model either
-    overpredicts the ages of old M dwarfs or underpredicts the ages of old G
-    and K dwarfs.
+However, the velocity dispersions of the oldest age groups increase \teff,
+    indicating the \citet{angus2019} gyrochronology model underpredicts the
+    the ages of late-K dwarfs relative to the ages of early K dwarfs at old
+    ages.
+An alternative explanation could be that the gyrochronology relation is
+    correct and {\it mass-dependent heating} is responsible for the greater
+    velocity dispersions of cooler stars.
 }
@@
 color.
-The color of the points corresponds to the age ranges specified in the legend,
-which also apply to the lines in the lower panel.
+The color of the points corresponds to the age ranges specified in the legend
+(in Gyr), which also apply to the lines in the lower panel.
 The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
@@
 with gyrochronal age across all temperatures.
-However, if the stars in each selected age group had the {\it same} age across
-the temperature range, their velocity dispersion would be a constant function
-of \teff.
-Although the youngest age groups have a relatively constant velocity
+The consistent velocity dispersion of young stars as a function of temperature
+shows that the Praesepe-calibrated gyrochronology relation accurately predicts
+the relative ages of {\it young} field stars.
+% To our knowledge, no gyrochronology relation has never been demonstrated to
+% correctly predict ages (either relative or absolute) for such cool or such
+% young field stars.
+% This is not particularly remarkable however, since these young stars are a
+% similar age to the Praesepe cluster, which was used to calibrate the
+% period-color relation.
+% However, if the stars in each selected age group had the {\it same} age across
+% the temperature range, their velocity dispersion would be a constant function
+% of \teff.
+However, although the youngest age groups have a relatively constant velocity
 dispersions across temperatures, the oldest age groups do not.
@@
 
-The consistent velocity dispersion of young stars as a function of temperature
-in figure \ref{fig:age_cut} shows that the Praesepe-calibrated gyrochronology
-relation accurately predicts the relative ages of {\it young} field stars.
-To our knowledge, no gyrochronology relation has never been demonstrated to
-correctly predict ages (either relative or absolute) for such cool or such
-young field stars.
-This is not particularly remarkable however, since these young stars are a
-similar age to the Praesepe cluster, which was used to calibrate the
-period-color relation.
-
 % Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
@@
 
-For now, we assume that the rise in velocity dispersions at cooler effective
+If we assume that the rise in velocity dispersions at cooler effective
 temperatures is caused by an inaccurate period-color relation rather than
-mass-dependent dynamical heating.
-This assumption allows us to estimate what the shape of the period-color
-relations {\it should} be.
+mass-dependent dynamical heating, we can estimate what the shape of the
+period-color relations {\it should} be.
 Figure \ref{fig:age_cut} indicates that the period-color relation flattens

### 2019-08-08 paper/results.tex
adding velocity dispersion as color

@@
 fields which {\it should} lead to more efficient angular momentum loss.
+
+\begin{figure}
+  \caption{
+The rotation periods vs effective temperatures of stars in the \mct\ sample,
+    colored by the velocity dispersions of stars in restricted period and
+    temperature ranges.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{dispersion_period_teff}
+\label{fig:dispersion_period_teff}
+\end{figure}
+
+Figure \ref{fig:dispersion_period_teff} shows the stars in our sample on a
+plot of period vs effective temperature, coloured by their (\vb) velocity
+dispersion, where velocity dispersion was calculated in period and temperature
+bins (made apparent by the visible grid-like structure).
+If we assume that mass dependent heating does {\it not} affect this sample,
+velocity dispersion can be interpreted as an age proxy and stars bins of
+similar colour in figure \ref{fig:dispersion_period_teff} are similar ages.
+Interpreted this way, it appears as though isochrones slope downwards in
+period-temperature space at old ages.
+This figure also suggests that the oldest stars with measured rotation periods
+are cool, which may be because cooler stars remain active for longer, and
+their rotation periods can therefore be measured at older ages.
+
 Unfortunately, we still cannot rule out the possibility that mass-dependent

### 2019-08-19 paper/abstract.tex
redrafting

@@
 
-Due to a lack of precise ages and rotation periods for old main-sequence
-stars, the rotational evolution of FGKM dwarfs is mostly unconstrained
-after around 2-3 billion years.
-However, recent evidence suggests that the relationship between rotation
-period and color/temperature/mass does not remain constant over large
-timescales, contradicting a long-held assumption that underlies many
-gyrochronology relations.
-In this work we use the kinematic ages of populations of K dwarfs to reveal
-the evolving relationship between rotation period and effective temperature.
-We find that groups of stars, selected to be the same age, do {\it not} have
+The rotational evolution of cool dwarfs is mostly unconstrained after around
+2-3 billion years, and the gyrochronology relations are poorly calibrated
+beyond this time due to a lack of precise ages and rotation periods for old
+main-sequence stars.
+% However, recent evidence suggests that the relationship between rotation
+% period and color/temperature/mass does not remain constant over large
+% timescales, contradicting a long-held assumption that underlies many
+% gyrochronology relations.
+In this work we use the velocities of populations of K dwarfs to reveal the
+rotational evolution of these cool stars at old ages.
+We find that groups of stars, selected to be the same age using a
+gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
 the same velocity dispersion at all temperatures; cooler stars have greater
 velocity dispersions than hotter stars.
-Two different scenarios could be responsible for producing this observation:
-either the rotation period-color relation changes over time, or stars of
-different masses experience different rates of dynamical heating.
+We interpret this to mean that these cooler stars are older than the hotter
+ones, which implies that the period-color relation of Praesepe is not
+appropriate for old stars.
+Futhermore, our results indicate that the slope of the period-color relation
+actually {\it inverts}: at young ages early K stars rotate more rapidly than
+late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
+These observations could be reproduced by mass-dependent dynamical heating,
+where lower-mass stars experience a greater rate of dynamical heating than
+higher-mass stars, however we find no evidence for mass-dependent heating
+within K and M dwarfs in the \kepler\ field.
+
 % We show however, that if mass-dependent heating is responsible, the
@@
 % around 1 Gyr.
-If we assume that mass-dependent heating does not affect the data, our results
-indicate that the slope of the period-color relation actually {\it inverts}:
-at young ages early K stars rotate more rapidly than late K stars, but beyond
-$\sim$ 2 Gyr they rotate more slowly.
+% If we assume that mass-dependent heating does not affect the data,
 

### 2019-08-19 paper/conclusion.tex
redrafting

@@
 dispersion across all temperatures.
-This indicates that either the \citep{angus2019} period-color relation is
-wrong, or that mass-dependent dynamical heating is affecting this data set.
-We presented three pieces of evidence that suggest mass-dependent heating
-cannot be responsible for all of the increasing velocity dispersion.
+% This indicates that either the \citep{angus2019} period-color relation is
+% incorrect, or that mass-dependent dynamical heating is affecting this data set.
+% We presented three pieces of evidence that suggest mass-dependent heating
+% cannot be responsible for all of the increasing velocity dispersion.
 
-\begin{itemize}
+% \begin{itemize}
 
-\item{
-        No strong evidence for mass-dependent heating has been found among M, L
-and T-type stars.
-These low-mass stars live longer than the current age of the Universe and
-        therefore do not suffer from a mass-age degeneracy
-        \citep{faherty2009}.
+% \item{
+%         No strong evidence for mass-dependent heating has been found among M, L
+% and T-type stars.
+% These low-mass stars live longer than the current age of the Universe and
+%         therefore do not suffer from a mass-age degeneracy
+%         \citep{faherty2009}.
+% }
+
+% \item{
+%         Literature measurements of the (\vz) AVR exponent are slightly higher
+% than the value measured from our data set, but calculated using more massive
+% stars.
+% Mass-dependent dynamical heating should however produce a {\it smaller} AVR
+%         exponent for higher-mass stars.
+% We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
+%         subject to different selection biases than those used to calculate
+%         AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
+%         drawing direct comparisons between them does not provide a reliable
+%         test.
+% }
+
+% \item{
+%         The differences in the (\vb) AVR exponent between the hottest and
+% coolest K dwarfs in our sample are extremely large, spanning values between
+% 0.4 and 0.75, yet the difference in mass is less than a factor of two.
+% Such a large spread in AVR exponents cannot be explained by mass-dependent
+%         heating alone.
+% }
+
+% \end{itemize}
+
+\racomment{Words about the significance of an inverted period-color relation.}
+
+\racomment{How does the weakened van Saders braking law fit into all this?}
+
+\racomment{Talk about the fact that the stars in each bin are not necessarily
+the same age.}
+
+\racomment{Dispersion could be dependent on the selection function.
+We cut stars at high b, because we find that these do lead to a bias.
+We find no evidence for increased velocity dispersion with decreasing teff in
+either vb or vz.
+We find that sigma clipping is necessary for vb but not vz and do not find
+that the outliers are distributed towards higher galactic latitudes.
 }
 
-\item{
-        Literature measurements of the (\vz) AVR exponent are slightly higher
-than the value measured from our data set, but calculated using more massive
-stars.
-Mass-dependent dynamical heating should however produce a {\it smaller} AVR
-        exponent for higher-mass stars.
-We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
-        subject to different selection biases than those used to calculate
-        AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
-        drawing direct comparisons between them does not provide a reliable
-        test.
-}
+% \racomment{Trevor suggest that stars forming in clusters could be the reason
+% why young stars are so dynamically cool.}
 
-\item{
-        The differences in the (\vb) AVR exponent between the hottest and
-coolest K dwarfs in our sample are extremely large, spanning values between
-0.4 and 0.75, yet the difference in mass is less than a factor of two.
-Such a large spread in AVR exponents cannot be explained by mass-dependent
-        heating alone.
-}
-
-\end{itemize}
-
-\racomment{Words about the significance of an inverted period-color relation.}
+% \racomment{Are there any asteroseismic stars in this sample that could help
+% pin down the age?}

### 2019-08-19 paper/intro.tex
less cringy
redrafting

@@
 \subsection{Gyrochronology}
-It has long been known that the rotation periods of FGKM dwarfs increase over
+It is well established that the rotation periods of FGKM dwarfs increase over
 time \citep{skumanich1972}.
-This characteristic of main-sequence stars allows them to be dated via their
-rotation periods in a practice known as gyrochronology, which is useful
-since the ages of main-sequence stars are extremely difficult to measure via
-the traditional age-dating method of isochrone placement.
-It is also well established that stars of the same age but different {\it
-masses} have different rotation periods \racomment{(citation)}, thought to be
-caused by the deeper convective zones, and therefore stronger magnetic dynamos
-(and more efficient magnetic braking) in lower-mass stars.
+This characteristic of main-sequence (MS) stars allows them to be dated via
+their rotation periods in a practice known as gyrochronology, which is useful
+since the ages of MS stars are extremely difficult to measure via the
+traditional age-dating method of isochrone placement.
+It is also well established that stars of the same age but different masses
+have different rotation periods \racomment{(citation)}, thought to be caused
+by the deeper convective zones, and therefore stronger magnetic dynamos (and
+more efficient magnetic braking) in lower-mass stars.
 However, an underlying assumption behind many empirical gyrochronology
-relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
-angus2015, angus2019} is that the relationships between rotation period and
-photometric color\footnote{As the directly observable quantity, color is
-usually used as a mass-proxy, and empirical gyrochronology relations are
-usually calibrated in color, rather than mass or effective temperature.
-}, and rotation period and age are {\it separable}.
-In other words, the period-color relation is the same at all ages and the
-period-age relation is the same at all colors.
+relations is that the relationships between rotation period and photometric
+color\footnote{As the directly observable quantity, color is usually used as a
+mass-proxy, and empirical gyrochronology relations are usually calibrated in
+color, rather than mass or effective temperature.
+}, and rotation period and age are {\it separable}, meaning that the
+period-color relation is the same at all ages and the period-age relation is
+the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
+meibom2011, angus2015, angus2019}.
 It was recently shown that old field stars rotate more rapidly than a simple,
@@
 to reproduce the data \citep{vansaders2016, vansaders2018}.
-An even more recent analysis of middle-aged open clusters provides further
-indication that the exponent of the period-age relation is mass-dependent.
-New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
-reveal a flattened relationship between rotation period and color
-\citep{curtis2019}.
-In this cluster, the G dwarfs rotate at the same rate as the K dwarfs.
-
-Unfortunately, the open clusters with rotation period measurements are mostly
-young -- currently the oldest is 2.5 Gyr (NGC 6819).
-Without rotation periods for precisely dated old stars, it is extremely
-difficult to calibrate the relationship between rotation period and color at
-old ages.
-For this reason, we used a population-based stellar age indicator, velocity
-dispersion, to investigate the period-color relations at old ages.
+An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
+further indication that the exponent of the period-age relation is
+mass and time-dependent.
+This cluster has a flattened relationship between rotation period and color:
+the G dwarfs rotate at the same rate as the K dwarfs \citep{curtis2019}.
 % Since a separable period-age and period-color relation provided an adequate
@@
 
+In this paper, we tested the \citet{angus2019} gyrochronology relation, which
+was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
+color) and the period-age relation of Praesepe and the Sun.
+The large number of Praesepe members with precise rotation periods from the
+\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
+through early M, makes it the ideal cluster for calibrating the period-color
+relation of stars at 650 Myrs.
+% This relation was calibrated by fitting a 5th-order polynomial to the relation
+% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
+% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
+% and the Sun.
+This relation accurately describes the rotation periods of F and G stars up to
+around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
+rotation periods), but over-predicts the rotation periods of K dwarfs in the
+1.1 Gyr NGC 6811 cluster.
+No reliable age estimates exist for K dwarfs with rotation periods older than
+1.1 Gyr, so the rotational evolution of these stars beyond this age is a
+mystery.
+% In this paper, we shed light on the rotational evolution of old K dwarfs using
+% population-based dynamical ages.
+% Unfortunately, open clusters with rotation period measurements are mostly
+% young -- currently the oldest is 2.5 Gyr (NGC 6819).
+% Without rotation periods for precisely dated old stars, it is extremely
+% difficult to calibrate the relationship between rotation period and color at
+% old ages.
+For this reason, we used a population-based stellar age indicator, velocity
+dispersion, to investigate the period-color relations at old ages.
+
 \subsection{Kinematics as an age proxy}
 
-Stars are thought to be born in the thin disk of the Milky Way, orbiting the
-center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
-$v_z$).
+Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
+the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
+or $v_z$).
+% because the MW's gas disk is thin and its gas moves in circular
+% orbits.
 Young stars have relatively small vertical velocities, but gain momentum in
@@
 Although the cause of orbital heating is not well understood, interactions
-with giant molecular clouds and spiral arms are thought to play an important
-role \racomment{(citations)}.
+with giant molecular clouds, spiral arms and the galactic bar are thought to
+play an important role \citep[see][for a review of secular evolution in the
+MW]{sellwood2014}.
 Although the velocity of any individual star will only provide a weak age
@@
 % in this sense, {\it actions} are the natural quantities to use as age proxies.
-Vertical action is a better age indicator than vertical velocity, however both
-vertical action ($J_z$) and vertical velocity (\vz/W) can only be calculated
-with full 6-dimensional position and velocity information.
+Vertical action is a better age indicator than vertical velocity
+\citep[although still only a weak age indicator for individual
+stars,][]{beane2018}, however both vertical action ($J_z$) and vertical velocity
+(\vz/W) can only be calculated with full 6-dimensional position and velocity
+information.
 Unfortunately, most stars with measured rotation periods do not have radial
@@
 ($\sim$11th-18th magnitudes).
-For this reason, we used an alternative age proxy: velocity in the direction
-of galactic latitude, \vb.
+For this reason, we used velocity in the direction of galactic latitude, \vb.
 The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
-\vb\ is a close approximation to \vz.
+\vb\ is a close (although not perfect -- see section \ref{sec:results})
+approximation to \vz.
 
@@
 
-In this paper, we tested the \citet{angus2019} gyrochronology relation, which
-was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
-color) and the period-age relation of Praesepe and the Sun.
-The large number of Praesepe members with precise rotation periods from the
-\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
-through early M, makes it the ideal cluster for calibrating the period-color
-relation of stars at 650 Myrs.
-% This relation was calibrated by fitting a 5th-order polynomial to the relation
-% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
-% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
-% and the Sun.
-Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
-Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
-members of NGC 6819, a 2.5 Gyr open cluster.
-However there are no rotation periods for K or M dwarfs in this cluster -- its
-coolest members with rotation periods are G dwarfs.
-The period-color relation of Praesepe is nearly identical to the period-color
-relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
-\citep{douglas2016, douglas2017, rebull2017}.
-However, Praesepe and the Hyades do {\it not} have the same period-color
-relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
-The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
-appears as though the K stars have `stalled' -- their spin-down has been
-halted.
-The \citet{angus2019} gyrochronology relation assumes that the rotation
-period-color relation of Praesepe is applicable to stars of all ages: the same
-polynomial relation fit to Praesepe is used to describe the period-color
-relation for all stars.
-However, the NGC 6811 cluster suggests that this assumption does not hold for
-K dwarfs past around 1 Gyr.
-If the period-color relation of the \citet{angus2019} gyrochronology model
-{\it were} a perfect model for the rotational evolution of stars, then groups
-of stars selected to be similar ages using this relation should fall on an
-isochrone on a CMD.
-% and have the same velocity dispersion across all colors.
-Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
-variations in metallicity and extinction, blur out the main sequence enough
-that differences between stars in different age and rotation bins are not
-easily discernable on the CMD, although there is still a general age and
-rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
-\ref{fig:age_gradient}.
-For this reason, we chose to use {\it kinematic} `isochrones', instead of
-magnitude and color-based isochrones: although kinematics as an age indicator
-is not necessarily as well calibrated to an absolute age scale as CMD
-position, it can be very sensitive to {\it differences} in the ages of stellar
-populations.
+% Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
+% Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
+% members of NGC 6819, a 2.5 Gyr open cluster.
+% However there are no rotation periods for K or M dwarfs in this cluster -- its
+% coolest members with rotation periods are G dwarfs.
+% The period-color relation of Praesepe is nearly identical to the period-color
+% relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
+% \citep{douglas2016, douglas2017, rebull2017}.
+% However, Praesepe and the Hyades do {\it not} have the same period-color
+% relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
+% The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
+% appears as though the K stars have `stalled' -- their spin-down has been
+% halted.
 
-Isochrones and stellar evolution tracks are highly dependent on choices made
-about input physics and assumptions.
-% and have often been calibrated using
-% different types of stars.
-As a result, different sets of models can have very different shapes on the
-CMD, particularly at low masses.
-In fact, only empirical models, not physical ones, are currently able to
-reproduce the CMD positions of M dwarfs.
-Instead of relying on CMD position to age-date groups of stars, we opted to
-explore age trends via kinematics.
-Kinematic age-dating has the advantage of being relatively model independent,
-or at least, having a very simple model: that velocity dispersion increases
-over time.
-This means that it is relatively easy to rank groups of stars by age: older
-groups have a larger velocity dispersion.
-% It is less easy to rank-order groups of stars by age on the CMD because
-% stellar evolution is highly mass and metallicity dependent, and because
-% different stellar evolution models predict different ages for the same CMD
-% position.
-% This does not mean that it is not possible to calibrate gyrochronology using
-% isochrones, however it is not the focus of this study.
+% The \citet{angus2019} gyrochronology relation assumes that the rotation
+% period-color relation of Praesepe is applicable to stars of all ages: the same
+% polynomial relation fit to Praesepe is used to describe the period-color
+% relation for all stars.
+% However, the NGC 6811 cluster suggests that this assumption does not hold for
+% K dwarfs past around 1 Gyr.
+% If the period-color relation of the \citet{angus2019} gyrochronology model
+% {\it were} a perfect model for the rotational evolution of stars, then groups
+% of stars selected to be similar ages using this relation should fall on an
+% isochrone on a CMD.
+% % and have the same velocity dispersion across all colors.
+% Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
+% variations in metallicity and extinction, blur out the main sequence enough
+% that differences between stars in different age and rotation bins are not
+% easily discernable on the CMD, although there is still a general age and
+% rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
+% \ref{fig:age_gradient}.
+% For this reason, we chose to use {\it kinematic} `isochrones', instead of
+% magnitude and color-based isochrones: although kinematics as an age indicator
+% is not necessarily as well calibrated to an absolute age scale as CMD
+% position, it can be very sensitive to {\it differences} in the ages of stellar
+% populations.
+
+% Isochrones and stellar evolution tracks are highly dependent on choices made
+% about input physics and assumptions.
+% % and have often been calibrated using
+% % different types of stars.
+% As a result, different sets of models can have very different shapes on the
+% CMD, particularly at low masses.
+% In fact, only empirical models, not physical ones, are currently able to
+% reproduce the CMD positions of M dwarfs.
+% Instead of relying on CMD position to age-date groups of stars, we opted to
+% explore age trends via kinematics.
+% Kinematic age-dating has the advantage of being relatively model independent,
+% or at least, having a very simple model: that velocity dispersion increases
+% over time.
+% This means that it is relatively easy to rank groups of stars by age: older
+% groups have a larger velocity dispersion.

### 2019-08-19 paper/method.tex
redrafting

@@
 % Limit: 500 words
-\section{Method}
 
 \subsection{The data}
+\label{sec:the_data}
 
-We crossmatched the \mct\ catalog of stellar rotation periods, measured from
-\kepler\ light curves, with the \gaia\ DR2 catalog.
+We used the publicly available \kepler-\gaia\ DR2 crossmatched
+catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
+stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
+DR2 catalog.
 Reddening and extinction from dust was calculated for each star using the
-Bayestar dust map implemented in the dustmaps {\it Python} package
-\citep{green2018}.
+Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
+\citep{green2018}, and {\tt astropy} \citep{astropy2013}.
 We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
@@
 %           301.13205924630165, -25.923997443169355]
+\label{eq:curtis}
 \end{equation}
@@
 
-\begin{figure}
-  \caption{
-A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
-    extinction-corrected magnitudes, colored by rotation period.
-We excluded photometric binaries and subgiants from our analysis by removing
-stars above the two dashed lines.
-% The rotation periods of binaries and subgiants do not follow a Skumanich-like
-% braking law.
-% The rotation period gradient across the main sequence is visible by eye in
-%     this figure: young, rapidly rotating stars are located below the old,
-%     slowly rotating stars.
-% since we rely on a simple scaling
-% between rotation period and age to make the argument that the rotation period
-% is not caused by incorrect period measurements.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{CMD_cuts}
-\label{fig:CMD_cuts}
-\end{figure}
-To explore the age of this stellar population from a rotation-period
-standpoint, it was first necessary to remove visual binaries and subgiants
-from the sample.
-The rotation-period evolution of these two types of stars is generally
-different to that of single stars which more usually follow a Skumanich-like
-spin-down law.
+% \begin{figure}
+%   \caption{
+% A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
+%     extinction-corrected magnitudes, colored by rotation period.
+% We excluded photometric binaries and subgiants from our analysis by removing
+% stars above the two dashed lines.
+% % The rotation periods of binaries and subgiants do not follow a Skumanich-like
+% % braking law.
+% % The rotation period gradient across the main sequence is visible by eye in
+% %     this figure: young, rapidly rotating stars are located below the old,
+% %     slowly rotating stars.
+% % since we rely on a simple scaling
+% % between rotation period and age to make the argument that the rotation period
+% % is not caused by incorrect period measurements.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{CMD_cuts}
+% \label{fig:CMD_cuts}
+% \end{figure}
+% To explore the age of this stellar population from a rotation standpoint, it
+% was first necessary to remove visual binaries and subgiants from the sample.
+% The rotational evolution of these two types of stars is generally different to
+% that of single stars which more usually follow a Skumanich-like spin-down law.
 % Tidal and magnetic interactions between the two components of a binary system
@@
 We removed visual binaries and subgiants from the sample by applying cuts to
-the color-magnitude diagram (CMD), shown in figure \ref{fig:CMD_cuts}.
-We fit a 6th-order polynomial to the main sequence and raised it by 0.22 dex,
-to approximate the division between single stars and visual binaries.
-We removed all stars above this line from the sample.
+the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
+We fit a 6th-order polynomial to the main sequence and raised it by 0.27 dex
+to approximate the division between single stars and visual binaries and
+removed all stars above this line from the sample.
 We also removed subgiants by eliminating stars brighter than 6th magnitude in
@@
   \caption{
-Dereddened main sequence stars with \mct\ rotation periods on the \gaia\ CMD
-    with visual binaries removed.  Points are colored by their gyrochronal
+Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
+We excluded visual binaries by removing stars above the dashed line.
+    Points are colored by their gyrochronal
     age, according to the
@@
   \centering
-    \includegraphics[width=1\textwidth]{age_gradient}
+    \includegraphics[width=1\textwidth]{age_gradient_combo}
 \label{fig:age_gradient}
@@
 
-We removed stars with negative parallaxes and parallax signal-to-noise ratios
-below 10 from the sample.
-We also removed a small number of stars fainter than 16th magnitude.
-We used the {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy}
-\citep{astropy2013, astropy2018} {\it Python} packages to calculate stellar
-velocities.
+The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
+astropy2018} {\it Python} packages were used to calculate stellar velocities.
 {\tt Pyia} has built-in functionality for calculating velocity samples from
@@
 between these properties.
-We removed stars with absolute \vb\ uncertainties greater than 1 \kms\ from
-the sample.
+We removed stars with negative parallaxes, parallax signal-to-noise ratios
+less than 10, stars fainter than 16th magnitude, stars with absolute \vb\
+uncertainties greater than 1 \kms\, and stars with galactic latitudes greater
+than 15\degrees\ from the sample.
+Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
+and because latitude is correlated with stellar mass (lower mass stars are
+older and tend to be preferentially located at high $b$), we found that
+including stars with $b > 15$\degrees\ led to a larger number of \vb\ outliers
+at low stellar mass.
+15\degrees\ was found to be an adequate compromise between maintaining a close
+relationship between \vb\ and \vz\ and keeping a large enough sample size to
+provide meaningful results.
+We also found that the anti-correlation between stellar mass and $b$ is
+negligable for stars cooler than 5000 K.
+For this reason, and because the rotational evolution of old, G stars does not
+follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
+restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
+dwarfs.
+
+To explore the relationship between rotation period, \teff\ and age, we
+selected groups of stars within different age ranges \citep[where age was
+calculated using the][gyrochronology relation]{angus2019}, and calculated the
+velocity dispersion: the standard deviation of velocities, \sigmavb, as a
+function of effective temperature for each age group.
+We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
+and temperature bin to remove non-Gaussian outliers.
+Without sigma-clipping, we found that a small number of high velocity
+outliers at the low-temperature end of our sample substantially raised the
+velocity dispersion for cooler stars.
+Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+throughout this paper we show rotation periods as a function of {\it effective
+temperature}.
+We chose to use effective temperatures and not colors in this analysis as it
+is the linear quantity and therefore easier to divide into bins of roughly
+equal numbers of stars.
 

### 2019-08-19 paper/period_gap.tex
redrafting

% zip ms_files.zip ms.tex abstract.tex intro.tex method.tex results.tex conclusion.tex Praesepe.pdf variance.pdf simulated_CMD.pdf rotation_model_praesepe.pdf simulation_results.pdf NGC6819.pdf NGC6819_results.pdf hz.bib

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
% \documentclass[a4paper,fleqn,usenatbib,useAMS]{mnras}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}

\AtBeginDocument{\let\textlabel\label}

\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\etal}{{\it et al.}}

\newcommand{\kepler}{{\it Kepler}}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\tess}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\lsst}{{\it LSST}}
\newcommand{\Wfirst}{{\it WFIRST}}
\newcommand{\wfirst}{{\it WFIRST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\plato}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\panstarrs}{{\it PanSTARRS}}
\newcommand{\LAMOST}{{\it LAMOST}}

\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\prot}{$P_{\mathrm{rot}}$}
\newcommand{\pmega}{$\bar{\omega}$}
\newcommand{\mj}{$m_j$}
\newcommand{\mh}{$m_h$}
\newcommand{\mk}{$m_k$}
\newcommand{\mx}{$m_x$}
\newcommand{\logg}{log(g)}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\degrees}{$^\circ$}
\newcommand{\vz}{$v_{\bf z}$}
\newcommand{\vb}{$v_{\bf b}$}
\newcommand{\kms}{$kms^{-1}$}
\newcommand{\sigmavb}{$\sigma_{v{\bf b}}$}
\newcommand{\sigmavz}{$\sigma_{v{\bf z}}$}

% \newcommand{\nsim_stars}{841}

\newcommand{\amnh}{1}
\newcommand{\cca}{2}
\newcommand{\hawaii}{3}

\newcommand{\sd}{{\tt stardate}}
\newcommand{\gcolor}{$G_{BP} - G_{RP}$}
\newcommand{\mcp}{\citep{mcquillan2014}}
\newcommand{\mct}{\citet{mcquillan2014}}
\newcommand{\bvector}{${\bf b}$}
\newcommand{\python}{{\it Python}}

\newcommand{\racomment}[1]{{\color{blue}#1}}

\begin{document}

\title{Exploring the ages of rotating stars using galactic dynamics: a novel
approach to calibrating gyrochronology}

\author{%
    Ruth Angus,
    Angus Beane,
    Adrian Price-Whelan,
    Jennifer van Saders,
    Elisabeth Newton,
    Jason Curtis,
    Travis Berger,
    Dan Foreman-Mackey,
    Lauren Anderson,
    Megan Bedell,
    Jackie Faherty,
    Rocio Kiman,
    Melissa Ness}

% \altaffiltext{\amnh}{American Museum of Natural History, Central Park West,
% Manhattan, NY, USA}
% \altaffiltext{\cca}{Center for Computational Astrophysics, Flatiron Institute,
% 162 5th Avenue, Manhattan, NY, USA}
% \altaffiltext{\hawaii}{Institute for Astronomy, University of Hawai'i at
% M\={a}noa, Honolulu, HI, USA}

% abstract.
\begin{abstract}

The distribution of rotation periods of K and M stars, measured from light
curves obtained from the \kepler\ spacecraft, has a sharp mass-dependent gap
at around 10-20 days.
This gap traces a line of constant age and constant Rossby number in the
rotation period-effective temperature plane, indicating that the cause could
be related to a discontinuity in either the local star formation history, or
the magnetic braking evolution of stars.
A third explanation for the rotation period gap is measurement error caused by
confounders such as binary companions or aliasing.
For example, the lower rotation sequence could be a reflection of the upper
sequence, caused by incorrect measurements at half the true period.
In this paper, we rule out the possibility that this gap could be caused by
incorrect period measurements or binary companions, by showing that the
rapidly rotating stars are dynamically young.
\end{abstract}

\section{Introduction}
\subsection{The rotation periods of field stars}
The rotation periods of middle-aged FGKM stars are mostly determined by their
mass, age and evolutionary stage, and this is due, for the most part, to
angular momentum loss through magnetic braking.
Although stars are born with a range of rotation periods, a steep dependence
of angular momentum loss, $J$ angular frequency, $\omega$, $J \propto
\omega^3$ causes stellar rotation periods to converge after a characteristic
timescale.
This timescale is on the order of a few tens of millions of years for F and G
stars, but hundreds of millions of years for K and M stars.
Once stars exceed this age, they steadily lose angular momentum via magnetized
stellar winds that remove angular momentum from the star at a radius that is
determined by the magnetic field strength.
The Rossby number, the ratio of rotation period to convective turnover
timescale, is linked to the strength of stellar magnetic dynamos.
Large-scale polodial and toroidal stellar magnetic fields are thought to be
generated by the $\alpha-\Omega$ mechanism, whereby convective motion of
plasma and differential stellar rotation cause the winding up of magnetic
field lines, generating a strong magnetic field \racomment{(citations)}.
In general, the ratio between a star's rotation period and its characteristic
timescale of convective motion (the Rossby number) determines the strength of
the magnetic field \racomment{(citation)}.
Stars that rotate quickly and have deep convection zones, therefore long
convective turnover times, have small Rossby numbers and strong magnetic
fields.
Young M dwarfs typically have the strongest magnetic fields.
Slowly rotating stars with shallow convective zones, such as old F and G
dwarfs, have weak magnetic fields \racomment{(citations)}.
Since magnetic field strength depends on stellar rotation period and mass via
the Rossby number, it follows that angular momentum loss driven by magnetized
stellar winds also depends on Rossby number.

The first large-scale stellar rotation period catalog, generated from
analyzing light curves from the \kepler\ spacecraft is the
\citet{mcquillan2014} catalog.
It contains around 34,000 rotation periods of FGKM dwarfs and subgiants,
measured by calculating autocorrelation functions (ACFs) of light curves.
This catalog revealed a gap in the rotation periods of K and M dwarfs: a strip
in the rotation vs effective temperature plane that is under-populated
compared to the surrounding parts of parameter space.
This gap can be seen in figure \ref{fig:the_gap}.
\begin{figure}
  \caption{
The rotation periods of 18,259 FGKM dwarf stars, measured by \mct, vs. their
effective temperatures.
We applied cuts to the \gaia\ color-magnitude diagram of these stars (figure
\ref{fig:CMD_cuts}) in order to remove equal-mass binaries and subgiants from
the sample.
The rotation period gap can be seen as an almost-horizontal gap toward the
right of this figure.
}
  \centering
    \includegraphics[width=1\textwidth]{the_gap}
\label{fig:the_gap}
\end{figure}
The rotation period gap is most prevalent for low-mass stars, of spectral type
K and M, but was recently shown to extend through G, F and A types as well
\citep{davenport2017}.

Three explanations for the origin of this have been proposed.
Firstly, since the gap falls along a rotational/gyrochronal isochrone, it
follows that it may be caused by a discontinuous age distribution within the
sample.
The gap could be created if there exist two populations of stars with
different age distributions, with the distribution of young ages peaking at
around 800 Myr, and the distribution of old ages peaking at around 2-3 billion
years.
This idea was investigated by \mct, who suggested that the two populations
might be the thin and thick disk of the Milky Way.
However, they found no sharp change in kinematic properties between the two
samples.
In addition, the age distributions of the thin and think disks are expected to
be around ... and ... years respectively: older than the majority of stars in
the \mct\ sample \racomment{(citations)}.

The second explanation for the rotation period gap is that stars do not lose
angular momentum uniformly over time: instead they undergo a period of rapid
spin down at an age of around 1.1 Gyr, or a Rossby number of around...
\racomment{(Calculate this!)}.
Since the rotation period gap is located along a line of constant Rossby
number \racomment{(what is this number?)}, and since the rate of stellar
angular momentum loss is known to be related to magnetic field strength (this
is the principle behind gyrochronology \citep[\eg][]{kawaler1989,
pinsonneault1989, barnes2003, barnes2007, angus2015, vansaders2016,
vansaders2018, angus2019}), it follows that the gap could be caused
by some phenomenon related to magnetic braking.

As of May 2019, the \citet{mcquillan2014} catalog has been used more than 250
times, for a range of studies spanning fields from exoplanets to white dwarfs,
to galactic archaeology.
Since it underpins so many astronomical studies, a large number of incorrectly
measured rotation periods in this catalog would undermine several of the
studies built upon it.

\subsection{Using kinematics to solve the mystery}

Why does proper motion in galactic latitude work for this problem?

\bibliographystyle{plainnat}
\bibliography{bib.bib}
\end{document}

### 2019-08-19 paper/results.tex
redrafting

--- 2019-08-03 paper/results.tex
@@
 \section{Results}
-
-\subsection{Selecting groups of coeval stars}
-
-To explore the relationship between rotation period, \teff\ and age, we
-selected groups of stars within different age ranges \citep[where age was
-calculated using the][gyrochronology relation]{angus2019}, and calculated the
-velocity dispersion: the standard deviation of velocities, \sigmavb, as a
-function of effective temperature for each age group.
-Since the processes that produce dynamical heating in the galactic disk are
-expected to generate Gaussian-distributed velocities, we performed
-sigma-clipping on stars in each age and temperature bin, to remove
-non-Gaussian outliers.
-Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-throughout this paper we show rotation periods as a function of {\it effective
-temperature}.
-We chose to use effective temperatures and not colors in this analysis as it
-is the linear quantity and therefore easier to divide into bins of roughly
-equal numbers of stars.
+\label{sec:results}
 
@@
     The full catalog, with subgiants and visual binaries removed is shown in
-    grey, and stars selected to be in different age groups (between 3500 and
+    grey, and stars selected to be in different age groups (between \tmin\ and
     5000 K) are overlayed in color.
@@
     constant as a function of \teff.
-However, the velocity dispersions of the oldest age groups increase \teff,
-    indicating the \citet{angus2019} gyrochronology model underpredicts the
-    the ages of late-K dwarfs relative to the ages of early K dwarfs at old
-    ages.
-An alternative explanation could be that the gyrochronology relation is
-    correct and {\it mass-dependent heating} is responsible for the greater
-    velocity dispersions of cooler stars.
+However, the velocity dispersions of the oldest age groups increase with
+    \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
+    the the ages of late-K dwarfs relative to the ages of early K dwarfs at
+    old ages.
+% An alternative explanation could be that the gyrochronology relation is
+%     correct and {\it mass-dependent heating} is responsible for the greater
+%     velocity dispersions of cooler stars.
 }
@@
 as a function of effective temperature.
-We only included stars within a temperature range of 5000 - 3500 K in our
-analysis, as hotter stars are more likely to have stopped magnetic braking
-\citep{vansaders2016}, which could bias the results.
+% We only included stars within a temperature range of 5000 - 3500 K in our
+% analysis, as hotter stars are more likely to have stopped magnetic braking
+% \citep{vansaders2016}, which could bias the results.
 M dwarfs were not included in our analysis because such faint stars cannot be
@@
 with gyrochronal age across all temperatures.
-The consistent velocity dispersion of young stars as a function of temperature
+The constant velocity dispersion of young stars as a function of temperature
 shows that the Praesepe-calibrated gyrochronology relation accurately predicts
@@
 % of \teff.
-However, although the youngest age groups have a relatively constant velocity
-dispersions across temperatures, the oldest age groups do not.
-This indicates {\it either} that the shape of the period-color relation does
-not remain constant over time, \ie\ it flattens out, {\it or} that cool stars
-experience more efficient dynamical heating than hot stars.
+However, although velocity dispersion is constant across temperature for young
+stars, it increases as a function of temperature for old stars.
+This indicates that the shape of the period-color relation flattens out over
+time.
 
-Mass-dependent dynamical heating could occur because lower-mass stars
-experience greater velocity changes when gravitationally perturbed.
-However, mass-dependent orbital heating has not yet been unambiguously
-observed in the galactic disk because of the strong anti-correlation between
-stellar mass and stellar age.
+An alternative explanation for this phenomenon is that lower-mass stars
+experience greater velocity changes when gravitationally perturbed and are
+more efficiently dynamically heated.
+Mass-dependent orbital heating has not been unambiguously observed in the
+galactic disk because of the strong anti-correlation between stellar mass and
+stellar age.
 Less massive stars do indeed have larger velocity dispersions, however they
-are also older on average \racomment{(citations)}.
+are also older on average.
 This mass-age degeneracy is highly reduced in M dwarfs because their
@@
 \citep{faherty2009}.
+To investigate further, we selected late K and M dwarfs observed by both
+\kepler\ and \gaia, whose MS lifetimes exceed around 11 Gyrs and are therefore
+representative of the initial mass function.
+% This allowed us to remove the degeneracy between mass and age that would
+% confound evidence for mass-dependent heating among more massive stars.
+% In a sample of massive stars with short MS lifetimes, lower-mass stars
+% typically have greater velocity dispersion, but this could be either because
+% dynamical heating is more efficient in lower mass stars, or because lower mass
+% stars tend to be older because they live longer.
+% In a sample of late K and M dwarfs, whose MS lifetime is longer than the age
+% of the Universe, we can assume that none have evolved into giants and any
+% additional heating seen in the lower-mass stars must be caused by
+% mass-dependent heating.
+We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
+greater than 1.2 (corresponding to an effective temperature $\lesssim$
+4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+We also eliminated visual binaries by removing stars above a 6th order
+polynomial, fit to the MS on the \gaia\ CMD.
+% shown in figure \ref{fig:mdwarf_CMD}.
+We then applied the same quality cuts as described in section
+\ref{sec:the_data}.
+% \begin{figure}
+%   \caption{
+% A \gaia\ color-magnitude diagram, showing all \kepler\ targets.
+% We selected K and M dwarfs by applying color and magnitude cuts at \gcolor\ =
+%     1.2 (\teff\ $\sim$ 4800 K) and $G$ = 4.
+% We eliminated visual binaries by fitting a 6th order polynomial to the MS and
+%     removing stars above it.
+% Selected K and M dwarfs are shown as darker points.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{mdwarf_CMD}
+% \label{fig:mdwarf_CMD}
+% \end{figure}
+To search for evidence of mass-dependent heating we calculated the (\vb)
+velocity dispersion of stars in effective temperature bins.
+Again, we applied sigma clipping to remove high velocity outliers before
+calculating the standard deviation of stars in each bin.
+% A slight excess of non-Gaussian outliers at lower temperatures leads to a
+% marginal increase in velocity dispersion at low temperatures when sigma
+% clipping is not performed.
+% Figure \ref{fig:vb_vs_color} shows velocity and velocity dispersion as a
+% function of \gaia\ \gcolor\ color.
+% There is no clear trend between velocity dispersion and color.
+Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
+as a function of effective temperature (calculated by transforming dereddened
+\gaia\ colors using equation \ref{eq:curtis}).
+Velocity dispersion very slightly {\it decreases} with decreasing temperature,
+(the opposite of the trend expected for mass-dependent heating) however the
+slope is only inconsistent with zero at the 1.3 $\sigma$ level.
+This trend may be due to a selection bias: cooler stars are fainter and
+therefore typically closer, with smaller heights above the galactic plane and
+smaller velocities.
+The essential point here is that the increasing velocity dispersion with age
+and decreasing temperature that is evident in figure \ref{fig:age_cut} is not
+a result of an intrinsically larger velocity dispersion for low-mass stars.
+We performed the same analysis on the 537 stars in this sample with RVs in
+{\it vertical} velocity (\vz) and found the slope of the velocity
+dispersion-temperature relation was consistent with zero.
+% \begin{figure}
+%   \caption{
+%       Top: Stellar velocity (\vb) as a function of \gaia\ \gcolor\ color for
+%       \kepler\ K and M dwarfs.
+% Vertical lines indicate different color-groupings used to calculate velocity
+%     dispersion.
+% Pink stars were not included in velocity dispersion calculations as they were
+%     removed as outliers during a sigma clipping process.
+% Bottom: velocity dispersion as a function of color.
+% There is no obvious trend between velocity dispersion and stellar color
+%     indicating that mass-dependent heating does not significantly effect
+%     low-mass stars in the \kepler\ field.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{vb_vs_color}
+% \label{fig:vb_vs_color}
+% \end{figure}
 
-The \vb\ AVR is not directly comparable to the \vz\ AVR, so, to draw a
-comparison with results from the literature, we calculated a \vz\ AVR for
-the subset of 290 stars in our sample with \gaia\ RVs.
-We measured an AVR exponent of 0.47 $\pm$ 0.1, which falls within the range of
-values (0.45-0.53) reported from measurements of F and G stars in the Solar
-neighborhood \citep{holmberg2009, aumer2009, aumer2016}.
-It is a little lower than the values of 0.56 $\pm$ 0.14 (for low-z) and 0.51
-$\pm$ 0.15 (for high-z stars), calculated using \LAMOST\ \racomment{(LAMOST
-citation)} K giants \citep{yu2018}.
-K giants are more massive than K dwarfs \racomment{(how much more massive?)},
-so this slightly higher value contradicts the mass-dependent heating
-hypothesis.
-\racomment{Add some words about selection functions...}
-
-To investigate further, we calculated the exponent of the (\vb) AVR for each
-temperature bin in figure \ref{fig:age_cut}.
 \begin{figure}
   \caption{
-      The exponent of the (\vb) AVR as a function of effective temperature.
+      Top: Stellar velocity (\vb) as a function of \teff\ for
+      \kepler\ K and M dwarfs.
+Vertical lines indicate different \teff-groupings used to calculate velocity
+    dispersion.
+Pink stars were not included in velocity dispersion calculations as they were
+    either removed as outliers during a sigma clipping process, or they lie at
+    the sparcely populated, extremely cool end of the temperature range.
+    Velocity dispersion and \teff\ are slightly correlated,
+    indicating that mass-dependent heating does not significantly affect
+    low-mass stars in the \kepler\ field.
 }
   \centering
-    \includegraphics[width=1\textwidth]{AVR_exponent}
-\label{fig:AVR_exponent}
+    \includegraphics[width=1\textwidth]{vb_vs_teff}
+\label{fig:vb_vs_teff}
 \end{figure}
-Figure \ref{fig:AVR_exponent} shows the significant rise the in AVR exponent
-as a function of effective temperature.
-The hottest stars in our sample (\teff\ = 4833-5000 K) have a (\vb) AVR
-exponent of 0.39 $\pm$ 0.02 and the coolest (\teff\ = 3500-3667 K) have an
-exponent of 0.74 $\pm$ 0.02.
-The masses of stars in our sample differ by less than a factor of two,
-so mass-dependent heating cannot be entirely responsible for this large spread
-in heating rates.
-\racomment{Need to look into this further.}
+Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+strongly affect the \mct\ sample of rotating \kepler\ stars.
+For this reason, we assume that age difference is the only significant cause
+of velocity dispersion differences between groups.
+In other words, (\vb) velocity dispersion is a reliable age proxy for the
+\mct\ sample.
+
+% The \vb\ AVR is not directly comparable to the \vz\ AVR, so, to draw a
+% comparison with results from the literature, we calculated a \vz\ AVR for
+% the subset of 290 cool stars in the \mct\ sample with \gaia\ RVs.
+% We measured an AVR exponent of 0.47 $\pm$ 0.1, which falls within the range of
+% values (0.45-0.53) reported from measurements of F and G stars in the Solar
+% neighborhood \citep{holmberg2009, aumer2009, aumer2016}.
+% It is a little lower than the values of 0.56 $\pm$ 0.14 (for low-z) and 0.51
+% $\pm$ 0.15 (for high-z stars), calculated using \LAMOST\ \racomment{(LAMOST
+% citation)} K giants \citep{yu2018}.
+% K giants are more massive than K dwarfs \racomment{(how much more massive?)},
+% so this slightly higher value contradicts the mass-dependent heating
+% hypothesis.
+% \racomment{Add some words about selection functions...}
+
+% To investigate further, we calculated the exponent of the (\vb) AVR for each
+% temperature bin in figure \ref{fig:age_cut}.
+% \begin{figure}
+%   \caption{
+%       The exponent of the (\vb) AVR as a function of effective temperature.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{AVR_exponent}
+% \label{fig:AVR_exponent}
+% \end{figure}
+% Figure \ref{fig:AVR_exponent} shows the significant rise the in AVR exponent
+% as a function of effective temperature.
+% The hottest stars in our sample (\teff\ = 4833-5000 K) have a (\vb) AVR
+% exponent of 0.39 $\pm$ 0.02 and the coolest (\teff\ = 3500-3667 K) have an
+% exponent of 0.74 $\pm$ 0.02.
+% The masses of stars in our sample differ by less than a factor of two,
+% so mass-dependent heating cannot be entirely responsible for this large spread
+% in heating rates.
+% \racomment{Need to look into this further.}
 
@@
 
-If we assume that the rise in velocity dispersions at cooler effective
-temperatures is caused by an inaccurate period-color relation rather than
-mass-dependent dynamical heating, we can estimate what the shape of the
-period-color relations {\it should} be.
-Figure \ref{fig:age_cut} indicates that the period-color relation flattens
-out, so we applied the same analysis to groups of stars with similar rotation
-periods, equivalent to a completely flat period-color relation.
-The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
-in different period ranges plotted in different colors.
-The bottom panel shows the velocity dispersion of each group as a function of
-effective temperature.
+% If we assume that the rise in velocity dispersions at cooler effective
+% temperatures is caused by an inaccurate period-color relation rather than
+% mass-dependent dynamical heating, we can estimate what the shape of the
+% period-color relations {\it should} be.
+% Figure \ref{fig:age_cut} indicates that the period-color relation flattens
+% out, so we applied the same analysis to groups of stars with similar rotation
+% periods, equivalent to a completely flat period-color relation.
+% The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
+% in different period ranges plotted in different colors.
+% The bottom panel shows the velocity dispersion of each group as a function of
+% effective temperature.
+% \begin{figure}
+%   \caption{
+% This figure is similar to figure \ref{fig:age_cut}, with stars divided into
+%     period groups rather than age groups.
+% The velocity dispersion is more constant across effective temperatures for the
+%     most slowly rotating stars, compared to the stars selected with the
+%     \citet{angus2019} gyrochronology model, indicating that the gyrochronology
+%     models flatten out, and possibly even invert, at old ages.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{period_cut}
+% \label{fig:period_cut}
+% \end{figure}
+% Once again, the velocity dispersion increases with rotation period overall.
+% For the most rapidly rotating groups of stars, velocity dispersion decreases
+% with \teff\ as expected given the positively sloped period-color relation of
+% Praesepe and other young clusters: late K dwarfs rotate more slowly than early
+% K dwarfs of the same age.
+% Between rotation periods of 15 and 25 days, the temperature dependence of the
+% velocity dispersion starts to disappear, indicating that the period-color
+% relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
+% of the same age.
+% At long rotation periods, the velocity dispersion still increases with
+% effective temperature, although the increase is much more modest compared to
+% the oldest stars in figure \ref{fig:age_cut}.
+
+% If we assume the rise in velocity dispersion at cooler effective temperatures
+% is caused by an inaccurate period-color relation rather than mass-dependent
+% dynamical heating, we can generate a graphical representation of what the
+% relations {\it should} look like.
 \begin{figure}
   \caption{
-This figure is similar to figure \ref{fig:age_cut}, with stars divided into
-    period groups rather than age groups.
-The velocity dispersion is more constant across effective temperatures for the
-    most slowly rotating stars, compared to the stars selected with the
-    \citet{angus2019} gyrochronology model, indicating that the gyrochronology
-    models flatten out, and possibly even invert, at old ages.
+Rotation period vs effective temperature for stars in the \mct\ sample,
+    colored by the velocity dispersions of stars in restricted period and
+    temperature ranges.
 }
   \centering
-    \includegraphics[width=1\textwidth]{period_cut}
-\label{fig:period_cut}
+    \includegraphics[width=1\textwidth]{dispersion_period_teff}
+\label{fig:dispersion_period_teff}
 \end{figure}
-Once again, the velocity dispersion increases with rotation period overall.
-For the most rapidly rotating groups of stars, velocity dispersion decreases
-with \teff\ as expected given the positively sloped period-color relation of
-Praesepe and other young clusters: late K dwarfs rotate more slowly than early
-K dwarfs of the same age.
-Between rotation periods of 15 and 25 days, the temperature dependence of the
-velocity dispersion starts to disappear, indicating that the period-color
-relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
-of the same age.
-At long rotation periods, the velocity dispersion still increases with
-effective temperature, although the increase is much more modest compared to
-the oldest stars in figure \ref{fig:age_cut}.
-This suggests that rotation period may {\it decrease} with effective
-temperature, \ie\ late K dwarfs may rotate {\it more} rapidly than early K
-dwarfs.
-This would be a paradigm shift for gyrochronology, stellar spin-down rate is
-thought to be directly tied to magnetic field strength.
-The deeper convection zones of later type stars generate stronger magnetic
-fields which {\it should} lead to more efficient angular momentum loss.
-Unfortunately, we still cannot rule out the possibility that mass-dependent
-heating is acting on these stars which could be responsible for some, if not
-all, of the increased velocity dispersion at cooler effective temperatures.
+Figure \ref{fig:dispersion_period_teff} shows the stars in our sample of
+rotating dwarfs on a
+plot of period vs effective temperature, coloured by their (\vb) velocity
+dispersion.
+Here, velocity dispersion was calculated for groups of stars in
+uniformly-spaced bins in $\log_{10}$(period) and temperature.
+If we assume that mass dependent heating doesn't affect this sample, and \vb\
+at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
+dispersion can be interpreted as an age proxy and stars plotted in a similar
+color in figure \ref{fig:dispersion_period_teff} are similar ages.
+Interpreted this way, it appears as though rotation period {\it decreases}
+with decreasing effective temperature at old ages.  This is a paradigm shift
+for gyrochronology because stellar spin-down rate is thought to be directly
+tied to magnetic field strength, and the deeper convection zones of cooler
+stars generate stronger magnetic fields which {\it should} lead to more
+efficient angular momentum loss.
+
+Something like this has been observed in the 1.1 Gyr NGC 6811 cluster, where
+the rotation periods of mid-K dwarfs are faster than expected; their
+rotational evolution appears to have stalled.
+NGC 6811 straddles the rotation period gap and may be the `missing link' of
+gyrochronology that connects the two periods of stellar spin-down: the early
+stage where the period-color relation for K dwarfs has a positive slope and
+the late stage where it has a negative slope.
+In this case the period gap may delineate the transition between these two
+regimes and is the point at which stellar magnetic dynamos likely undergo a
+dramatic structural shift.
+
+Figure \ref{fig:dispersion_period_teff} also indicates that the oldest stars
+with measured rotation periods are cool, which may be because cooler stars
+remain active for longer, and their rotation periods can therefore be measured
+at older ages.
+
+% Unfortunately, we still cannot rule out the possibility that mass-dependent
+% heating is acting on these stars which could be responsible for some, if not
+% all, of the increased velocity dispersion at cooler effective temperatures.

### 2019-08-20 paper/conclusion.tex
adding period gap section and some rewriting

@@
 
+We examined the gyrochronology relations using the velocity dispersions of
+field stars with measured rotation periods.
 We found that old groups of K dwarfs selected to be coeval using the
@@
 dispersion across all temperatures.
+It appears that the period-\teff\ relation changes shape over time: rotation
+period is negatively correlated with \teff\ at young ages, and positively
+correlated at old ages.
+The relation appears to start flattening out after $\sim$1 Gyr (see figure
+\ref{fig:age_cut}), potentially around the same age, or just older than the
+period gap which is located at a (gyro) age of around $\sim$1.1 Gyr.
+This gap may represent a significant transitional epoch in the magnetic
+behavior of stars.
+Finally, when velocity dispersion is interpreted as an age proxy, it appears
+that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
+suggests that lower-mass stars remain active for longer and their rotation
+periods can be measured at older ages.
 % This indicates that either the \citep{angus2019} period-color relation is
@@
 
-\racomment{Words about the significance of an inverted period-color relation.}
+% \racomment{Words about the significance of an inverted period-color relation.}
 
-\racomment{How does the weakened van Saders braking law fit into all this?}
+% \racomment{How does the weakened van Saders braking law fit into all this?}
 
-\racomment{Talk about the fact that the stars in each bin are not necessarily
-the same age.}
+% \racomment{Talk about the fact that the stars in each bin are not necessarily
+% the same age.}
 
-\racomment{Dispersion could be dependent on the selection function.
-We cut stars at high b, because we find that these do lead to a bias.
-We find no evidence for increased velocity dispersion with decreasing teff in
-either vb or vz.
-We find that sigma clipping is necessary for vb but not vz and do not find
-that the outliers are distributed towards higher galactic latitudes.
-}
+% \racomment{Dispersion could be dependent on the selection function.
+% We cut stars at high b, because we find that these do lead to a bias.
+% We find no evidence for increased velocity dispersion with decreasing teff in
+% either vb or vz.
+% We find that sigma clipping is necessary for vb but not vz and do not find
+% that the outliers are distributed towards higher galactic latitudes.
+% }
 

### 2019-08-20 paper/method.tex
adding period gap section and some rewriting

@@
 relationship between \vb\ and \vz\ and keeping a large enough sample size to
-provide meaningful results.
+provide meaningful results as we found that reducing this to 10\degrees did
+not significantly affect our results.
 We also found that the anti-correlation between stellar mass and $b$ is
@@
 dwarfs.
-
-To explore the relationship between rotation period, \teff\ and age, we
-selected groups of stars within different age ranges \citep[where age was
-calculated using the][gyrochronology relation]{angus2019}, and calculated the
-velocity dispersion: the standard deviation of velocities, \sigmavb, as a
-function of effective temperature for each age group.
-We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
-and temperature bin to remove non-Gaussian outliers.
-Without sigma-clipping, we found that a small number of high velocity
-outliers at the low-temperature end of our sample substantially raised the
-velocity dispersion for cooler stars.
-Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-throughout this paper we show rotation periods as a function of {\it effective
-temperature}.
-We chose to use effective temperatures and not colors in this analysis as it
-is the linear quantity and therefore easier to divide into bins of roughly
-equal numbers of stars.
 

### 2019-08-20 paper/results.tex
adding period gap section and some rewriting

@@
 \label{sec:results}
+
+\subsection{Velocity dispersion of coeval groups}
+
+To explore the relationship between rotation period, \teff\ and age, we
+selected groups of stars within different age ranges \citep[where age was
+calculated using the][gyrochronology relation]{angus2019}, and calculated the
+velocity dispersion: the standard deviation of velocities, \sigmavb, as a
+function of effective temperature for each age group.
+We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
+and temperature bin to remove non-Gaussian outliers.
+Without sigma-clipping, we found that a small number of high velocity
+outliers at the low-temperature end of our sample substantially raised the
+velocity dispersion for cooler stars.
+Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+throughout this paper we show rotation periods as a function of {\it effective
+temperature}.
+We chose to use effective temperatures and not colors in this analysis as it
+is the linear quantity and therefore easier to divide into bins of roughly
+equal numbers of stars.
 
@@
 time.
+
+\subsection{Mass-dependent heating}
 
@@
 \mct\ sample.
+
+\subsection{The period gap}
+An unsolved mystery within gyrochronology is the origin of the rotation period
+gap, first identified by \citet{mcquillan2013}.
+This gap can be seen in figure \ref{fig:age_cut} as an under-density of points
+between the 0.7-1.0 and 1.0-1.5 Gyr age ranges and roughly follows a line of
+constant gyrochronal age of around 1.1 Gyr \citep[according to
+the gyrochronology relation of][]{angus2019}.
+Several explanations for its origin have been proposed, including a
+discontinuous star formation history \citep{mcquillan2013, davenport2017,
+davenport2018}, a rapid change in magnetic field structure
+\citep{reinhold2019}, and erroneous rotation period measurements that are
+incorrect by a factor of two \citep{koen2018}.
+The latter explanation can be ruled out because stars below the gap have
+smaller velocity dispersions than the stars above the gap, indicating that
+they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+evident in figure \ref{fig:age_cut}.
+For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+velocity dispersion increases with \teff.
+The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+dispersion as the hottest stars in the age range above which indicates that
+the period-\teff\ relations are {\it flat} at these rotation periods.
+It may be a coincidence that the gyrochronology relations seem to only flatten
+off {\it above} the period gap, or it may be that the period gap is located at
+magnetic transition period.
+We lack a sufficient quantity of data to investigate in detail but new field
+star rotation periods from the \ktwo\ and \tess\ missions may be able to
+validate or rule out this hypothesis in the future.
 
@@
 % relations {\it should} look like.
+
+\subsection{The period-\teff\ relations, revealed}
+
 \begin{figure}
@@
 \end{figure}
-Figure \ref{fig:dispersion_period_teff} shows the stars in our sample of
-rotating dwarfs on a
-plot of period vs effective temperature, coloured by their (\vb) velocity
-dispersion.
+Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
+our sample, coloured by (\vb) velocity dispersion.
 Here, velocity dispersion was calculated for groups of stars in
@@
 color in figure \ref{fig:dispersion_period_teff} are similar ages.
-Interpreted this way, it appears as though rotation period {\it decreases}
-with decreasing effective temperature at old ages.  This is a paradigm shift
+Interpreted this way, it appears that rotation period {\it decreases} with
+decreasing effective temperature at old ages.  This is a paradigm shift
 for gyrochronology because stellar spin-down rate is thought to be directly
@@
 efficient angular momentum loss.
-
-Something like this has been observed in the 1.1 Gyr NGC 6811 cluster, where
+A similar phenomenon has been observed in the 1.1 Gyr NGC 6811 cluster, where
 the rotation periods of mid-K dwarfs are faster than expected; their
-rotational evolution appears to have stalled.
-NGC 6811 straddles the rotation period gap and may be the `missing link' of
-gyrochronology that connects the two periods of stellar spin-down: the early
-stage where the period-color relation for K dwarfs has a positive slope and
-the late stage where it has a negative slope.
+rotational evolution appears to have stalled, and the period-\teff\ relation
+is flat \citep{curtis2019}.
+NGC 6811 straddles the rotation period gap and may be the `missing link' that
+connects two epoch of stellar spin-down: an early stage where the
+period-\teff\ relation for K dwarfs has a negative slope and a late stage
+where it has a positive slope.
 In this case the period gap may delineate the transition between these two
@@
 
-Figure \ref{fig:dispersion_period_teff} also indicates that the oldest stars
-with measured rotation periods are cool, which may be because cooler stars
-remain active for longer, and their rotation periods can therefore be measured
-at older ages.
+Figure \ref{fig:dispersion_period_teff} also shows that the (kinematically)
+oldest stars with measured rotation periods are cooler than 4500 K.
+This may be because cooler stars remain active for longer, and their rotation
+periods can therefore be measured at older ages.
 

### 2019-08-29 calc_velocities.py
trying to calculate v_l

@@ -35,5 +35,54 @@ def calc_vb(pandas_df, nsamples=1000):
     pm_b = np.array(vels.pm_b.value)
 
     gal = c_samples.galactic
-    v_b = (gal.pm_b * gal.distance).to(u.km/u.s, u.dimensionless_angles())
+    v_b = (gal.pm_l * gal.distance).to(u.km/u.s, u.dimensionless_angles())
     return pm_b, np.array(v_b.value)
+
+
+def calc_vl(pandas_df, nsamples=1000):
+    """
+    Calculate velocity samples in the l direction from a pandas DataFrame.
+
+    Args:
+        df (DataFrame): pandas dataframe containing Gaia columns
+
+    Returns:
+        pm_l (array): samples of velocity in the l direction.
+            Shape = nstars x nsamples
+
+    """
+
+    df = Table.from_pandas(pandas_df)
+    g = GaiaData(df)
+    g_samples = g.get_error_samples(size=nsamples,
+                                    rnd=np.random.RandomState(seed=42))
+    c_samples = g_samples.get_skycoord()
+    vels = c_samples.transform_to(coord.Galactic)
+    # pm_l = np.array(vels.pm_l.value)
+
+    gal = c_samples.galactic
+    v_l = (gal.pm_l * gal.distance).to(u.km/u.s, u.dimensionless_angles())
+    return np.array(v_l.value)
+
+
+def calc_vz(pandas_df, nsamples=1000):
+    """
+    Calculate W samples from a pandas DataFrame.
+
+    Args:
+        df (DataFrame): pandas dataframe containing Gaia columns
+
+    Returns:
+        W (array): samples of W velocity.
+            Shape = nstars x nsamples
+
+    """
+
+    df = Table.from_pandas(pandas_df)
+    g = GaiaData(df)
+    g_samples = g.get_error_samples(size=nsamples,
+                                    rnd=np.random.RandomState(seed=42))
+    c_samples = g_samples.get_skycoord()
+    vels = c_samples.transform_to(coord.Galactocentric)
+    W = vels.v_z.value
+    return np.mean(W, axis=1), np.std(W, axis=1)

### 2019-08-29 paper/intro.tex
some fairly small edits and rewrites. adding some references

@@
 time \citep{skumanich1972}.
-This characteristic of main-sequence (MS) stars allows them to be dated via
-their rotation periods in a practice known as gyrochronology, which is useful
-since the ages of MS stars are extremely difficult to measure via the
-traditional age-dating method of isochrone placement.
+This characteristic of main-sequence (MS) stars allows them to be dated using
+their rotation periods, via gyrochronology \citep[\eg][]{kawaler1989,
+pinsonneault1989, barnes2003, barnes2007, barnes2010, meibom2011, meibom2015},
+which is particularly useful since the ages of MS stars are difficult to
+measure via isochrone placement.
 It is also well established that stars of the same age but different masses
-have different rotation periods \racomment{(citation)}, thought to be caused
-by the deeper convective zones, and therefore stronger magnetic dynamos (and
-more efficient magnetic braking) in lower-mass stars.
+have different rotation periods \citep[\eg][]{kraft1967, matt2012}, thought to
+be caused by the deeper convective zones, and therefore stronger magnetic
+dynamos (and more efficient magnetic braking) in lower-mass stars.
 However, an underlying assumption behind many empirical gyrochronology
 relations is that the relationships between rotation period and photometric
-color\footnote{As the directly observable quantity, color is usually used as a
-mass-proxy, and empirical gyrochronology relations are usually calibrated in
-color, rather than mass or effective temperature.
-}, and rotation period and age are {\it separable}, meaning that the
+color\footnote{As the directly observable quantity, color is often used as a
+mass proxy, and empirical gyrochronology relations are usually calibrated in
+color, rather than mass or effective temperature.},
+and rotation period and age are {\it separable}, meaning that the
 period-color relation is the same at all ages and the period-age relation is
@@
 meibom2011, angus2015, angus2019}.
+Not all gyrochronology models follow this form, particularly those motivated
+based on stellar structure and evolution theory or with a Rossby number,
+rather than rotation period dependence \citep[\eg][]{barnes2010, matt2012,
+vansaders2016]}.
 It was recently shown that old field stars rotate more rapidly than a simple,
@@
 vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
-modification to the classical \citet{skumanich1972} spin-down law is required
-to reproduce the data \citep{vansaders2016, vansaders2018}.
+modification to the classical \prot\ $\propto t^{\frac{1}{2}}$ spin-down law
+\citep{skumanich1972} is required to reproduce the data \citep{vansaders2016,
+vansaders2018}.
 An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
 further indication that the exponent of the period-age relation is
-mass and time-dependent.
-This cluster has a flattened relationship between rotation period and color:
-the G dwarfs rotate at the same rate as the K dwarfs \citep{curtis2019}.
+mass and time-dependent \citep{curtis2019}.
+This cluster has a flattened relationship between rotation period and color
+and the G dwarfs rotate at the same rate as the K dwarfs.
 % Since a separable period-age and period-color relation provided an adequate
@@
 
-In this paper, we tested the \citet{angus2019} gyrochronology relation, which
-was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
-color) and the period-age relation of Praesepe and the Sun.
+In this paper, we tested the \citet{angus2019} gyrochronology relation, a
+separable relation, calibrated using the period-color relation of Praesepe (in
+\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
 The large number of Praesepe members with precise rotation periods from the
@@
 1.1 Gyr NGC 6811 cluster.
-No reliable age estimates exist for K dwarfs with rotation periods older than
-1.1 Gyr, so the rotational evolution of these stars beyond this age is a
-mystery.
+Very few reliable age estimates exist for K dwarfs with rotation periods older
+than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
+evolution of middle-aged K dwarfs is a mystery\footnote{The oldest M dwarfs
+with reliable ages and rotation periods are members of Praesepe ($\sim$ 650
+Myrs), so even less is known about their rotational evolution.}.
 % In this paper, we shed light on the rotational evolution of old K dwarfs using
@@
 For this reason, we used a population-based stellar age indicator, velocity
-dispersion, to investigate the period-color relations at old ages.
+dispersion, to investigate the period-color relations at old ages in the
+field.
 
@@
 the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
-or $v_z$).
+or $v_z$), just like the star-forming molecular gas observed in the disk today
+\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
 % because the MW's gas disk is thin and its gas moves in circular
@@
 Young stars have relatively small vertical velocities, but gain momentum in
-the vertical direction over time.
+the vertical direction over time \citep[\eg][]{nordstrom2004, holmberg2007,
+holmberg2009, aumer2009, casagrande2011}.
 Although the cause of orbital heating is not well understood, interactions
@@
 % west2011}.
-Since the age-velocity dispersion relations (AVRs) are themselves still
-actively being calibrated, it is difficult to directly compare gyrochronal
-ages with kinematic ones.
+The age-velocity dispersion relations (AVRs) are still actively being
+calibrated, so it is difficult to directly compare gyrochronal ages with
+kinematic ones.
 However, regardless of the exact relation between velocity dispersion and
@@
 Vertical action is a better age indicator than vertical velocity
-\citep[although still only a weak age indicator for individual
-stars,][]{beane2018}, however both vertical action ($J_z$) and vertical velocity
-(\vz/W) can only be calculated with full 6-dimensional position and velocity
-information.
+\citep{ting2019}, although still only a weak age indicator for individual
+stars \citep{beane2018}, however both vertical action ($J_z$) and vertical
+velocity (\vz/W) can only be calculated with full 6-dimensional position and
+velocity information.
 Unfortunately, most stars with measured rotation periods do not have radial
@@
 The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
-\vb\ is a close (although not perfect -- see section \ref{sec:results})
+\vb\ is a close (although imperfect -- see section \ref{sec:results})
 approximation to \vz.

### 2019-08-29 paper/method.tex
some fairly small edits and rewrites. adding some references

@@
 stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
-DR2 catalog.
+DR2 catalog of parallaxes, proper motions and apparent magnitudes.
 Reddening and extinction from dust was calculated for each star using the
@@
 % of subgiants drive rapid spin-down through conservation of angular momentum.
-We removed visual binaries and subgiants from the sample by applying cuts to
+Visual binaries and subgiants were removed from the sample by applying cuts to
 the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
-We fit a 6th-order polynomial to the main sequence and raised it by 0.27 dex
-to approximate the division between single stars and visual binaries and
-removed all stars above this line from the sample.
-We also removed subgiants by eliminating stars brighter than 6th magnitude in
-\gaia\ G-band.
+A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
+approximate the division between single stars and visual binaries.
+All stars above this line were removed from the sample.
+Subgiants were also removed by eliminating stars brighter than 6th magnitude
+in \gaia\ G-band.
 \begin{figure}
@@
 astropy2018} {\it Python} packages were used to calculate stellar velocities.
-{\tt Pyia} has built-in functionality for calculating velocity samples from
-the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
+{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
+covariance matrix via Monte Carlo sampling.
 It therefore not only incorporates uncertainties on the \gaia\ positions
@@
 between these properties.
-We removed stars with negative parallaxes, parallax signal-to-noise ratios
-less than 10, stars fainter than 16th magnitude, stars with absolute \vb\
-uncertainties greater than 1 \kms\, and stars with galactic latitudes greater
-than 15\degrees\ from the sample.
+Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
+stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
+greater than 1 \kms\, and stars with galactic latitudes greater than
+15\degrees\ were removed from the sample.
 Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
 and because latitude is correlated with stellar mass (lower mass stars are
-older and tend to be preferentially located at high $b$), we found that
-including stars with $b > 15$\degrees\ led to a larger number of \vb\ outliers
-at low stellar mass.
+older and tend to be preferentially located at high $b$, although this trends
+is greatly reduced for stars cooler than 5000 K), we found that excluding
+stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
+stellar mass.
 15\degrees\ was found to be an adequate compromise between maintaining a close
-relationship between \vb\ and \vz\ and keeping a large enough sample size to
-provide meaningful results as we found that reducing this to 10\degrees did
-not significantly affect our results.
-We also found that the anti-correlation between stellar mass and $b$ is
-negligable for stars cooler than 5000 K.
-For this reason, and because the rotational evolution of old, G stars does not
-follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
-restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
-dwarfs.
+relationship between \vb\ and \vz\ and retaining a sample large enough to
+provide meaningful results.
+Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
+trends in the data.
+% We also found that the anti-correlation between stellar mass and $b$ is
+% negligable for stars cooler than 5000 K.
+% For this reason, and because the rotational evolution of old, G stars does not
+% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
+% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
+% dwarfs.
 

### 2019-08-29 paper/results.tex
some fairly small edits and rewrites. adding some references

@@
 calculated using the][gyrochronology relation]{angus2019}, and calculated the
-velocity dispersion: the standard deviation of velocities, \sigmavb, as a
-function of effective temperature for each age group.
+velocity dispersion: the standard deviation of velocities in the direction of
+increasing galactic latitude, \sigmavb, as a function of effective temperature
+for each age group.
 We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
@@
 Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-throughout this paper we show rotation periods as a function of {\it effective
-temperature}.
-We chose to use effective temperatures and not colors in this analysis as it
-is the linear quantity and therefore easier to divide into bins of roughly
-equal numbers of stars.
+throughout this paper we show rotation periods as a function of effective
+temperature, \teff.
+We chose to use \teff and not photometric color in this analysis as it is the
+linear quantity and therefore easier to divide into bins of roughly equal
+numbers of stars.
 
@@
     grey, and stars selected to be in different age groups (between \tmin\ and
-    5000 K) are overlayed in color.
+    \tmax\ K) are overlayed in color.
 These age groups were selected using the \citet{angus2019} gyrochronology
@@
 % \citep{vansaders2016}, which could bias the results.
-M dwarfs were not included in our analysis because such faint stars cannot be
-observed at large heights above the plane (because of the low galactic
-latitude of the \kepler\ field, stars at high-Z are more distant), which
-introduces a mass-dependent velocity bias: cooler populations of stars are
-skewed towards lower velocity dispersions.
-The coolest temperature bins in figure \ref{fig:age_cut} have low velocity
-dispersions, indicating that this effect may already become important at
-temperatures lower than $\sim$ 4000 K.
+Late M dwarfs were not included in our analysis because such faint stars
+cannot be observed at large heights above the plane (because of the low
+galactic latitude of the \kepler\ field, stars at high-Z are more distant),
+which introduces a mass-dependent velocity bias: cooler populations of stars
+are skewed towards lower velocity dispersions.
+The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
+have low velocity dispersions, indicating that this effect may already become
+important at temperatures lower than $\sim$ 4000 K.
 
 Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
-with gyrochronal age across all temperatures.
+with gyrochronal age across all temperatures, implying that both velocity
+dispersion and rotation periods increase with age as expected.
 The constant velocity dispersion of young stars as a function of temperature
@@
 % of \teff.
-However, although velocity dispersion is constant across temperature for young
-stars, it increases as a function of temperature for old stars.
-This indicates that the shape of the period-color relation flattens out over
-time.
+In contrast however, velocity dispersion increases as a function of
+temperature for old stars, meaning the \citet{angus2019} gyrochronology
+relation underpredicts the ages of old, late-K dwarfs (or overpredicts the
+ages of old early-K dwarfs).
+This indicates that the relationship between period and photometric
+color or \teff\ flattens out over time.
+In other words, the lines of constant age sweeping diagonally upwards in the
+top panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
 
@@
 and decreasing temperature that is evident in figure \ref{fig:age_cut} is not
-a result of an intrinsically larger velocity dispersion for low-mass stars.
-We performed the same analysis on the 537 stars in this sample with RVs in
+a result of an intrinsically larger velocity dispersion for low-mass stars,
+nor is it a result of the selection function.
+Since we use \vb, not \vz\ in this analysis, it is possible that the
+anti-correlation between \teff\ and galactic latitude, $b$, could influence
+these results: stars at higher latitudes would have additional velocity
+components in the ${\bf x}$ and ${\bf y}$ directions, which could increase
+\vb\ but not \vz.
+However, since the relationship between \sigmavb\ and \teff\ is positively,
+not negatively correlated for cool stars in the \kepler\ field, this effect is
+probably too small to influence our results.
+We performed the same analysis on the 537 stars in this sample with RVs using
 {\it vertical} velocity (\vz) and found the slope of the velocity
@@
     the sparcely populated, extremely cool end of the temperature range.
-    Velocity dispersion and \teff\ are slightly correlated,
-    indicating that mass-dependent heating does not significantly affect
-    low-mass stars in the \kepler\ field.
+    Velocity dispersion and \teff\ are slightly positively correlated, likely
+    due to a brightness-related selection bias, indicating that mass-dependent
+    heating does not significantly affect low-mass stars in the \kepler\
+    field.
 }
@@
 \mct\ sample.
-
-\subsection{The period gap}
-An unsolved mystery within gyrochronology is the origin of the rotation period
-gap, first identified by \citet{mcquillan2013}.
-This gap can be seen in figure \ref{fig:age_cut} as an under-density of points
-between the 0.7-1.0 and 1.0-1.5 Gyr age ranges and roughly follows a line of
-constant gyrochronal age of around 1.1 Gyr \citep[according to
-the gyrochronology relation of][]{angus2019}.
-Several explanations for its origin have been proposed, including a
-discontinuous star formation history \citep{mcquillan2013, davenport2017,
-davenport2018}, a rapid change in magnetic field structure
-\citep{reinhold2019}, and erroneous rotation period measurements that are
-incorrect by a factor of two \citep{koen2018}.
-The latter explanation can be ruled out because stars below the gap have
-smaller velocity dispersions than the stars above the gap, indicating that
-they are kinematically younger \citep{mcquillan2013, davenport2018}, as
-evident in figure \ref{fig:age_cut}.
-For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
-\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
-temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
-velocity dispersion increases with \teff.
-The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
-dispersion as the hottest stars in the age range above which indicates that
-the period-\teff\ relations are {\it flat} at these rotation periods.
-It may be a coincidence that the gyrochronology relations seem to only flatten
-off {\it above} the period gap, or it may be that the period gap is located at
-magnetic transition period.
-We lack a sufficient quantity of data to investigate in detail but new field
-star rotation periods from the \ktwo\ and \tess\ missions may be able to
-validate or rule out this hypothesis in the future.
 
@@
   \centering
-    \includegraphics[width=1\textwidth]{dispersion_period_teff}
+    \includegraphics[width=1\textwidth]{vplot}
 \label{fig:dispersion_period_teff}
@@
 % all, of the increased velocity dispersion at cooler effective temperatures.
+
+\subsection{The period gap}
+An unsolved mystery within gyrochronology is the origin of the rotation period
+gap, first identified by \citet{mcquillan2013}.
+This gap can be seen in figure \ref{fig:age_cut} as an under-density of points
+between the 0.7-1.0 and 1.0-1.5 Gyr age ranges and roughly follows a line of
+constant gyrochronal age of around 1.1 Gyr \citep[according to
+the gyrochronology relation of][]{angus2019}.
+Several explanations for its origin have been proposed, including a
+discontinuous star formation history \citep{mcquillan2013, davenport2017,
+davenport2018}, a rapid change in magnetic field structure
+\citep{reinhold2019}, and erroneous rotation period measurements that are
+incorrect by a factor of two \citep{koen2018}.
+The latter explanation can be ruled out because stars below the gap have
+smaller velocity dispersions than the stars above the gap, indicating that
+they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+evident in figure \ref{fig:age_cut}.
+For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+velocity dispersion increases with \teff.
+The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+dispersion as the hottest stars in the age range above which indicates that
+the period-\teff\ relations are {\it flat} at these rotation periods.
+It may be a coincidence that the gyrochronology relations seem to only flatten
+off {\it above} the period gap, or it may be that the period gap is located at
+magnetic transition period.
+We lack a sufficient quantity of data to investigate in detail but new field
+star rotation periods from the \ktwo\ and \tess\ missions may be able to
+validate or rule out this hypothesis in the future.

### 2019-08-29 data.py
photometric teffs as a function

@@ -1,0 +1,191 @@
+# Converting Exploring data into a script.
+
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from astropy.io import fits
+import astropy.utils as au
+from astropy.coordinates import SkyCoord
+from dustmaps.bayestar import BayestarQuery
+import astropy.units as units
+from tools import getDust
+from stardate.lhf import age_model
+from calc_velocities import calc_vb, calc_vz, calc_vl
+import astropy.units as u
+from astropy.coordinates import ICRS
+from astropy.coordinates import Galactic
+from astropy.table import Table
+from pyia import GaiaData
+import astropy.coordinates as coord
+from photometric_teff import bprp_to_teff
+
+plotpar = {'axes.labelsize': 30,
+           'font.size': 30,
+           'legend.fontsize': 15,
+           'xtick.labelsize': 30,
+           'ytick.labelsize': 30,
+           'text.usetex': True}
+plt.rcParams.update(plotpar)
+
+
+print("Load McQuillan data")
+mc = pd.read_csv("Table_1_Periodic.txt")
+
+print("Loading Gaia catalog")
+with fits.open("kepler_dr2_1arcsec.fits") as data:
+    gaia = pd.DataFrame(data[1].data, dtype="float64")
+
+gaia_mc = pd.merge(mc, gaia, on="kepid", how="left")
+
+# S/N cuts
+sn = gaia_mc.parallax.values/gaia_mc.parallax_error.values
+
+m = (sn > 10)
+m &= (gaia_mc.parallax.values > 0) * np.isfinite(gaia_mc.parallax.values)
+m &= gaia_mc.astrometric_excess_noise.values < 5
+gaia_mc = gaia_mc.iloc[m]
+
+print("Loading Dustmaps")
+bayestar = BayestarQuery(max_samples=2, version='bayestar2019')
+
+print("Calculating Ebv")
+coords = SkyCoord(gaia_mc.ra.values*units.deg, gaia_mc.dec.values*units.deg,
+                  distance=gaia_mc.r_est.values*units.pc)
+
[…]

### 2019-08-29 photometric_teff.py

@@ -1,0 +1,21 @@
+# Calculate photometric Teff from Gaia color
+# This should be dereddened colour!
+
+
+def bprp_to_teff(bp, rp):
+    """
+    Calculate photometric Teff from Gaia color (use dereddened color!)
+
+    Args:
+        bp (array): Gaia G_BP colour.
+        rp (array): Gaia G_RP colour.
+
+    Returns:
+        teffs (array): Photometric effective temperatures.
+
+    """
+
+    coeffs = [8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
+            -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
+            301.13205924630165, -25.923997443169355]
+    return np.polyval(coeffs[::-1], bp - rp)

### 2019-08-29 data.py
switching to bp-rp rather than bp, rp

@@ -113,7 +113,7 @@ flag[~m] = np.ones(len(flag[~m]))
 gaia_mc["flag"] = flag
 
 # Calculate photometric Teff
-teffs = bprp_to_teff(gaia_mc.bp_dered, gaia_mc.rp_dered)
+teffs = bprp_to_teff(gaia_mc.bp_dered - gaia_mc.rp_dered)
 gaia_mc["color_teffs"] = teffs
 
 print("Calculating gyro ages")

### 2019-08-29 photometric_teff.py

@@ -2,13 +2,12 @@
 # This should be dereddened colour!
 
 
-def bprp_to_teff(bp, rp):
+def bprp_to_teff(bprp):
     """
     Calculate photometric Teff from Gaia color (use dereddened color!)
 
     Args:
-        bp (array): Gaia G_BP colour.
-        rp (array): Gaia G_RP colour.
+        bprp (array): Gaia G_BP colour minus Gaia G_RP colour.
 
     Returns:
         teffs (array): Photometric effective temperatures.
@@ -18,4 +17,4 @@ def bprp_to_teff(bp, rp):
     coeffs = [8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
             -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
             301.13205924630165, -25.923997443169355]
-    return np.polyval(coeffs[::-1], bp - rp)
+    return np.polyval(coeffs[::-1], bprp)

### 2019-08-30 paper/discussion.tex
updating draft and adding figures
redrafting
restructuring and moving some results sections to discussion

@@
 \section{Discussion}
+\label{sec:discussion}
 
-What if the M dwarfs just have larger velocity dispersion because they're less
-massive?
+\subsection{Mass-dependent heating}
+
+An alternative explanation for this phenomenon is that lower-mass stars
+experience greater velocity changes when gravitationally perturbed and are
+more efficiently dynamically heated.
+Mass-dependent orbital heating has not been unambiguously observed in the
+galactic disk because of the strong anti-correlation between stellar mass and
+stellar age.
+Less massive stars do indeed have larger velocity dispersions, however they
+are also older on average.
+This mass-age degeneracy is highly reduced in M dwarfs because their
+main-sequence lifetimes are longer than the age of the Universe, however no
+evidence for mass-dependent heating has been detected in these low mass stars
+\citep{faherty2009}.
+
+To investigate, we selected late K and M dwarfs observed by both \kepler\ and
+\gaia, whose MS lifetimes exceed around 11 Gyrs and are therefore
+representative of the initial mass function.
+We could not perform this analysis on the sample of stars with measured
+rotation periods, because only stars with {\it detectable} rotation periods
+appear in the \mct\ catalog.
+Since lower-mass stars stay active for longer it is likely that there are more
+old stars (with larger velocity dispersions) at low masses in this sample.
+% This allowed us to remove the degeneracy between mass and age that would
+% confound evidence for mass-dependent heating among more massive stars.
+% In a sample of massive stars with short MS lifetimes, lower-mass stars
+% typically have greater velocity dispersion, but this could be either because
+% dynamical heating is more efficient in lower mass stars, or because lower mass
+% stars tend to be older because they live longer.
+% In a sample of late K and M dwarfs, whose MS lifetime is longer than the age
+% of the Universe, we can assume that none have evolved into giants and any
+% additional heating seen in the lower-mass stars must be caused by
+% mass-dependent heating.
+We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
+greater than 1.2 (corresponding to an effective temperature $\lesssim$
+4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+We also eliminated visual binaries by removing stars above a 6th order
+polynomial, fit to the MS on the \gaia\ CMD.
+% shown in figure \ref{fig:mdwarf_CMD}.
+We then applied the quality cuts described in section \ref{sec:the_data}.
+% \begin{figure}
+%   \caption{
+% A \gaia\ color-magnitude diagram, showing all \kepler\ targets.
+% We selected K and M dwarfs by applying color and magnitude cuts at \gcolor\ =
+%     1.2 (\teff\ $\sim$ 4800 K) and $G$ = 4.
+% We eliminated visual binaries by fitting a 6th order polynomial to the MS and
+%     removing stars above it.
+% Selected K and M dwarfs are shown as darker points.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{mdwarf_CMD}
+% \label{fig:mdwarf_CMD}
+% \end{figure}
+To search for evidence of mass-dependent heating we calculated the (\vb)
+velocity dispersion of stars in effective temperature bins.
+Again, sigma clipping was used to remove high velocity outliers before
+calculating the standard deviation of stars in each bin.
+% A slight excess of non-Gaussian outliers at lower temperatures leads to a
+% marginal increase in velocity dispersion at low temperatures when sigma
+% clipping is not performed.
+% Figure \ref{fig:vb_vs_color} shows velocity and velocity dispersion as a
+% function of \gaia\ \gcolor\ color.
+% There is no clear trend between velocity dispersion and color.
+
+Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
+as a function of effective temperature (calculated by transforming dereddened
+\gaia\ colors using equation \ref{eq:curtis}).
+Velocity dispersion very slightly {\it decreases} with decreasing temperature,
+(the opposite of the trend expected for mass-dependent heating) however the
+slope is only inconsistent with zero at the 1.3 $\sigma$ level.
+This trend may be due to a selection bias: cooler stars are fainter and
+therefore typically closer, with smaller heights above the galactic plane and
+smaller velocities.
+The essential point here is that the increasing velocity dispersion with age
+and decreasing temperature that is evident in figure \ref{fig:age_cut} is not
+a result of an intrinsically larger velocity dispersion for low-mass stars.
+We performed the same analysis on the 537 stars in this sample with RVs using
+{\it vertical} velocity (\vz) and found the slope of the velocity
+dispersion-temperature relation was consistent with zero.
+
+Since we used \vb, not \vz\ in our analysis, it is possible that the
+anti-correlation between \teff\ and galactic latitude, $b$, could influence
+our results: stars at higher latitudes would have additional velocity
+components in the ${\bf x}$ and ${\bf y}$ directions, which could increase
+\vb\ but not \vz.
+However, since the relationship between \sigmavb\ and \teff\ is positively,
+not negatively correlated for cool stars in the \kepler\ field, this effect is
+probably too small to influence our results.
+% \begin{figure}
+%   \caption{
+%       Top: Stellar velocity (\vb) as a function of \gaia\ \gcolor\ color for
+%       \kepler\ K and M dwarfs.
+% Vertical lines indicate different color-groupings used to calculate velocity
+%     dispersion.
+% Pink stars were not included in velocity dispersion calculations as they were
+%     removed as outliers during a sigma clipping process.
+% Bottom: velocity dispersion as a function of color.
+% There is no obvious trend between velocity dispersion and stellar color
+%     indicating that mass-dependent heating does not significantly effect
+%     low-mass stars in the \kepler\ field.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{vb_vs_color}
+% \label{fig:vb_vs_color}
+% \end{figure}
+
+\begin{figure}
+  \caption{
+      Top: Stellar velocity (\vb) as a function of \teff\ for
+      \kepler\ K and M dwarfs.
+Vertical lines indicate different \teff-groupings used to calculate velocity
+    dispersion.
+Pink stars were not included in velocity dispersion calculations as they were
+    either removed as outliers during a sigma clipping process, or they lie at
+    the sparcely populated, extremely cool end of the temperature range.
+    Velocity dispersion and \teff\ are slightly positively correlated, likely
+    due to a brightness-related selection bias, indicating that mass-dependent
+    heating does not significantly affect low-mass stars in the \kepler\
+    field.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{vb_vs_teff}
+\label{fig:vb_vs_teff}
+\end{figure}
+Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+strongly affect the \mct\ sample of rotating \kepler\ stars.
+For this reason, we assume that age difference is the major cause of velocity
+dispersion differences between groups.
+In other words, (\vb) velocity dispersion is a reliable age proxy for the
+\mct\ sample.
+
+% The \vb\ AVR is not directly comparable to the \vz\ AVR, so, to draw a
+% comparison with results from the literature, we calculated a \vz\ AVR for
+% the subset of 290 cool stars in the \mct\ sample with \gaia\ RVs.
+% We measured an AVR exponent of 0.47 $\pm$ 0.1, which falls within the range of
+% values (0.45-0.53) reported from measurements of F and G stars in the Solar
+% neighborhood \citep{holmberg2009, aumer2009, aumer2016}.
+% It is a little lower than the values of 0.56 $\pm$ 0.14 (for low-z) and 0.51
+% $\pm$ 0.15 (for high-z stars), calculated using \LAMOST\ \racomment{(LAMOST
+% citation)} K giants \citep{yu2018}.
+% K giants are more massive than K dwarfs \racomment{(how much more massive?)},
+% so this slightly higher value contradicts the mass-dependent heating
+% hypothesis.
+% \racomment{Add some words about selection functions...}
+
+% To investigate further, we calculated the exponent of the (\vb) AVR for each
+% temperature bin in figure \ref{fig:age_cut}.
+% \begin{figure}
+%   \caption{
+%       The exponent of the (\vb) AVR as a function of effective temperature.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{AVR_exponent}
+% \label{fig:AVR_exponent}
+% \end{figure}
+% Figure \ref{fig:AVR_exponent} shows the significant rise the in AVR exponent
+% as a function of effective temperature.
+% The hottest stars in our sample (\teff\ = 4833-5000 K) have a (\vb) AVR
+% exponent of 0.39 $\pm$ 0.02 and the coolest (\teff\ = 3500-3667 K) have an
+% exponent of 0.74 $\pm$ 0.02.
+% The masses of stars in our sample differ by less than a factor of two,
+% so mass-dependent heating cannot be entirely responsible for this large spread
+% in heating rates.
+% \racomment{Need to look into this further.}
+
+% If mass-dependent heating is strongly affecting our data, the exponent of the
+% AVR should increase with decreasing effective temperature, \ie\ the heating
+% rate should be greater for lower-mass stars.
+% Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
+% decreasing effective temperature, however, the AVR exponent calculated using
+% more massive stars from the Geneva Copenhagen Survey (GCS)
+% \citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
+% Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
+% in our study.
+% If mass-dependent orbital heating is the main cause the rise in velocity
+% dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
+% AVR exponent of the more massive GCS stars should fall {\it below} the AVR
+% exponent for the most massive stars in our sample.
+% In other words, the dynamical heating rate should be lowest for the highest
+% mass stars and highest for the lowest mass stars.
+% % nordstrom2004, jorgensen2005,
+% These AVR exponents were calculated with different populations of stars,
+% subject to different selection effects, so directly comparing them comes with
+% some risk.
+% The most important difference is that the GCS AVR is calculated using \vz,
+% however our AVRs are calculated using \vb\ since most stars in our sample do
+% not have radial velocities.
+% The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
+% latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
+% close approximation to \vz.
+% Only 290 out of the 6820 stars in our sample, plotted in color in figure
+% \ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
+% that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
+% between 3500 and 5000 K.
+% For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
+% AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
+% The similarity of these two AVR exponents suggests that directly comparing
+% these two quantities {\it may} be a reasonable approach.
+% For now, given the large difference between the (\vz) AVR exponent of the GCS
+% sample and the (\vb) AVR exponent of the hottest stars in our sample, we
+% assume that the increased velocity dispersion at cooler temperatures is mostly
+% caused by incorrect age-grouping due to an incorrect period-color relation at
+% old ages, and that any mass-dependent heating, while it may contribute at a
+% low level to this result, is not the dominant driver.
+% However, differentiating the effects of mass-dependent heating and the shape
+% of the gyrochronology relations is certainly warranted in a follow-up study.
+% }
+
+% Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
+% the \citet{angus2019} gyrochronology relation, compared with the ages of star
+% groups, predicted with the \citet{holmberg2009} AVR.
+% \citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
+% relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
+% dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
+% dispersion across all temperatures, of our sample.
+% \begin{figure}
+%   \caption{
+%       Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
+%     predicted by the \citet{angus2019} gyrochronology relation, for stars in
+%     different temperature ranges.
+% Points are colored by the effective temperature in the center of the \teff\
+% bin.
+% The dashed line shows the y=x relation.
+% Predicted kinematic ages fan out over gyrochronal time, either because the AVR
+%     is temperature-dependent or because the gyrochronology relations have an
+%     age-dependent period-color relation.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{age_comparison}
+% \label{fig:age_comparison}
+% \end{figure}
+% Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
+% the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
+% predicted by the \citet{angus2019} gyrochronology relation, for stars in
+% different temperature ranges.
+% The predicted kinematic ages fan out over gyrochronal time, either because the
+% AVR is temperature-dependent or because the gyrochronology relations have an
+% age-dependent period-color relation.
+% Figure \ref{fig:age_comparion} also suggests that dynamical heating does not
+% begin until after around ... Gyr:
+
+% Alternatively, the ages of these young stars could be under-predicted by
+% gyrochronology.
+% These young stars have similar rotation periods to the Praesepe cluster, which
+% was used to calibrate the gyrochronology relation used in this analysis.
+% The ages of these young stars should, therefore, be the most accurate of the
+% entire sample.
+% However, they are tied to the age of the Praesepe cluster, whose age is not
+% accurately known.
+% \citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
+% kinematic age-prediction for these stars is around 400 Myrs.
+% Other studies have assigned a much older age of 800 Myrs to Praesepe
+% \citep{brandt2015}.
+% If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
+% calibration was inaccurate, the entire age {\it scale} would be incorrect.
+% However, this is not what figure \ref{fig:age_comparion} shows -- it shows
+% that the {\it relative} age of these youngest stars are under-predicted by
+% kinematics or over-predicted by gyrochronology.
+
+% If we assume that the rise in velocity dispersions at cooler effective
+% temperatures is caused by an inaccurate period-color relation rather than
+% mass-dependent dynamical heating, we can estimate what the shape of the
+% period-color relations {\it should} be.
+% Figure \ref{fig:age_cut} indicates that the period-color relation flattens
+% out, so we applied the same analysis to groups of stars with similar rotation
+% periods, equivalent to a completely flat period-color relation.
+% The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
+% in different period ranges plotted in different colors.
+% The bottom panel shows the velocity dispersion of each group as a function of
+% effective temperature.
+% \begin{figure}
+%   \caption{
+% This figure is similar to figure \ref{fig:age_cut}, with stars divided into
+%     period groups rather than age groups.
+% The velocity dispersion is more constant across effective temperatures for the
+%     most slowly rotating stars, compared to the stars selected with the
+%     \citet{angus2019} gyrochronology model, indicating that the gyrochronology
+%     models flatten out, and possibly even invert, at old ages.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{period_cut}
+% \label{fig:period_cut}
+% \end{figure}
+% Once again, the velocity dispersion increases with rotation period overall.
+% For the most rapidly rotating groups of stars, velocity dispersion decreases
+% with \teff\ as expected given the positively sloped period-color relation of
+% Praesepe and other young clusters: late K dwarfs rotate more slowly than early
+% K dwarfs of the same age.
+% Between rotation periods of 15 and 25 days, the temperature dependence of the
+% velocity dispersion starts to disappear, indicating that the period-color
+% relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
+% of the same age.
+% At long rotation periods, the velocity dispersion still increases with
+% effective temperature, although the increase is much more modest compared to
+% the oldest stars in figure \ref{fig:age_cut}.
+
+% If we assume the rise in velocity dispersion at cooler effective temperatures
+% is caused by an inaccurate period-color relation rather than mass-dependent
+% dynamical heating, we can generate a graphical representation of what the
+% relations {\it should} look like.
+
+% Aside from the assumption that mass-dependent dynamical heating is {\it not}
+% the main cause of the increased velocity dispersion for cooler stars, a number
+% of assumptions went into this analysis which should be addressed before a
+% strong conclusion about the nature of stellar spin down is drawn.
+% There are some potential confounders that could produce trends between
+% velocity dispersion and effective temperature, of which the main ones are
+% listed below.
+% \begin{itemize}
+
+    % \item{{\bf Extinction.}
+% Although we dereddened stars before their ages were estimated, if the
+    %     reddening values were underestimated, the gyrochronal ages of stars
+    %     would be under-estimated.
+% This effect would be largest for the late Ks and early Ms, where the shape of
+    %     the period-color relation is more steeply sloped, and stars can more
+    %     easily move into the wrong age bin with a small perturbation in color
+    %     or temperature.
+% Old stars would be incorrectly assigned young ages, and this would happen more
+    %     often for cooler stars.
+% This effect could lead to an increase in velocity dispersion with \teff, even
+% if the gyrochronology relation was perfect.}
+
+    % \item{{\bf Mass-dependent orbital heating.}
+% A key assumption going into this analysis is that the orbits of all stars are
+% heated at the same rate, regardless of their mass.
+% However, if the heating mechanism leads to preferential heating of lower mass
+% stars, those stars would have a larger velocity dispersion, not caused by
+% ageing.
+% No strong evidence has yet been provided to show a mass-dependence in orbital
+% heating, however this is an extremely difficult exercise because mass and age
+% are so strongly correlated.
+% The mass-age degeneracy is reduced in the lowest mass stars which represent
+    %     the initial mass function of the Solar neighborhood.
+% No mass-dependent heating was found by \citet{faherty2009} who looked
+% at the velocity dispersions of low-mass stars.
+% Another argument against the increased velocity dispersion as a function of
+    %     \teff\ being caused by mass-dependent orbital heating, is that heating
+    %     acts most strongly at young ages and most weakly at old ages, \ie\ on
+    %     a logarithmic timescale.
+% This is related to the fact that stars are more difficult to scatter once they
+% are on orbits with a large out-of-plane component and they spend more time out
+% of the mid-plane.
+% Our data show the opposite trend: stars show little mass-dependent heating
+    %     before around 1 Gyr.
+% Moreover, the stars in the Geneva-Copenhagen study, used to calibrate the AVR
+    %     \citep{holmberg2009} are F and G stars, around 1-3 times as massive as
+    %     the K stars in our sample.
+% % nordstrom2004, jorgensen2005,
+% The ages for our sample of K stars, predicted by gyrochronology agree with the
+    %     AVR calibrated using F and G stars, which implies that either
+    %     mass-dependent heating is a weak effect, or, mass-dependent heating is
+    %     strong and, by coincidence, gyrochronology underpredicts the ages of
+    %     the early K dwarfs by just the right amount that they still appear to
+    %     agree with the \citet{holmberg2009} AVR.
+% We therefore expect that the increased velocity dispersion at cooler
+    %     temperatures is mostly caused by incorrect age-grouping, \ie\ the
+    %     period-color relation is incorrect for old, low-mass stars, and that
+    %     any mass-dependent heating, while it may contribute at a low level to
+    %     this result, is not the dominant cause.
+% }
+
+    % \item{{\bf Selection bias.}
+% The \gaia\
+% catalog is incompletebecause they are too faint.
+% This is simply due to the fact that \kepler's field of view was aimed at low
+    %     galactic latitudes, so nearby \kepler\ stars are close to the galactic
+    %     plane.
+% The heights of stars above the galactic plane, and correspondingly their
+    %     velocities and ages, are therefore correlated with distance.
+% Almost no high velocity stars appear in our sample at temperatures cooler than
+% 3500, which is why the coolest stars were excluded from our sample, however
+% the selection function could still affect stars hotter than 3500 K, and is
+% suggested by the decrease in velocity dispersions for cool stars of all ages
+% in figure \ref{fig:age_cut}.
+% However, this selection effect would act in opposition to the observed trend.
+% }
+
+% \item{{\bf Weakened magnetic braking.}
+% The rotation periods of old stars are faster than predicted by Skumanich-like
+% magnetic braking \citep{angus2015, vansaders2016, metcalfe2019}.
+% Magnetic braking is expected to become inefficient once stars approach a
+% Rossby number (the ratio of rotation period to convective turnover time) of
+% around 2 \citep{vansaders2016, vansaders2018}.
+% Since hot stars have shallow convection zones and shorter overturn timescales,
+% they reach $Ro = 2$ and stop spinning down at shorter rotation periods than
+% cool stars.
+% For this reason, some of the hotter stars in the sample may already have
+% stopped spinning down, and could be much older than a traditional
+% gyrochronology relation (like the simple Praesepe-based relation used in this
+% analysis) would suggest.
+% However, this effect would work to cancel out the trend observed in figure
+% \ref{fig:age_cut} because hotter groups of stars would contain more old stars
+% with large velocities.
+% }
+
+% \item{Stellar or exoplanet companions.}
+% Hotter stars are more likely to be binaries.
+% This would make synchronized some stars seem younger, more likely for massive
+% stars, although really only likely at short rotation periods.
+
+% \end{itemize}
+
+\subsection{The period gap}
+The origin of the rotation period gap, first identified
+by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+\ref{fig:dispersion_period_teff} remains a mystery.
+This gap can be seen as an under-density of points between the 0.7-1.0 and
+1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
+of constant gyrochronal age of around 1.1 Gyr \citep[according to the
+gyrochronology relation of][]{angus2019}, as shown in figure
+\ref{fig:dispersion_period_teff}.
+Several explanations for the gap's origin have been proposed, including a
+discontinuous star formation history \citep{mcquillan2013, davenport2017,
+davenport2018}, a rapid change in magnetic field structure
+\citep{reinhold2019}, and erroneous rotation period measurements that are
+incorrect by a factor of two \citep{koen2018}.
+The latter explanation can be ruled out because stars below the gap have
+smaller velocity dispersions than the stars above the gap, indicating that
+they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+evident in figure \ref{fig:age_cut}.
+For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+velocity dispersion increases with \teff.
+The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+dispersion as the hottest stars in the age range above which indicates that
+the period-\teff\ relations are {\it flat} at these rotation periods.
+This is also visible in figure \ref{fig:dispersion_period_teff}.
+Below the gap, velocity dispersion within a given period range appears to {\it
+decrease} with decreasing temperature.
+The opposite appears to be true above the gap.
+
+In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
+are faster than expected; their rotational evolution appears to have stalled,
+and the period-\teff\ relation is flat \citep{curtis2019}.
+The rotation periods of the K dwarfs in this cluster are plotted in figure
+\ref{fig:dispersion_period_teff}.
+Although the G dwarfs in this cluster fall on the 1.1 Gyr gyrochronology
+model, the K dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
+% If we assume that the gap is located at a single (gyrochronal) age, its age is
+% remarkably similar to the NGC 6811 cluster (measured from MS turn-off) of 1.1
+% Gyr \citep{meibom2011}.
+NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
+K dwarfs lie below it.
+This cluster may be the `missing link' that connects two epoch of stellar
+spin-down: an early stage where the period-\teff\ relation for K dwarfs has a
+negative slope and a late stage where it has a positive slope.
+In this case the period gap may delineate the transition between these two
+regimes and is the point at which stellar magnetic dynamos likely undergo a
+dramatic structural shift at an age of $\sim$ 1.1 Gyr.
+
+It may be a coincidence that the gyrochronology relations seem to only flatten
+off {\it above} the period gap, or it may be that the period gap is located at
+magnetic transition period.
+We lack a sufficient quantity of data to investigate in detail but new field
+star rotation periods from the \ktwo\ and \tess\ missions may be able to
+validate or rule out this hypothesis in the future.

### 2019-08-30 paper/results.tex
restructuring and moving some results sections to discussion

@@
 
-\subsection{Mass-dependent heating}
-
-An alternative explanation for this phenomenon is that lower-mass stars
-experience greater velocity changes when gravitationally perturbed and are
-more efficiently dynamically heated.
-Mass-dependent orbital heating has not been unambiguously observed in the
-galactic disk because of the strong anti-correlation between stellar mass and
-stellar age.
-Less massive stars do indeed have larger velocity dispersions, however they
-are also older on average.
-This mass-age degeneracy is highly reduced in M dwarfs because their
-main-sequence lifetimes are longer than the age of the Universe, however no
-evidence for mass-dependent heating has been detected in these low mass stars
-\citep{faherty2009}.
-To investigate further, we selected late K and M dwarfs observed by both
-\kepler\ and \gaia, whose MS lifetimes exceed around 11 Gyrs and are therefore
-representative of the initial mass function.
-% This allowed us to remove the degeneracy between mass and age that would
-% confound evidence for mass-dependent heating among more massive stars.
-% In a sample of massive stars with short MS lifetimes, lower-mass stars
-% typically have greater velocity dispersion, but this could be either because
-% dynamical heating is more efficient in lower mass stars, or because lower mass
-% stars tend to be older because they live longer.
-% In a sample of late K and M dwarfs, whose MS lifetime is longer than the age
-% of the Universe, we can assume that none have evolved into giants and any
-% additional heating seen in the lower-mass stars must be caused by
-% mass-dependent heating.
-We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
-greater than 1.2 (corresponding to an effective temperature $\lesssim$
-4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
-We also eliminated visual binaries by removing stars above a 6th order
-polynomial, fit to the MS on the \gaia\ CMD.
-% shown in figure \ref{fig:mdwarf_CMD}.
-We then applied the same quality cuts as described in section
-\ref{sec:the_data}.
-% \begin{figure}
-%   \caption{
-% A \gaia\ color-magnitude diagram, showing all \kepler\ targets.
-% We selected K and M dwarfs by applying color and magnitude cuts at \gcolor\ =
-%     1.2 (\teff\ $\sim$ 4800 K) and $G$ = 4.
-% We eliminated visual binaries by fitting a 6th order polynomial to the MS and
-%     removing stars above it.
-% Selected K and M dwarfs are shown as darker points.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{mdwarf_CMD}
-% \label{fig:mdwarf_CMD}
-% \end{figure}
-To search for evidence of mass-dependent heating we calculated the (\vb)
-velocity dispersion of stars in effective temperature bins.
-Again, we applied sigma clipping to remove high velocity outliers before
-calculating the standard deviation of stars in each bin.
-% A slight excess of non-Gaussian outliers at lower temperatures leads to a
-% marginal increase in velocity dispersion at low temperatures when sigma
-% clipping is not performed.
-% Figure \ref{fig:vb_vs_color} shows velocity and velocity dispersion as a
-% function of \gaia\ \gcolor\ color.
-% There is no clear trend between velocity dispersion and color.
-Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
-as a function of effective temperature (calculated by transforming dereddened
-\gaia\ colors using equation \ref{eq:curtis}).
-Velocity dispersion very slightly {\it decreases} with decreasing temperature,
-(the opposite of the trend expected for mass-dependent heating) however the
-slope is only inconsistent with zero at the 1.3 $\sigma$ level.
-This trend may be due to a selection bias: cooler stars are fainter and
-therefore typically closer, with smaller heights above the galactic plane and
-smaller velocities.
-The essential point here is that the increasing velocity dispersion with age
-and decreasing temperature that is evident in figure \ref{fig:age_cut} is not
-a result of an intrinsically larger velocity dispersion for low-mass stars,
-nor is it a result of the selection function.
-Since we use \vb, not \vz\ in this analysis, it is possible that the
-anti-correlation between \teff\ and galactic latitude, $b$, could influence
-these results: stars at higher latitudes would have additional velocity
-components in the ${\bf x}$ and ${\bf y}$ directions, which could increase
-\vb\ but not \vz.
-However, since the relationship between \sigmavb\ and \teff\ is positively,
-not negatively correlated for cool stars in the \kepler\ field, this effect is
-probably too small to influence our results.
-We performed the same analysis on the 537 stars in this sample with RVs using
-{\it vertical} velocity (\vz) and found the slope of the velocity
-dispersion-temperature relation was consistent with zero.
-% \begin{figure}
-%   \caption{
-%       Top: Stellar velocity (\vb) as a function of \gaia\ \gcolor\ color for
-%       \kepler\ K and M dwarfs.
-% Vertical lines indicate different color-groupings used to calculate velocity
-%     dispersion.
-% Pink stars were not included in velocity dispersion calculations as they were
-%     removed as outliers during a sigma clipping process.
-% Bottom: velocity dispersion as a function of color.
-% There is no obvious trend between velocity dispersion and stellar color
-%     indicating that mass-dependent heating does not significantly effect
-%     low-mass stars in the \kepler\ field.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{vb_vs_color}
-% \label{fig:vb_vs_color}
-% \end{figure}
-
-\begin{figure}
-  \caption{
-      Top: Stellar velocity (\vb) as a function of \teff\ for
-      \kepler\ K and M dwarfs.
-Vertical lines indicate different \teff-groupings used to calculate velocity
-    dispersion.
-Pink stars were not included in velocity dispersion calculations as they were
-    either removed as outliers during a sigma clipping process, or they lie at
-    the sparcely populated, extremely cool end of the temperature range.
-    Velocity dispersion and \teff\ are slightly positively correlated, likely
-    due to a brightness-related selection bias, indicating that mass-dependent
-    heating does not significantly affect low-mass stars in the \kepler\
-    field.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vb_vs_teff}
-\label{fig:vb_vs_teff}
-\end{figure}
-Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
-strongly affect the \mct\ sample of rotating \kepler\ stars.
-For this reason, we assume that age difference is the only significant cause
-of velocity dispersion differences between groups.
-In other words, (\vb) velocity dispersion is a reliable age proxy for the
-\mct\ sample.
-
-% The \vb\ AVR is not directly comparable to the \vz\ AVR, so, to draw a
-% comparison with results from the literature, we calculated a \vz\ AVR for
-% the subset of 290 cool stars in the \mct\ sample with \gaia\ RVs.
-% We measured an AVR exponent of 0.47 $\pm$ 0.1, which falls within the range of
-% values (0.45-0.53) reported from measurements of F and G stars in the Solar
-% neighborhood \citep{holmberg2009, aumer2009, aumer2016}.
-% It is a little lower than the values of 0.56 $\pm$ 0.14 (for low-z) and 0.51
-% $\pm$ 0.15 (for high-z stars), calculated using \LAMOST\ \racomment{(LAMOST
-% citation)} K giants \citep{yu2018}.
-% K giants are more massive than K dwarfs \racomment{(how much more massive?)},
-% so this slightly higher value contradicts the mass-dependent heating
-% hypothesis.
-% \racomment{Add some words about selection functions...}
-
-% To investigate further, we calculated the exponent of the (\vb) AVR for each
-% temperature bin in figure \ref{fig:age_cut}.
-% \begin{figure}
-%   \caption{
-%       The exponent of the (\vb) AVR as a function of effective temperature.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{AVR_exponent}
-% \label{fig:AVR_exponent}
-% \end{figure}
-% Figure \ref{fig:AVR_exponent} shows the significant rise the in AVR exponent
-% as a function of effective temperature.
-% The hottest stars in our sample (\teff\ = 4833-5000 K) have a (\vb) AVR
-% exponent of 0.39 $\pm$ 0.02 and the coolest (\teff\ = 3500-3667 K) have an
-% exponent of 0.74 $\pm$ 0.02.
-% The masses of stars in our sample differ by less than a factor of two,
-% so mass-dependent heating cannot be entirely responsible for this large spread
-% in heating rates.
-% \racomment{Need to look into this further.}
-
-% If mass-dependent heating is strongly affecting our data, the exponent of the
-% AVR should increase with decreasing effective temperature, \ie\ the heating
-% rate should be greater for lower-mass stars.
-% Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
-% decreasing effective temperature, however, the AVR exponent calculated using
-% more massive stars from the Geneva Copenhagen Survey (GCS)
-% \citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
-% Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
-% in our study.
-% If mass-dependent orbital heating is the main cause the rise in velocity
-% dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
-% AVR exponent of the more massive GCS stars should fall {\it below} the AVR
-% exponent for the most massive stars in our sample.
-% In other words, the dynamical heating rate should be lowest for the highest
-% mass stars and highest for the lowest mass stars.
-% % nordstrom2004, jorgensen2005,
-% These AVR exponents were calculated with different populations of stars,
-% subject to different selection effects, so directly comparing them comes with
-% some risk.
-% The most important difference is that the GCS AVR is calculated using \vz,
-% however our AVRs are calculated using \vb\ since most stars in our sample do
-% not have radial velocities.
-% The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
-% latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
-% close approximation to \vz.
-% Only 290 out of the 6820 stars in our sample, plotted in color in figure
-% \ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
-% that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
-% between 3500 and 5000 K.
-% For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
-% AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
-% The similarity of these two AVR exponents suggests that directly comparing
-% these two quantities {\it may} be a reasonable approach.
-% For now, given the large difference between the (\vz) AVR exponent of the GCS
-% sample and the (\vb) AVR exponent of the hottest stars in our sample, we
-% assume that the increased velocity dispersion at cooler temperatures is mostly
-% caused by incorrect age-grouping due to an incorrect period-color relation at
-% old ages, and that any mass-dependent heating, while it may contribute at a
-% low level to this result, is not the dominant driver.
-% However, differentiating the effects of mass-dependent heating and the shape
-% of the gyrochronology relations is certainly warranted in a follow-up study.
-% }
-
-% Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
-% the \citet{angus2019} gyrochronology relation, compared with the ages of star
-% groups, predicted with the \citet{holmberg2009} AVR.
-% \citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
-% relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
-% dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
-% dispersion across all temperatures, of our sample.
-% \begin{figure}
-%   \caption{
-%       Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
-%     predicted by the \citet{angus2019} gyrochronology relation, for stars in
-%     different temperature ranges.
-% Points are colored by the effective temperature in the center of the \teff\
-% bin.
-% The dashed line shows the y=x relation.
-% Predicted kinematic ages fan out over gyrochronal time, either because the AVR
-%     is temperature-dependent or because the gyrochronology relations have an
-%     age-dependent period-color relation.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{age_comparison}
-% \label{fig:age_comparison}
-% \end{figure}
-% Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
-% the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
-% predicted by the \citet{angus2019} gyrochronology relation, for stars in
-% different temperature ranges.
-% The predicted kinematic ages fan out over gyrochronal time, either because the
-% AVR is temperature-dependent or because the gyrochronology relations have an
-% age-dependent period-color relation.
-% Figure \ref{fig:age_comparion} also suggests that dynamical heating does not
-% begin until after around ... Gyr:
-
-% Alternatively, the ages of these young stars could be under-predicted by
-% gyrochronology.
-% These young stars have similar rotation periods to the Praesepe cluster, which
-% was used to calibrate the gyrochronology relation used in this analysis.
-% The ages of these young stars should, therefore, be the most accurate of the
-% entire sample.
-% However, they are tied to the age of the Praesepe cluster, whose age is not
-% accurately known.
-% \citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
-% kinematic age-prediction for these stars is around 400 Myrs.
-% Other studies have assigned a much older age of 800 Myrs to Praesepe
-% \citep{brandt2015}.
-% If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
-% calibration was inaccurate, the entire age {\it scale} would be incorrect.
-% However, this is not what figure \ref{fig:age_comparion} shows -- it shows
-% that the {\it relative} age of these youngest stars are under-predicted by
-% kinematics or over-predicted by gyrochronology.
-
-% If we assume that the rise in velocity dispersions at cooler effective
-% temperatures is caused by an inaccurate period-color relation rather than
-% mass-dependent dynamical heating, we can estimate what the shape of the
-% period-color relations {\it should} be.
-% Figure \ref{fig:age_cut} indicates that the period-color relation flattens
-% out, so we applied the same analysis to groups of stars with similar rotation
-% periods, equivalent to a completely flat period-color relation.
-% The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
-% in different period ranges plotted in different colors.
-% The bottom panel shows the velocity dispersion of each group as a function of
-% effective temperature.
-% \begin{figure}
-%   \caption{
-% This figure is similar to figure \ref{fig:age_cut}, with stars divided into
-%     period groups rather than age groups.
-% The velocity dispersion is more constant across effective temperatures for the
-%     most slowly rotating stars, compared to the stars selected with the
-%     \citet{angus2019} gyrochronology model, indicating that the gyrochronology
-%     models flatten out, and possibly even invert, at old ages.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{period_cut}
-% \label{fig:period_cut}
-% \end{figure}
-% Once again, the velocity dispersion increases with rotation period overall.
-% For the most rapidly rotating groups of stars, velocity dispersion decreases
-% with \teff\ as expected given the positively sloped period-color relation of
-% Praesepe and other young clusters: late K dwarfs rotate more slowly than early
-% K dwarfs of the same age.
-% Between rotation periods of 15 and 25 days, the temperature dependence of the
-% velocity dispersion starts to disappear, indicating that the period-color
-% relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
-% of the same age.
-% At long rotation periods, the velocity dispersion still increases with
-% effective temperature, although the increase is much more modest compared to
-% the oldest stars in figure \ref{fig:age_cut}.
-
-% If we assume the rise in velocity dispersion at cooler effective temperatures
-% is caused by an inaccurate period-color relation rather than mass-dependent
-% dynamical heating, we can generate a graphical representation of what the
-% relations {\it should} look like.
-
 \subsection{The period-\teff\ relations, revealed}
@@
   \caption{
-Rotation period vs effective temperature for stars in the \mct\ sample,
-    colored by the velocity dispersions of stars in restricted period and
-    temperature ranges.
+      Top: Rotation period vs effective temperature for stars in the \mct\
+    sample, colored by the velocity dispersions of stars over a grid in
+    $\log_{10}$(period) and \teff.
+    Bottom: the velocity dispersions of groups of stars, shown as a solid
+    grid for clarity.
+    The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
+    with the \citet{angus2019} gyrochronology relation which roughly traces
+    the rotation period gap.
+    The black dashed lines show a 650 Myr isochrone, indicating the location
+    and shape of the Praesepe cluster (to which this gyrochronology model was
+    calibrated).
+    The black points show the 1.1 Gyr NGC 6811 open cluster.
 }
@@
 our sample, coloured by (\vb) velocity dispersion.
-Here, velocity dispersion was calculated for groups of stars in
-uniformly-spaced bins in $\log_{10}$(period) and temperature.
+Velocity dispersion was calculated for groups of stars over a grid in
+$\log_{10}$(period) and temperature.
 If we assume that mass dependent heating doesn't affect this sample, and \vb\
@@
 color in figure \ref{fig:dispersion_period_teff} are similar ages.
-Interpreted this way, it appears that rotation period {\it decreases} with
-decreasing effective temperature at old ages.  This is a paradigm shift
-for gyrochronology because stellar spin-down rate is thought to be directly
-tied to magnetic field strength, and the deeper convection zones of cooler
-stars generate stronger magnetic fields which {\it should} lead to more
-efficient angular momentum loss.
-A similar phenomenon has been observed in the 1.1 Gyr NGC 6811 cluster, where
-the rotation periods of mid-K dwarfs are faster than expected; their
-rotational evolution appears to have stalled, and the period-\teff\ relation
-is flat \citep{curtis2019}.
-NGC 6811 straddles the rotation period gap and may be the `missing link' that
-connects two epoch of stellar spin-down: an early stage where the
-period-\teff\ relation for K dwarfs has a negative slope and a late stage
-where it has a positive slope.
-In this case the period gap may delineate the transition between these two
-regimes and is the point at which stellar magnetic dynamos likely undergo a
-dramatic structural shift.
+Interpreted this way, lines of constant age (isochrones) appear to follow the
+shape of the Praesepe-based gyrochronology model (black dotted line) at young
+ages.
+However, this does not seem to be true age old ages where it appears that
+rotation period {\it decreases} with decreasing effective temperatures.
+This is a paradigm shift for gyrochronology because stellar spin-down rate is
+thought to be directly tied to magnetic field strength, and the deeper
+convection zones of cooler stars generate stronger magnetic fields which {\it
+should} lead to more efficient angular momentum loss.
+% The shape of the upper envelope of rotation periods in the top panels of
+% figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff} also hints at
+% this.
+The shape of the period-\teff\ relations at old ages appears to follow the
+shape of the upper detection edge, with the so-called `M dwarf dip'
+\citep{vansaders2018} reflected in the lines of constant velocity dispersion
+(and presumed age) in the top panel of figure
+\ref{fig:dispersion_period_teff}.
+If the shape of that upper edge is created by a detection limit, our results
+suggest that stars cooler than $\sim$4500 K become inactive (and their
+rotation periods therefore become undetectable) at around the same age, but at
+an older age than stars hotter than $\sim$4500.
+The (kinematically) oldest stars in the \mct\ sample are cooler than 4500 K.
+This is probably because cooler stars remain active for longer, and their
+rotation periods can therefore be measured at older ages.
 
-Figure \ref{fig:dispersion_period_teff} also shows that the (kinematically)
-oldest stars with measured rotation periods are cooler than 4500 K.
-This may be because cooler stars remain active for longer, and their rotation
-periods can therefore be measured at older ages.
-
-% Unfortunately, we still cannot rule out the possibility that mass-dependent
-% heating is acting on these stars which could be responsible for some, if not
-% all, of the increased velocity dispersion at cooler effective temperatures.
-
-\subsection{The period gap}
-An unsolved mystery within gyrochronology is the origin of the rotation period
-gap, first identified by \citet{mcquillan2013}.
-This gap can be seen in figure \ref{fig:age_cut} as an under-density of points
-between the 0.7-1.0 and 1.0-1.5 Gyr age ranges and roughly follows a line of
-constant gyrochronal age of around 1.1 Gyr \citep[according to
-the gyrochronology relation of][]{angus2019}.
-Several explanations for its origin have been proposed, including a
-discontinuous star formation history \citep{mcquillan2013, davenport2017,
-davenport2018}, a rapid change in magnetic field structure
-\citep{reinhold2019}, and erroneous rotation period measurements that are
-incorrect by a factor of two \citep{koen2018}.
-The latter explanation can be ruled out because stars below the gap have
-smaller velocity dispersions than the stars above the gap, indicating that
-they are kinematically younger \citep{mcquillan2013, davenport2018}, as
-evident in figure \ref{fig:age_cut}.
-For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
-\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
-temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
-velocity dispersion increases with \teff.
-The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
-dispersion as the hottest stars in the age range above which indicates that
-the period-\teff\ relations are {\it flat} at these rotation periods.
-It may be a coincidence that the gyrochronology relations seem to only flatten
-off {\it above} the period gap, or it may be that the period gap is located at
-magnetic transition period.
-We lack a sufficient quantity of data to investigate in detail but new field
-star rotation periods from the \ktwo\ and \tess\ missions may be able to
-validate or rule out this hypothesis in the future.
+We reiterate that the velocity dispersions of the coolest stars may be
+affected by a selection bias -- these extremely faint stars are more difficult
+to detect at larger distances, larger heights above the plane, and therefore
+larger velocities.
+It is possible that some high velocity stars of temperatures cooler than
+$\sim$4000 K are missing from this sample, and the velocity dispersions may
+therefore appear lower than they truly are.
+This could be why the velocity dispersion appears to decrease towards the
+right of figure \ref{fig:dispersion_period_teff}.

### 2019-08-30 paper/intro.tex
adding some discussion about ages on the CMD and adding structural organization of paper

@@
 \ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
-through early M, makes it the ideal cluster for calibrating the period-color
+through early M, makes it a good cluster for calibrating the period-color
 relation of stars at 650 Myrs.
@@
 than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
-evolution of middle-aged K dwarfs is a mystery\footnote{The oldest M dwarfs
-with reliable ages and rotation periods are members of Praesepe ($\sim$ 650
-Myrs), so even less is known about their rotational evolution.}.
+evolution of middle-aged K dwarfs is difficult to characterize\footnote{The
+oldest M dwarfs with reliable ages and rotation periods are members of
+Praesepe ($\sim$ 650 Myrs), so even less is known about their rotational
+evolution.}.
+
+The gyrochronal ages of the stars in our sample were calculated with the
+\citep{angus2019} gyrochronology model, using de-reddened \Gaia\ \gcolor\
+color and rotation periods reported by \mct.
+These rotational ages are shown on the CMD in figure \ref{fig:age_gradient}.
+The stars with old rotational ages, plotted in yellow hues, predominantly lie
+along the upper edge of the MS, where stellar evolution models predict old
+stars to be, however the majority of stars with old rotational ages are bluer
+than \gcolor\ $\sim$ 1.5 dex.
+This provides a first hint that the \citet{angus2019} gyrochronology relation
+under-predicts the ages of low-mass stars.
+There is no reason to expect the oldest stars in this sample to be the bluer
+ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
+active for longer, so should therefore have measurable rotation periods at
+older ages.
+Since the \citet{angus2019} gyrochronology model, which is based on the
+period-color relation of Praesepe, does not predict old ages for any M dwarfs,
+it is probably under-predicting the ages of some of these low-mass stars.
+It is therefore likely that old M dwarfs rotate more rapidly than this
+Praesepe-based model predicts.
+In what follows we use a population-based stellar age indicator, velocity
+dispersion, to investigate the period-color relations at old ages in the
+field.
+
 % In this paper, we shed light on the rotational evolution of old K dwarfs using
@@
 % old ages.
-For this reason, we used a population-based stellar age indicator, velocity
-dispersion, to investigate the period-color relations at old ages in the
-field.
+% For this reason, we used a population-based stellar age indicator, velocity
+% dispersion, to investigate the period-color relations at old ages in the
+% field.
 
@@
 % groups have a larger velocity dispersion.
+
+This paper is laid out as follows: in section \ref{sec:the_data} we describe
+our sample selection process and the methods used to calculated stellar
+velocities.
+In section \ref{sec:results} we use stellar kinematics to investigate the
+relationship between rotation period, age and color/\teff\ in field stars.
+In section \ref{sec:mass-dependent-heating} we show that neither
+mass-dependent heating nor the selection function is likely to have a strong
+affect on our sample.
+Finally, in section \ref{sec:period_gap} we discuss the implications of our
+results.

### 2019-08-30 paper/abstract.tex
some fairly small edits and rewrites. adding some references
moving comments to comments

@@
 main-sequence stars.
-% However, recent evidence suggests that the relationship between rotation
-% period and color/temperature/mass does not remain constant over large
-% timescales, contradicting a long-held assumption that underlies many
-% gyrochronology relations.
-In this work we use the velocities of populations of K dwarfs to reveal the
-rotational evolution of these cool stars at old ages.
+However, recent evidence suggests that the relationship between rotation
+period and color/temperature/mass does not remain constant over large
+timescales, contradicting a long-held assumption that underlies many
+gyrochronology relations.
+We use the velocities of populations of cool dwarfs to reveal the rotational
+evolution of these stars at old ages.
 We find that groups of stars, selected to be the same age using a
@@
 within K and M dwarfs in the \kepler\ field.
-
-% We show however, that if mass-dependent heating is responsible, the
-% differential heating rate is not constant over logarithmic time intervals
-% which implies that different heating processes must be acting before and after
-% ages of around 1 Gyr.
-% Perhaps a more likely scenario is that the relation between rotation period
-% and color changes over time, undergoing a particularly sharp transition at
-% around 1 Gyr.
-% If we assume that mass-dependent heating does not affect the data,
-
-% (similar to the age of the rotation gap and the 1.1 Gyr NGC 6811
-% cluster, shown to have a different period-color relation to Praesepe).
-% We find that the ages of field stars with measured rotation periods do not
-% exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
-% braking efficiency.
-% We also show that stars rotating just below the rotation period gap are
-% dynamically young, ruling out the possibility that the rotation period gap is
-% caused by incorrect period measurements or binary companions.

### 2019-08-30 paper/intro.tex
moving comments to comments

@@
 and the G dwarfs rotate at the same rate as the K dwarfs.
-% Since a separable period-age and period-color relation provided an adequate
-% fit for the young stars, it was naturally assumed to apply to old stars too.
 
@@
 relation of stars at 650 Myrs.
-% This relation was calibrated by fitting a 5th-order polynomial to the relation
-% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
-% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
-% and the Sun.
 This relation accurately describes the rotation periods of F and G stars up to
@@
 In what follows we use a population-based stellar age indicator, velocity
-dispersion, to investigate the period-color relations at old ages in the
-field.
-
-% In this paper, we shed light on the rotational evolution of old K dwarfs using
-% population-based dynamical ages.
-% Unfortunately, open clusters with rotation period measurements are mostly
-% young -- currently the oldest is 2.5 Gyr (NGC 6819).
-% Without rotation periods for precisely dated old stars, it is extremely
-% difficult to calibrate the relationship between rotation period and color at
-% old ages.
-% For this reason, we used a population-based stellar age indicator, velocity
-% dispersion, to investigate the period-color relations at old ages in the
-% field.
+dispersion, to further investigate the gyrochronology relations at old ages in
+the field.
 
@@
 \citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
-% because the MW's gas disk is thin and its gas moves in circular
-% orbits.
 Young stars have relatively small vertical velocities, but gain momentum in
@@
 on the implied relative ages of populations.
-% Stellar velocities have a history of being used as an age proxy, with several
-% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
-% west2011}.
 The age-velocity dispersion relations (AVRs) are still actively being
@@
 
-% Vertical {\it actions} are better age indicators than velocities, because
-% actions are calculated by integrating angular momentum over the Milky Way's
-% potential, and are therefore position invariant -- \eg\ a star will have the
-% same action at periapsis and apoapsis.
-% In contrast, orbital velocities are different at periapsis and apoapsis -- so
-% in this sense, {\it actions} are the natural quantities to use as age proxies.
-Vertical action is a better age indicator than vertical velocity
+Vertical action is thought to be a better age indicator than vertical velocity
 \citep{ting2019}, although still only a weak age indicator for individual
@@
 
-% \subsection{The degeneracy between gyrochronology and mass-dependent heating}
-
-% Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
-% Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
-% members of NGC 6819, a 2.5 Gyr open cluster.
-% However there are no rotation periods for K or M dwarfs in this cluster -- its
-% coolest members with rotation periods are G dwarfs.
-% The period-color relation of Praesepe is nearly identical to the period-color
-% relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
-% \citep{douglas2016, douglas2017, rebull2017}.
-% However, Praesepe and the Hyades do {\it not} have the same period-color
-% relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
-% The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
-% appears as though the K stars have `stalled' -- their spin-down has been
-% halted.
-
-% The \citet{angus2019} gyrochronology relation assumes that the rotation
-% period-color relation of Praesepe is applicable to stars of all ages: the same
-% polynomial relation fit to Praesepe is used to describe the period-color
-% relation for all stars.
-% However, the NGC 6811 cluster suggests that this assumption does not hold for
-% K dwarfs past around 1 Gyr.
-% If the period-color relation of the \citet{angus2019} gyrochronology model
-% {\it were} a perfect model for the rotational evolution of stars, then groups
-% of stars selected to be similar ages using this relation should fall on an
-% isochrone on a CMD.
-% % and have the same velocity dispersion across all colors.
-% Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
-% variations in metallicity and extinction, blur out the main sequence enough
-% that differences between stars in different age and rotation bins are not
-% easily discernable on the CMD, although there is still a general age and
-% rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
-% \ref{fig:age_gradient}.
-% For this reason, we chose to use {\it kinematic} `isochrones', instead of
-% magnitude and color-based isochrones: although kinematics as an age indicator
-% is not necessarily as well calibrated to an absolute age scale as CMD
-% position, it can be very sensitive to {\it differences} in the ages of stellar
-% populations.
-
-% Isochrones and stellar evolution tracks are highly dependent on choices made
-% about input physics and assumptions.
-% % and have often been calibrated using
-% % different types of stars.
-% As a result, different sets of models can have very different shapes on the
-% CMD, particularly at low masses.
-% In fact, only empirical models, not physical ones, are currently able to
-% reproduce the CMD positions of M dwarfs.
-% Instead of relying on CMD position to age-date groups of stars, we opted to
-% explore age trends via kinematics.
-% Kinematic age-dating has the advantage of being relatively model independent,
-% or at least, having a very simple model: that velocity dispersion increases
-% over time.
-% This means that it is relatively easy to rank groups of stars by age: older
-% groups have a larger velocity dispersion.
-
 This paper is laid out as follows: in section \ref{sec:the_data} we describe
@@
 In section \ref{sec:results} we use stellar kinematics to investigate the
-relationship between rotation period, age and color/\teff\ in field stars.
+relationship between rotation period, age and color/\teff\ in the field.
 In section \ref{sec:mass-dependent-heating} we show that neither
 mass-dependent heating nor the selection function is likely to have a strong
-affect on our sample.
-Finally, in section \ref{sec:period_gap} we discuss the implications of our
-results.
+affect on our sample, and in section \ref{sec:period_gap} we discuss the
+implications of our results.

### 2019-08-30 paper/method.tex
moving comments to comments

@@
 
-% \begin{figure}
-%   \caption{
-% A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
-%     extinction-corrected magnitudes, colored by rotation period.
-% We excluded photometric binaries and subgiants from our analysis by removing
-% stars above the two dashed lines.
-% % The rotation periods of binaries and subgiants do not follow a Skumanich-like
-% % braking law.
-% % The rotation period gradient across the main sequence is visible by eye in
-% %     this figure: young, rapidly rotating stars are located below the old,
-% %     slowly rotating stars.
-% % since we rely on a simple scaling
-% % between rotation period and age to make the argument that the rotation period
-% % is not caused by incorrect period measurements.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{CMD_cuts}
-% \label{fig:CMD_cuts}
-% \end{figure}
-% To explore the age of this stellar population from a rotation standpoint, it
-% was first necessary to remove visual binaries and subgiants from the sample.
-% The rotational evolution of these two types of stars is generally different to
-% that of single stars which more usually follow a Skumanich-like spin-down law.
-% Tidal and magnetic interactions between the two components of a binary system
-% can influence the rotation periods of both stars, and the expanding envelopes
-% of subgiants drive rapid spin-down through conservation of angular momentum.
 Visual binaries and subgiants were removed from the sample by applying cuts to
@@
 
-% \begin{figure}
-%   \caption{
-% the rotation periods of stars in the \citet{mcquillan2014} sample vs.
-% effective temperature.
-% black points are dwarf stars identified as non-photometric binaries, hotter
-% than 4800 k.
-% blue circle points are non-photometric binary dwarfs, cooler than 4800 k, with
-% a rotation period and \gaia\ color indicating they are older than 1.1 gyr.
-% orange squares are stars that with rotation periods that fall just below the
-% gap: they have rotation-ages between 0.5 and 1.1 gyrs.
-% green triangles are stars with rotation periods faster than the main envelop
-% of stars.
-% these are probably binaries whose rotation periods are synchronized to their
-% orbits and have been spun-up via tidal interactions.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{period_teff}
-% \label{fig:period_teff}
-% \end{figure}
+\subsection{Establishing \vb\ as an age proxy}
+\label{sec:mass-dependent-heating}
+
+If lower-mass stars experience greater velocity changes when gravitationally
+perturbed and are dynamically heated more efficiently than higher-mass stars,
+that means velocity dispersion is a function of {\it both} age and mass and
+cannot be straightforwardly interpreted as an age proxy.
+Mass-dependent dynamical heating has not been unambiguously observed in the
+galactic disk because of the strong anti-correlation between stellar mass and
+stellar age.
+Less massive stars do indeed have larger velocity dispersions, however they
+are also older on average.
+This mass-age degeneracy is highly reduced in M dwarfs because their
+main-sequence lifetimes are longer than the age of the Universe, however no
+evidence for mass-dependent heating has been found in M dwarfs
+\citep{faherty2009}.
+
+To investigate whether mass-dependent heating could be acting on the \kepler\
+sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
+whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
+the initial mass function.
+We could not perform this analysis on the \mct\ sample, because only stars
+with {\it detectable} rotation periods appear there.
+Since lower-mass stars stay active for longer it is likely that there are more
+old stars (with larger velocity dispersions) at low masses in this sample.
+We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
+greater than 1.2 (corresponding to an effective temperature $\lesssim$
+4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+We also eliminated visual binaries by removing stars above a 6th order
+polynomial, fit to the MS on the \gaia\ CMD.
+We then applied the quality cuts described in section \ref{sec:the_data}.
+To search for evidence of mass-dependent heating we calculated the (\vb)
+velocity dispersion of stars in effective temperature bins.
+Again, sigma clipping was used to remove high velocity outliers before
+calculating the standard deviation of stars in each bin.
+
+Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
+as a function of effective temperature (calculated by transforming dereddened
+\gaia\ colors using equation \ref{eq:curtis}).
+Velocity dispersion very slightly {\it decreases} with decreasing temperature,
+(the opposite of the trend expected for mass-dependent heating) however the
+slope is only inconsistent with zero at the 1.3 $\sigma$ level.
+This trend may be due to a selection bias: cooler stars are fainter and
+therefore typically closer, with smaller heights above the galactic plane and
+smaller velocities.
+The essential point here is that we do not find  evidence for mass-dependent
+heating acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be
+used as an age proxy.
+We performed the same analysis on the 537 stars in this sample with RVs using
+{\it vertical} velocity (\vz) and found the slope of the velocity
+dispersion-temperature relation was consistent with zero.
+
+Since we used \vb, not \vz\ in our analysis, it is possible that an
+anti-correlation between \teff\ and galactic latitude, $b$, could influence
+our results: stars at higher latitudes have additional velocity components in
+the ${\bf x}$ and ${\bf y}$ directions, which could increase \vb\ but not \vz.
+Again however, since the relationship between \sigmavb\ and \teff\ is
+positively, not negatively correlated for cool stars in the \kepler\ field,
+this effect is probably too small to influence our results.
+
+\begin{figure}
+  \caption{
+      Top: Stellar velocity (\vb) as a function of \teff\ for
+      \kepler\ K and M dwarfs.
+Vertical lines indicate different \teff-groupings used to calculate velocity
+    dispersion.
+Pink stars were not included in velocity dispersion calculations as they were
+    either removed as outliers during a sigma clipping process, or they lie at
+    the sparcely populated, extremely cool end of the temperature range.
+    Velocity dispersion and \teff\ are slightly positively correlated, likely
+    due to a brightness-related selection bias, indicating that mass-dependent
+    heating does not significantly affect low-mass stars in the \kepler\
+    field.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{vb_vs_teff}
+\label{fig:vb_vs_teff}
+\end{figure}
+Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+strongly affect the \mct\ sample of rotating \kepler\ stars.
+For this reason, we assume that age difference is the major cause of velocity
+dispersion differences between groups.
+In other words, (\vb) velocity dispersion is a reliable age proxy for the
+\mct\ sample.

### 2019-08-30 paper/results.tex
moving comments to comments

@@
 % word limit: 500
-\section{Results}
+\section{Results and discussion}
 \label{sec:results}
@@
 $\log_{10}$(period) and temperature.
-If we assume that mass dependent heating doesn't affect this sample, and \vb\
-at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
+If we assume that mass dependent heating does not strongly affect this sample,
+and \vb\ at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
 dispersion can be interpreted as an age proxy and stars plotted in a similar
@@
 Interpreted this way, lines of constant age (isochrones) appear to follow the
-shape of the Praesepe-based gyrochronology model (black dotted line) at young
-ages.
-However, this does not seem to be true age old ages where it appears that
-rotation period {\it decreases} with decreasing effective temperatures.
+shape of the Praesepe-based gyrochronology model (black solid and dotted
+lines) at young ages.
+However, at old ages it appears that rotation period {\it decreases} with
+decreasing effective temperatures.
 This is a paradigm shift for gyrochronology because stellar spin-down rate is
@@
 should} lead to more efficient angular momentum loss.
-% The shape of the upper envelope of rotation periods in the top panels of
-% figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff} also hints at
-% this.
 The shape of the period-\teff\ relations at old ages appears to follow the
@@
 The (kinematically) oldest stars in the \mct\ sample are cooler than 4500 K.
-This is probably because cooler stars remain active for longer, and their
+This is presumably because cooler stars remain active for longer, and their
 rotation periods can therefore be measured at older ages.
 
-We reiterate that the velocity dispersions of the coolest stars may be
-affected by a selection bias -- these extremely faint stars are more difficult
-to detect at larger distances, larger heights above the plane, and therefore
-larger velocities.
+The velocity dispersions of the coolest stars may be affected by a selection
+bias -- these extremely faint stars are more difficult to detect at larger
+distances, larger heights above the plane, and therefore larger velocities.
 It is possible that some high velocity stars of temperatures cooler than
@@
 right of figure \ref{fig:dispersion_period_teff}.
+
+\subsection{The period gap}
+\label{sec:period_gap}
+
+The origin of the rotation period gap, first identified
+by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+\ref{fig:dispersion_period_teff} still remains a mystery.
+This gap can be seen as an under-density of points between the 0.7-1.0 and
+1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
+of constant gyrochronal age of around 1.1 Gyr \citep[according to the
+gyrochronology relation of][]{angus2019}, as shown in figure
+\ref{fig:dispersion_period_teff}.
+Several explanations for the gap's origin have been proposed, including a
+discontinuous star formation history \citep{mcquillan2013, davenport2017,
+davenport2018}, a rapid change in magnetic field structure
+\citep{reinhold2019}, and erroneous rotation period measurements that are
+incorrect by a factor of two \citep{koen2018}.
+The latter explanation can be ruled out because stars below the gap have
+smaller velocity dispersions than the stars above the gap, indicating that
+they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
+For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+velocity dispersion increases with \teff.
+The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+dispersion as the hottest stars in the age range above which indicates that
+the period-\teff\ relations are {\it flat} at these rotation periods.
+This is also visible in figure \ref{fig:dispersion_period_teff}.
+Below the gap, velocity dispersion within a given period range appears to {\it
+decrease} with decreasing temperature.
+The opposite appears to be true above the gap.
+
+In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
+are faster than expected; their rotational evolution appears to have stalled,
+and the period-\teff\ relation is flat \citep{curtis2019}.
+The rotation periods of the K dwarfs in this cluster are plotted in figure
+\ref{fig:dispersion_period_teff}.
+Although the G dwarfs in this cluster fall on the 1.1 Gyr gyrochronology
+model, the K dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
+NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
+K dwarfs lie below it.
+This cluster may be the `missing link' that connects two epoch of stellar
+spin-down: an early stage where the period-\teff\ relation for K dwarfs has a
+negative slope and a late stage where it has a positive slope.
+In this case the period gap may delineate the transition between these two
+regimes and is the point at which stellar magnetic dynamos likely undergo a
+dramatic structural shift at an age of $\sim$ 1.1 Gyr.
+
+It may be a coincidence that the gyrochronology relations seem to only flatten
+off {\it above} the period gap, or it may be that the period gap is located at
+magnetic transition period.
+We lack a sufficient quantity of data to investigate in detail but new field
+star rotation periods from the \ktwo\ and \tess\ missions may be able to
+validate or rule out this hypothesis in the future.

### 2019-09-03 paper/results.tex
small edits and moving period gap section back to discussion

--- 2019-08-30 paper/results.tex
@@
 \subsection{Velocity dispersion of coeval groups}
+\label{sec:age_cut}
 
 To explore the relationship between rotation period, \teff\ and age, we
-selected groups of stars within different age ranges \citep[where age was
-calculated using the][gyrochronology relation]{angus2019}, and calculated the
-velocity dispersion: the standard deviation of velocities in the direction of
-increasing galactic latitude, \sigmavb, as a function of effective temperature
-for each age group.
+selected groups of stars within different gyrochronal age ranges \citep[where
+age was calculated using the][gyrochronology relation]{angus2019}, and
+calculated the velocity dispersion (\sigmavb), the standard deviation of \vb\
+velocities, as a function of effective temperature for each age group.
 We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
@@
 temperature, \teff.
-We chose to use \teff and not photometric color in this analysis as it is the
-linear quantity and therefore easier to divide into bins of roughly equal
-numbers of stars.
+We chose to use \teff\ as the abscissa because it is easier to divide stars
+into bins of roughly equal numbers in \teff-space than in color-space.
 
@@
     \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
-    the the ages of late-K dwarfs relative to the ages of early K dwarfs at
-    old ages.
+    the the ages of late-K dwarfs relative to the ages of late G and early K
+    dwarfs at old ages.
 % An alternative explanation could be that the gyrochronology relation is
@@
 % \citep{vansaders2016}, which could bias the results.
-Late M dwarfs were not included in our analysis because such faint stars
-cannot be observed at large heights above the plane (because of the low
-galactic latitude of the \kepler\ field, stars at high-Z are more distant),
-which introduces a mass-dependent velocity bias: cooler populations of stars
-are skewed towards lower velocity dispersions.
-The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
-have low velocity dispersions, indicating that this effect may already become
-important at temperatures lower than $\sim$ 4000 K.
-
+% Late M dwarfs were not included in our analysis because such faint stars
+% cannot be observed at large heights above the plane (because of the low
+% galactic latitude of the \kepler\ field, stars at high-Z are more distant),
+% which introduces a mass-dependent velocity bias: cooler populations of stars
+% are skewed towards lower velocity dispersions.
+% The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
+% have low velocity dispersions, indicating that this effect may already become
+% important at temperatures lower than $\sim$ 4000 K.
 Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
@@
 the relative ages of {\it young} field stars.
-% To our knowledge, no gyrochronology relation has never been demonstrated to
+% To our knowledge, no gyrochronology relation has previously been demonstrated to
 % correctly predict ages (either relative or absolute) for such cool or such
@@
 temperature for old stars, meaning the \citet{angus2019} gyrochronology
-relation underpredicts the ages of old, late-K dwarfs (or overpredicts the
-ages of old early-K dwarfs).
-This indicates that the relationship between period and photometric
-color or \teff\ flattens out over time.
-In other words, the lines of constant age sweeping diagonally upwards in the
-top panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
+relation either under-predicts the ages of old, late-K dwarfs, or
+over-predicts the ages of old early-K and late-G dwarfs\footnote{Since \vb\
+velocity dispersion only provides relative and not absolute ages, it is
+difficult to tell whether the ages of cool stars are being under-predicted,
+the ages of hot stars being are over-predicted, or both.}.
+This indicates that the relationship between rotation period and photometric
+color or \teff\ flattens out over time, and possibly even inverts.
+The lines of constant age (isochrones) sweeping diagonally upwards in the top
+panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
 
 \subsection{The period-\teff\ relations, revealed}
+\label{sec:the_reveal}
 
@@
       Top: Rotation period vs effective temperature for stars in the \mct\
-    sample, colored by the velocity dispersions of stars over a grid in
-    $\log_{10}$(period) and \teff.
+    sample, colored by the velocity dispersions of stars calculated over a
+    grid in $\log_{10}$(period) and \teff.
     Bottom: the velocity dispersions of groups of stars, shown as a solid
     grid for clarity.
+    The hatched area indicates the temperature regime where selection biases
+    could play an important role, so these velocity dispersions should be
+    interpreted with caution.
     The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
-    with the \citet{angus2019} gyrochronology relation which roughly traces
+    with the \citet{angus2019} gyrochronology relation, which roughly traces
     the rotation period gap.
@@
 Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
-our sample, coloured by (\vb) velocity dispersion.
-Velocity dispersion was calculated for groups of stars over a grid in
-$\log_{10}$(period) and temperature.
-If we assume that mass dependent heating doesn't affect this sample, and \vb\
-at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
-dispersion can be interpreted as an age proxy and stars plotted in a similar
-color in figure \ref{fig:dispersion_period_teff} are similar ages.
+our sample, coloured by (\vb) velocity dispersion, where \sigmavb\ was
+calculated for groups of stars over a grid in $\log_{10}$(period) and
+temperature.
+If we assume that mass dependent heating does not strongly affect this sample
+and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
+velocity dispersion can be interpreted as an age proxy and stars plotted in a
+similar color in figure \ref{fig:dispersion_period_teff} are similar ages.
 Interpreted this way, lines of constant age (isochrones) appear to follow the
-shape of the Praesepe-based gyrochronology model (black dotted line) at young
-ages.
-However, this does not seem to be true age old ages where it appears that
-rotation period {\it decreases} with decreasing effective temperatures.
-This is a paradigm shift for gyrochronology because stellar spin-down rate is
-thought to be directly tied to magnetic field strength, and the deeper
-convection zones of cooler stars generate stronger magnetic fields which {\it
-should} lead to more efficient angular momentum loss.
-% The shape of the upper envelope of rotation periods in the top panels of
-% figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff} also hints at
-% this.
+shape of the Praesepe-based gyrochronology model (black solid and dotted
+lines) at young ages.
+However, at older ages it appears that the relation between rotation period
+and \teff\ flattens out, until eventually rotation period {\it decreases} with
+decreasing effective temperature at a given age.
+
 The shape of the period-\teff\ relations at old ages appears to follow the
-shape of the upper detection edge, with the so-called `M dwarf dip'
-\citep{vansaders2018} reflected in the lines of constant velocity dispersion
-(and presumed age) in the top panel of figure
+shape of the upper detection edge.
+The so-called `M dwarf dip' \citep{vansaders2018} is reflected in the lines of
+constant velocity dispersion (and presumed age) in the top panel of figure
 \ref{fig:dispersion_period_teff}.
-If the shape of that upper edge is created by a detection limit, our results
-suggest that stars cooler than $\sim$4500 K become inactive (and their
-rotation periods therefore become undetectable) at around the same age, but at
-an older age than stars hotter than $\sim$4500.
-The (kinematically) oldest stars in the \mct\ sample are cooler than 4500 K.
-This is probably because cooler stars remain active for longer, and their
-rotation periods can therefore be measured at older ages.
+If the shape of the upper edge of rotation period measurements is created by a
+detection limit, it could indicate the rotation periods at which stars of
+different temperatures become relatively inactive (and their rotation periods
+therefore become undetectable).
+% Once inactive, stars would have little photometric variability induced by
+% star spots and their rotation periods would be difficult to measure, so may
+% not feature in the \mct\ catalog.
+If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
+$\sim$4500 K become inactive at a similar age, and stay active longer than
+stars hotter than $\sim$4500 on average.
 
-We reiterate that the velocity dispersions of the coolest stars may be
-affected by a selection bias -- these extremely faint stars are more difficult
-to detect at larger distances, larger heights above the plane, and therefore
-larger velocities.
+% The groups of stars with the largest velocity dispersions are cooler than 4500
+% K, presumably because cooler stars remain active for longer, and their
+% rotation periods can therefore be measured at older ages.
+
+The velocity dispersions of the coolest stars may be affected by a selection
+bias -- these extremely faint stars are more difficult to detect at larger
+distances, larger heights above the plane, and therefore larger velocities.
 It is possible that some high velocity stars of temperatures cooler than
@@
 right of figure \ref{fig:dispersion_period_teff}.
+
+% \subsection{The period gap}
+% \label{sec:period_gap}
+
+% The origin of the rotation period gap, first identified
+% by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+% \ref{fig:dispersion_period_teff} still remains a mystery.
+% This gap can be seen as an under-density of points between the 0.7-1.0 and
+% 1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
+% of constant gyrochronal age of around 1.1 Gyr \citep[according to the
+% gyrochronology relation of][]{angus2019}, as shown in figure
+% \ref{fig:dispersion_period_teff}.
+% Several explanations for the gap's origin have been proposed, including a
+% discontinuous star formation history \citep{mcquillan2013, davenport2017,
+% davenport2018}, a rapid change in magnetic field structure
+% \citep{reinhold2019}, and erroneous rotation period measurements that are
+% incorrect by a factor of two \citep{koen2018}.
+% The latter explanation can be ruled out because stars below the gap have
+% smaller velocity dispersions than the stars above the gap, indicating that
+% they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+% evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
+% For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+% \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+% temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+% velocity dispersion increases with \teff.
+% The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+% dispersion as the hottest stars in the age range above which indicates that
+% the period-\teff\ relations are {\it flat} at these rotation periods.
+% This is also visible in figure \ref{fig:dispersion_period_teff}.
+% Below the gap, velocity dispersion within a given period range appears to {\it
+% decrease} with decreasing temperature.
+% The opposite appears to be true above the gap.
+% It could be that the gap is positioned at a significant Rossby number/age at
+% which stellar magnetic dynamos go through a transition.
+% Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
+% magnetic braking is more efficient for stars with deeper convection zones.
+% Once stars reach this critical age or Rossby number their magnetic fields
+% undergo some radical transition, which produces the gap in the rotation
+% period-\teff\ plane.
+% After this transition, magnetic braking efficiency no longer increases with
+% decreasing mass.
+% Of course, it may be a coincidence that the gyrochronology relations seem to
+% only flatten off above the period gap and we lack a sufficient quantity of
+% data to do more than speculate here.
+% New rotation periods from the \ktwo\ and \tess\ missions may be able to
+% validate or rule out this hypothesis in the future.
+
+% In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
+% are faster than expected; their rotational evolution appears to have stalled,
+% and they have similar rotation periods to the 650 Myr Praesepe cluster
+% \citep{curtis2019}.
+% The rotation periods of the K dwarfs in this cluster are plotted in figure
+% \ref{fig:dispersion_period_teff}.
+% Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
+% dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
+% NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
+% K dwarfs lie below it.
+% This cluster may be the `missing link' that connects two epoch of stellar
+% spin-down.
+% % , however without more observations of middle-aged open clusters it
+% % : an early stage where the period-\teff\ relation for cool dwarfs has
+% % a negative slope and a late stage where it has a positive slope.
+% % In this case the period gap may delineate the transition between these two
+% % regimes and is the point at which stellar magnetic dynamos likely undergo a
+% % dramatic structural shift at an age of $\sim$ 1.1 Gyr.

### 2019-09-05 paper/acknowledgements.tex
somechanges

@@
-This work was partly developed at the KITP conference 'Better stars, better
-planets'.
+This work was partly developed at the 2019 KITP conference 'Better stars,
+better planets'.
+Parts of this project are based on ideas explored at the Gaia sprints at the
+Flatiron Institute in New York City, 2016 and MPIA, Heidelberg, 2017.
+
+This work made use of the gaia-kepler.fun crossmatch database created by Megan
+Bedell.
 
@@
 
-This work made use of the gaia-kepler.fun crossmatch database created by Megan
-Bedell
-
 This work has made use of data from the European Space Agency (ESA) mission
-{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
-Data Processing and Analysis Consortium (DPAC,
-\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
-has been provided by national institutions, in particular the institutions
-participating in the {\it Gaia} Multilateral Agreement.
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it
+Gaia} Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
+Funding for the DPAC has been provided by national institutions, in particular
+the institutions participating in the {\it Gaia} Multilateral Agreement.

### 2019-09-05 paper/conclusion.tex
moving comments to comments
somechanges

--- 2019-08-01 paper/conclusion.tex
@@
 
-We found that old groups of K dwarfs selected to be coeval using the
+We examined the gyrochronology relations using the velocity dispersions of
+field stars with measured rotation periods.
+We found that old groups of cool dwarfs selected to be coeval using the
 \citet{angus2019} gyrochronology relation do {\it not} have the same velocity
 dispersion across all temperatures.
-This indicates that either the \citep{angus2019} period-color relation is
-wrong, or that mass-dependent dynamical heating is affecting this data set.
-We presented three pieces of evidence that suggest mass-dependent heating
-cannot be responsible for all of the increasing velocity dispersion.
+This implies that the \citet{angus2019} relation, which is based on the
+period-color relation of the 650 Myr Praesepe cluster, does not correctly
+describe the period-age-color/\teff\ relation for old stars.
+It appears that the period-color/\teff\ relation changes shape over time, \ie\
+the period-color/\teff\ relation and the period-age relation are
+non-separable.
+At ages younger than $\sim$1 Gyr, rotation period is anti-correlated with
+\teff: cooler stars spin more slowly than hotter stars of the same age.
+However, at old ages the opposite is true: cooler stars spin more rapidly than
+hotter stars of the same age.
+The period-color/\teff\ relation appears to start flattening out after $\sim$1
+Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
+older than the period gap which is located at a gyrochronal age of around
+$\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
+This gap may represent a significant transitional epoch in the magnetic
+behavior of stars.
+Finally, when velocity dispersion is interpreted as an age proxy, it appears
+that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
+suggests that lower-mass stars remain active for longer and their rotation
+periods can be measured at older ages.
+We outlined a number of scenarios which could provide alternative
+explainations for these observations, including incorrect dust corrections for
+the lowest-mass stars and an excess of companions increasing the velocity
+dispersion for these stars.
 
-\begin{itemize}
-
-\item{
-        No strong evidence for mass-dependent heating has been found among M, L
-and T-type stars.
-These low-mass stars live longer than the current age of the Universe and
-        therefore do not suffer from a mass-age degeneracy
-        \citep{faherty2009}.
-}
-
-\item{
-        Literature measurements of the (\vz) AVR exponent are slightly higher
-than the value measured from our data set, but calculated using more massive
-stars.
-Mass-dependent dynamical heating should however produce a {\it smaller} AVR
-        exponent for higher-mass stars.
-We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
-        subject to different selection biases than those used to calculate
-        AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
-        drawing direct comparisons between them does not provide a reliable
-        test.
-}
-
-\item{
-        The differences in the (\vb) AVR exponent between the hottest and
-coolest K dwarfs in our sample are extremely large, spanning values between
-0.4 and 0.75, yet the difference in mass is less than a factor of two.
-Such a large spread in AVR exponents cannot be explained by mass-dependent
-        heating alone.
-}
-
-\end{itemize}
-
-\racomment{Words about the significance of an inverted period-color relation.}
+If the period-color/\teff\ relation does invert at old ages as our results
+suggest, this would be a paradigm shift for gyrochronology.
+Stellar spin-down rate is thought to be directly tied to magnetic field
+strength, and the deeper convection zones of cooler stars generate stronger
+magnetic fields which {\it should} lead to more efficient angular momentum
+loss.
+However, the micro- and macro-physics of stellar structure and evolution and
+magnetic dynamo models are extremely complicated and a lot is still unknown
+about the magnetic behavior of stars.
+Observations like these can provide useful constraints for physical models,
+and may help to reveal new physical processes at work in stars like our own
+Sun and other planet hosts.

### 2019-09-05 paper/discussion.tex
somechanges

--- 2019-07-23 paper/discussion.tex
@@
 \section{Discussion}
+\label{sec:discussion}
 
-What if the M dwarfs just have larger velocity dispersion because they're less
-massive?
+\subsection{The period gap}
+\label{sec:period_gap}
+
+The origin of the rotation period gap, first identified
+by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+\ref{fig:dispersion_period_teff} still remains a mystery.
+This gap can be seen as an under-density of points between the 0.7-1.0 and
+1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
+of constant gyrochronal age of around 1.1 Gyr \citep[according to the
+gyrochronology relation of][]{angus2019}, as shown in figure
+\ref{fig:dispersion_period_teff}.
+Several explanations for the gap's origin have been proposed, including a
+discontinuous star formation history \citep{mcquillan2013, davenport2017,
+davenport2018}, a rapid change in magnetic field structure
+\citep{reinhold2019}, and erroneous rotation period measurements that are
+incorrect by a factor of two \citep{koen2018}.
+The latter explanation can be ruled out because stars below the gap have
+smaller velocity dispersions than the stars above the gap, indicating that
+they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
+For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+velocity dispersion increases with \teff.
+The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+dispersion as the hottest stars in the age range above which indicates that
+the period-\teff\ relations are {\it flat} at these rotation periods.
+This is also visible in figure \ref{fig:dispersion_period_teff}.
+Below the gap, velocity dispersion within a given period range appears to {\it
+decrease} with decreasing temperature.
+The opposite appears to be true above the gap.
+It could be that the gap is positioned at a significant Rossby number/age at
+which stellar magnetic dynamos go through a transition.
+Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
+magnetic braking is more efficient for stars with deeper convection zones.
+Once stars reach this critical age or Rossby number their magnetic fields
+undergo some radical transition, which produces the gap in the rotation
+period-\teff\ plane.
+After this transition, magnetic braking efficiency no longer increases with
+decreasing mass.
+Of course, it may be a coincidence that the gyrochronology relations seem to
+only flatten off above the period gap and we lack a sufficient quantity of
+data to do more than speculate here.
+New rotation periods from the \ktwo\ and \tess\ missions may be able to
+validate or rule out this hypothesis in the future.
+
+In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
+are faster than expected; their rotational evolution appears to have stalled,
+and they have similar rotation periods to the 650 Myr Praesepe cluster
+\citep{curtis2019}.
+The rotation periods of the K dwarfs in this cluster are plotted in figure
+\ref{fig:dispersion_period_teff}.
+Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
+dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
+NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
+K dwarfs lie below it.
+This cluster may be the `missing link' that connects two epoch of stellar
+spin-down.
+However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
+model is, on average, a good model for stars younger than $\sim$1 Gyr. This
+model has a very different shape to the NGC 6811 cluster.
+This would suggest that stars in the field do follow the period-color/\teff\
+relation of Praesepe, at least up to around 1 Gyr.
+We do not find evidence of the K dwarf stalling, which is observed in NGC
+6811, in the field.
+% , however without more observations of middle-aged open clusters it : an
+% early stage where the period-\teff\ relation for cool dwarfs has a negative
+% slope and a late stage where it has a positive slope.
+% In this case the period gap may delineate the transition between these two
+% regimes and is the point at which stellar magnetic dynamos likely undergo a
+% dramatic structural shift at an age of $\sim$ 1.1 Gyr.
+
+% \subsection{Caveats}
+% \label{sec:caveats}
+
+% There are some alternative explanations for the trends seen in figures
+% \ref{fig:age_cut} and \ref{fig:}
+% We list these potential issues below.
+% \begin{itemize}
+% \item{
+% The selection function for these data is extremely complicated, etc.
+% }
+% \item{
+% Redenning for the coolest stars.
+% }
+% \item{
+% Planets/companions for the coolest stars.
+% }
+% \item{
+% Subgiant contamination/weakened magnetic braking.
+% }
+% \end{itemize}

### 2019-09-05 paper/intro.tex
somechanges

@@
 relation of stars at 650 Myrs.
-This relation accurately describes the rotation periods of F and G stars up to
-around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
-rotation periods), but over-predicts the rotation periods of K dwarfs in the
-1.1 Gyr NGC 6811 cluster.
+However, although this relation accurately describes the rotation periods of F
+and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
+with available rotation periods), it over-predicts the rotation periods of K
+dwarfs in the 1.1 Gyr NGC 6811 cluster.
 Very few reliable age estimates exist for K dwarfs with rotation periods older
 than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
-evolution of middle-aged K dwarfs is difficult to characterize\footnote{The
-oldest M dwarfs with reliable ages and rotation periods are members of
-Praesepe ($\sim$ 650 Myrs), so even less is known about their rotational
-evolution.}.
+evolution of middle-aged K dwarfs is largely unknown.
+The oldest M dwarfs with reliable ages and rotation periods are members of
+Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
+evolution.
 
-The gyrochronal ages of the stars in our sample were calculated with the
-\citep{angus2019} gyrochronology model, using de-reddened \Gaia\ \gcolor\
-color and rotation periods reported by \mct.
-These rotational ages are shown on the CMD in figure \ref{fig:age_gradient}.
-The stars with old rotational ages, plotted in yellow hues, predominantly lie
+In this paper, gyrochronal ages of cool field dwarfs in the \kepler\ field
+were calculated with the \citep{angus2019} gyrochronology model, using
+de-reddened \Gaia\ \gcolor\ color and rotation periods reported by \mct\ (we
+explain our data selection process in section \ref{sec:the_data}).
+These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
+figure \ref{fig:age_gradient}.
+The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
 along the upper edge of the MS, where stellar evolution models predict old
-stars to be, however the majority of stars with old rotational ages are bluer
-than \gcolor\ $\sim$ 1.5 dex.
-This provides a first hint that the \citet{angus2019} gyrochronology relation
-under-predicts the ages of low-mass stars.
+stars to be, however the majority of these `old' stars are bluer than \gcolor\
+$\sim$ 1.5 dex.
+This provides the first hint that the \citet{angus2019} gyrochronology
+relation under-predicts the ages of low-mass stars.
 There is no reason to expect the oldest stars in this sample to be the bluer
@@
 Since the \citet{angus2019} gyrochronology model, which is based on the
-period-color relation of Praesepe, does not predict old ages for any M dwarfs,
-it is probably under-predicting the ages of some of these low-mass stars.
-It is therefore likely that old M dwarfs rotate more rapidly than this
-Praesepe-based model predicts.
-In what follows we use a population-based stellar age indicator, velocity
-dispersion, to further investigate the gyrochronology relations at old ages in
+period-color relation of Praesepe, predicts the oldest stars in this sample to
+be K dwarfs, it is probably either under-predicting M dwarf ages or
+over-predicting K dwarf ages.
+% In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
+% slowly than the model predicts.
+In what follows, we use a population-based stellar age indicator, velocity
+dispersion, to investigate the gyrochronology relations at old ages in
 the field.
@@
 Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
-the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
-or $v_z$), just like the star-forming molecular gas observed in the disk today
+the galaxy with a low out-of-plane, or vertical, velocity ($W$, or \vz),
+just like the star-forming molecular gas observed in the disk today
 \citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
-Young stars have relatively small vertical velocities, but gain momentum in
-the vertical direction over time \citep[\eg][]{nordstrom2004, holmberg2007,
-holmberg2009, aumer2009, casagrande2011}.
-Although the cause of orbital heating is not well understood, interactions
+On average, the vertical velocities of stars increase over time
+\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
+casagrande2011}.
+Although the cause of dynamical heating is not well understood, interactions
 with giant molecular clouds, spiral arms and the galactic bar are thought to
@@
 on the implied relative ages of populations.
-The age-velocity dispersion relations (AVRs) are still actively being
-calibrated, so it is difficult to directly compare gyrochronal ages with
-kinematic ones.
-However, regardless of the exact relation between velocity dispersion and
-stellar age, it is expected to be a monotonic relationship, therefore velocity
-dispersion can be used effectively to {\it rank} groups of stars by age.
+% The age-velocity dispersion relations (AVRs) are still actively being
 
-Vertical action is thought to be a better age indicator than vertical velocity
-\citep{ting2019}, although still only a weak age indicator for individual
-stars \citep{beane2018}, however both vertical action ($J_z$) and vertical
-velocity (\vz/W) can only be calculated with full 6-dimensional position and
-velocity information.
+% Vertical action is thought to be a better age indicator than vertical
+% velocity \citep{ting2019}, although still only a weak age indicator for
+% individual stars \citep{beane2018}, however both vertical action ($J_z$) and
+% vertical
+Vertical velocity, \vz, can only be calculated with full 6-dimensional
+position and velocity information.
 Unfortunately, most stars with measured rotation periods do not have radial
@@
 approximation to \vz.
+Because we use \vb\ rather than \vz\, it is difficult to directly compare
+gyrochronal ages with kinematic ones using an age-velocity dispersion relation
+(AVR).
+However, regardless of direction, velocity dispersion is expected to
+monotonically increase over time, and can therefore be used effectively to
+{\it rank} groups of stars by age.
 
-This paper is laid out as follows: in section \ref{sec:the_data} we describe
-our sample selection process and the methods used to calculated stellar
+This paper is laid out as follows: in section \ref{sec:method} we describe our
+sample selection process and the methods used to calculated stellar
 velocities.
+We also establish that \vb\ velocity dispersion, \sigmavb, can be used as an
+age proxy by demonstrating that neither mass-dependent heating nor the
+selection function is likely to have a strong affect on our sample.
 In section \ref{sec:results} we use stellar kinematics to investigate the
 relationship between rotation period, age and color/\teff\ in the field.
-In section \ref{sec:mass-dependent-heating} we show that neither
-mass-dependent heating nor the selection function is likely to have a strong
-affect on our sample, and in section \ref{sec:period_gap} we discuss the
-implications of our results.
+We show that the period-color relation of Praesepe is not applicable to
+old stars in section \ref{sec:age_cut}, and reveal the true shape of the
+period-color relations in section \ref{sec:the_reveal}.
+We discuss a possible connection with the rotation period gap in section
+\ref{sec:period_gap}.

### 2019-09-05 paper/method.tex
some changes

@@
 % Limit: 500 words
+
+\section{Method}
+\label{sec:method}
 
@@
 Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
-trends in the data.
+trends in the data, and intermediate cuts did not significantly change the
+results.
 % We also found that the anti-correlation between stellar mass and $b$ is
@@
 perturbed and are dynamically heated more efficiently than higher-mass stars,
-that means velocity dispersion is a function of {\it both} age and mass and
-cannot be straightforwardly interpreted as an age proxy.
+velocity dispersion would be a function of {\it both} age and mass and cannot
+be straightforwardly interpreted as an age proxy.
+So, in order to establish whether \sigmavb\ can be used as an age proxy, we
+searched for signs of mass-dependent heating within the \kepler\ field.
+
 Mass-dependent dynamical heating has not been unambiguously observed in the
@@
 velocity dispersion of stars in effective temperature bins.
-Again, sigma clipping was used to remove high velocity outliers before
-calculating the standard deviation of stars in each bin.
+Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
+outliers before calculating the standard deviation of stars in each bin.
 
@@
 Velocity dispersion very slightly {\it decreases} with decreasing temperature,
-(the opposite of the trend expected for mass-dependent heating) however the
+the opposite of the trend expected for mass-dependent heating, however the
 slope is only inconsistent with zero at the 1.3 $\sigma$ level.
@@
 smaller velocities.
-The essential point here is that we do not find  evidence for mass-dependent
-heating acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be
-used as an age proxy.
+The essential point is that we do not find evidence for mass-dependent heating
+acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be used as
+an age proxy.
 We performed the same analysis on the 537 stars in this sample with RVs using
@@
 anti-correlation between \teff\ and galactic latitude, $b$, could influence
-our results: stars at higher latitudes have additional velocity components in
-the ${\bf x}$ and ${\bf y}$ directions, which could increase \vb\ but not \vz.
+our results.
+Stars at higher latitudes have additional velocity components in the ${\bf x}$
+and ${\bf y}$ directions, which could increase \vb\ but not \vz.
 Again however, since the relationship between \sigmavb\ and \teff\ is
@@
 \end{figure}
-Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
-strongly affect the \mct\ sample of rotating \kepler\ stars.
-For this reason, we assume that age difference is the major cause of velocity
-dispersion differences between groups.
-In other words, (\vb) velocity dispersion is a reliable age proxy for the
-\mct\ sample.
+
+% Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+% strongly affect the \mct\ sample of rotating \kepler\ stars.
+% For this reason, we assume that age difference is the major cause of velocity
+% dispersion differences between groups.
+% In other words, (\vb) velocity dispersion is a reliable age proxy for the
+% \mct\ sample.

### 2019-09-05 photometric_teff.py

@@ -1,6 +1,8 @@
 # Calculate photometric Teff from Gaia color
 # This should be dereddened colour!
 
+import numpy as np
+
 
 def bprp_to_teff(bprp):
     """

### 2019-09-05 paper/discussion.tex
some updates to the discussion

@@
 evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
+
 For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
@@
 The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
-dispersion as the hottest stars in the age range above which indicates that
-the period-\teff\ relations are {\it flat} at these rotation periods.
-This is also visible in figure \ref{fig:dispersion_period_teff}.
+dispersion as the mid-temperature stars in the age range above, and the
+hottest stars in the age range above {\it that}, which indicates that the
+period-\teff\ relations are {\it flat} at rotation periods between $sim$15-25
+days, for 5500 K $<$ \teff\ $<$ 4000 K.
 Below the gap, velocity dispersion within a given period range appears to {\it
@@
 The opposite appears to be true above the gap.
-It could be that the gap is positioned at a significant Rossby number/age at
-which stellar magnetic dynamos go through a transition.
+
+The gap may be positioned at a significant Rossby number/age at which stellar
+magnetic dynamos go through a transition.
 Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
@@
 Once stars reach this critical age or Rossby number their magnetic fields
-undergo some radical transition, which produces the gap in the rotation
-period-\teff\ plane.
-After this transition, magnetic braking efficiency no longer increases with
-decreasing mass.
+undergo some transition, which produces the gap in the rotation period-\teff\
+plane and after this transition, magnetic braking efficiency no longer
+increases with decreasing mass.
 Of course, it may be a coincidence that the gyrochronology relations seem to
@@
 \citep{curtis2019}.
-The rotation periods of the K dwarfs in this cluster are plotted in figure
+The rotation periods of the K dwarfs in NGC 6811 are plotted in figure
 \ref{fig:dispersion_period_teff}.
@@
 However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
-model is, on average, a good model for stars younger than $\sim$1 Gyr. This
-model has a very different shape to the NGC 6811 cluster.
-This would suggest that stars in the field do follow the period-color/\teff\
-relation of Praesepe, at least up to around 1 Gyr.
-We do not find evidence of the K dwarf stalling, which is observed in NGC
-6811, in the field.
+model is, on average, a {\it good} model for stars younger than $\sim$1 Gyr
+% and this model has a very different shape to the NGC 6811 cluster.
+suggesting that stars in the field follow the period-color/\teff\ relation of
+Praesepe, at least up to around 1 Gyr.
+We do not see strong evidence of K dwarf stalling in the field.
 % , however without more observations of middle-aged open clusters it : an

### 2019-09-05 paper/discussion.tex
typo

@@
 hottest stars in the age range above {\it that}, which indicates that the
-period-\teff\ relations are {\it flat} at rotation periods between $sim$15-25
+period-\teff\ relations are {\it flat} at rotation periods between $\sim$15-25
 days, for 5500 K $<$ \teff\ $<$ 4000 K.

### 2019-10-10 dispersion.py
modifying MAD

@@ -69,7 +69,7 @@ def dispersion(x, y, bins, method):
         if method == "std":
             d[i] = np.std(y[m])
         if method == "mad":
-            d[i] = np.median(np.abs(y[m])) # - np.median(y[m])))
+            d[i] = 1.5*np.median(np.abs(y[m] - np.median(y[m])))
 
         # Calculate kurtosis
         k[i] = sps.kurtosis(y[m])

### 2019-10-10 paper/abstract.tex
somechanges
adding a bunch of paper files -- not sure why these weren't staged?

@@
 
-The rotational evolution of cool dwarfs is mostly unconstrained after around
-2-3 billion years, and the gyrochronology relations are poorly calibrated
-beyond this time due to a lack of precise ages and rotation periods for old
+The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
+billion years due to a lack of precise ages and rotation periods for old
 main-sequence stars.
-However, recent evidence suggests that the relationship between rotation
-period and color/temperature/mass does not remain constant over large
-timescales, contradicting a long-held assumption that underlies many
-gyrochronology relations.
-We use the velocities of populations of cool dwarfs to reveal the rotational
-evolution of these stars at old ages.
-We find that groups of stars, selected to be the same age using a
-gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
-the same velocity dispersion at all temperatures; cooler stars have greater
-velocity dispersions than hotter stars.
-We interpret this to mean that these cooler stars are older than the hotter
-ones, which implies that the period-color relation of Praesepe is not
-appropriate for old stars.
-Futhermore, our results indicate that the slope of the period-color relation
-actually {\it inverts}: at young ages early K stars rotate more rapidly than
-late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
-These observations could be reproduced by mass-dependent dynamical heating,
-where lower-mass stars experience a greater rate of dynamical heating than
-higher-mass stars, however we find no evidence for mass-dependent heating
-within K and M dwarfs in the \kepler\ field.
+% However, recent evidence suggests that the relationship between rotation
+% period and color/temperature/mass does not remain constant over large
+% timescales, contradicting a long-held assumption that underlies many
+% gyrochronology relations.
+In this work, we use the velocities of low-mass \kepler\ dwarfs to reveal
+their rotational evolution and demonstrate that kinematics have potential to
+be a useful tool for calibrating gyrochronology.
+At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
+to the Praesepe cluster, accurately predicts the relative ages of stars.
+At these young ages, lower-mass stars lose angular momentum more quickly than
+higher mass stars and therefore rotate more slowly at the same age.
+However, at old ages, late G and early K dwarfs rotate {\it faster} than late
+K dwarfs of the same age.
+This implies that low-mass stars lose angular momentum more slowly than
+higher-mass stars after a certain age or Rossby number.
+Our results, based on field star rotation periods, align with recent findings
+from the rotation periods of stars in middle-aged open clusters and
+theoretical models that vary the rate of surface-to-core angular momentum
+transport.
+Finally, we find no evidence for mass-dependent heating in a sample of K and M
+dwarfs in the \kepler\ field.
+
+% We find that groups of stars, selected to be the same age using a
+% gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
+% the same velocity dispersion at all temperatures; cooler stars have greater
+% velocity dispersions than hotter stars.
+% We interpret this to mean that these cooler stars are older than the hotter
+% ones, which implies that the period-color relation of Praesepe is not
+% appropriate for old stars.
+% Futhermore, our results indicate that the slope of the period-color relation
+% actually {\it inverts}: at young ages early K stars rotate more rapidly than
+% late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
+% These observations could be reproduced by mass-dependent dynamical heating,
+% where lower-mass stars experience a greater rate of dynamical heating than
+% higher-mass stars, however we find no evidence for mass-dependent heating
+% within K and M dwarfs in the \kepler\ field.

### 2019-10-10 paper/intro.tex
adding a bunch of paper files -- not sure why these weren't staged?

@@
 It is well established that the rotation periods of FGKM dwarfs increase over
-time \citep{skumanich1972}.
-This characteristic of main-sequence (MS) stars allows them to be dated using
-their rotation periods, via gyrochronology \citep[\eg][]{kawaler1989,
-pinsonneault1989, barnes2003, barnes2007, barnes2010, meibom2011, meibom2015},
-which is particularly useful since the ages of MS stars are difficult to
-measure via isochrone placement.
+time \citep{skumanich1972}, however the exact nature and form of rotational
+evolution is still being actively studied.
+Once rotational evolution is well understood, rotation periods can be used to
+precisely and accurately date stars via gyrochronology
+\citep[\eg][]{kawaler1989, pinsonneault1989, barnes2003, barnes2007,
+barnes2010, meibom2011, meibom2015}.
 It is also well established that stars of the same age but different masses
-have different rotation periods \citep[\eg][]{kraft1967, matt2012}, thought to
-be caused by the deeper convective zones, and therefore stronger magnetic
-dynamos (and more efficient magnetic braking) in lower-mass stars.
-However, an underlying assumption behind many empirical gyrochronology
-relations is that the relationships between rotation period and photometric
-color\footnote{As the directly observable quantity, color is often used as a
-mass proxy, and empirical gyrochronology relations are usually calibrated in
-color, rather than mass or effective temperature.},
-and rotation period and age are {\it separable}, meaning that the
-period-color relation is the same at all ages and the period-age relation is
-the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
-meibom2011, angus2015, angus2019}.
-Not all gyrochronology models follow this form, particularly those motivated
-based on stellar structure and evolution theory or with a Rossby number,
-rather than rotation period dependence \citep[\eg][]{barnes2010, matt2012,
-vansaders2016]}.
-It was recently shown that old field stars rotate more rapidly than a simple,
-separable gyrochronology relation would predict \citep{angus2015,
-vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
-modification to the classical \prot\ $\propto t^{\frac{1}{2}}$ spin-down law
-\citep{skumanich1972} is required to reproduce the data \citep{vansaders2016,
-vansaders2018}.
-An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
-further indication that the exponent of the period-age relation is
-mass and time-dependent \citep{curtis2019}.
-This cluster has a flattened relationship between rotation period and color
-and the G dwarfs rotate at the same rate as the K dwarfs.
+have different rotation periods \citep[\eg][]{kraft1967, matt2012}, which is
+thought to be caused by the variation in convection zone depth.
+For example, lower mass stars with deeper convections zones are expected to
+have stronger magnetic dynamos and lose angular momentum more rapidly.
+However, as new rotation period measurements have become available, a more
+complicated picture of stellar rotational evolution is emerging.
+% However, an underlying assumption behind many empirical gyrochronology
+% relations is that the relationships between rotation period and photometric
+% color\footnote{As the directly observable quantity, color is often used as a
+% mass proxy, and empirical gyrochronology relations are usually calibrated in
+% color, rather than mass or effective temperature.},
+% and rotation period and age are {\it separable}, meaning that the
+% period-color relation is the same at all ages and the period-age relation is
+% the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
+% meibom2011, angus2015, angus2019}.
+% Not all gyrochronology models follow this form, particularly those based on
+% stellar structure and evolution theory or with a Rossby number, rather than
+% rotation period dependence \citep[\eg][]{barnes2010, matt2012,
+% vansaders2016]}.
+For example, it was recently shown that old field stars rotate more rapidly
+than models predicted \citep{angus2015, vansaders2016, vansaders2018,
+metcalfe2019}, and a mass-dependent modification to the classical \prot\
+$\propto t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} was required to
+reproduce the data \citep{vansaders2016, vansaders2018}.
+Even more recent analyses of the 1.1 Gyr open cluster, NGC 6811, provides
+further indication that the exponent of the period-age relation is mass and
+time-dependent \citep{curtis2019 spada2019}.
+This cluster has a flattened relationship between rotation period and
+color: the G dwarfs rotate at the same rate as the K dwarfs.
+New theoretical models that vary the rate of magnetic braking and the
+redistribution of angular momentum in the interior of stars are able to
+successfully reproduce the flattened period-color relation seen in old open
+clusters \citep{spada2019}.
+With the advent of large time-series photometry surveys, providing large
+numbers of precise rotation period measurements over the last few years, we
+now know that the rotational evolution of stars is a complicated process.
+However, relying on open clusters (which are typically young) and
+asteroseismic stars (which are typically G-type stars or more massive) to
+calibrate gyrochronology limits the mass and age coverage of the calibration
+sample.
 
 In this paper, we tested the \citet{angus2019} gyrochronology relation, a
-separable relation, calibrated using the period-color relation of Praesepe (in
-\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
+separable relation in color and age, calibrated using the period-color
+relation of Praesepe (in \gaia\ \gcolor\ color) and the period-age relation of
+Praesepe and the Sun.
 The large number of Praesepe members with precise rotation periods from the
@@
 
-In this paper, gyrochronal ages of cool field dwarfs in the \kepler\ field
-were calculated with the \citep{angus2019} gyrochronology model, using
-de-reddened \Gaia\ \gcolor\ color and rotation periods reported by \mct\ (we
-explain our data selection process in section \ref{sec:the_data}).
+We calculated gyrochronal ages of cool field dwarfs in the \kepler\ field
+using the \citep{angus2019} gyrochronology model, with de-reddened \Gaia\
+\gcolor\ color and rotation periods reported by \mct\ (we explain our data
+selection process in section \ref{sec:the_data}).
 These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in

### 2019-10-10 paper/results.tex
adding a bunch of paper files -- not sure why these weren't staged?

@@
 dispersion and rotation periods increase with age as expected.
-The constant velocity dispersion of young stars as a function of temperature
+The flat velocity dispersion of young stars as a function of temperature
 shows that the Praesepe-calibrated gyrochronology relation accurately predicts
@@
 % of \teff.
-In contrast however, velocity dispersion increases as a function of
-temperature for old stars, meaning the \citet{angus2019} gyrochronology
-relation either under-predicts the ages of old, late-K dwarfs, or
-over-predicts the ages of old early-K and late-G dwarfs\footnote{Since \vb\
-velocity dispersion only provides relative and not absolute ages, it is
-difficult to tell whether the ages of cool stars are being under-predicted,
-the ages of hot stars being are over-predicted, or both.}.
-This indicates that the relationship between rotation period and photometric
+If the gyrochronology relations worked at all ages and temperatures, the
+bottom panel of figure \ref{fig:age_cut} would show a flat relationship
+between velocity dispersion and \teff at all ages.
+However, velocity dispersion {\it increases} as a function of temperature for
+old stars, meaning the \citet{angus2019} gyrochronology relation either
+under-predicts the ages of old, late-K dwarfs, or over-predicts the ages of
+old early-K and late-G dwarfs\footnote{Since \vb\ velocity dispersion only
+provides relative and not absolute ages, it is difficult to tell whether the
+ages of cool stars are being under-predicted, the ages of hot stars being are
+over-predicted, or both.}.
+This suggests that the relationship between rotation period and photometric
 color or \teff\ flattens out over time, and possibly even inverts.
@@
 shape of the upper detection edge.
-The so-called `M dwarf dip' \citep{vansaders2018} is reflected in the lines of
-constant velocity dispersion (and presumed age) in the top panel of figure
+The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
+period catalog characterized by a dearth of slowly rotating M dwarfs between
+$\sim$2750 and 4250 K, is reflected in the lines of constant velocity
+dispersion (and presumed age) in the top panel of figure
 \ref{fig:dispersion_period_teff}.

### 2019-10-11 paper/intro.tex
modifying paper text and experimenting with various cuts (the overall results are the same)

@@
 
-\subsection{Gyrochronology}
-It is well established that the rotation periods of FGKM dwarfs increase over
-time \citep{skumanich1972}, however the exact nature and form of rotational
-evolution is still being actively studied.
-Once rotational evolution is well understood, rotation periods can be used to
-precisely and accurately date stars via gyrochronology
-\citep[\eg][]{kawaler1989, pinsonneault1989, barnes2003, barnes2007,
-barnes2010, meibom2011, meibom2015}.
-It is also well established that stars of the same age but different masses
-have different rotation periods \citep[\eg][]{kraft1967, matt2012}, which is
-thought to be caused by the variation in convection zone depth.
-For example, lower mass stars with deeper convections zones are expected to
-have stronger magnetic dynamos and lose angular momentum more rapidly.
-However, as new rotation period measurements have become available, a more
-complicated picture of stellar rotational evolution is emerging.
+\subsection{Gyrochronology} It is well established that the rotation periods
+of FGKM dwarfs increase over time \citep[\eg][]{schatzmann, weber,
+skumanich1972 kawaler1989, pinsonneault1989, barnes2003, barnes2007,
+barnes2010, meibom2011, meibom2015}, however the exact nature and form of
+rotational evolution is still being actively studied.
+% Once the relationship between rotation period, mass and age is well
+% calibrated, rotation periods can be used to precisely and accurately date
+% stars via gyrochronology \citep[\eg][]{kawaler1989, pinsonneault1989,
+% barnes2003, barnes2007, barnes2010, meibom2011, meibom2015}.
+% It is also well established that stars of the same age but different masses
+In particular, as new rotation period measurements for stars with precise ages
+become available, the mass-dependence of the relationship between rotation and
+age appears to be more complicated than once thought.
+From both observations and theoretical modelling, stars of different masses
+are expected to have different rotation periods \citep[\eg][]{kraft1967,
+matt2012}, which is thought to be mainly driven by variation in convection
+zone depth.
+For example in the Praesepe cluster, M dwarfs spin more slowly than G dwarfs,
+presumably because lower mass stars with deeper convections zones are have
+stronger magnetic dynamos and lose angular momentum more rapidly.
+However, new rotation period measurements show that variation in convection
+zone depth cannot be solely responsible for mass-dependent rotational
+evolution.
+% revealing that rotational
+% evolution is clearly a more complicated process than once thought.
 % However, an underlying assumption behind many empirical gyrochronology
@@
 % vansaders2016]}.
-For example, it was recently shown that old field stars rotate more rapidly
-than models predicted \citep{angus2015, vansaders2016, vansaders2018,
-metcalfe2019}, and a mass-dependent modification to the classical \prot\
-$\propto t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} was required to
-reproduce the data \citep{vansaders2016, vansaders2018}.
-Even more recent analyses of the 1.1 Gyr open cluster, NGC 6811, provides
-further indication that the exponent of the period-age relation is mass and
+For example, old field stars seem rotate more rapidly than models predicted
+\citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}, and a
+mass-dependent modification to the classical \prot\ $\propto t^{\frac{1}{2}}$
+spin-down law \citep{skumanich1972} is required to reproduce the data
+\citep{vansaders2016, vansaders2018}.
+Recent analyses of the 1.1 Gyr open cluster, NGC 6811, provide further
+indication that the exponent of the period-age relation is mass and
 time-dependent \citep{curtis2019 spada2019}.
@@
 color: the G dwarfs rotate at the same rate as the K dwarfs.
-New theoretical models that vary the rate of magnetic braking and the
-redistribution of angular momentum in the interior of stars are able to
-successfully reproduce the flattened period-color relation seen in old open
-clusters \citep{spada2019}.
-With the advent of large time-series photometry surveys, providing large
-numbers of precise rotation period measurements over the last few years, we
-now know that the rotational evolution of stars is a complicated process.
-However, relying on open clusters (which are typically young) and
-asteroseismic stars (which are typically G-type stars or more massive) to
-calibrate gyrochronology limits the mass and age coverage of the calibration
+New semi-empirical models that vary the redistribution of angular momentum in
+the interiors of stars are able to reproduce this flattened period-color
+relation \citep{spada2019}.
+% With the advent of large time-series photometry surveys, providing large
+% numbers of precise rotation period measurements over the last few years.
+We now know that the rotational evolution of stars is a complicated process,
+however relying on open clusters (which are typically young) and asteroseismic
+stars (which are typically G-type stars or more massive) to calibrate
+gyrochronology relations limits the mass and age coverage of the calibration
 sample.
-
-In this paper, we tested the \citet{angus2019} gyrochronology relation, a
-separable relation in color and age, calibrated using the period-color
-relation of Praesepe (in \gaia\ \gcolor\ color) and the period-age relation of
-Praesepe and the Sun.
-The large number of Praesepe members with precise rotation periods from the
-\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
-through early M, makes it a good cluster for calibrating the period-color
-relation of stars at 650 Myrs.
-However, although this relation accurately describes the rotation periods of F
-and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
-with available rotation periods), it over-predicts the rotation periods of K
-dwarfs in the 1.1 Gyr NGC 6811 cluster.
-Very few reliable age estimates exist for K dwarfs with rotation periods older
-than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
-evolution of middle-aged K dwarfs is largely unknown.
-The oldest M dwarfs with reliable ages and rotation periods are members of
-Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
-evolution.
-
-We calculated gyrochronal ages of cool field dwarfs in the \kepler\ field
-using the \citep{angus2019} gyrochronology model, with de-reddened \Gaia\
-\gcolor\ color and rotation periods reported by \mct\ (we explain our data
-selection process in section \ref{sec:the_data}).
-These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
-figure \ref{fig:age_gradient}.
-The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
-along the upper edge of the MS, where stellar evolution models predict old
-stars to be, however the majority of these `old' stars are bluer than \gcolor\
-$\sim$ 1.5 dex.
-This provides the first hint that the \citet{angus2019} gyrochronology
-relation under-predicts the ages of low-mass stars.
-There is no reason to expect the oldest stars in this sample to be the bluer
-ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
-active for longer, so should therefore have measurable rotation periods at
-older ages.
-Since the \citet{angus2019} gyrochronology model, which is based on the
-period-color relation of Praesepe, predicts the oldest stars in this sample to
-be K dwarfs, it is probably either under-predicting M dwarf ages or
-over-predicting K dwarf ages.
-% In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
-% slowly than the model predicts.
-In what follows, we use a population-based stellar age indicator, velocity
-dispersion, to investigate the gyrochronology relations at old ages in
-the field.
+In this paper, we use stellar velocity dispersions to gain insight into the
+rotational evolution of field stars.
+This technique provides relative ages for groups of stars that cover a broad
+range of masses and ages, allowing us to observe the rotational evolution of
+stars in an average sense.
 
@@
 Vertical velocity, \vz, can only be calculated with full 6-dimensional
-position and velocity information.
-Unfortunately, most stars with measured rotation periods do not have radial
-velocity (RV) measurements because they are relatively faint \kepler\ targets
-($\sim$11th-18th magnitudes).
-For this reason, we used velocity in the direction of galactic latitude, \vb.
-The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
-\vb\ is a close (although imperfect -- see section \ref{sec:results})
-approximation to \vz.
+position and velocity information, and unfortunately most stars with measured
+rotation periods do not have radial velocity (RV) measurements because they
+are relatively faint \kepler\ targets ($\sim$11th-18th magnitudes).
+For this reason, we used velocity in the direction of galactic latitude, \vb,
+to mimic \vz.
+The \kepler\ field is positioned at low galactic latitude
+(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect -- see section
+\ref{sec:results}) approximation to \vz.
 Because we use \vb\ rather than \vz\, it is difficult to directly compare
@@
 \ref{sec:period_gap}.
+
+% We also tested the \citet{angus2019} gyrochronology relation, a separable
+% relation in color and age, calibrated using the period-color relation of
+% Praesepe (in \gaia\ \gcolor\ color) and the period-age relation of Praesepe
+% and the Sun.
+% The large number of Praesepe members with precise rotation periods from the
+% \ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
+% through early M, makes it a good cluster for calibrating the period-color
+% relation of stars at 650 Myrs.
+% However, although this relation accurately describes the rotation periods of F
+% and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
+% with available rotation periods), it over-predicts the rotation periods of K
+% dwarfs in the 1.1 Gyr NGC 6811 cluster.
+% Very few reliable age estimates exist for K dwarfs with rotation periods older
+% than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
+% evolution of middle-aged K dwarfs is largely unknown.
+% The oldest M dwarfs with reliable ages and rotation periods are members of
+% Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
+% evolution.
+
+% We calculated gyrochronal ages of cool field dwarfs in the \kepler\ field
+% using the \citep{angus2019} gyrochronology model, with de-reddened \Gaia\
+% \gcolor\ color and rotation periods reported by \mct\ (we explain our data
+% selection process in section \ref{sec:the_data}).
+% These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
+% figure \ref{fig:age_gradient}.
+% The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
+% along the upper edge of the MS, where stellar evolution models predict old
+% stars to be, however the majority of these `old' stars are bluer than \gcolor\
+% $\sim$ 1.5 dex.
+% This suggests that the \citet{angus2019} gyrochronology relation
+% under-predicts the ages of low-mass stars.
+% There is no reason to expect the oldest stars in this sample to be the bluer
+% ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
+% active for longer, so should therefore have measurable rotation periods at
+% older ages.
+% Since the \citet{angus2019} gyrochronology model, which is based on the
+% period-color relation of Praesepe, predicts the oldest stars in this sample to
+% be K dwarfs, it is probably either under-predicting M dwarf ages or
+% over-predicting K dwarf ages.
+% % In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
+% % slowly than the model predicts.
+% In what follows, we use a population-based stellar age indicator, velocity
+% dispersion, to investigate the gyrochronology relations at old ages in
+% the field.

### 2019-10-11 paper/results.tex
modifying paper text and experimenting with various cuts (the overall results are the same)

@@
 
-To explore the relationship between rotation period, \teff\ and age, we
-selected groups of stars within different gyrochronal age ranges \citep[where
-age was calculated using the][gyrochronology relation]{angus2019}, and
+To explore the relationship between rotation period, \teff\ and age for field
+K dwarfs, we calculated gyrochronal ages using dereddened \citep[extinction
+was calculated using the][Python package]{green2018} \Gaia\ photometry
+($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
+We used the \citet{angus2019} gyrochronology relation which is a separable
+relation in \gcolor\ color and age, calibrated using the period-color relation
+of Praesepe and the period-age relation of Praesepe and the Sun.
+The large number of Praesepe members with precise rotation periods from the
+\ktwo\ mission \citep{howell2013, douglas2017, rebull2017}, spanning spectral
+types F through early M, makes it a good cluster for calibrating the
+period-color relation of stars at 650 Myrs.
+However, this is an extremely simple model and although it accurately
+describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
+of NGC 6819 -- the oldest cluster with available rotation periods), it
+over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
+cluster.
+These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
+figure \ref{fig:age_gradient}.
+The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
+along the upper edge of the MS, where stellar evolution models predict old
+stars to be, however the majority of these `old' stars are bluer than \gcolor\
+$\sim$ 1.5 dex.
+This suggests that the \citet{angus2019} gyrochronology relation
+under-predicts the ages of low-mass stars because there is no reason to expect
+the oldest stars in this sample to be the bluer ones: M dwarfs are, on
+average, older than K dwarfs and are expected to remain active for longer, so
+should therefore have measurable rotation periods at older ages.
+
+To further explore the relationship between rotation period, \teff\ and age,
+we selected groups of stars within different gyrochronal age ranges and
 calculated the velocity dispersion (\sigmavb), the standard deviation of \vb\

### 2019-10-14 paper/intro.tex
simplifying language and better explanations

@@
 
-\subsection{Gyrochronology} It is well established that the rotation periods
-of FGKM dwarfs increase over time \citep[\eg][]{schatzmann, weber,
-skumanich1972 kawaler1989, pinsonneault1989, barnes2003, barnes2007,
-barnes2010, meibom2011, meibom2015}, however the exact nature and form of
-rotational evolution is still being actively studied.
+\subsection{Gyrochronology}
+
+It is well established that magnetized stellar winds cause the rotation
+periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
+weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
+calibrated, the relationship between period, mass and age can be used to
+age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
+meibom2011, meibom2015}.
+% However, the exact nature of rotational evolution is still being actively
+% studied.
 % Once the relationship between rotation period, mass and age is well
@@
 % It is also well established that stars of the same age but different masses
-In particular, as new rotation period measurements for stars with precise ages
-become available, the mass-dependence of the relationship between rotation and
-age appears to be more complicated than once thought.
-From both observations and theoretical modelling, stars of different masses
-are expected to have different rotation periods \citep[\eg][]{kraft1967,
-matt2012}, which is thought to be mainly driven by variation in convection
-zone depth.
-For example in the Praesepe cluster, M dwarfs spin more slowly than G dwarfs,
-presumably because lower mass stars with deeper convections zones are have
-stronger magnetic dynamos and lose angular momentum more rapidly.
-However, new rotation period measurements show that variation in convection
-zone depth cannot be solely responsible for mass-dependent rotational
-evolution.
+However, new photometric rotation periods measured from \kepler\ light curves
+reveal that rotational evolution is more complicated than previously thought.
+For example, the M dwarfs in the $\sim$ 700 Myr Praesepe cluster spin more
+slowly than the G dwarfs -- in theory because lower-mass stars have deeper
+convections zones which generate stronger magnetic fields and more efficient
+magnetic braking \citep{schatzman1962, weber1967}.
+However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
+same} rate as early-K dwarfs \citep{curtis2019}.
+In other words, convection zone depth cannot be the only variable that affects
+stellar spin-down rate.
+New semi-empirical models that vary the rate of angular momentum
+redistribution in the interiors of stars are able to reproduce this flattened
+period-color relation \citep{spada2019}.
+These models suggest that mass and age-dependent coupling between the cores
+and envelopes of stars has a significant impact on their surface rotation
+rates.
+
+Another example of unexpected rotational evolution is seen in old field stars
+which appear to rotate more rapidly than classical gyrochronology models
+predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
+A mass-dependent modification to the classical \prot\ $\propto
+t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
+the observations \citep{vansaders2016, vansaders2018}.
 % revealing that rotational
@@
 % vansaders2016]}.
-For example, old field stars seem rotate more rapidly than models predicted
-\citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}, and a
-mass-dependent modification to the classical \prot\ $\propto t^{\frac{1}{2}}$
-spin-down law \citep{skumanich1972} is required to reproduce the data
-\citep{vansaders2016, vansaders2018}.
-Recent analyses of the 1.1 Gyr open cluster, NGC 6811, provide further
-indication that the exponent of the period-age relation is mass and
-time-dependent \citep{curtis2019 spada2019}.
-This cluster has a flattened relationship between rotation period and
-color: the G dwarfs rotate at the same rate as the K dwarfs.
-New semi-empirical models that vary the redistribution of angular momentum in
-the interiors of stars are able to reproduce this flattened period-color
-relation \citep{spada2019}.
+% For example, old field stars seem rotate more rapidly than models predicted
+% \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}, and a
+% mass-dependent modification to the classical \prot\ $\propto t^{\frac{1}{2}}$
+% spin-down law \citep{skumanich1972} is required to reproduce the data
+% \citep{vansaders2016, vansaders2018}.
+% In addition, recent analyses of the 1.1 Gyr open cluster, NGC 6811, provide
+% further indication that rotational evolution is mass and time-dependent
+% \citep{curtis2019 spada2019}.
+% This cluster has a flattened relationship between rotation period and
+% color: the early-K dwarfs rotate at the same rate as the late-K dwarfs.
 % With the advent of large time-series photometry surveys, providing large
 % numbers of precise rotation period measurements over the last few years.
-We now know that the rotational evolution of stars is a complicated process,
-however relying on open clusters (which are typically young) and asteroseismic
-stars (which are typically G-type stars or more massive) to calibrate
-gyrochronology relations limits the mass and age coverage of the calibration
-sample.
-In this paper, we use stellar velocity dispersions to gain insight into the
-rotational evolution of field stars.
-This technique provides relative ages for groups of stars that cover a broad
-range of masses and ages, allowing us to observe the rotational evolution of
-stars in an average sense.
 
-\subsection{Kinematics as an age proxy}
+The rotational evolution of stars is clearly a complicated process, and to
+fully calibrate the gyrochronology relations we need reliable ages for large
+numbers of field stars which span a range of ages.
+% however relying on open clusters (which are typically young) and asteroseismic
+% stars (which are typically G-type stars or more massive) to calibrate
+% gyrochronology relations limits the mass and age coverage of the calibration
+% sample.
+In this paper, we use the velocity dispersions of field stars to gain insight
+into their rotational evolution.
+% This technique provides relative ages for groups of stars that cover a broad
+% range of masses and ages, allowing us to observe the rotational evolution of
+% stars in an average sense.
+
+\subsection{Using kinematics as an age proxy}
 
@@
 on average, that group is old or young relative to other groups.
-In this work we compare the velocity dispersions of groups of stars to
-ascertain which groups are older and which younger and draw conclusions based
-on the implied relative ages of populations.
-% The age-velocity dispersion relations (AVRs) are still actively being
+In this work we compare the velocity dispersions of groups of field stars in
+the Galactic thin disk to ascertain which groups are older and which younger
+and draw conclusions based on the implied relative ages.
+% The age-velocity dispersion reations (AVRs) are still actively being
 
@@
 For this reason, we used velocity in the direction of galactic latitude, \vb,
-to mimic \vz.
+to approximate \vz.
 The \kepler\ field is positioned at low galactic latitude
@@
 \ref{sec:results}) approximation to \vz.
-Because we use \vb\ rather than \vz\, it is difficult to directly compare
-gyrochronal ages with kinematic ones using an age-velocity dispersion relation
-(AVR).
+Because we use \vb\ rather than \vz\, we cannot calculate absolute kinematic
+ages using an age-velocity dispersion relation (AVR).
 However, regardless of direction, velocity dispersion is expected to
-monotonically increase over time, and can therefore be used effectively to
-{\it rank} groups of stars by age.
+monotonically increase over time, and can therefore be used to {\it rank}
+groups of stars by age.
 
@@
 age proxy by demonstrating that neither mass-dependent heating nor the
-selection function is likely to have a strong affect on our sample.
-In section \ref{sec:results} we use stellar kinematics to investigate the
-relationship between rotation period, age and color/\teff\ in the field.
-We show that the period-color relation of Praesepe is not applicable to
-old stars in section \ref{sec:age_cut}, and reveal the true shape of the
-period-color relations in section \ref{sec:the_reveal}.
-We discuss a possible connection with the rotation period gap in section
-\ref{sec:period_gap}.
+selection function seems to strongly affect on our sample.
+In section \ref{sec:results} we use kinematics to investigate the
+relationship between rotation period, age and color/\teff\ in the field and
+interpret our results in section \ref{sec:discussion}.
+% We show that the period-color relation of Praesepe is not applicable to
+% old stars in section \ref{sec:age_cut}, and reveal the true shape of the
+% period-color relations in section \ref{sec:the_reveal}.
+% We discuss a possible connection with the rotation period gap in section
+% \ref{sec:period_gap}.
 

### 2019-10-14 paper/results.tex
some improvements but many more needed

--- 2019-10-10 paper/results.tex
@@
 
-To explore the relationship between rotation period, \teff\ and age, we
-selected groups of stars within different gyrochronal age ranges \citep[where
-age was calculated using the][gyrochronology relation]{angus2019}, and
-calculated the velocity dispersion (\sigmavb), the standard deviation of \vb\
-velocities, as a function of effective temperature for each age group.
-We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
-and temperature bin to remove non-Gaussian outliers.
+To explore the relationship between rotation period, \teff\ and age for field
+K dwarfs, we calculated gyrochronal ages using dereddened \Gaia\ photometry
+($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
+We used the \citet{angus2019} gyrochronology relation which is a simple,
+separable relation in \gcolor\ color and age, calibrated using the
+period-color relation of Praesepe and the period-age relation of Praesepe and
+the Sun.
+The large number of Praesepe members with precise rotation periods from the
+\ktwo\ mission \citep{howell2014, douglas2017, rebull2017}, spanning spectral
+types F through early M, makes it a good cluster for calibrating the
+period-color relation of stars at young ages.
+However, this is an extremely simple model and although it accurately
+describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
+of NGC 6819 -- the oldest cluster with available rotation periods), it
+over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
+cluster.
+These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
+figure \ref{fig:age_gradient}.
+The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
+along the upper edge of the MS, where stellar evolution models predict old
+stars to be, however the majority of these `old' stars are bluer than \gcolor\
+$\sim$ 1.5 dex.
+This suggests that either old M dwarfs are missing from the \mct\ catalog, or
+the \citet{angus2019} gyrochronology relation under-predicts the ages of
+low-mass stars.
+Given that lower-mass stars stay active for longer than higher-mass stars
+\citep[\eg][]{west2011}, and are therefore more likely to have measurable
+rotation periods at old ages, the latter scenario seems more likely.
+
+To further explore the relationship between rotation period, \teff\ and age,
+we first removed high and low velocity outliers by performing 3$\sigma$
+sigma-clipping on the \vb\ velocities.
 Without sigma-clipping, we found that a small number of high velocity
@@
 velocity dispersion for cooler stars.
+We then selected groups of stars within different gyrochronal age ranges and
+calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
+of effective temperature for each age group.
 Ages were calculated using dereddened \gaia\ \gcolor\ color, however
 throughout this paper we show rotation periods as a function of effective
-temperature, \teff.
-We chose to use \teff\ as the abscissa because it is easier to divide stars
-into bins of roughly equal numbers in \teff-space than in color-space.
+temperature, \teff, because it is easier to divide stars into bins of roughly
+equal numbers in \teff-space than in color-space.
 
@@
 The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
-(excluding visual binaries and subgiants) in grey, with coeval groups shown in
-color.
+(excluding visual binaries and subgiants) in grey, with coeval groups
+\citep[according to the][relation]{angus2019} shown in color.
 The color of the points corresponds to the age ranges specified in the legend
@@
 with gyrochronal age across all temperatures, implying that both velocity
-dispersion and rotation periods increase with age as expected.
+dispersion and rotation period increases with age as expected.
 The flat velocity dispersion of young stars as a function of temperature
@@
 % of \teff.
-If the gyrochronology relations worked at all ages and temperatures, the
-bottom panel of figure \ref{fig:age_cut} would show a flat relationship
-between velocity dispersion and \teff at all ages.
+If \citet{angus2019} gyrochronology relation worked at all ages and
+temperatures, the bottom panel of figure \ref{fig:age_cut} would show a flat
+relationship between velocity dispersion and \teff at all ages.
 However, velocity dispersion {\it increases} as a function of temperature for
@@
 color or \teff\ flattens out over time, and possibly even inverts.
-The lines of constant age (isochrones) sweeping diagonally upwards in the top
-panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
+% The lines of constant age (isochrones) sweeping diagonally upwards in the top
+% panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
 
@@
     \includegraphics[width=1\textwidth]{vplot}
-\label{fig:dispersion_period_teff}
+\label{fig:vplot}
 \end{figure}
@@
 period catalog characterized by a dearth of slowly rotating M dwarfs between
-$\sim$2750 and 4250 K, is reflected in the lines of constant velocity
+$\sim$3750 and 4250 K, is reflected in the lines of constant velocity
 dispersion (and presumed age) in the top panel of figure
@@
 If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
-$\sim$4500 K become inactive at a similar age, and stay active longer than
-stars hotter than $\sim$4500 on average.
+$\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.
 
@@
 right of figure \ref{fig:dispersion_period_teff}.
+
+\begin{figure}
+  \caption{
+    Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
+    x-axis.
+    Masses are from the \kepler\ input catalog.
+    The white lines show the \citet{spada2019} rotational evolution models at
+    0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
+    period.
+    These models include age and mass-dependent coupling between the stellar
+    core and envelope.
+    The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
+    an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
+    decreases with decreasing mass.
+    A similar phenomenon is visible in the velocity dispersions of field stars
+    shown as colored points in the background of this figure.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{masses_with_models}
+\label{fig:masses_with_models}
+\end{figure}
 

### 2019-10-15 paper/abstract.tex
many paper edits

@@
 to the Praesepe cluster, accurately predicts the relative ages of stars.
-At these young ages, lower-mass stars lose angular momentum more quickly than
-higher mass stars and therefore rotate more slowly at the same age.
-However, at old ages, late G and early K dwarfs rotate {\it faster} than late
-K dwarfs of the same age.
-This implies that low-mass stars lose angular momentum more slowly than
-higher-mass stars after a certain age or Rossby number.
-Our results, based on field star rotation periods, align with recent findings
-from the rotation periods of stars in middle-aged open clusters and
+At these young ages, lower-mass stars spin more slowly than higher-mass stars
+because their stronger magnetic fields lead to more efficient angular momentum
+loss.
+However, at old ages we find that late G and early K dwarfs rotate {\it
+faster} than late K dwarfs of the same age.
+% These oThis may be a feature of angular momentum redistribution in the stellar
+% interior.
+% This implies that low-mass stars lose angular momentum more slowly than
+% higher-mass stars after a certain age or Rossby number.
+These results, based on field star rotation periods, align with recent
+findings from the rotation periods of stars in middle-aged open clusters and
 theoretical models that vary the rate of surface-to-core angular momentum
-transport.
+transport as a function of time and mass.
 Finally, we find no evidence for mass-dependent heating in a sample of K and M

### 2019-10-15 paper/conclusion.tex
many paper edits

@@
 
-We examined the gyrochronology relations using the velocity dispersions of
-field stars with measured rotation periods.
+We examined the rotational evolution of \kepler\ field stars using their
+velocity dispersions as a proxy for age.
 We found that old groups of cool dwarfs selected to be coeval using the
@@
 describe the period-age-color/\teff\ relation for old stars.
-It appears that the period-color/\teff\ relation changes shape over time, \ie\
-the period-color/\teff\ relation and the period-age relation are
-non-separable.
-At ages younger than $\sim$1 Gyr, rotation period is anti-correlated with
-\teff: cooler stars spin more slowly than hotter stars of the same age.
-However, at old ages the opposite is true: cooler stars spin more rapidly than
-hotter stars of the same age.
-The period-color/\teff\ relation appears to start flattening out after $\sim$1
-Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
-older than the period gap which is located at a gyrochronal age of around
-$\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
-This gap may represent a significant transitional epoch in the magnetic
-behavior of stars.
+It appears that the period-color/\teff\ relation changes shape over time in a
+way that qualitatively agrees with theoretical that which include a
+mass-dependent core-envelope angular momentum transport \citep{spada2019}.
+At young ages, rotation period is anti-correlated with \teff: cooler stars
+spin more slowly than hotter stars of the same age.
+However, at intermediate ages the relation flattens out and K dwarfs rotate at
+the same rate, regardless of mass.
+At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
+dwarfs of the same age.
+We speculate that the rotation period gap \citep{mcquillan2014} may separate
+a young regime where stellar rotation periods decrease with increasing mass
+from an old regime where periods increase with increasing mass, however more
+data are needed to provide a conclusive result.
+% The period-color/\teff\ relation seems to start flattening out after $\sim$1
+% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
+% older than the period gap which is located at a gyrochronal age of around
+% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
 Finally, when velocity dispersion is interpreted as an age proxy, it appears
 that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
-suggests that lower-mass stars remain active for longer and their rotation
-periods can be measured at older ages.
-We outlined a number of scenarios which could provide alternative
-explainations for these observations, including incorrect dust corrections for
-the lowest-mass stars and an excess of companions increasing the velocity
-dispersion for these stars.
+suggests that lower-mass stars remain active for longer, allowing their
+rotation periods to be measured at older ages.
+% We outlined a number of scenarios which could provide alternative
+% explanations for these observations, including incorrect dust corrections for
+% the lowest-mass stars and an excess of companions increasing the velocity
+% dispersion for these stars.
 
-If the period-color/\teff\ relation does invert at old ages as our results
-suggest, this would be a paradigm shift for gyrochronology.
-Stellar spin-down rate is thought to be directly tied to magnetic field
-strength, and the deeper convection zones of cooler stars generate stronger
-magnetic fields which {\it should} lead to more efficient angular momentum
-loss.
-However, the micro- and macro-physics of stellar structure and evolution and
-magnetic dynamo models are extremely complicated and a lot is still unknown
-about the magnetic behavior of stars.
-Observations like these can provide useful constraints for physical models,
-and may help to reveal new physical processes at work in stars like our own
-Sun and other planet hosts.
+% If the period-color/\teff\ relation does invert at old ages as our results
+% suggest, this would be a paradigm shift for gyrochronology.
+% Stellar spin-down rate is thought to be directly tied to magnetic field
+% strength, and the deeper convection zones of cooler stars generate stronger
+% magnetic fields which {\it should} lead to more efficient angular momentum
+% loss.
+% However, the micro- and macro-physics of stellar structure and evolution and
+% magnetic dynamo models are extremely complicated and a lot is still unknown
+% about the magnetic behavior of stars.
+% Observations like these can provide useful constraints for physical models,
+% and may help to reveal new physical processes at work in stars like our own
+% Sun and other planet hosts.

### 2019-10-15 paper/discussion.tex
modifying paper text and experimenting with various cuts (the overall results are the same)
many paper edits

--- 2019-09-05 paper/discussion.tex
@@
 
-\subsection{The period gap}
-\label{sec:period_gap}
+% \subsection{The period gap}
+% \label{sec:period_gap}
 
-The origin of the rotation period gap, first identified
-by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+The results presented above indicate that stars of spectral type ranging from
+late G to late K follow a braking law that changes over time.
+These results provide further evidence for `stalled' rotational evolution of K
+dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
+by models that vary angular momentum transport between stellar core and
+envelope with time and mass \citep{spada2019}.
+
+Our results indicate that, at young ages, stellar rotation period decreases
+with mass.
+This is probably because lower-mass stars with deeper convection zones have
+stronger magnetic fields, larger Alfven radii and therefore experience greater
+angular momentum loss rate.
+According to the \citet{spada2019} models, there is minimal transportation of
+angular momentum from the surface to the core of the star at these young ages,
+so the surface slows down but the core keeps spinning rapidly.
+At intermediate ages, rotation periods appear constant with mass, and at late
+ages rotation period increases with mass for GK dwarfs.
+The explanation for this phenomenon, according to the \citet{spada2019}
+models, is that lower-mass stars are still braking more efficiently at these
+intermediate and old ages but their cores are more tightly coupled with their
+envelopes, so angular momentum resurfaces and prevents the stellar envelopes
+from spinning-down rapidly.
+
+The origin of the rotation period gap, first identified by
+\citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
 \ref{fig:dispersion_period_teff} still remains a mystery.
@@
 discontinuous star formation history \citep{mcquillan2013, davenport2017,
-davenport2018}, a rapid change in magnetic field structure
-\citep{reinhold2019}, and erroneous rotation period measurements that are
-incorrect by a factor of two \citep{koen2018}.
-The latter explanation can be ruled out because stars below the gap have
-smaller velocity dispersions than the stars above the gap, indicating that
-they are kinematically younger \citep{mcquillan2013, davenport2018}, as
-evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
+davenport2018} and a change in magnetic field structure causing a brief period
+where rotational variability is reduced and rotation periods cannot be
+measured \citep{reinhold2019}.
+Our results indicate there might be a slight difference in rotational
+evolution below and above the gap.
+Stars below the gap appear to be in the regime where wind-braking is dominant
+(rotation period decreases with increasing mass at a given age) and stars
+above the gap are in a regime where rotational coupling is dominant (rotation
+period increases with increasing mass at a given age).
+This can be seen in figures \ref{fig:age_cut} and \ref{fig:vplot}.
 For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
@@
 velocity dispersion increases with \teff.
-The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
-dispersion as the hottest stars in the age range above which indicates that
-the period-\teff\ relations are {\it flat} at these rotation periods.
-This is also visible in figure \ref{fig:dispersion_period_teff}.
-Below the gap, velocity dispersion within a given period range appears to {\it
-decrease} with decreasing temperature.
+In figure \ref{fig:age_cut}, velocity dispersion within a given period range
+appears to {\it decrease} with decreasing temperature below the gap.
 The opposite appears to be true above the gap.
-It could be that the gap is positioned at a significant Rossby number/age at
-which stellar magnetic dynamos go through a transition.
-Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
-magnetic braking is more efficient for stars with deeper convection zones.
-Once stars reach this critical age or Rossby number their magnetic fields
-undergo some radical transition, which produces the gap in the rotation
-period-\teff\ plane.
-After this transition, magnetic braking efficiency no longer increases with
-decreasing mass.
-Of course, it may be a coincidence that the gyrochronology relations seem to
-only flatten off above the period gap and we lack a sufficient quantity of
-data to do more than speculate here.
-New rotation periods from the \ktwo\ and \tess\ missions may be able to
-validate or rule out this hypothesis in the future.
+This suggests that perhaps stars transition from a wind braking regime to a
+rotational coupling regime as they pass across the gap.
+However, it is unclear how this process could {\it create} the gap.
+A gap could be created if stars rapidly lose angular momentum over a short
+period of time, leading to a dearth of stars in the gap.
+However, a sudden redistribution of angular momentum from the core to the
+surface, should cause stellar rotation periods to {\it decrease}, not
+increase.
+In addition, this theory is not supported by observations of the NGC 6811
+cluster whose K dwarf members have already undergone stalled braking below and
+before they reach the gap.
+The gap may be associated with a wind braking-rotational coupling transition,
+however our results only provide weak evidence for this, so we can only
+speculate until more data become available.
 
-In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
-are faster than expected; their rotational evolution appears to have stalled,
-and they have similar rotation periods to the 650 Myr Praesepe cluster
-\citep{curtis2019}.
-The rotation periods of the K dwarfs in this cluster are plotted in figure
-\ref{fig:dispersion_period_teff}.
-Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
-dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
-NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
-K dwarfs lie below it.
-This cluster may be the `missing link' that connects two epoch of stellar
-spin-down.
-However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
-model is, on average, a good model for stars younger than $\sim$1 Gyr. This
-model has a very different shape to the NGC 6811 cluster.
-This would suggest that stars in the field do follow the period-color/\teff\
-relation of Praesepe, at least up to around 1 Gyr.
-We do not find evidence of the K dwarf stalling, which is observed in NGC
-6811, in the field.
-% , however without more observations of middle-aged open clusters it : an
-% early stage where the period-\teff\ relation for cool dwarfs has a negative
-% slope and a late stage where it has a positive slope.
-% In this case the period gap may delineate the transition between these two
-% regimes and is the point at which stellar magnetic dynamos likely undergo a
-% dramatic structural shift at an age of $\sim$ 1.1 Gyr.
+% The \citet{spada2019} models do predict that stars transition from a regime
+% where wind-braking is dominant to a regime where rotational coupling is
+% domimant at different times for different masses.
+% However, these models are based on observations of NGC 6811, whose K dwarfs
+% already show signs of stalling by 1.1 Gyr.
+% For stellar masses of 0.7 M$_\odot$, this transition takes place at $\sim$ 1
+% Gyr.
+% It happens at 500 Myr for 0.9 M$_\odot$ and 5 Gyr for 0.5 M$_\odot$
+% \citep[see][for details]{spada2019}.
+% We propose a new theory for the origin of the rotation gap as the transition
+% between the wind-braking and rotational coupling dominated regimes.
+% The gap could be caused by rapidly transitioning across the gap
+% as
+
+% These results are extremely tentative and should be interpreted with caution:
+% more data are needed to confirm this behavior.
+
+% For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+% \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+% temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+% velocity dispersion increases with \teff.
+% The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+% dispersion as the mid-temperature stars in the age range above, and the
+% hottest stars in the age range above {\it that}, which indicates that the
+% period-\teff\ relations are {\it flat} at rotation periods between $\sim$15-25
+% days, for 5500 K $<$ \teff\ $<$ 4000 K.
+% Below the gap, velocity dispersion within a given period range appears to {\it
+% decrease} with decreasing temperature.
+% The opposite appears to be true above the gap.
+
+% The gap may be positioned at a significant Rossby number/age at which stellar
+% magnetic dynamos go through a transition.
+% Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
+% magnetic braking is more efficient for stars with deeper convection zones.
+% Once stars reach this critical age or Rossby number their magnetic fields
+% undergo some transition, which produces the gap in the rotation period-\teff\
+% plane and after this transition, magnetic braking efficiency no longer
+% increases with decreasing mass.
+% Of course, it may be a coincidence that the gyrochronology relations seem to
+% only flatten off above the period gap and we lack a sufficient quantity of
+% data to do more than speculate here.
+% New rotation periods from the \ktwo\ and \tess\ missions may be able to
+% validate or rule out this hypothesis in the future.
+
+% In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
+% are faster than expected; their rotational evolution appears to have stalled,
+% and they have similar rotation periods to the 650 Myr Praesepe cluster
+% \citep{curtis2019}.
+% The rotation periods of the K dwarfs in NGC 6811 are plotted in figure
+% \ref{fig:dispersion_period_teff}.
+% Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
+% dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
+% NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
+% K dwarfs lie below it.
+% This cluster may be the `missing link' that connects two epoch of stellar
+% spin-down.
+% However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
+% model is, on average, a {\it good} model for stars younger than $\sim$1 Gyr
+% % and this model has a very different shape to the NGC 6811 cluster.
+% suggesting that stars in the field follow the period-color/\teff\ relation of
+% Praesepe, at least up to around 1 Gyr.
+% We do not see strong evidence of K dwarf stalling in the field.
+% % , however without more observations of middle-aged open clusters it : an
+% % early stage where the period-\teff\ relation for cool dwarfs has a negative
+% % slope and a late stage where it has a positive slope.
+% % In this case the period gap may delineate the transition between these two
+% % regimes and is the point at which stellar magnetic dynamos likely undergo a
+% % dramatic structural shift at an age of $\sim$ 1.1 Gyr.
 

### 2019-10-15 paper/method.tex
adding a bunch of paper files -- not sure why these weren't staged?
modifying paper text and experimenting with various cuts (the overall results are the same)
many paper edits

@@
 greater than 1 \kms\, and stars with galactic latitudes greater than
-15\degrees\ were removed from the sample.
-Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
-and because latitude is correlated with stellar mass (lower mass stars are
-older and tend to be preferentially located at high $b$, although this trends
-is greatly reduced for stars cooler than 5000 K), we found that excluding
-stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
-stellar mass.
-15\degrees\ was found to be an adequate compromise between maintaining a close
-relationship between \vb\ and \vz\ and retaining a sample large enough to
-provide meaningful results.
-Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
-trends in the data, and intermediate cuts did not significantly change the
-results.
+15\degrees\ (justification provided below) were removed from the sample.
+% Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
+% and because latitude is correlated with stellar mass (lower mass stars are
+% older and tend to be preferentially located at high $b$, although this trends
+% is greatly reduced for stars cooler than 5000 K), we found that excluding
+% stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
+% stellar mass.
+% 15\degrees\ was found to be an adequate compromise between maintaining a close
+% relationship between \vb\ and \vz\ and retaining a sample large enough to
+% provide meaningful results.
+% Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
+% trends in the data, and intermediate cuts did not significantly change the
+% results.
 % We also found that the anti-correlation between stellar mass and $b$ is
@@
 
-\subsection{Establishing \vb\ as an age proxy}
+\subsection{Validating \vb\ dispersion as an age proxy}
 \label{sec:mass-dependent-heating}
+
+There are two main reasons why \vb\ velocity dispersion may not be a good age
+proxy.
+Firstly, mass-dependent heating may act on the sample, meaning that velocity
+dispersion depends on both age and mass, so cannot be interpreted as a simple
+age proxy.
+Secondly, since stars in the \kepler\ field have a range of galactic
+latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
+stars, introducing a velocity bias for high latitude stars (which are more
+likely to be cooler and older).
+In this section we demonstrate that neither of these problems seem to be a
+significant issue for our data.
 
@@
 searched for signs of mass-dependent heating within the \kepler\ field.
-
 Mass-dependent dynamical heating has not been unambiguously observed in the
@@
 evidence for mass-dependent heating has been found in M dwarfs
-\citep{faherty2009}.
+\citep[\eg][]{faherty2009, newton2016}.
 
@@
 We could not perform this analysis on the \mct\ sample, because only stars
-with {\it detectable} rotation periods appear there.
-Since lower-mass stars stay active for longer it is likely that there are more
-old stars (with larger velocity dispersions) at low masses in this sample.
+with {\it detectable} rotation periods are included in it.
+Since lower-mass stars stay active for longer it is likely that the oldest
+stars in this sample have low masses.
 We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
@@
 polynomial, fit to the MS on the \gaia\ CMD.
-We then applied the quality cuts described in section \ref{sec:the_data}.
+We then applied the quality cuts described above in section
+\ref{sec:the_data}.
 To search for evidence of mass-dependent heating we calculated the (\vb)
@@
 outliers before calculating the standard deviation of stars in each bin.
+These high velocity outliers may be very old late K and M dwarfs, or they
+result from using \vb\ instead of \vz, which introduces additional velocity
+scatter.
 
@@
 an age proxy.
-We performed the same analysis on the 537 stars in this sample with RVs using
-{\it vertical} velocity (\vz) and found the slope of the velocity
-dispersion-temperature relation was consistent with zero.
+We also examined the {\it vertical} velocities of the 537 stars in this sample
+with RV measurements.
+Again, we found no evidence for mass-dependent heating: the slope of the
+velocity dispersion-temperature relation was consistent with zero.
 
-Since we used \vb, not \vz\ in our analysis, it is possible that an
-anti-correlation between \teff\ and galactic latitude, $b$, could influence
-our results.
-Stars at higher latitudes have additional velocity components in the ${\bf x}$
-and ${\bf y}$ directions, which could increase \vb\ but not \vz.
-Again however, since the relationship between \sigmavb\ and \teff\ is
-positively, not negatively correlated for cool stars in the \kepler\ field,
-this effect is probably too small to influence our results.
+At a galactic latitude of zero, $v_b=v_z$, however for increasing values of
+$b$, this equivalence becomes an approximation that grows noisier with $b$.
+To test the validity of the \vb$\sim$\vz\ approximation over a range of
+latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
+(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
+We downloaded stars from four pointings in the \kepler\ field with galactic
+latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
+a limiting magnitude of 16 dex.
+We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
+with a standard deviation (calculated as 1.5 $\times$ the median absolute
+deviation) that increased with increasing $b$.
+The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
+$, 4kms${^-1}$ at 10$^\circ$, 6kms${^-1}$ at 15$^\circ$, and 9kms${^-1}$ at
+20$^\circ$.
+Since we are concerned with velocity {\it dispersions}, rather than velocities
+themselves, we also compared the \vb\ and \vz\ velocity dispersions for stars
+downloaded from the GUMS simulation.
+For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
+consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
+latitudes the two quantities became significantly different.
+For this reason we proceeded by only including stars with galactic latitudes
+less than 15\degrees\ in our analysis.
+
+% Since we use \vb, not \vz\ in our analysis, it is possible that the sample
+% selection function could influence our results.
+% For example, higher-mass stars tend to be younger and reside closer to the
+% galactic disk mid-plane (low galactic latitudes).
+% Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
+% lower-mass stars typically have greater velocity dispersions due to .
+% \teff\ and galactic latitude, $b$,
+% could result in
+% Stars at higher latitudes have additional velocity components in the ${\bf x}$
+% and ${\bf y}$ directions, which could increase \vb\ but not \vz.
+% Again however, since the relationship between \sigmavb\ and \teff\ is
+% positively, not negatively correlated for cool stars in the \kepler\ field,
+% this effect is probably too small to influence our results.
 

### 2019-11-05 paper/abstract.tex
changes to main_figure and paper

@@
 % gyrochronology relations.
-In this work, we use the velocities of low-mass \kepler\ dwarfs to reveal
-their rotational evolution and demonstrate that kinematics have potential to
-be a useful tool for calibrating gyrochronology.
+In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
+proxy, to reveal their rotational evolution and demonstrate that kinematics
+could be a useful tool for calibrating gyrochronology.
 At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated

### 2019-11-05 paper/discussion.tex
changes to main_figure and paper

@@
 late G to late K follow a braking law that changes over time.
+In particular, the relationship between rotation period and effective
+temperature appears to flatten out and eventually invert.
 These results provide further evidence for `stalled' rotational evolution of K
@@
 
-Our results indicate that, at young ages, stellar rotation period decreases
-with mass.
+The velocity dispersions of stars in the \mct\ sample provide the following
+picture of rotational evolution.
+At young ages, stellar rotation period decreases with mass.
 This is probably because lower-mass stars with deeper convection zones have
@@
 so the surface slows down but the core keeps spinning rapidly.
-At intermediate ages, rotation periods appear constant with mass, and at late
-ages rotation period increases with mass for GK dwarfs.
-The explanation for this phenomenon, according to the \citet{spada2019}
-models, is that lower-mass stars are still braking more efficiently at these
-intermediate and old ages but their cores are more tightly coupled with their
-envelopes, so angular momentum resurfaces and prevents the stellar envelopes
-from spinning-down rapidly.
+At intermediate ages, rotation period is constant with mass, and at late ages
+rotation period {\it increases} with mass for GK dwarfs.
+The explanation for this, according to the \citet{spada2019} models, is that
+lower-mass stars are still braking more efficiently at these intermediate and
+old ages but their cores are more tightly coupled to their envelopes, allowing
+angular momentum transport between the two interior layers.
+Angular momentum resurfaces and prevents the stellar envelopes from
+spinning-down rapidly.
 
-The origin of the rotation period gap, first identified by
-\citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
-\ref{fig:dispersion_period_teff} still remains a mystery.
-This gap can be seen as an under-density of points between the 0.7-1.0 and
-1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
-of constant gyrochronal age of around 1.1 Gyr \citep[according to the
-gyrochronology relation of][]{angus2019}, as shown in figure
-\ref{fig:dispersion_period_teff}.
-Several explanations for the gap's origin have been proposed, including a
-discontinuous star formation history \citep{mcquillan2013, davenport2017,
-davenport2018} and a change in magnetic field structure causing a brief period
-where rotational variability is reduced and rotation periods cannot be
-measured \citep{reinhold2019}.
-Our results indicate there might be a slight difference in rotational
-evolution below and above the gap.
-Stars below the gap appear to be in the regime where wind-braking is dominant
-(rotation period decreases with increasing mass at a given age) and stars
-above the gap are in a regime where rotational coupling is dominant (rotation
-period increases with increasing mass at a given age).
-This can be seen in figures \ref{fig:age_cut} and \ref{fig:vplot}.
-For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
-\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
-temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
-velocity dispersion increases with \teff.
-In figure \ref{fig:age_cut}, velocity dispersion within a given period range
-appears to {\it decrease} with decreasing temperature below the gap.
-The opposite appears to be true above the gap.
-This suggests that perhaps stars transition from a wind braking regime to a
-rotational coupling regime as they pass across the gap.
-However, it is unclear how this process could {\it create} the gap.
-A gap could be created if stars rapidly lose angular momentum over a short
-period of time, leading to a dearth of stars in the gap.
-However, a sudden redistribution of angular momentum from the core to the
-surface, should cause stellar rotation periods to {\it decrease}, not
-increase.
-In addition, this theory is not supported by observations of the NGC 6811
-cluster whose K dwarf members have already undergone stalled braking below and
-before they reach the gap.
-The gap may be associated with a wind braking-rotational coupling transition,
-however our results only provide weak evidence for this, so we can only
-speculate until more data become available.
+% The origin of the rotation period gap, first identified by
+% \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+% \ref{fig:dispersion_period_teff} still remains a mystery.
+% This gap can be seen as an under-density of points between the 0.7-1.0 and
+% 1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
+% of constant gyrochronal age of around 1.1 Gyr \citep[according to the
+% gyrochronology relation of][]{angus2019}, as shown in figure
+% \ref{fig:dispersion_period_teff}.
+% Several explanations for the gap's origin have been proposed, including a
+% discontinuous star formation history \citep{mcquillan2013, davenport2017,
+% davenport2018} and a change in magnetic field structure causing a brief period
+% where rotational variability is reduced and rotation periods cannot be
+% measured \citep{reinhold2019}.
+% Our results indicate there might be a slight difference in rotational
+% evolution below and above the gap.
+% Stars below the gap appear to be in the regime where wind-braking is dominant
+% (rotation period decreases with increasing mass at a given age) and stars
+% above the gap are in a regime where rotational coupling is dominant (rotation
+% period increases with increasing mass at a given age).
+% This can be seen in figures \ref{fig:age_cut} and \ref{fig:vplot}.
+% For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+% \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+% temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+% velocity dispersion increases with \teff.
+% In figure \ref{fig:age_cut}, velocity dispersion within a given period range
+% appears to {\it decrease} with decreasing temperature below the gap.
+% The opposite appears to be true above the gap.
+% This suggests that perhaps stars transition from a wind braking regime to a
+% rotational coupling regime as they pass across the gap.
+% However, it is unclear how this process could {\it create} the gap.
+% A gap could be created if stars rapidly lose angular momentum over a short
+% period of time, leading to a dearth of stars in the gap.
+% However, a sudden redistribution of angular momentum from the core to the
+% surface, should cause stellar rotation periods to {\it decrease}, not
+% increase.
+% In addition, this theory is not supported by observations of the NGC 6811
+% cluster whose K dwarf members have already undergone stalled braking below and
+% before they reach the gap.
+% The gap may be associated with a wind braking-rotational coupling transition,
+% however our results only provide weak evidence for this, so we can only
+% speculate until more data become available.
 

### 2019-11-05 paper/intro.tex
changes to main_figure and paper

@@
 
-It is well established that magnetized stellar winds cause the rotation
-periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
-weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
-calibrated, the relationship between period, mass and age can be used to
-age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
-meibom2011, meibom2015}.
-% However, the exact nature of rotational evolution is still being actively
-% studied.
-% Once the relationship between rotation period, mass and age is well
-% calibrated, rotation periods can be used to precisely and accurately date
-% stars via gyrochronology \citep[\eg][]{kawaler1989, pinsonneault1989,
-% barnes2003, barnes2007, barnes2010, meibom2011, meibom2015}.
-% It is also well established that stars of the same age but different masses
-However, new photometric rotation periods measured from \kepler\ light curves
-reveal that rotational evolution is more complicated than previously thought.
-For example, the M dwarfs in the $\sim$ 700 Myr Praesepe cluster spin more
+Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
+strong magnetic fields and spin more slowly over time through magnetic braking
+\citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
+pinsonneault1989}.
+Although stars are typically born with random rotation periods, ranging from 1
+to 10 days, observations of young open clusters reveal that their rotation
+periods converge onto a unique sequence by $\sim$500-700 million years
+\citep[\eg][]{irwin2009}.
+After this time, it is thought that the rotation period of a star is
+determined, to first order, by its color and age alone.
+This is the principle behind gyrochronology -- the method of inferring a
+star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
+barnes2010, meibom2011, meibom2015}.
+% It is well established that magnetized stellar winds cause the rotation
+% periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
+% weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
+% calibrated, the relationship between period, mass and age can be used to
+% age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
+% meibom2011, meibom2015}.
+However, new photometric rotation periods made available by the \kepler\
+\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
+\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
+meibom2015, curtis2019} reveal that rotational evolution is more complicated
+than previously thought.
+For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
 slowly than the G dwarfs -- in theory because lower-mass stars have deeper
 convections zones which generate stronger magnetic fields and more efficient
-magnetic braking \citep{schatzman1962, weber1967}.
+magnetic braking.
 However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
@@
 period-color relation \citep{spada2019}.
-These models suggest that mass and age-dependent coupling between the cores
-and envelopes of stars has a significant impact on their surface rotation
-rates.
+These models suggest that mass and age-dependent angular momentum transport
+between the cores and envelopes of stars has a significant impact on their
+surface rotation rates.
 
@@
 t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
-the observations \citep{vansaders2016, vansaders2018}.
-% revealing that rotational
-% evolution is clearly a more complicated process than once thought.
-% However, an underlying assumption behind many empirical gyrochronology
-% relations is that the relationships between rotation period and photometric
-% color\footnote{As the directly observable quantity, color is often used as a
-% mass proxy, and empirical gyrochronology relations are usually calibrated in
-% color, rather than mass or effective temperature.},
-% and rotation period and age are {\it separable}, meaning that the
-% period-color relation is the same at all ages and the period-age relation is
-% the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
-% meibom2011, angus2015, angus2019}.
-% Not all gyrochronology models follow this form, particularly those based on
-% stellar structure and evolution theory or with a Rossby number, rather than
-% rotation period dependence \citep[\eg][]{barnes2010, matt2012,
-% vansaders2016]}.
-% For example, old field stars seem rotate more rapidly than models predicted
-% \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}, and a
-% mass-dependent modification to the classical \prot\ $\propto t^{\frac{1}{2}}$
-% spin-down law \citep{skumanich1972} is required to reproduce the data
-% \citep{vansaders2016, vansaders2018}.
-% In addition, recent analyses of the 1.1 Gyr open cluster, NGC 6811, provide
-% further indication that rotational evolution is mass and time-dependent
-% \citep{curtis2019 spada2019}.
-% This cluster has a flattened relationship between rotation period and
-% color: the early-K dwarfs rotate at the same rate as the late-K dwarfs.
-% With the advent of large time-series photometry surveys, providing large
-% numbers of precise rotation period measurements over the last few years.
+these observations.
+To fit magnetic braking models to these data, a cessation of magnetic braking
+is required after stars reach a Rossby number (ratio of rotation period to
+convective turnover time) of around 2 \citep{vansaders2016, vansaders2018}.
 
@@
 into their rotational evolution.
-% This technique provides relative ages for groups of stars that cover a broad
-% range of masses and ages, allowing us to observe the rotational evolution of
-% stars in an average sense.
+Velocity dispersion provides relative ages for groups of stars, allowing us to
+observe the rotational evolution of stars over a range of masses and ages.
 
@@
 Although the velocity of any individual star will only provide a weak age
-constraint, the velocity dispersion of a group of stars can indicate whether,
-on average, that group is old or young relative to other groups.
+constraint, the velocity {\it dispersion} of a {\it group} of stars can
+indicate whether, on average, that group is old or young relative to other
+groups.
 In this work we compare the velocity dispersions of groups of field stars in

### 2019-11-05 paper/method.tex
changes to main_figure and paper

@@
 in \gaia\ G-band.
+
+The dwarf stars in the \mct\ sample are shown on a \gaia\ color-magnitude
+diagram (CMD) in figure \ref{fig:age_gradient}.
+The stars are colored by their gyrochronal age, calculated using the
+\citep{angus2019} gyrochronology relation.
+The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
+along the upper edge of the MS, where stellar evolution models predict old
+stars to be, however the majority of these `old' stars are bluer than \gcolor\
+$\sim$ 1.5 dex.
+This suggests that either old M dwarfs are missing from the \mct\ catalog, or
+the \citet{angus2019} gyrochronology relation under-predicts the ages of
+low-mass stars.
+Given that lower-mass stars stay active for longer than higher-mass stars
+\citep[\eg][]{west2011}, and are therefore more likely to have measurable
+rotation periods at old ages, the latter scenario seems more likely.
 \begin{figure}
@@
 We could not perform this analysis on the \mct\ sample, because only stars
-with {\it detectable} rotation periods are included in it.
-Since lower-mass stars stay active for longer it is likely that the oldest
-stars in this sample have low masses.
+with {\it detectable} rotation periods are included in it and since lower-mass
+stars stay active for longer it is likely that the oldest stars in this sample
+have low masses.
 We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
@@
 We also eliminated visual binaries by removing stars above a 6th order
-polynomial, fit to the MS on the \gaia\ CMD.
+polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
+figure \ref{fig:age_gradient}).
 We then applied the quality cuts described above in section
@@
 
-Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
-as a function of effective temperature (calculated by transforming dereddened
-\gaia\ colors using equation \ref{eq:curtis}).
+Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
+function of effective temperature \footnote{calculated by transforming
+dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
+\kepler\ dwarf sample.
 Velocity dispersion very slightly {\it decreases} with decreasing temperature,
@@
 smaller velocities.
-The essential point is that we do not find evidence for mass-dependent heating
-acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be used as
-an age proxy.
+The essential point however, is that we do not see evidence for mass-dependent
+heating acting on stars in the \kepler\ field, indicating that \sigmavb\ {\it
+can} be used as an age proxy.
 We also examined the {\it vertical} velocities of the 537 stars in this sample
@@
 % \mct\ sample.
+
+To explore the relationship between rotation period, \teff\ and age, we
+removed high and low velocity outliers from the \mct\ sample by performing
+3$\sigma$ sigma-clipping on the \vb\ velocities.
+Without sigma-clipping, we found that a small number of high velocity
+outliers at the low-temperature end of our sample substantially raised the
+velocity dispersion for cooler stars, however the overall trends remained the
+same.
+We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
+3500 K to avoid biases caused by the selection function at the faint, cool
+end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
+
+% We then selected groups of stars within different gyrochronal age ranges and
+% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
+% of effective temperature for each age group.
+% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+% throughout this paper we show rotation periods as a function of effective
+% temperature, \teff, because it is easier to divide stars into bins of roughly
+% equal numbers in \teff-space than in color-space.

### 2019-11-05 paper/results.tex
changes to main_figure and paper

--- 2019-10-10 paper/results.tex
@@
 
-\subsection{Velocity dispersion of coeval groups}
-\label{sec:age_cut}
+% \subsection{Velocity dispersion of coeval groups}
+% \label{sec:age_cut}
 
-To explore the relationship between rotation period, \teff\ and age, we
-selected groups of stars within different gyrochronal age ranges \citep[where
-age was calculated using the][gyrochronology relation]{angus2019}, and
-calculated the velocity dispersion (\sigmavb), the standard deviation of \vb\
-velocities, as a function of effective temperature for each age group.
-We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
-and temperature bin to remove non-Gaussian outliers.
-Without sigma-clipping, we found that a small number of high velocity
-outliers at the low-temperature end of our sample substantially raised the
-velocity dispersion for cooler stars.
-Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-throughout this paper we show rotation periods as a function of effective
-temperature, \teff.
-We chose to use \teff\ as the abscissa because it is easier to divide stars
-into bins of roughly equal numbers in \teff-space than in color-space.
+% To explore the relationship between rotation period, \teff\ and age for field
+% K dwarfs, we calculated gyrochronal ages using dereddened \Gaia\ photometry
+% ($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
+% We used the \citet{angus2019} gyrochronology relation which is a simple,
+% separable relation in \gcolor\ color and age, calibrated using the
+% period-color relation of Praesepe and the period-age relation of Praesepe and
+% the Sun.
+% The large number of Praesepe members with precise rotation periods from the
+% \ktwo\ mission \citep{howell2014, douglas2017, rebull2017}, spanning spectral
+% types F through early M, makes it a good cluster for calibrating the
+% period-color relation of stars at young ages.
+% However, this is an extremely simple model and although it accurately
+% describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
+% of NGC 6819 -- the oldest cluster with available rotation periods), it
+% over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
+% cluster.
+% These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
+% figure \ref{fig:age_gradient}.
+% The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
+% along the upper edge of the MS, where stellar evolution models predict old
+% stars to be, however the majority of these `old' stars are bluer than \gcolor\
+% $\sim$ 1.5 dex.
+% This suggests that either old M dwarfs are missing from the \mct\ catalog, or
+% the \citet{angus2019} gyrochronology relation under-predicts the ages of
+% low-mass stars.
+% Given that lower-mass stars stay active for longer than higher-mass stars
+% \citep[\eg][]{west2011}, and are therefore more likely to have measurable
+% rotation periods at old ages, the latter scenario seems more likely.
 
-\begin{figure}
-  \caption{
-Top: rotation period vs effective temperature for stars in the \mct\
-    catalog.
-    The full catalog, with subgiants and visual binaries removed is shown in
-    grey, and stars selected to be in different age groups (between \tmin\ and
-    \tmax\ K) are overlayed in color.
-These age groups were selected using the \citet{angus2019} gyrochronology
-    relation.
-The legend in the center of the figure lists the age range, in Gyr, of each
-    group.
-Bottom: velocity dispersion vs effective temperature for each age
-    group.
-The color of the line corresponds to the color of the group shown in the top
-    panel.
-If the gyrochronal model were correct at all ages, and the stars in each group
-    were the same age across temperatures, the velocity dispersion would be
-    constant as a function of \teff.
-However, the velocity dispersions of the oldest age groups increase with
-    \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
-    the the ages of late-K dwarfs relative to the ages of late G and early K
-    dwarfs at old ages.
-% An alternative explanation could be that the gyrochronology relation is
-%     correct and {\it mass-dependent heating} is responsible for the greater
-%     velocity dispersions of cooler stars.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{age_cut}
-\label{fig:age_cut}
-\end{figure}
-The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
-(excluding visual binaries and subgiants) in grey, with coeval groups shown in
-color.
-The color of the points corresponds to the age ranges specified in the legend
-(in Gyr), which also apply to the lines in the lower panel.
-The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
-as a function of effective temperature.
-% We only included stars within a temperature range of 5000 - 3500 K in our
-% analysis, as hotter stars are more likely to have stopped magnetic braking
-% \citep{vansaders2016}, which could bias the results.
-% Late M dwarfs were not included in our analysis because such faint stars
-% cannot be observed at large heights above the plane (because of the low
-% galactic latitude of the \kepler\ field, stars at high-Z are more distant),
-% which introduces a mass-dependent velocity bias: cooler populations of stars
-% are skewed towards lower velocity dispersions.
-% The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
-% have low velocity dispersions, indicating that this effect may already become
-% important at temperatures lower than $\sim$ 4000 K.
-Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
-with gyrochronal age across all temperatures, implying that both velocity
-dispersion and rotation periods increase with age as expected.
-The flat velocity dispersion of young stars as a function of temperature
-shows that the Praesepe-calibrated gyrochronology relation accurately predicts
-the relative ages of {\it young} field stars.
-% To our knowledge, no gyrochronology relation has previously been demonstrated to
-% correctly predict ages (either relative or absolute) for such cool or such
-% young field stars.
-% This is not particularly remarkable however, since these young stars are a
-% similar age to the Praesepe cluster, which was used to calibrate the
-% period-color relation.
-% However, if the stars in each selected age group had the {\it same} age across
-% the temperature range, their velocity dispersion would be a constant function
-% of \teff.
-If the gyrochronology relations worked at all ages and temperatures, the
-bottom panel of figure \ref{fig:age_cut} would show a flat relationship
-between velocity dispersion and \teff at all ages.
-However, velocity dispersion {\it increases} as a function of temperature for
-old stars, meaning the \citet{angus2019} gyrochronology relation either
-under-predicts the ages of old, late-K dwarfs, or over-predicts the ages of
-old early-K and late-G dwarfs\footnote{Since \vb\ velocity dispersion only
-provides relative and not absolute ages, it is difficult to tell whether the
-ages of cool stars are being under-predicted, the ages of hot stars being are
-over-predicted, or both.}.
-This suggests that the relationship between rotation period and photometric
-color or \teff\ flattens out over time, and possibly even inverts.
-The lines of constant age (isochrones) sweeping diagonally upwards in the top
-panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
+% To further explore the relationship between rotation period, \teff\ and age,
+% we first removed high and low velocity outliers by performing 3$\sigma$
+% sigma-clipping on the \vb\ velocities.
+% Without sigma-clipping, we found that a small number of high velocity
+% outliers at the low-temperature end of our sample substantially raised the
+% velocity dispersion for cooler stars.
+% We then selected groups of stars within different gyrochronal age ranges and
+% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
+% of effective temperature for each age group.
+% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+% throughout this paper we show rotation periods as a function of effective
+% temperature, \teff, because it is easier to divide stars into bins of roughly
+% equal numbers in \teff-space than in color-space.
+
+% \begin{figure}
+%   \caption{
+% Top: rotation period vs effective temperature for stars in the \mct\
+%     catalog.
+%     The full catalog, with subgiants and visual binaries removed is shown in
+%     grey, and stars selected to be in different age groups (between \tmin\ and
+%     \tmax\ K) are overlayed in color.
+% These age groups were selected using the \citet{angus2019} gyrochronology
+%     relation.
+% The legend in the center of the figure lists the age range, in Gyr, of each
+%     group.
+% Bottom: velocity dispersion vs effective temperature for each age
+%     group.
+% The color of the line corresponds to the color of the group shown in the top
+%     panel.
+% If the gyrochronal model were correct at all ages, and the stars in each group
+%     were the same age across temperatures, the velocity dispersion would be
+%     constant as a function of \teff.
+% However, the velocity dispersions of the oldest age groups increase with
+%     \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
+%     the the ages of late-K dwarfs relative to the ages of late G and early K
+%     dwarfs at old ages.
+% % An alternative explanation could be that the gyrochronology relation is
+% %     correct and {\it mass-dependent heating} is responsible for the greater
+% %     velocity dispersions of cooler stars.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{age_cut}
+% \label{fig:age_cut}
+% \end{figure}
+% The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
+% (excluding visual binaries and subgiants) in grey, with coeval groups
+% \citep[according to the][relation]{angus2019} shown in color.
+% The color of the points corresponds to the age ranges specified in the legend
+% (in Gyr), which also apply to the lines in the lower panel.
+% The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
+% as a function of effective temperature.
+% % We only included stars within a temperature range of 5000 - 3500 K in our
+% % analysis, as hotter stars are more likely to have stopped magnetic braking
+% % \citep{vansaders2016}, which could bias the results.
+% % Late M dwarfs were not included in our analysis because such faint stars
+% % cannot be observed at large heights above the plane (because of the low
+% % galactic latitude of the \kepler\ field, stars at high-Z are more distant),
+% % which introduces a mass-dependent velocity bias: cooler populations of stars
+% % are skewed towards lower velocity dispersions.
+% % The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
+% % have low velocity dispersions, indicating that this effect may already become
+% % important at temperatures lower than $\sim$ 4000 K.
+% Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
+% with gyrochronal age across all temperatures, implying that both velocity
+% dispersion and rotation period increases with age as expected.
+% The flat velocity dispersion of young stars as a function of temperature
+% shows that the Praesepe-calibrated gyrochronology relation accurately predicts
+% the relative ages of {\it young} field stars.
+% % To our knowledge, no gyrochronology relation has previously been demonstrated to
+% % correctly predict ages (either relative or absolute) for such cool or such
+% % young field stars.
+% % This is not particularly remarkable however, since these young stars are a
+% % similar age to the Praesepe cluster, which was used to calibrate the
+% % period-color relation.
+% % However, if the stars in each selected age group had the {\it same} age across
+% % the temperature range, their velocity dispersion would be a constant function
+% % of \teff.
+% If \citet{angus2019} gyrochronology relation worked at all ages and
+% temperatures, the bottom panel of figure \ref{fig:age_cut} would show a flat
+% relationship between velocity dispersion and \teff at all ages.
+% However, velocity dispersion {\it increases} as a function of temperature for
+% old stars, meaning the \citet{angus2019} gyrochronology relation either
+% under-predicts the ages of old, late-K dwarfs, or over-predicts the ages of
+% old early-K and late-G dwarfs\footnote{Since \vb\ velocity dispersion only
+% provides relative and not absolute ages, it is difficult to tell whether the
+% ages of cool stars are being under-predicted, the ages of hot stars being are
+% over-predicted, or both.}.
+% This suggests that the relationship between rotation period and photometric
+% color or \teff\ flattens out over time, and possibly even inverts.
+% % The lines of constant age (isochrones) sweeping diagonally upwards in the top
+% % panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
 
@@
     \includegraphics[width=1\textwidth]{vplot}
-\label{fig:dispersion_period_teff}
+\label{fig:vplot}
 \end{figure}
-Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
-our sample, coloured by (\vb) velocity dispersion, where \sigmavb\ was
-calculated for groups of stars over a grid in $\log_{10}$(period) and
-temperature.
+Figure \ref{fig:vplot} shows rotation period versus effective temperature for
+the \mct\ sample, coloured by the standard deviation of their (\vb)
+velocities, where \sigmavb\ was calculated for groups of stars over a grid in
+$\log_{10}$(period) and temperature.
 If we assume that mass dependent heating does not strongly affect this sample
@@
 velocity dispersion can be interpreted as an age proxy and stars plotted in a
-similar color in figure \ref{fig:dispersion_period_teff} are similar ages.
-Interpreted this way, lines of constant age (isochrones) appear to follow the
-shape of the Praesepe-based gyrochronology model (black solid and dotted
-lines) at young ages.
+similar color in figure \ref{fig:vplot} are similar ages.
+
+Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
+gyrochronal age across all temperatures, implying that both velocity
+dispersion and rotation period increases with age as expected.
+Lines of constant age (isochrones) appear to follow the shape of the
+Praesepe-based gyrochronology model (black dotted line) at young ages.
 However, at older ages it appears that the relation between rotation period
@@
 
-The shape of the period-\teff\ relations at old ages appears to follow the
-shape of the upper detection edge.
-The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
-period catalog characterized by a dearth of slowly rotating M dwarfs between
-$\sim$2750 and 4250 K, is reflected in the lines of constant velocity
-dispersion (and presumed age) in the top panel of figure
-\ref{fig:dispersion_period_teff}.
-If the shape of the upper edge of rotation period measurements is created by a
-detection limit, it could indicate the rotation periods at which stars of
-different temperatures become relatively inactive (and their rotation periods
-therefore become undetectable).
-% Once inactive, stars would have little photometric variability induced by
-% star spots and their rotation periods would be difficult to measure, so may
-% not feature in the \mct\ catalog.
-If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
-$\sim$4500 K become inactive at a similar age, and stay active longer than
-stars hotter than $\sim$4500 on average.
+Figure \ref{fig:masses_with_models} shows the same as figure \ref{fig:vplot},
+however mass is plotted on the x-axis instead of temperature and the models
+are those of \citet{spada2019}.
+The masses of these stars are taken from the \kepler\ input catalog
+\citep{KIC}.
+The \citet{spada2019} models appear to qualitatively agree with the data:
+their rotation period-\teff\ relation flattens out over time and eventually
+inverts over a narrow range of temperatures.
+The models reproduce the `dip' feature seen in the data.
 
-% The groups of stars with the largest velocity dispersions are cooler than 4500
-% K, presumably because cooler stars remain active for longer, and their
-% rotation periods can therefore be measured at older ages.
+% Each star is assigned the same velocity dispersion as in figure
+% \ref{fig:vplot}, and because stars in the same temperature group are not
+% necessarily in the same mass group, which is why stars of the same mass and period do not nece
 
-The velocity dispersions of the coolest stars may be affected by a selection
-bias -- these extremely faint stars are more difficult to detect at larger
-distances, larger heights above the plane, and therefore larger velocities.
-It is possible that some high velocity stars of temperatures cooler than
-$\sim$4000 K are missing from this sample, and the velocity dispersions may
-therefore appear lower than they truly are.
-This could be why the velocity dispersion appears to decrease towards the
-right of figure \ref{fig:dispersion_period_teff}.
+% The shape of the period-\teff\ relations at old ages appears to follow the
+% shape of the upper detection edge.
+% The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
+% period catalog characterized by a dearth of slowly rotating M dwarfs between
+% $\sim$3750 and 4250 K, is reflected in the lines of constant velocity
+% dispersion (and presumed age) in the top panel of figure \ref{fig:vplot}.
+% If the shape of the upper edge of rotation period measurements is created by a
+% detection limit, it could indicate the rotation periods at which stars of
+% different temperatures become relatively inactive (and their rotation periods
+% therefore become undetectable).
+% % Once inactive, stars would have little photometric variability induced by
+% % star spots and their rotation periods would be difficult to measure, so may
+% % not feature in the \mct\ catalog.
+% If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
+% $\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.
+
+% % The groups of stars with the largest velocity dispersions are cooler than 4500
+% % K, presumably because cooler stars remain active for longer, and their
+% % rotation periods can therefore be measured at older ages.
+
+% The velocity dispersions of the coolest stars may be affected by a selection
+% bias -- these extremely faint stars are more difficult to detect at larger
+% distances, larger heights above the plane, and therefore larger velocities.
+% It is possible that some high velocity stars of temperatures cooler than
+% $\sim$4000 K are missing from this sample, and the velocity dispersions may
+% therefore appear lower than they truly are.
+% This could be why the velocity dispersion appears to decrease towards the
+% right of figure \ref{fig:dispersion_period_teff}.
+
+\begin{figure}
+  \caption{
+    Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
+    x-axis.
+    Masses are from the \kepler\ input catalog.
+    The white lines show the \citet{spada2019} rotational evolution models at
+    0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
+    period.
+    These models include age and mass-dependent coupling between the stellar
+    core and envelope.
+    The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
+    an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
+    decreases with decreasing mass.
+    A similar phenomenon is visible in the velocity dispersions of field stars
+    shown as colored points in the background of this figure.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{masses_with_models}
+\label{fig:masses_with_models}
+\end{figure}
 

### 2019-11-05 calc_velocities.py

@@ -35,7 +35,7 @@ def calc_vb(pandas_df, nsamples=1000):
     pm_b = np.array(vels.pm_b.value)
 
     gal = c_samples.galactic
-    v_b = (gal.pm_l * gal.distance).to(u.km/u.s, u.dimensionless_angles())
+    v_b = (gal.pm_b * gal.distance).to(u.km/u.s, u.dimensionless_angles())
     return pm_b, np.array(v_b.value)
 
 

### 2019-11-05 kinematics_and_rotation/__init__.py

@@ -1,6 +1,5 @@
 #!/usr/bin/env python
 # -*- coding: utf-8 -*-
 
-from .version import __version__
 from .dispersion import *
 from .sigma_clip import *

### 2019-11-05 kinematics_and_rotation/sigma_clip.py

@@ -95,7 +95,7 @@ def interval_sigma_clip(interval, x, y, nsigma=3):
     """
 
     nints = int((max(x) - min(x)) / interval)
-    intervals =
+    # intervals =
 
     full_mask = []
     n = 0

### 2019-12-08 paper/abstract.tex
letterizing

@@
 could be a useful tool for calibrating gyrochronology.
-At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
-to the Praesepe cluster, accurately predicts the relative ages of stars.
-At these young ages, lower-mass stars spin more slowly than higher-mass stars
-because their stronger magnetic fields lead to more efficient angular momentum
-loss.
-However, at old ages we find that late G and early K dwarfs rotate {\it
-faster} than late K dwarfs of the same age.
+% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
+% to the Praesepe cluster, accurately predicts the relative ages of stars.
+We find that, at young ages, lower-mass stars spin more slowly than
+higher-mass stars because their stronger magnetic fields lead to more
+efficient angular momentum loss.
+This reflects the behavior of young open clusters such as the Pleaides,
+Praesepe, and the Hyades.
+However, at older ages we find that late G and early K dwarfs rotate at the
+same rate as late K dwarfs of the same age, and eventually rotate faster.
 % These oThis may be a feature of angular momentum redistribution in the stellar
@@
 % higher-mass stars after a certain age or Rossby number.
-These results, based on field star rotation periods, align with recent
-findings from the rotation periods of stars in middle-aged open clusters and
-theoretical models that vary the rate of surface-to-core angular momentum
-transport as a function of time and mass.
+These results align with recent findings from the rotation periods of stars in
+middle-aged open clusters and theoretical models that vary the rate of
+surface-to-core angular momentum transport as a function of time and mass.
 Finally, we find no evidence for mass-dependent heating in a sample of K and M

### 2019-12-08 paper/discussion.tex
letterizing

@@
 picture of rotational evolution.
-At young ages, stellar rotation period decreases with mass.
-This is probably because lower-mass stars with deeper convection zones have
-stronger magnetic fields, larger Alfven radii and therefore experience greater
-angular momentum loss rate.
+At young ages, stellar rotation period decreases with mass, likely because
+lower-mass stars with deeper convection zones have stronger magnetic fields,
+larger Alfv\'en radii and therefore experience greater angular momentum loss
+rate.
 According to the \citet{spada2019} models, there is minimal transportation of

### 2019-12-08 paper/intro.tex
letterizing

@@
 periods converge onto a unique sequence by $\sim$500-700 million years
-\citep[\eg][]{irwin2009}.
-After this time, it is thought that the rotation period of a star is
-determined, to first order, by its color and age alone.
+\citep[\eg][]{irwin2009, gallet2013}.
+After this time, the rotation period of a star is thought to be determined, to
+first order, by its color and age alone.
 This is the principle behind gyrochronology -- the method of inferring a
@@
 For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
-slowly than the G dwarfs -- in theory because lower-mass stars have deeper
-convections zones which generate stronger magnetic fields and more efficient
-magnetic braking.
+slowly than the G dwarfs.
+In theory this is because lower-mass stars have deeper convections zones which
+generate stronger magnetic fields and more efficient magnetic braking.
 However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
@@
 surface rotation rates.
-
 Another example of unexpected rotational evolution is seen in old field stars
@@
 To fit magnetic braking models to these data, a cessation of magnetic braking
-is required after stars reach a Rossby number (ratio of rotation period to
-convective turnover time) of around 2 \citep{vansaders2016, vansaders2018}.
+is required after stars reach a Rossby number (Ro; the ratio of rotation
+period to convective turnover time) of around 2 \citep{vansaders2016,
+vansaders2018}.
 
-The rotational evolution of stars is clearly a complicated process, and to
-fully calibrate the gyrochronology relations we need reliable ages for large
-numbers of field stars which span a range of ages.
+The rotational evolution of stars is clearly a complicated process and, to
+fully calibrate the gyrochronology relations we need a large sample of
+reliable ages for stars spanning a range of ages and masses.
 % however relying on open clusters (which are typically young) and asteroseismic
@@
 % sample.
-In this paper, we use the velocity dispersions of field stars to gain insight
-into their rotational evolution.
-Velocity dispersion provides relative ages for groups of stars, allowing us to
-observe the rotational evolution of stars over a range of masses and ages.
+In this paper, we use the velocity dispersions of field stars to qualitatively
+explore the rotational evolution of GKM dwarfs, and show that kinematics could
+provide a gyrochronology calibration sample.
 
@@
 rotation periods do not have radial velocity (RV) measurements because they
-are relatively faint \kepler\ targets ($\sim$11th-18th magnitudes).
+are relatively faint \kepler\ targets ($\sim$12th-17th magnitudes).
 For this reason, we used velocity in the direction of galactic latitude, \vb,

### 2019-12-08 paper/method.tex
letterizing

@@
 latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
-a limiting magnitude of 16 dex.
+a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
+The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
+\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
+Galaxy.
+We subtracted this offset and examined the residuals of the \vz\-\vb\
+relationship to investigate the variance as a function of Galactic latitude.
 We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
@@
 deviation) that increased with increasing $b$.
+Figure \ref{fig:vb_vz} shows how \vz\ transforms to \vb\ over a range of
+Galactic latitudes.
 The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
@@
 20$^\circ$.
+Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
+over different Galactic latitudes.
+Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
+fits are shown as dashed blue lines.
+The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
+the variance increases with increasing Galactic latitude.
+\begin{figure}
+  \caption{
+This figure demonstrates the relationship between \vb\ and \vz\, after
+    subtracting a straight-line-fit, for stars in
+    the \kepler\ field, based on the GUMS simulation.
+The panels show a kernel density estimator (KDE) (black solid line) for
+    the \vz-\vb\ residuals of stars in the GUMS simulation at four different
+    Galactic latitudes.
+Blue dashed lines show Gaussian fits to these KDEs.
+The distributions are close to Gaussians, but with slightly heavy tails.
+The standard deviations of the Gaussian fits increase with Galactic latitude.
+This figure illustrates the fact using \vb\ instead of \vz\ artificially
+    increases the velocity dispersion, especially at high latitudes.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{vb_vz}
+\label{fig:vb_vz}
+\end{figure}
+
 Since we are concerned with velocity {\it dispersions}, rather than velocities
-themselves, we also compared the \vb\ and \vz\ velocity dispersions for stars
-downloaded from the GUMS simulation.
+themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
+function of temperature for
+stars downloaded from the GUMS simulation.
 For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
@@
 less than 15\degrees\ in our analysis.
+Although it seems that the transformation between \vz\ and \vb\ does not {\it
+strongly} affect our results, we cannot rule out the possibility that it
+introduces systematic biases into the velocity dispersions we present here.
+In \gaia\ DR3, RVs will be available for most stars in this sample, providing
+an opportunity to validate (or correct) the results presented here.
 
@@
 % equal numbers in \teff-space than in color-space.
+
+In this paper, we do not attempt to convert velocity dispersion (\sigmavb)
+into an age via an age-velocity dispersion relation (AVR)
+\citep[\eg][]{holmberg2009}.
+Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
+careful analysis will be needed to calculate absolute ages.
+For example, the velocity distributions could be modeled as a mixture of
+Gaussians in order to account for the additional velocity dispersion caused by
+the \vz-\vb\ transformation.
+The RVs of most of these stars will become available in \Gaia\ DR3, allowing
+calculations of \vz, which can be used to calculate more reliable ages via an
+AVR.

### 2019-12-08 kinematics_and_rotation/__init__.py

@@ -3,3 +3,5 @@
 
 from .dispersion import *
 from .sigma_clip import *
+from .photometric_teff import *
+from .teff_bv import *

### 2019-12-08 paper/abstract.tex
small edits to text

@@
 % to the Praesepe cluster, accurately predicts the relative ages of stars.
-We find that, at young ages, lower-mass stars spin more slowly than
-higher-mass stars because their stronger magnetic fields lead to more
-efficient angular momentum loss.
-This reflects the behavior of young open clusters such as the Pleaides,
-Praesepe, and the Hyades.
+We find that the period-color relationship of the Praesepe cluster is not
+repeated in stars of older ages: a polynomial gyrochronology model, fit to
+Praesepe stars, under-predicts the ages of late K dwarfs in the field.
+In general we find that lower-mass stars spin more slowly than higher-mass
+stars at young ages.
+This is likely because their stronger magnetic fields lead to more efficient
+angular momentum loss and reflects the behavior of young open clusters such as
+the Pleaides, Praesepe, and the Hyades.
 However, at older ages we find that late G and early K dwarfs rotate at the
-same rate as late K dwarfs of the same age, and eventually rotate faster.
+{\it same rate} or faster than late K dwarfs of the same age.
 % These oThis may be a feature of angular momentum redistribution in the stellar

### 2019-12-08 paper/intro.tex
small edits to text

@@
 Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
-strong magnetic fields and spin more slowly over time through magnetic braking
+strong magnetic fields and slowly lose angular momentum via magnetic braking
 \citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
@@
 first order, by its color and age alone.
-This is the principle behind gyrochronology -- the method of inferring a
+This is the principle behind gyrochronology, the method of inferring a
 star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
@@
 \citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
-meibom2015, curtis2019} reveal that rotational evolution is more complicated
-than previously thought.
+meibom2015, curtis2019} have revealed that rotational evolution is more
+complicated than previously thought.
 For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
@@
 % vertical
-Vertical velocity, \vz, can only be calculated with full 6-dimensional
-position and velocity information, and unfortunately most stars with measured
-rotation periods do not have radial velocity (RV) measurements because they
-are relatively faint \kepler\ targets ($\sim$12th-17th magnitudes).
+Although {\it vertical} velocity, \vz, is a well-established age proxy, it can
+only be calculated with full 6-dimensional position and velocity information.
+Unfortunately most field stars with measured rotation periods do not have
+radial velocity (RV) measurements because they are relatively faint \kepler\
+targets ($\sim$12th-16th magnitudes).
 For this reason, we used velocity in the direction of galactic latitude, \vb,
@@
 The \kepler\ field is positioned at low galactic latitude
-(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect -- see section
+(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see section
 \ref{sec:results}) approximation to \vz.

### 2019-12-08 paper/method.tex
small edits to text

@@
 rotation periods at old ages, the latter scenario seems more likely.
+The \citep{angus2019} gyrochronology relation is a simple polynomial model,
+fit to the period-color relation of Praesepe.
+Since this relation predicts that the oldest stars in the \mct\ sample are
+late-G and early-K dwarfs, it is probably under-predicting the ages of late-K
+and early-M dwarfs.
 \begin{figure}
@@
     A general age gradient is visible across the main sequence.
+    Since the \citet{angus2019} relation predicts that the oldest stars in
+    the \mct\ sample are late-G and early-K dwarfs, it is probably
+    under-predicting the ages of late-K and early-M dwarfs.
 }
@@
 latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
-stars, introducing a velocity bias for high latitude stars (which are more
+stars, and introduce a velocity bias for high latitude stars (which are more
 likely to be cooler and older).
@@
 If lower-mass stars experience greater velocity changes when gravitationally
-perturbed and are dynamically heated more efficiently than higher-mass stars,
-velocity dispersion would be a function of {\it both} age and mass and cannot
-be straightforwardly interpreted as an age proxy.
+perturbed, and are dynamically heated more efficiently than higher-mass stars,
+velocity dispersion would be a function of both age {\it and} mass and cannot
+be interpreted as a straightforward age proxy.
 So, in order to establish whether \sigmavb\ can be used as an age proxy, we
@@
 This mass-age degeneracy is highly reduced in M dwarfs because their
-main-sequence lifetimes are longer than the age of the Universe, however no
-evidence for mass-dependent heating has been found in M dwarfs
+main-sequence lifetimes are longer than the age of the Universe, and no
+evidence for mass-dependent heating has previously been found in M dwarfs
 \citep[\eg][]{faherty2009, newton2016}.
 
-To investigate whether mass-dependent heating could be acting on the \kepler\
-sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
-whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
-the initial mass function.
-We could not perform this analysis on the \mct\ sample, because only stars
-with {\it detectable} rotation periods are included in it and since lower-mass
-stars stay active for longer it is likely that the oldest stars in this sample
-have low masses.
+To further investigate whether mass-dependent heating could be acting on the
+\kepler\ sample, we selected late K and M dwarfs observed by both \kepler\ and
+\gaia, whose MS lifetimes exceed around 11 Gyrs and are therefore
+representative of the initial mass function\footnote{We could not perform this
+analysis on the \mct\ sample, because it only includes stars with {\it
+detectable} rotation periods, and since lower-mass stars stay active for
+longer it is likely that it contains a strong mass-age correlation.}.
 We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
@@
 The essential point however, is that we do not see evidence for mass-dependent
-heating acting on stars in the \kepler\ field, indicating that \sigmavb\ {\it
-can} be used as an age proxy.
-We also examined the {\it vertical} velocities of the 537 stars in this sample
-with RV measurements.
-Again, we found no evidence for mass-dependent heating: the slope of the
+heating acting on stars in the \kepler\ field, indicating that velocity
+dispersion {\it can} be used as an age proxy.
+This analysis was performed using \vb\ but we also examined the {\it vertical}
+velocities of the 537 stars in this sample with RV measurements.
+Again, no evidence was found for mass-dependent heating: the slope of the
 velocity dispersion-temperature relation was consistent with zero.

### 2019-12-08 paper/method.tex
updating main figure

@@
 velocity dispersion-temperature relation was consistent with zero.
+\begin{figure}
+  \caption{
+      Top: Stellar velocity (\vb) as a function of \teff\ for
+      \kepler\ K and M dwarfs.
+Vertical lines indicate different \teff-groupings used to calculate velocity
+    dispersion.
+Pink stars were not included in velocity dispersion calculations as they were
+    either removed as outliers during a sigma clipping process, or they lie at
+    the sparcely populated, extremely cool end of the temperature range.
+    Velocity dispersion and \teff\ are slightly positively correlated, likely
+    due to a brightness-related selection bias, indicating that mass-dependent
+    heating does not significantly affect low-mass stars in the \kepler\
+    field.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{vb_vs_teff}
+\label{fig:vb_vs_teff}
+\end{figure}
 
-At a galactic latitude of zero, $v_b=v_z$, however for increasing values of
-$b$, this equivalence becomes an approximation that grows noisier with $b$.
+Having found no strong evidence for mass-dependent heating, we next tested
+the validity of \vb\ as a proxy for \vz\ in more detail.
+At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
+of $b$, this equivalence becomes an approximation that grows noisier with $b$.
 To test the validity of the \vb$\sim$\vz\ approximation over a range of
@@
 Galaxy.
-We subtracted this offset and examined the residuals of the \vz\-\vb\
+We subtracted this offset and examined the residuals of the \vz\ -- \vb\
 relationship to investigate the variance as a function of Galactic latitude.
 We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
-with a standard deviation (calculated as 1.5 $\times$ the median absolute
-deviation) that increased with increasing $b$.
-Figure \ref{fig:vb_vz} shows how \vz\ transforms to \vb\ over a range of
-Galactic latitudes.
+with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
 The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
-$, 4kms${^-1}$ at 10$^\circ$, 6kms${^-1}$ at 15$^\circ$, and 9kms${^-1}$ at
+$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
 20$^\circ$.
-Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
-over different Galactic latitudes.
-Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
-fits are shown as dashed blue lines.
-The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
-the variance increases with increasing Galactic latitude.
+
+% Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
+% over different Galactic latitudes.
+% Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
+% fits are shown as dashed blue lines.
+% The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
+% the variance increases with increasing Galactic latitude.
 \begin{figure}
   \caption{
-This figure demonstrates the relationship between \vb\ and \vz\, after
-    subtracting a straight-line-fit, for stars in
-    the \kepler\ field, based on the GUMS simulation.
+This figure demonstrates the variance in the relationship between \vb\ and
+    \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
 The panels show a kernel density estimator (KDE) (black solid line) for
-    the \vz-\vb\ residuals of stars in the GUMS simulation at four different
-    Galactic latitudes.
+    the \vz -- \vb\ residuals of stars in the GUMS simulation at four
+    different Galactic latitudes.
 Blue dashed lines show Gaussian fits to these KDEs.
-The distributions are close to Gaussians, but with slightly heavy tails.
+The distributions are close to Gaussian, with slightly heavy tails.
 The standard deviations of the Gaussian fits increase with Galactic latitude.
-This figure illustrates the fact using \vb\ instead of \vz\ artificially
-    increases the velocity dispersion, especially at high latitudes.
+This figure illustrates how using \vb\ instead of \vz\ artificially
+    increases velocity dispersion, especially at high latitudes.
 }
@@
 
-\begin{figure}
-  \caption{
-      Top: Stellar velocity (\vb) as a function of \teff\ for
-      \kepler\ K and M dwarfs.
-Vertical lines indicate different \teff-groupings used to calculate velocity
-    dispersion.
-Pink stars were not included in velocity dispersion calculations as they were
-    either removed as outliers during a sigma clipping process, or they lie at
-    the sparcely populated, extremely cool end of the temperature range.
-    Velocity dispersion and \teff\ are slightly positively correlated, likely
-    due to a brightness-related selection bias, indicating that mass-dependent
-    heating does not significantly affect low-mass stars in the \kepler\
-    field.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vb_vs_teff}
-\label{fig:vb_vs_teff}
-\end{figure}
-
 % Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
@@
 % \mct\ sample.
+
+Because of the noisy relationship between \vb\ and \vz\, in this paper we do
+not attempt to convert velocity dispersion (\sigmavb) into an age via an
+age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
+Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
+careful analysis that includes formal modeling of the \vb\ -- \vz\
+relationship will be needed to calculate absolute ages.
+% For example, the velocity distributions could be modeled as a mixture of
+% Gaussians in order to account for the additional velocity dispersion caused by
+% the \vz-\vb\ transformation.
+The RVs of most of these stars will become available in \Gaia\ DR3, allowing
+calculations of \vz, which can be used to calculate more reliable ages via an
+AVR.
 
@@
 % equal numbers in \teff-space than in color-space.
-
-In this paper, we do not attempt to convert velocity dispersion (\sigmavb)
-into an age via an age-velocity dispersion relation (AVR)
-\citep[\eg][]{holmberg2009}.
-Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
-careful analysis will be needed to calculate absolute ages.
-For example, the velocity distributions could be modeled as a mixture of
-Gaussians in order to account for the additional velocity dispersion caused by
-the \vz-\vb\ transformation.
-The RVs of most of these stars will become available in \Gaia\ DR3, allowing
-calculations of \vz, which can be used to calculate more reliable ages via an
-AVR.

### 2019-12-08 paper/results.tex
updating main figure

@@
 
-\begin{figure}
-  \caption{
-      Top: Rotation period vs effective temperature for stars in the \mct\
-    sample, colored by the velocity dispersions of stars calculated over a
-    grid in $\log_{10}$(period) and \teff.
-    Bottom: the velocity dispersions of groups of stars, shown as a solid
-    grid for clarity.
-    The hatched area indicates the temperature regime where selection biases
-    could play an important role, so these velocity dispersions should be
-    interpreted with caution.
-    The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
-    with the \citet{angus2019} gyrochronology relation, which roughly traces
-    the rotation period gap.
-    The black dashed lines show a 650 Myr isochrone, indicating the location
-    and shape of the Praesepe cluster (to which this gyrochronology model was
-    calibrated).
-    The black points show the 1.1 Gyr NGC 6811 open cluster.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vplot}
-\label{fig:vplot}
-\end{figure}
+To explore the relationship between rotation period, \teff\ and velocity
+dispersion, we first removed high and low velocity outliers from the \mct\
+sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
+Without sigma-clipping, we found that a small number of high velocity
+outliers at the low-temperature end of our sample substantially raised the
+velocity dispersion for cooler stars, however the overall trends remain the
+same with, or without sigma-clipping.
+We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
+3500 K to avoid biases caused by the selection function at the faint, cool
+end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
+
 Figure \ref{fig:vplot} shows rotation period versus effective temperature for
@@
 similar color in figure \ref{fig:vplot} are similar ages.
-
+\begin{figure}
+  \caption{
+    Top: Rotation period vs effective temperature for stars in the \mct\
+    sample, colored by the velocity dispersions of stars calculated over a
+    grid in $\log_{10}$(period) and \teff.
+    Black lines show isochrones that project the rotation-color relation of
+    Praesepe to longer rotation periods over time \citep{angus2019}.
+    These isochrones do not appear to reflect the evolution of field stars at
+    long rotation periods/old ages: they do not trace lines of constant
+    velocity dispersion.
+    Isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
+    and bottom panels.
+    Bottom: Same as top panel with rotation period vs {\it mass}
+    \citep[from the Kepler Input Catalog][]{brown2011}.
+    White lines show isochrones from a model that includes mass and
+    age-dependent angular momentum transport between the core and envelope
+    \citep{spada2019}.
+    Qualitatively, these isochrones appear to reflect the evolution of field
+    stars at long rotation periods/old ages: they trace lines of constant
+    velocity dispersion.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{main_figure}
+\label{fig:main_figure}
+\end{figure}
 Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
-gyrochronal age across all temperatures, implying that both velocity
+rotation period across all temperatures, implying that both velocity
 dispersion and rotation period increases with age as expected.

### 2019-12-08 paper/conclusion.tex
small changes -- still need to go over wording in discussion section

@@
 
-We examined the rotational evolution of \kepler\ field stars using their
-velocity dispersions as a proxy for age.
-We found that old groups of cool dwarfs selected to be coeval using the
-\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
-dispersion across all temperatures.
-This implies that the \citet{angus2019} relation, which is based on the
+We examined the rotational evolution of \kepler\ field stars by interpreting
+their velocity dispersions as a proxy for age.
+% We found that old groups of cool dwarfs, selected to be coeval using the
+% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
+% dispersion across all temperatures.
+We found that the \citet{angus2019} relation, which is based on the
 period-color relation of the 650 Myr Praesepe cluster, does not correctly
@@
 It appears that the period-color/\teff\ relation changes shape over time in a
-way that qualitatively agrees with theoretical that which include a
+way that qualitatively agrees with theoretical models which include a
 mass-dependent core-envelope angular momentum transport \citep{spada2019}.
@@
 dwarfs of the same age.
-We speculate that the rotation period gap \citep{mcquillan2014} may separate
-a young regime where stellar rotation periods decrease with increasing mass
-from an old regime where periods increase with increasing mass, however more
-data are needed to provide a conclusive result.
+% We speculate that the rotation period gap \citep{mcquillan2014} may separate
+% a young regime where stellar rotation periods decrease with increasing mass
+% from an old regime where periods increase with increasing mass, however more
+% data are needed to provide a conclusive result.
 % The period-color/\teff\ relation seems to start flattening out after $\sim$1

### 2019-12-08 paper/results.tex
small changes -- still need to go over wording in discussion section

@@
 % word limit: 500
-\section{Results}
+\section{Results and Discussion}
 \label{sec:results}
@@
 The masses of these stars are taken from the \kepler\ input catalog
-\citep{KIC}.
+\citep{brown2011}.
 The \citet{spada2019} models appear to qualitatively agree with the data:
@@
 inverts over a narrow range of temperatures.
-The models reproduce the `dip' feature seen in the data.
+
+These results indicate that stars of spectral type ranging from late G to late
+K follow a braking law that changes over time.
+In particular, the relationship between rotation period and effective
+temperature appears to flatten out and eventually invert.
+These results provide further evidence for `stalled' rotational evolution of K
+dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
+by models that vary angular momentum transport between stellar core and
+envelope with time and mass \citep{spada2019}.
+
+The velocity dispersions of stars in the \mct\ sample provide the following
+picture of rotational evolution.
+At young ages, stellar rotation period decreases with mass, likely because
+lower-mass stars with deeper convection zones have stronger magnetic fields,
+larger Alfv\'en radii and therefore experience greater angular momentum loss
+rate.
+According to the \citet{spada2019} model, there is minimal transportation of
+angular momentum from the surface to the core of the star at these young ages,
+so the surface slows down but the core keeps spinning rapidly.
+At intermediate ages, rotation period is constant with mass, and at late ages
+rotation period {\it increases} with mass for GK dwarfs.
+The explanation for this, according to the \citet{spada2019} model, is that
+lower-mass stars are still braking more efficiently at these intermediate and
+old ages but their cores are more tightly coupled to their envelopes, allowing
+angular momentum transport between the two interior layers.
+Angular momentum resurfaces and prevents the stellar envelopes from
+spinning-down rapidly.
 
@@
 
-\begin{figure}
-  \caption{
-    Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
-    x-axis.
-    Masses are from the \kepler\ input catalog.
-    The white lines show the \citet{spada2019} rotational evolution models at
-    0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
-    period.
-    These models include age and mass-dependent coupling between the stellar
-    core and envelope.
-    The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
-    an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
-    decreases with decreasing mass.
-    A similar phenomenon is visible in the velocity dispersions of field stars
-    shown as colored points in the background of this figure.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{masses_with_models}
-\label{fig:masses_with_models}
-\end{figure}
+% \begin{figure}
+%   \caption{
+%     Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
+%     x-axis.
+%     Masses are from the \kepler\ input catalog.
+%     The white lines show the \citet{spada2019} rotational evolution models at
+%     0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
+%     period.
+%     These models include age and mass-dependent coupling between the stellar
+%     core and envelope.
+%     The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
+%     an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
+%     decreases with decreasing mass.
+%     A similar phenomenon is visible in the velocity dispersions of field stars
+%     shown as colored points in the background of this figure.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{masses_with_models}
+% \label{fig:masses_with_models}
+% \end{figure}
 

### 2019-12-12 paper/abstract.tex
adding period gap section

@@
 % gyrochronology relations.
-In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
-proxy, to reveal their rotational evolution and demonstrate that kinematics
-could be a useful tool for calibrating gyrochronology.
+In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
+an age proxy, to reveal their rotational evolution and demonstrate that
+kinematics could be a useful tool for calibrating gyrochronology.
+Firstly, we find no evidence for mass-dependent heating in a sample of K and M
+dwarfs in the \kepler\ field.
 % At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
@@
 stars at young ages.
-This is likely because their stronger magnetic fields lead to more efficient
+This is probably because their stronger magnetic fields lead to more efficient
 angular momentum loss and reflects the behavior of young open clusters such as
@@
 surface-to-core angular momentum transport as a function of time and mass.
-Finally, we find no evidence for mass-dependent heating in a sample of K and M
-dwarfs in the \kepler\ field.
+We also find that the kinematically oldest \kepler\ stars with measured
+rotation periods are late-K and early-M dwarfs, indicating that these stars
+maintain spotted surfaces and stay magnetically active longer than more
+massive stars.
+Finally, we find that the most rapidly rotating K-dwarfs are likely to be
+synchronized binaries.
 

### 2019-12-12 paper/conclusion.tex
adding period gap section

--- 2019-10-15 paper/conclusion.tex
@@
 
-We examined the rotational evolution of \kepler\ field stars using their
-velocity dispersions as a proxy for age.
-We found that old groups of cool dwarfs selected to be coeval using the
-\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
-dispersion across all temperatures.
-This implies that the \citet{angus2019} relation, which is based on the
+In this paper, we demonstrated that the dispersion of velocities in the
+direction of galactic latitude, \vb, can be used as an age proxy, by showing
+that there is no strong evidence for mass-dependent heating in low-mass
+\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
+main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
+increase with decreasing mass.
+Although {\it vertical} velocity, \vz, is the quantity that most strongly
+traces time-dependent orbital heating in the disc of the Galaxy, most stars
+with measured rotation periods do not yet have radial velocities, so we used
+velocity in the direction of Galactic latitude, \vb\, as a proxy for \vz.
+Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
+\vz\ introduces an additional velocity dispersion, which increases with
+increasing Galactic latitude.
+For this reason we did not attempt to convert \vb\ dispersions into ages using
+an age-velocity dispersion relation.
+However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
+confirmed that using \vb\ instead of \vz\ does not introduce any
+mass-dependent velocity dispersion bias into the sample.
+We therefore assumed that \vb\ velocity dispersion can be used to accurately
+rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
+on average, older than a group of stars with a small velocity dispersion.
+
+% We found that old groups of cool dwarfs, selected to be coeval using the
+% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
+% dispersion across all temperatures.
+We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
+the evolution of stellar rotation period as a function of effective
+temperature and age.
+We found that the \citet{angus2019} relation, which is based on the
 period-color relation of the 650 Myr Praesepe cluster, does not correctly
-describe the period-age-color/\teff\ relation for old stars.
-It appears that the period-color/\teff\ relation changes shape over time in a
-way that qualitatively agrees with theoretical that which include a
-mass-dependent core-envelope angular momentum transport \citep{spada2019}.
+describe the period-age-\teff\ relation for old stars.
 At young ages, rotation period is anti-correlated with \teff: cooler stars
@@
 dwarfs of the same age.
+We showed that the period-\teff\ relations change shape over time in a way
+that qualitatively agrees with theoretical models which include a
+mass-dependent core-envelope angular momentum transport \citep{spada2019}.
+
+% The period-color/\teff\ relation seems to start flattening out after $\sim$1
+% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
+% older than the period gap which is located at a gyrochronal age of around
+% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
+We also found that the oldest stars in the \mct\ catalog are cooler than 4500
+K, which suggests that lower-mass stars remain active for longer, allowing
+their rotation periods to be measured at older ages.
 We speculate that the rotation period gap \citep{mcquillan2014} may separate
@@
 data are needed to provide a conclusive result.
-% The period-color/\teff\ relation seems to start flattening out after $\sim$1
-% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
-% older than the period gap which is located at a gyrochronal age of around
-% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
-Finally, when velocity dispersion is interpreted as an age proxy, it appears
-that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
-suggests that lower-mass stars remain active for longer, allowing their
-rotation periods to be measured at older ages.
+The velocity dispersions of stars increase smoothly across the rotation period
+gap, indicating that the gap does not separate two distinct populations.
+Finally, we used kinematics to reveal a population of synchronized binaries
+with rotation periods less than around 10 days.
+
 % We outlined a number of scenarios which could provide alternative

### 2019-12-12 paper/results.tex
adding period gap section

@@
 % cluster.
-% These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
-% figure \ref{fig:age_gradient}.
-% The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
-% along the upper edge of the MS, where stellar evolution models predict old
-% stars to be, however the majority of these `old' stars are bluer than \gcolor\
-% $\sim$ 1.5 dex.
-% This suggests that either old M dwarfs are missing from the \mct\ catalog, or
-% the \citet{angus2019} gyrochronology relation under-predicts the ages of
-% low-mass stars.
-% Given that lower-mass stars stay active for longer than higher-mass stars
-% \citep[\eg][]{west2011}, and are therefore more likely to have measurable
-% rotation periods at old ages, the latter scenario seems more likely.
 
-% To further explore the relationship between rotation period, \teff\ and age,
-% we first removed high and low velocity outliers by performing 3$\sigma$
-% sigma-clipping on the \vb\ velocities.
-% Without sigma-clipping, we found that a small number of high velocity
-% outliers at the low-temperature end of our sample substantially raised the
-% velocity dispersion for cooler stars.
 % We then selected groups of stars within different gyrochronal age ranges and
@@
 end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
+Finally, we removed stars with short rotation periods by cutting out stars
+with gyrochronal ages less than 0.5 Gyr.
 
-Figure \ref{fig:vplot} shows rotation period versus effective temperature for
-the \mct\ sample, coloured by the standard deviation of their (\vb)
-velocities, where \sigmavb\ was calculated for groups of stars over a grid in
-$\log_{10}$(period) and temperature.
+The top panel of figure \ref{fig:vplot} shows rotation period versus effective
+temperature for the \mct\ sample, coloured by the standard deviation of their
+(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
+grid in $\log_{10}$(period) and temperature.
 If we assume that mass dependent heating does not strongly affect this sample
 and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
-velocity dispersion can be interpreted as an age proxy and stars plotted in a
+velocity dispersion can be interpreted as an age proxy, and stars plotted in a
 similar color in figure \ref{fig:vplot} are similar ages.
@@
     grid in $\log_{10}$(period) and \teff.
-    Black lines show isochrones that project the rotation-color relation of
+    Black lines show isochrones from a gyrochronology model that projects the
+    rotation-color relation of
     Praesepe to longer rotation periods over time \citep{angus2019}.
@@
     \citep{spada2019}.
-    Qualitatively, these isochrones appear to reflect the evolution of field
+    Qualitatively, these isochrones reflect the evolution of field
     stars at long rotation periods/old ages: they trace lines of constant
-    velocity dispersion.
+    velocity dispersion by reproducing periods of `stalled' rotational
+    evolution for K-dwarfs.
 }
@@
     \includegraphics[width=1\textwidth]{main_figure}
-\label{fig:main_figure}
+\label{fig:vplot}
 \end{figure}
@@
 dispersion and rotation period increases with age as expected.
-Lines of constant age (isochrones) appear to follow the shape of the
-Praesepe-based gyrochronology model (black dotted line) at young ages.
-However, at older ages it appears that the relation between rotation period
-and \teff\ flattens out, until eventually rotation period {\it decreases} with
-decreasing effective temperature at a given age.
+Black lines show isochrones from the \citep{angus2019} gyrochronology model,
+which projects the rotation-color relation of Praesepe to longer rotation
+periods over time.
+These isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
+At the youngest ages, these isochrones seem to describe the data well: the
+palest yellow (youngest) stars with the lowest velocity dispersions all fall
+close to the 0.5 Gyr isochrone.
+% Lines of constant age (isochrones) appear to follow the shape of the
+% Praesepe-based gyrochronology model (black dotted line) at young ages.
+% However, at older ages it appears that the relation between rotation period
+% and \teff\ flattens out, until eventually rotation period {\it decreases} with
+% decreasing effective temperature at a given age.
+However, although the 0.5 Gyr and 1 Gyr isochrones do appear to trace constant
+velocity dispersion/age among the field stars, by 1.5 Gyrs the isochrones
+start to {\it cross} different velocity dispersion regimes.
+For example, the 1.5 Gyr isochrone lies on top of stars with velocity
+dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
+\kms\ velocity dispersions at 4000-4500K.
+The isochrones older than 1.5 Gyr also cross a range of velocity dispersions.
+If these were true isochrones however, they should follow lines of constant
+velocity dispersion.
+At ages older than around 1 Gyr, it appears that isochrones should have a more
+flattened, or even inverted relationship between rotation period and effective
+temperature than these Praesepe-based models.
+% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
+% rotate more rapidly than higher-mass stars.
+% This is the opposite of the trend that is observed in young open clusters such
+% as Praesepe and the Hyades.
 
-Figure \ref{fig:masses_with_models} shows the same as figure \ref{fig:vplot},
-however mass is plotted on the x-axis instead of temperature and the models
-are those of \citet{spada2019}.
-The masses of these stars are taken from the \kepler\ input catalog
-\citep{brown2011}.
+The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
+function of rotation period and {\it mass}, obtained from the Kepler Input
+Catalog \citep{brown2011}, with isochrones from the \citep{spada2019}
+gyrochronology model shown as white lines.
+These isochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
+Although these models do not trace lines of constant velocity dispersion at
+0.5 and 1 Gyr, they do appear to reproduce trends in the data at ages of 1.5
+Gyr and older.
 The \citet{spada2019} models appear to qualitatively agree with the data:
 their rotation period-\teff\ relation flattens out over time and eventually
-inverts over a narrow range of temperatures.
+inverts.
 
-These results indicate that stars of spectral type ranging from late G to late
-K follow a braking law that changes over time.
+The results shown in figure \ref{fig:vplot} indicate that stars of spectral
+type ranging from late G to late K follow a braking law that changes over
+time.
 In particular, the relationship between rotation period and effective
@@
 envelope with time and mass \citep{spada2019}.
-
 The velocity dispersions of stars in the \mct\ sample provide the following
@@
 At intermediate ages, rotation period is constant with mass, and at late ages
-rotation period {\it increases} with mass for GK dwarfs.
-The explanation for this, according to the \citet{spada2019} model, is that
+rotation period {\it increases} with mass for K-dwarfs.
+The interpretation of this, according to the \citet{spada2019} model, is that
 lower-mass stars are still braking more efficiently at these intermediate and
 old ages but their cores are more tightly coupled to their envelopes, allowing
-angular momentum transport between the two interior layers.
+angular momentum transport between the two layers.
 Angular momentum resurfaces and prevents the stellar envelopes from
-spinning-down rapidly.
+spinning-down rapidly, and this effect is strongest for late K-dwarfs with
+effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.
+
+It has been demonstrated that lower-mass stars remain magnetically active
+longer than more massive stars, \citep{west2008, kiman2019}.
+If the detectability of a rotation period is considered to be a magnetic
+activity proxy, then our results provide further evidence for a mass-dependent
+activity lifetime.
+Figure \ref{fig:vplot} shows that the groups of stars with the largest
+velocity dispersions are cooler than 4500 K.
+This implies that the oldest stars with detectable rotation periods, are
+cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
+massive stars.
+
+\subsection{The period gap and synchronized binaries}
+\label{sec:gap}
+
+There is a sharp gap in the population of rotation periods, which lies just
+above the 1 Gyr isochrone in the upper panel of figure \ref{fig:vplot}, whose
+origin is unknown and is the subject of much speculation \citep{mcquillan2014,
+davenport2018, reinhold2019}.
+The Praesepe-based model appears to be valid below the gap but not above.
+Although this may be coincidental (and more data would be needed to confirm a
+connection) the gap may indeed separate a young regime where stellar cores are
+decoupled from their envelopes from an old regime where these layers are more
+tightly coupled.
+If so, this could indicate that the phenomenon responsible for changing the
+shape of isochrones in rotation-\teff\ space is related to the phenomenon that
+produces the gap.
+
+\begin{figure}
+  \caption{
+      Top: rotation period vs. effective temperate for stars in the \mct\
+    sample, separated into three groups. Blue circles
+      show stars with rotation periods longer than the
+    period gap, orange squares show stars with rotation periods shorter than
+    the gap, but longer than the lower edge of the main rotation period
+    distribution, and green triangles show stars with rotation periods shorter
+    than this lower edge.
+    Stars were separated into these three groups using \citet{angus2019}
+    gyrochronology models, with the scheme shown in the legend.
+    Only stars cooler than 5000 K are plotted in
+    the bottom panel in order to isolate populations above and below the
+    period gap, which only extends up to temperatures of $\sim$4600 K.
+    Bottom: the velocities of these groups of stars (in the direction of
+    Galactic latitude, $b$) are shown as a function of rotation period.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{gap}
+\label{fig:gap}
+\end{figure}
+
+Figure \ref{fig:gap} shows the velocity dispersions of stars in the \mct\
+sample, with stars subdivided into three groups: those that rotate more
+quickly than the major rotation period distribution (green triangles), those
+with rotation periods shorter than the gap (orange squares), and those with
+rotation periods longer than the gap (blue circles).
+Stars were separated into these three groups using \citet{angus2019}
+gyrochronology models, according to the scheme shown in the legend.
+Only stars cooler than 5000 K are included in the bottom panel in order to
+isolate populations above and below the period gap, which only extends up to a
+temperature of $\sim$4600 K.
+In general, velocity dispersion increases with rotation period because both
+quantities increase with age.
+There is a smooth transition in velocity dispersion between stars with
+rotation periods below and above the gap (orange squares to blue circles),
+suggesting that these groups are part of the same galactic population.
+Previously, only the overall velocity dispersions of all stars above and below
+the gap have been compared, leading to the assumption that these groups belong
+to two distinct populations \citep{mcquillan2014}.
+The velocity dispersion of stars with rotation periods shorter than the lower
+edge of the rotation period distribution (green triangles) is relatively
+large.
+This is an indication that many of these are synchronized binaries.
+The rotation periods of stars in synchronized binaries are tidally locked to
+the orbital period of the binary system and since synchronized binaries have
+short orbital periods this can produce old, seemingly isolated stars, with
+short rotation periods.
+The large velocity dispersions of the most rapidly rotating stars indicates
+that some fraction of these stars are old but rotating rapidly, and therefore
+likely to be synchronized binaries.
+Figure \ref{fig:gap} indicates that there is an increased probability of stars
+with rotation periods less than $\sim$10 days being synchronized binaries.
+This result is in agreement with a recent study which found that a large
+fraction of visual binaries were rapid rotators, and the probability of a star
+being a synchronized binary system substantially increased at rotation periods
+of around 7 days \citep{simonian2019}.
+
+% We calculated the overall velocity dispersions of stars as a function of
+% rotation period and gyrochronal age.
 
@@
 % $\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.
-
-% % The groups of stars with the largest velocity dispersions are cooler than 4500
-% % K, presumably because cooler stars remain active for longer, and their
-% % rotation periods can therefore be measured at older ages.
 
@@
 % % regimes and is the point at which stellar magnetic dynamos likely undergo a
+% \subsection{Velocity dispersion of coeval groups}
+% \label{sec:age_cut}
 % % dramatic structural shift at an age of $\sim$ 1.1 Gyr.

### 2019-12-12 paper/abstract.tex
small edits

--- 2019-12-08 paper/abstract.tex
@@
 % gyrochronology relations.
-In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
-proxy, to reveal their rotational evolution and demonstrate that kinematics
-could be a useful tool for calibrating gyrochronology.
+In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
+an age proxy, to reveal their rotational evolution and demonstrate that
+kinematics could be a useful tool for calibrating gyrochronology.
 % At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
 % to the Praesepe cluster, accurately predicts the relative ages of stars.
-We find that, at young ages, lower-mass stars spin more slowly than
-higher-mass stars because their stronger magnetic fields lead to more
-efficient angular momentum loss.
-This reflects the behavior of young open clusters such as the Pleaides,
-Praesepe, and the Hyades.
-However, at older ages we find that late G and early K dwarfs rotate at the
-same rate as late K dwarfs of the same age, and eventually rotate faster.
+We find that the period-\teff\ relationship of the Praesepe cluster does not
+apply to stars older than around 1 Gyr.
+% : a polynomial gyrochronology model, fit
+% to Praesepe stars, under-predicts the ages of late K dwarfs in the field.
+Although late-K dwarfs spin more slowly than early-K dwarfs when they are
+young, at old ages we find that late-G and early-K dwarfs rotate at the {\it
+same rate} or faster than late-K dwarfs of the same age.
+% This is probably because their stronger magnetic fields lead to more efficient
+% angular momentum loss and reflects the behavior of young open clusters such as
+% the Pleaides, Praesepe, and the Hyades.
+% However, at older ages we find that late G and early K dwarfs rotate at the
+% {\it same rate} or faster than late K dwarfs of the same age.
 % These oThis may be a feature of angular momentum redistribution in the stellar
@@
 % higher-mass stars after a certain age or Rossby number.
-These results align with recent findings from the rotation periods of stars in
-middle-aged open clusters and theoretical models that vary the rate of
-surface-to-core angular momentum transport as a function of time and mass.
-Finally, we find no evidence for mass-dependent heating in a sample of K and M
-dwarfs in the \kepler\ field.
+This result agrees qualitatively with semi-empirical models that vary the rate
+of surface-to-core angular momentum transport as a function of time and mass.
+It also aligns with recent observations of stars in the NGC 6811 cluster,
+which indicate that K dwarfs experience an epoch of stalled spin-down.
+We find that the oldest \kepler\ stars with measured rotation periods are
+late-K and early-M dwarfs, indicating that these stars maintain spotted
+surfaces and stay magnetically active longer than more massive stars.
+Finally, based on their kinematics, we find that many rapidly rotating
+K-dwarfs are likely to be synchronized binaries.
+% Firstly, we
+% find no evidence for mass-dependent heating in a sample of K and M dwarfs in
+% the \kepler\ field.
 

### 2019-12-12 paper/method.tex
small edits

@@
 
-To explore the relationship between rotation period, \teff\ and age, we
-removed high and low velocity outliers from the \mct\ sample by performing
-3$\sigma$ sigma-clipping on the \vb\ velocities.
-Without sigma-clipping, we found that a small number of high velocity
-outliers at the low-temperature end of our sample substantially raised the
-velocity dispersion for cooler stars, however the overall trends remained the
-same.
-We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
-3500 K to avoid biases caused by the selection function at the faint, cool
-end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
-
 % We then selected groups of stars within different gyrochronal age ranges and

### 2019-12-12 paper/results.tex
small edits

@@
 with gyrochronal ages less than 0.5 Gyr.
+After these cuts, 6879 stars were included in the sample.
 
@@
 It has been demonstrated that lower-mass stars remain magnetically active
-longer than more massive stars, \citep{west2008, kiman2019}.
+longer than more massive stars, \citep[\eg][]{west2008, kiman2019}.
 If the detectability of a rotation period is considered to be a magnetic

### 2020-02-18 paper/abstract.tex
old changes

@@
 % to Praesepe stars, under-predicts the ages of late K dwarfs in the field.
-Although late-K dwarfs spin more slowly than early-K dwarfs when they are
-young, at old ages we find that late-G and early-K dwarfs rotate at the {\it
-same rate} or faster than late-K dwarfs of the same age.
+Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
+they are young, at old ages we find that late-G and early-K dwarfs rotate at
+the {\it same rate} or faster than late-K dwarfs of the same age.
 % This is probably because their stronger magnetic fields lead to more efficient

### 2020-02-18 paper/intro.tex
old changes

@@
 pinsonneault1989}.
-Although stars are typically born with random rotation periods, ranging from 1
-to 10 days, observations of young open clusters reveal that their rotation
-periods converge onto a unique sequence by $\sim$500-700 million years
+Although stars are born with random rotation periods, from 1 to 10 days,
+observations of young open clusters reveal that their rotation periods
+converge onto a unique sequence by $\sim$500-700 million years
 \citep[\eg][]{irwin2009, gallet2013}.
@@
 complicated than previously thought.
-For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
-slowly than the G dwarfs.
+For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
+spin more slowly than the G dwarfs.
 In theory this is because lower-mass stars have deeper convections zones which
@@
 Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
-the galaxy with a low out-of-plane, or vertical, velocity ($W$, or \vz),
+the Galaxy with a low out-of-plane, or vertical, velocity (\vz),
 just like the star-forming molecular gas observed in the disk today
@@
 casagrande2011}.
-Although the cause of dynamical heating is not well understood, interactions
-with giant molecular clouds, spiral arms and the galactic bar are thought to
-play an important role \citep[see][for a review of secular evolution in the
-MW]{sellwood2014}.
+Although the cause of vertical dynamical heating is not well understood,
+interactions with giant molecular clouds, spiral arms and the galactic bar are
+thought to play an important role \citep[see][for a review of secular
+evolution in the MW]{sellwood2014}.
 Although the velocity of any individual star will only provide a weak age
-constraint, the velocity {\it dispersion} of a {\it group} of stars can
-indicate whether, on average, that group is old or young relative to other
-groups.
+constraint because velocity is dependent on the position of a star in its
+orbit, the velocity {\it dispersion} of a {\it group} of stars can, in
+principle, average over this effect and indicate whether that group is old or
+young relative to other groups.
+Vertical velocity dispersion is thought to increase monotonically with age.
 In this work we compare the velocity dispersions of groups of field stars in
@@
 only be calculated with full 6-dimensional position and velocity information.
+In fact, with full 6D phase space and an assumed Galactic potential, it is
+possible to calculate vertical {\it action}, which may be an even better age
+indicator \citep{beane2018, ting2019}.
 Unfortunately most field stars with measured rotation periods do not have
@@
 (b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see section
-\ref{sec:results}) approximation to \vz.
+\ref{sec:method}) approximation to \vz.
 Because we use \vb\ rather than \vz\, we cannot calculate absolute kinematic
-ages using an age-velocity dispersion relation (AVR).
+ages using a published age-velocity dispersion relation (AVR), but this could
+be addressed in follow-up work that also takes into account the Kepler
+selection function.
 However, regardless of direction, velocity dispersion is expected to
-monotonically increase over time, and can therefore be used to {\it rank}
-groups of stars by age.
+monotonically increase over time \citep[\eg][]{holmberg2009}, and can
+therefore be used to {\it rank} groups of stars by age.
 
 This paper is laid out as follows: in section \ref{sec:method} we describe our
-sample selection process and the methods used to calculated stellar
+sample selection process and the methods used to calculate stellar
 velocities.
@@
 age proxy by demonstrating that neither mass-dependent heating nor the
-selection function seems to strongly affect on our sample.
+selection function is observed to strongly affect our sample.
 In section \ref{sec:results} we use kinematics to investigate the
 relationship between rotation period, age and color/\teff\ in the field and
-interpret our results in section \ref{sec:discussion}.
+interpret our results.
 % We show that the period-color relation of Praesepe is not applicable to
@@
 % Praesepe (in \gaia\ \gcolor\ color) and the period-age relation of Praesepe
-% and the Sun.
-% The large number of Praesepe members with precise rotation periods from the
+% and the Sun.  The large number of Praesepe members with precise rotation
+% periods from the
 % \ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F

### 2020-02-18 code/data.py

@@ -29,10 +29,10 @@ plt.rcParams.update(plotpar)
 
 
 print("Load McQuillan data")
-mc = pd.read_csv("Table_1_Periodic.txt")
+mc = pd.read_csv("../Table_1_Periodic.txt")
 
 print("Loading Gaia catalog")
-with fits.open("kepler_dr2_1arcsec.fits") as data:
+with fits.open("../kepler_dr2_1arcsec.fits") as data:
     gaia = pd.DataFrame(data[1].data, dtype="float64")
 
 gaia_mc = pd.merge(mc, gaia, on="kepid", how="left")

### 2020-02-18 paper/method.tex
some comments from elisabeth and gus

@@
 The stars are colored by their gyrochronal age, calculated using the
-\citep{angus2019} gyrochronology relation.
+\citet{angus2019} gyrochronology relation.
 The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
@@
 $\sim$ 1.5 dex.
-This suggests that either old M dwarfs are missing from the \mct\ catalog, or
-the \citet{angus2019} gyrochronology relation under-predicts the ages of
-low-mass stars.
+The lack of chronologically old M dwarfs suggests that either old M dwarfs are
+missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
+relation under-predicts the ages of low-mass stars.
 Given that lower-mass stars stay active for longer than higher-mass stars
@@
 rotation periods at old ages, the latter scenario seems more likely.
-The \citep{angus2019} gyrochronology relation is a simple polynomial model,
+The \citet{angus2019} gyrochronology relation is a simple polynomial model,
 fit to the period-color relation of Praesepe.
@@
 and early-M dwarfs.
+This is a common problem with gyrochronology models: since suitable calibrator
+stars are small in number, models cannot be completely calibrated across a
+broad range of parameter space.
+Plotting gyrochronal ages on a color-magnitude diagram reveals the
+deficiencies in the \citet{angus2019} model used to calculate them, however
+imperfections are expected given that no published gyrochronology model is
+thought to be perfectly calibrated.
 \begin{figure}

### 2020-02-19 code/data.py
switching subcut to 4

@@ -104,7 +104,7 @@ w = np.linalg.solve(ATA, np.dot(AT, y))
 
 minb, maxb, extra = 0, 2.2, .27
 xs = np.linspace(minb, maxb, 1000)
-subcut = 5.5
+subcut = 4.
 
 m = (minb < x) * (x < maxb)
 m &= (y < np.polyval(w, x) - extra) + (subcut > y)

### 2020-02-19 code/data.py
saving data as gaia_mc3.csv with new subcut

@@ -188,4 +188,4 @@ gaia_mc["v_dec"] = vdec.value
 gaia_mc["v_b"] = v_b
 
 print("Saving file")
-gaia_mc.to_csv("gaia_mc2.csv")
+gaia_mc.to_csv("gaia_mc3.csv")

### 2020-02-19 paper/method.tex
Elisabeth's comments and some of Gus's

--- 2019-12-12 paper/method.tex
@@
 diagram (CMD) in figure \ref{fig:age_gradient}.
-The stars are colored by their gyrochronal age, calculated using the
-\citep{angus2019} gyrochronology relation.
-The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
+In the bottom panel, the stars are colored by their gyrochronal age,
+calculated using the \citet{angus2019} gyrochronology relation.
+The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
 along the upper edge of the MS, where stellar evolution models predict old
@@
 $\sim$ 1.5 dex.
-This suggests that either old M dwarfs are missing from the \mct\ catalog, or
-the \citet{angus2019} gyrochronology relation under-predicts the ages of
-low-mass stars.
+The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
+are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
+relation under-predicts the ages of low-mass stars.
 Given that lower-mass stars stay active for longer than higher-mass stars
@@
 and early-M dwarfs.
+This is a typical feature of empirically calibrated gyrochronology models:
+since there are no (or at least very few) old M dwarfs with rotation periods,
+the models are poorly calibrated for these stars.
 \begin{figure}
   \caption{
-Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
-We excluded visual binaries by removing stars above the dashed line.
-    Points are colored by their gyrochronal
-    age, according to the
-    \citet{angus2019} gyrochronology relation.
+      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
+    on a \gaia\ CMD.
+    We removed visual binaries and subgiants from the sample by excluding
+    stars above the dashed lines.
+    Bottom: a zoom-in of the top panel, with stars colored by their
+    gyrochronal age \citep{angus2019}, instead of their rotation period.
     A general age gradient is visible across the main sequence.
@@
   \centering
-    \includegraphics[width=1\textwidth]{age_gradient_combo}
+    % \includegraphics[width=1\textwidth]{age_gradient_combo}
+    \includegraphics[width=1\textwidth]{CMD_cuts_double}
 \label{fig:age_gradient}
@@
 {\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
-covariance matrix via Monte Carlo sampling.
-It therefore not only incorporates uncertainties on the \gaia\ positions
-parallaxes and proper motions, it also accounts for the {\it covariance}
-between these properties.
+covariance matrix via Monte Carlo sampling, thereby accounting for the
+ovariances between \gaia\ positions, parallaxes and proper motions.
 Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
@@
 
-\subsection{Validating \vb\ dispersion as an age proxy}
-\label{sec:mass-dependent-heating}
+% \subsection{Validating \vb\ dispersion as an age proxy}
+% \label{sec:mass-dependent-heating}
 
-There are two main reasons why \vb\ velocity dispersion may not be a good age
-proxy.
-Firstly, mass-dependent heating may act on the sample, meaning that velocity
-dispersion depends on both age and mass, so cannot be interpreted as a simple
-age proxy.
-Secondly, since stars in the \kepler\ field have a range of galactic
-latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
-stars, and introduce a velocity bias for high latitude stars (which are more
-likely to be cooler and older).
-In this section we demonstrate that neither of these problems seem to be a
-significant issue for our data.
+% There are two main reasons why \vb\ velocity dispersion may not be a good age
+% proxy.
+% Firstly, mass-dependent heating may act on the sample, meaning that velocity
+% dispersion depends on both age and mass, so cannot be interpreted as a simple
+% age proxy.
+% Secondly, since stars in the \kepler\ field have a range of galactic
+% latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
+% stars, and introduce a velocity bias for high latitude stars (which are more
+% likely to be cooler and older).
+% In this section we demonstrate that neither of these problems seem to be a
+% significant issue for our data.
 
-If lower-mass stars experience greater velocity changes when gravitationally
-perturbed, and are dynamically heated more efficiently than higher-mass stars,
-velocity dispersion would be a function of both age {\it and} mass and cannot
-be interpreted as a straightforward age proxy.
-So, in order to establish whether \sigmavb\ can be used as an age proxy, we
-searched for signs of mass-dependent heating within the \kepler\ field.
-Mass-dependent dynamical heating has not been unambiguously observed in the
-galactic disk because of the strong anti-correlation between stellar mass and
-stellar age.
-Less massive stars do indeed have larger velocity dispersions, however they
-are also older on average.
-This mass-age degeneracy is highly reduced in M dwarfs because their
-main-sequence lifetimes are longer than the age of the Universe, and no
-evidence for mass-dependent heating has previously been found in M dwarfs
-\citep[\eg][]{faherty2009, newton2016}.
+% In order to establish whether \sigmavb\ can be used as an age proxy, we
+% searched for signs of mass-dependent heating within the \kepler\ field.
+% Mass-dependent dynamical heating may result from lower-mass stars experiencing
+% greater velocity changes when gravitationally perturbed than more massive
+% stars.
+% It has not been unambiguously observed in the galactic disk because of the
+% strong anti-correlation between stellar mass and stellar age.
+% Less massive stars do indeed have larger velocity dispersions, however they
+% are also older on average.
+% This mass-age degeneracy is highly reduced in M dwarfs because their
+% main-sequence lifetimes are longer than the age of the Universe, and no
+% evidence for mass-dependent heating has previously been found in M dwarfs
+% \citep[\eg][]{faherty2009, newton2016}.
 
-To further investigate whether mass-dependent heating could be acting on the
-\kepler\ sample, we selected late K and M dwarfs observed by both \kepler\ and
-\gaia, whose MS lifetimes exceed around 11 Gyrs and are therefore
-representative of the initial mass function\footnote{We could not perform this
-analysis on the \mct\ sample, because it only includes stars with {\it
-detectable} rotation periods, and since lower-mass stars stay active for
-longer it is likely that it contains a strong mass-age correlation.}.
-We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
-greater than 1.2 (corresponding to an effective temperature $\lesssim$
-4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
-We also eliminated visual binaries by removing stars above a 6th order
-polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
-figure \ref{fig:age_gradient}).
-We then applied the quality cuts described above in section
-\ref{sec:the_data}.
-To search for evidence of mass-dependent heating we calculated the (\vb)
-velocity dispersion of stars in effective temperature bins.
-Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
-outliers before calculating the standard deviation of stars in each bin.
-These high velocity outliers may be very old late K and M dwarfs, or they
-result from using \vb\ instead of \vz, which introduces additional velocity
-scatter.
+% To investigate whether mass-dependent heating could be acting on the \kepler\
+% sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
+% whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
+% the initial mass function.
+% We could not perform this analysis on the \mct\ sample, because it only
+% includes stars with {\it detectable} rotation periods, and since lower-mass
+% stars stay active for longer it is likely that it contains a strong mass-age
+% correlation.
+% We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
+% greater than 1.2 (corresponding to an effective temperature $\lesssim$
+% 4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+% We also eliminated visual binaries by removing stars above a 6th order
+% polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
+% figure \ref{fig:age_gradient}).
+% We then applied the quality cuts described above in section
+% \ref{sec:the_data}.
+% To search for evidence of mass-dependent heating we calculated the (\vb)
+% velocity dispersion of stars in effective temperature bins.
+% Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
+% outliers before calculating the standard deviation of stars in each bin.
+% These high velocity outliers may be very old late K and M dwarfs, or they
+% result from using \vb\ instead of \vz, which introduces additional velocity
+% scatter.
 
-Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
-function of effective temperature \footnote{calculated by transforming
-dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
-\kepler\ dwarf sample.
-Velocity dispersion very slightly {\it decreases} with decreasing temperature,
-the opposite of the trend expected for mass-dependent heating, however the
-slope is only inconsistent with zero at the 1.3 $\sigma$ level.
-This trend may be due to a selection bias: cooler stars are fainter and
-therefore typically closer, with smaller heights above the galactic plane and
-smaller velocities.
-The essential point however, is that we do not see evidence for mass-dependent
-heating acting on stars in the \kepler\ field, indicating that velocity
-dispersion {\it can} be used as an age proxy.
-This analysis was performed using \vb\ but we also examined the {\it vertical}
-velocities of the 537 stars in this sample with RV measurements.
-Again, no evidence was found for mass-dependent heating: the slope of the
-velocity dispersion-temperature relation was consistent with zero.
-\begin{figure}
-  \caption{
-      Top: Stellar velocity (\vb) as a function of \teff\ for
-      \kepler\ K and M dwarfs.
-Vertical lines indicate different \teff-groupings used to calculate velocity
-    dispersion.
-Pink stars were not included in velocity dispersion calculations as they were
-    either removed as outliers during a sigma clipping process, or they lie at
-    the sparcely populated, extremely cool end of the temperature range.
-    Velocity dispersion and \teff\ are slightly positively correlated, likely
-    due to a brightness-related selection bias, indicating that mass-dependent
-    heating does not significantly affect low-mass stars in the \kepler\
-    field.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vb_vs_teff}
-\label{fig:vb_vs_teff}
-\end{figure}
+% Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
+% function of effective temperature \footnote{calculated by transforming
+% dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
+% \kepler\ dwarf sample.
+% Velocity dispersion very slightly {\it decreases} with decreasing temperature,
+% the opposite of the trend expected for mass-dependent heating, however the
+% slope is only inconsistent with zero at the 1.3 $\sigma$ level.
+% This trend may be due to a selection bias: cooler stars are fainter and
+% therefore typically closer, with smaller heights above the galactic plane and
+% smaller velocities.
+% The essential point however, is that we do not see evidence for mass-dependent
+% heating acting on stars in the \kepler\ field, indicating that velocity
+% dispersion {\it can} be used as an age proxy.
+% This analysis was performed using \vb\ but we also examined the {\it vertical}
+% velocities of the 537 stars in this sample with RV measurements.
+% Again, no evidence was found for mass-dependent heating: the slope of the
+% velocity dispersion-temperature relation was consistent with zero.
+% \begin{figure}
+%   \caption{
+%       Top: Stellar velocity (\vb) as a function of \teff\ for
+%       \kepler\ K and M dwarfs.
+% Vertical lines indicate different \teff-groupings used to calculate velocity
+%     dispersion.
+% Pink stars were not included in velocity dispersion calculations as they were
+%     either removed as outliers during a sigma clipping process, or they lie at
+%     the sparcely populated, extremely cool end of the temperature range.
+%     Velocity dispersion and \teff\ are slightly positively correlated, likely
+%     due to a brightness-related selection bias, indicating that mass-dependent
+%     heating does not significantly affect low-mass stars in the \kepler\
+%     field.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{vb_vs_teff}
+% \label{fig:vb_vs_teff}
+% \end{figure}
 
-Having found no strong evidence for mass-dependent heating, we next tested
-the validity of \vb\ as a proxy for \vz\ in more detail.
-At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
-of $b$, this equivalence becomes an approximation that grows noisier with $b$.
-To test the validity of the \vb$\sim$\vz\ approximation over a range of
-latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
-(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
-We downloaded stars from four pointings in the \kepler\ field with galactic
-latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
-a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
-The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
-\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
-Galaxy.
-We subtracted this offset and examined the residuals of the \vz\ -- \vb\
-relationship to investigate the variance as a function of Galactic latitude.
-We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
-with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
-The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
-$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
-20$^\circ$.
+% Having found no strong evidence for mass-dependent heating, we next tested
+% the validity of \vb\ as a proxy for \vz\ in more detail.
+% At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
+% of $b$, this equivalence becomes an approximation that grows noisier with $b$.
+% To test the validity of the \vb$\sim$\vz\ approximation over a range of
+% latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
+% (GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
+% We downloaded stars from four pointings in the \kepler\ field with galactic
+% latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
+% a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
+% The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
+% \vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
+% Galaxy.
+% We subtracted this offset and examined the residuals of the \vz\ -- \vb\
+% relationship to investigate the variance as a function of Galactic latitude.
+% We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
+% with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
+% The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
+% $, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
+% 20$^\circ$.
 
-% Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
-% over different Galactic latitudes.
-% Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
-% fits are shown as dashed blue lines.
-% The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
-% the variance increases with increasing Galactic latitude.
-\begin{figure}
-  \caption{
-This figure demonstrates the variance in the relationship between \vb\ and
-    \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
-The panels show a kernel density estimator (KDE) (black solid line) for
-    the \vz -- \vb\ residuals of stars in the GUMS simulation at four
-    different Galactic latitudes.
-Blue dashed lines show Gaussian fits to these KDEs.
-The distributions are close to Gaussian, with slightly heavy tails.
-The standard deviations of the Gaussian fits increase with Galactic latitude.
-This figure illustrates how using \vb\ instead of \vz\ artificially
-    increases velocity dispersion, especially at high latitudes.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vb_vz}
-\label{fig:vb_vz}
-\end{figure}
+% % Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
+% % over different Galactic latitudes.
+% % Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
+% % fits are shown as dashed blue lines.
+% % The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
+% % the variance increases with increasing Galactic latitude.
+% \begin{figure}
+%   \caption{
+% This figure demonstrates the variance in the relationship between \vb\ and
+%     \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
+% The panels show a kernel density estimator (KDE) (black solid line) for
+%     the \vz -- \vb\ residuals of stars in the GUMS simulation at four
+%     different Galactic latitudes.
+% Blue dashed lines show Gaussian fits to these KDEs.
+% The distributions are close to Gaussian, with slightly heavy tails.
+% The standard deviations of the Gaussian fits increase with Galactic latitude.
+% This figure illustrates how using \vb\ instead of \vz\ artificially
+%     increases velocity dispersion, especially at high latitudes.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{vb_vz}
+% \label{fig:vb_vz}
+% \end{figure}
 
-Since we are concerned with velocity {\it dispersions}, rather than velocities
-themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
-function of temperature for
-stars downloaded from the GUMS simulation.
-For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
-consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
-latitudes the two quantities became significantly different.
-For this reason we proceeded by only including stars with galactic latitudes
-less than 15\degrees\ in our analysis.
-Although it seems that the transformation between \vz\ and \vb\ does not {\it
-strongly} affect our results, we cannot rule out the possibility that it
-introduces systematic biases into the velocity dispersions we present here.
-In \gaia\ DR3, RVs will be available for most stars in this sample, providing
-an opportunity to validate (or correct) the results presented here.
+% Since we are concerned with velocity {\it dispersions}, rather than velocities
+% themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
+% function of temperature for
+% stars downloaded from the GUMS simulation.
+% For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
+% consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
+% latitudes the two quantities became significantly different.
+% For this reason we proceeded by only including stars with galactic latitudes
+% less than 15\degrees\ in our analysis.
+% Although it seems that the transformation between \vz\ and \vb\ does not {\it
+% strongly} affect our results, we cannot rule out the possibility that it
+% introduces systematic biases into the velocity dispersions we present here.
+% In \gaia\ DR3, RVs will be available for most stars in this sample, providing
+% an opportunity to validate (or correct) the results presented here.
 
-% Since we use \vb, not \vz\ in our analysis, it is possible that the sample
-% selection function could influence our results.
-% For example, higher-mass stars tend to be younger and reside closer to the
-% galactic disk mid-plane (low galactic latitudes).
-% Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
-% lower-mass stars typically have greater velocity dispersions due to .
-% \teff\ and galactic latitude, $b$,
-% could result in
-% Stars at higher latitudes have additional velocity components in the ${\bf x}$
-% and ${\bf y}$ directions, which could increase \vb\ but not \vz.
-% Again however, since the relationship between \sigmavb\ and \teff\ is
-% positively, not negatively correlated for cool stars in the \kepler\ field,
-% this effect is probably too small to influence our results.
+% % Since we use \vb, not \vz\ in our analysis, it is possible that the sample
+% % selection function could influence our results.
+% % For example, higher-mass stars tend to be younger and reside closer to the
+% % galactic disk mid-plane (low galactic latitudes).
+% % Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
+% % lower-mass stars typically have greater velocity dispersions due to .
+% % \teff\ and galactic latitude, $b$,
+% % could result in
+% % Stars at higher latitudes have additional velocity components in the ${\bf x}$
+% % and ${\bf y}$ directions, which could increase \vb\ but not \vz.
+% % Again however, since the relationship between \sigmavb\ and \teff\ is
+% % positively, not negatively correlated for cool stars in the \kepler\ field,
+% % this effect is probably too small to influence our results.
 
-% Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
-% strongly affect the \mct\ sample of rotating \kepler\ stars.
-% For this reason, we assume that age difference is the major cause of velocity
-% dispersion differences between groups.
-% In other words, (\vb) velocity dispersion is a reliable age proxy for the
-% \mct\ sample.
+% % Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+% % strongly affect the \mct\ sample of rotating \kepler\ stars.
+% % For this reason, we assume that age difference is the major cause of velocity
+% % dispersion differences between groups.
+% % In other words, (\vb) velocity dispersion is a reliable age proxy for the
+% % \mct\ sample.
 
-Because of the noisy relationship between \vb\ and \vz\, in this paper we do
-not attempt to convert velocity dispersion (\sigmavb) into an age via an
-age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
-Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
-careful analysis that includes formal modeling of the \vb\ -- \vz\
-relationship will be needed to calculate absolute ages.
-% For example, the velocity distributions could be modeled as a mixture of
-% Gaussians in order to account for the additional velocity dispersion caused by
-% the \vz-\vb\ transformation.
-The RVs of most of these stars will become available in \Gaia\ DR3, allowing
-calculations of \vz, which can be used to calculate more reliable ages via an
-AVR.
+% Because of the noisy relationship between \vb\ and \vz\, in this paper we do
+% not attempt to convert velocity dispersion (\sigmavb) into an age via an
+% age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
+% Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
+% careful analysis that includes formal modeling of the \vb\ -- \vz\
+% relationship will be needed to calculate absolute ages.
+% % For example, the velocity distributions could be modeled as a mixture of
+% % Gaussians in order to account for the additional velocity dispersion caused by
+% % the \vz-\vb\ transformation.
+% The RVs of most of these stars will become available in \Gaia\ DR3, allowing
+% calculations of \vz, which can be used to calculate more reliable ages via an
+% AVR.
 
-% We then selected groups of stars within different gyrochronal age ranges and
-% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
-% of effective temperature for each age group.
-% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-% throughout this paper we show rotation periods as a function of effective
-% temperature, \teff, because it is easier to divide stars into bins of roughly
-% equal numbers in \teff-space than in color-space.
+% % We then selected groups of stars within different gyrochronal age ranges and
+% % calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
+% % of effective temperature for each age group.
+% % Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+% % throughout this paper we show rotation periods as a function of effective
+% % temperature, \teff, because it is easier to divide stars into bins of roughly
+% % equal numbers in \teff-space than in color-space.

### 2020-02-19 paper/results.tex
Elisabeth's comments and some of Gus's

@@
 velocity dispersion for cooler stars, however the overall trends remain the
-same with, or without sigma-clipping.
+same without sigma-clipping.
 We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
@@
 end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
-Finally, we removed stars with short rotation periods by cutting out stars
-with gyrochronal ages less than 0.5 Gyr.
+Finally, we removed stars with rotation periods shorter than the bulk of
+periods, since this area of the period-\teff\ diagram is sparsely populated.
+We removed rapid rotators by cutting out stars with gyrochronal ages less than
+0.5 Gyr, because a 0.5 Gyr gyrochrone\footnote{A gyrochrone is a
+gyrochronological isochrone, or a line of constant age in period-\teff, or
+period-color space.} traces the bottom edge of the main population of rotation
+periods.
 After these cuts, 6879 stars were included in the sample.
@@
     grid in $\log_{10}$(period) and \teff.
-    Black lines show isochrones from a gyrochronology model that projects the
+    Black lines show gyrochrones from a gyrochronology model that projects the
     rotation-color relation of
     Praesepe to longer rotation periods over time \citep{angus2019}.
-    These isochrones do not appear to reflect the evolution of field stars at
+    These gyrochrones do not appear to reflect the evolution of field stars at
     long rotation periods/old ages: they do not trace lines of constant
     velocity dispersion.
-    Isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
+    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
     and bottom panels.
@@
     \citep[from the Kepler Input Catalog][]{brown2011}.
-    White lines show isochrones from a model that includes mass and
+    White lines show gyrochrones from a model that includes mass and
     age-dependent angular momentum transport between the core and envelope
     \citep{spada2019}.
-    Qualitatively, these isochrones reflect the evolution of field
+    Qualitatively, these gyrochrones reflect the evolution of field
     stars at long rotation periods/old ages: they trace lines of constant
@@
 Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
-rotation period across all temperatures, implying that both velocity
-dispersion and rotation period increases with age as expected.
-Black lines show isochrones from the \citep{angus2019} gyrochronology model,
+rotation period across all temperatures, implying that rotation period
+increases with age as expected.
+Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
 which projects the rotation-color relation of Praesepe to longer rotation
 periods over time.
-These isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
-At the youngest ages, these isochrones seem to describe the data well: the
-palest yellow (youngest) stars with the lowest velocity dispersions all fall
-close to the 0.5 Gyr isochrone.
+These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
+At the youngest ages, these gyrochrones describe the data well: the palest
+yellow (youngest) stars with the lowest velocity dispersions all fall close to
+the 0.5 Gyr gyrochrone.
 % Lines of constant age (isochrones) appear to follow the shape of the
@@
 % decreasing effective temperature at a given age.
-However, although the 0.5 Gyr and 1 Gyr isochrones do appear to trace constant
-velocity dispersion/age among the field stars, by 1.5 Gyrs the isochrones
+However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
+velocity dispersion/age among the field stars, by 1.5 Gyrs the gyrochrones
 start to {\it cross} different velocity dispersion regimes.
-For example, the 1.5 Gyr isochrone lies on top of stars with velocity
+For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
 dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
 \kms\ velocity dispersions at 4000-4500K.
-The isochrones older than 1.5 Gyr also cross a range of velocity dispersions.
+The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
 If these were true isochrones however, they should follow lines of constant
 velocity dispersion.
-At ages older than around 1 Gyr, it appears that isochrones should have a more
+At ages older than around 1 Gyr, it appears that gyrochrones should have a more
 flattened, or even inverted relationship between rotation period and effective
@@
 function of rotation period and {\it mass}, obtained from the Kepler Input
-Catalog \citep{brown2011}, with isochrones from the \citep{spada2019}
+Catalog \citep{brown2011}, with gyrochrones from the \citep{spada2019}
 gyrochronology model shown as white lines.
-These isochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
+These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
 Although these models do not trace lines of constant velocity dispersion at
@@
 picture of rotational evolution.
-At young ages, stellar rotation period decreases with mass, likely because
-lower-mass stars with deeper convection zones have stronger magnetic fields,
-larger Alfv\'en radii and therefore experience greater angular momentum loss
-rate.
+At young ages \citep[younger than around 1 Gyr but still old enough to be on
+the main sequence and have transitioned from the `I' sequence to the `C'
+sequence ][]{barnes2003}, stellar rotation period decreases with increasing
+mass.
+This is likely because lower-mass stars with deeper convection zones have
+stronger magnetic fields, larger Alfv\'en radii and therefore experience
+greater angular momentum loss rate.
 According to the \citet{spada2019} model, there is minimal transportation of
@@
 There is a sharp gap in the population of rotation periods, which lies just
-above the 1 Gyr isochrone in the upper panel of figure \ref{fig:vplot}, whose
+above the 1 Gyr gyrochrone in the upper panel of figure \ref{fig:vplot}, whose
 origin is unknown and is the subject of much speculation \citep{mcquillan2014,
@@
 If so, this could indicate that the phenomenon responsible for changing the
-shape of isochrones in rotation-\teff\ space is related to the phenomenon that
+shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
 produces the gap.
@@
 isolate populations above and below the period gap, which only extends up to a
-temperature of $\sim$4600 K.
+temperature of $\sim$4600 K in our sample \citep[Although][found that the gap
+extends to temperatures as hot as 6000 K]{davenport2017}.
 In general, velocity dispersion increases with rotation period because both
@@
 to two distinct populations \citep{mcquillan2014}.
-The velocity dispersion of stars with rotation periods shorter than the lower
-edge of the rotation period distribution (green triangles) is relatively
-large.
-This is an indication that many of these are synchronized binaries.
-The rotation periods of stars in synchronized binaries are tidally locked to
-the orbital period of the binary system and since synchronized binaries have
-short orbital periods this can produce old, seemingly isolated stars, with
-short rotation periods.
+On the other hand, the velocity dispersions of stars with rotation periods
+shorter than the lower edge of the rotation period distribution (green
+triangles) are not significantly smaller than the, presumed older,
+orange-colored stars.
+% This is an indication that many of these are synchronized binaries.
+% The rotation periods of stars in synchronized binaries are tidally locked to
+% the orbital period of the binary system and since synchronized binaries have
+% short orbital periods this can produce old, seemingly isolated stars, with
+% short rotation periods.
 The large velocity dispersions of the most rapidly rotating stars indicates
@@
 likely to be synchronized binaries.
+% Given the strong correlation between binary fraction and primary mass
+% \citep[\eg][]{price-whelan2020}, and the fact that M dwarfs take longer to
+% transition from the `I` (rapidly rotating) sequence to the `C' (slowly
+% rotating) sequence, it is likely that
 Figure \ref{fig:gap} indicates that there is an increased probability of stars
@@
 fraction of visual binaries were rapid rotators, and the probability of a star
-being a synchronized binary system substantially increased at rotation periods
-of around 7 days \citep{simonian2019}.
+being a synchronized binary system substantially increased below rotation
+periods of around 7 days \citep{simonian2019}.
+
+\subsection{Validating \vb\ dispersion as an age proxy}
+\label{sec:mass-dependent-heating}
+
+There are two main reasons why \vb\ velocity dispersion may not be a good age
+proxy.
+Firstly, mass-dependent heating may act on the sample, meaning that velocity
+dispersion depends on both age and mass, so cannot be interpreted as a simple
+age proxy.
+Secondly, since stars in the \kepler\ field have a range of galactic
+latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
+stars, and introduce a velocity bias for high latitude stars (which are more
+likely to be cooler and older).
+In this section we demonstrate that neither of these problems seem to be a
+significant issue for our data.
+
+In order to establish whether \sigmavb\ can be used as an age proxy, we
+searched for signs of mass-dependent heating within the \kepler\ field.
+Mass-dependent dynamical heating may result from lower-mass stars experiencing
+greater velocity changes when gravitationally perturbed than more massive
+stars.
+It has not been unambiguously observed in the galactic disk because of the
+strong anti-correlation between stellar mass and stellar age.
+Less massive stars do indeed have larger velocity dispersions, however they
+are also older on average.
+This mass-age degeneracy is highly reduced in M dwarfs because their
+main-sequence lifetimes are longer than the age of the Universe, and no
+evidence for mass-dependent heating has previously been found in M dwarfs
+\citep[\eg][]{faherty2009, newton2016}.
+
+To investigate whether mass-dependent heating could be acting on the \kepler\
+sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
+whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
+the initial mass function.
+We could not perform this analysis on the \mct\ sample, because it only
+includes stars with {\it detectable} rotation periods, and since lower-mass
+stars stay active for longer it is likely that it contains a strong mass-age
+correlation.
+We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
+greater than 1.2 (corresponding to an effective temperature $\lesssim$
+4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+We also eliminated visual binaries by removing stars above a 6th order
+polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
+figure \ref{fig:age_gradient}).
+We then applied the quality cuts described above in section
+\ref{sec:the_data}.
+To search for evidence of mass-dependent heating we calculated the (\vb)
+velocity dispersion of stars in effective temperature bins.
+Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
+outliers before calculating the standard deviation of stars in each bin.
+These extreme velocity outliers may be very old late K and M dwarfs, or they
+result from using \vb\ instead of \vz, which introduces additional velocity
+scatter.
+
+Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
+function of effective temperature \footnote{calculated by transforming
+dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
+\kepler\ dwarf sample.
+Velocity dispersion very slightly {\it decreases} with decreasing temperature,
+the opposite of the trend expected for mass-dependent heating, however the
+slope is only inconsistent with zero at 1.3 $\sigma$.
+This trend may be due to a selection bias: cooler stars are fainter and
+therefore typically closer, with smaller heights above the galactic plane and
+smaller velocities.
+The essential point however, is that we do not see evidence for mass-dependent
+heating acting on stars in the \kepler\ field, indicating that velocity
+dispersion {\it can} be used as an age proxy (with the caveat that there is
+still a chance, albeit a small one, that the opposing effects of the selection
+function and mass-dependent heating are working to cancel each other out).
+This analysis was performed using \vb\ but we also examined the {\it vertical}
+velocities of the 537 stars in this sample with RV measurements.
+Again, no evidence was found for mass-dependent heating: the slope of the
+velocity dispersion-temperature relation was consistent with zero.
+\begin{figure}
+  \caption{
+      Top: Stellar velocity (\vb) as a function of \teff\ for
+      \kepler\ K and M dwarfs.
+Vertical lines indicate different \teff-groupings used to calculate velocity
+    dispersion.
+Pink stars were not included in velocity dispersion calculations as they were
+    either removed as outliers during a sigma clipping process, or they lie at
+    the sparcely populated, extremely cool end of the temperature range.
+    Velocity dispersion and \teff\ are slightly positively correlated, likely
+    due to a brightness-related selection bias, indicating that mass-dependent
+    heating does not significantly affect low-mass stars in the \kepler\
+    field.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{vb_vs_teff}
+\label{fig:vb_vs_teff}
+\end{figure}
+
+Having found no strong evidence for mass-dependent heating, we next tested
+the validity of \vb\ as a proxy for \vz\ in more detail.
+At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
+of $b$, this equivalence becomes an approximation that grows noisier with $b$.
+To test the validity of the \vb$\sim$\vz\ approximation over a range of
+latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
+(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
+We downloaded stars from four pointings in the \kepler\ field with galactic
+latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
+a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
+The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
+\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
+Galaxy.
+We subtracted this offset and examined the residuals of the \vz\ -- \vb\
+relationship to investigate the variance as a function of Galactic latitude.
+We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
+with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
+The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
+$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
+20$^\circ$.
+
+% Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
+% over different Galactic latitudes.
+% Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
+% fits are shown as dashed blue lines.
+% The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
+% the variance increases with increasing Galactic latitude.
+\begin{figure}
+  \caption{
+This figure demonstrates the variance in the relationship between \vb\ and
+    \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
+The panels show a kernel density estimator (KDE) (black solid line) for
+    the \vz -- \vb\ residuals of stars in the GUMS simulation at four
+    different Galactic latitudes.
+Blue dashed lines show Gaussian fits to these KDEs.
+The distributions are close to Gaussian, with slightly heavy tails.
+The standard deviations of the Gaussian fits increase with Galactic latitude.
+This figure illustrates how using \vb\ instead of \vz\ artificially
+    increases velocity dispersion, especially at high latitudes.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{vb_vz}
+\label{fig:vb_vz}
+\end{figure}
+
+Since we are concerned with velocity {\it dispersions}, rather than velocities
+themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
+function of temperature for
+stars downloaded from the GUMS simulation.
+For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
+consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
+latitudes the two quantities became significantly different.
+For this reason we proceeded by only including stars with galactic latitudes
+less than 15\degrees\ in our analysis.
+Although we find that the transformation between \vz\ and \vb\ does not {\it
+strongly} affect our results, we cannot rule out the possibility that it
+introduces systematic biases into the velocity dispersions we present here.
+In \gaia\ DR3, RVs will be available for most stars in this sample, providing
+an opportunity to validate (or correct) the results presented here, and to
+work in action-space, rather than velocity-space.
+
+% Since we use \vb, not \vz\ in our analysis, it is possible that the sample
+% selection function could influence our results.
+% For example, higher-mass stars tend to be younger and reside closer to the
+% galactic disk mid-plane (low galactic latitudes).
+% Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
+% lower-mass stars typically have greater velocity dispersions due to .
+% \teff\ and galactic latitude, $b$,
+% could result in
+% Stars at higher latitudes have additional velocity components in the ${\bf x}$
+% and ${\bf y}$ directions, which could increase \vb\ but not \vz.
+% Again however, since the relationship between \sigmavb\ and \teff\ is
+% positively, not negatively correlated for cool stars in the \kepler\ field,
+% this effect is probably too small to influence our results.
+
+% Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+% strongly affect the \mct\ sample of rotating \kepler\ stars.
+% For this reason, we assume that age difference is the major cause of velocity
+% dispersion differences between groups.
+% In other words, (\vb) velocity dispersion is a reliable age proxy for the
+% \mct\ sample.
+
+Because of the noisy relationship between \vb\ and \vz\, in this paper we do
+not attempt to convert velocity dispersion (\sigmavb) into an age via an
+age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
+Although we find that \sigmavb\ can be used to rank groups of stars by age, a
+more careful analysis that includes formal modeling of the \vb\ -- \vz\
+relationship will be needed to calculate absolute ages.
+% For example, the velocity distributions could be modeled as a mixture of
+% Gaussians in order to account for the additional velocity dispersion caused by
+% the \vz-\vb\ transformation.
+The RVs of most of these stars will become available in \Gaia\ DR3, allowing
+calculations of \vz, which can be used to calculate more reliable ages via an
+AVR and suitable modeling of the selection function.
+
+% We then selected groups of stars within different gyrochronal age ranges and
+% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
+% of effective temperature for each age group.
+% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+% throughout this paper we show rotation periods as a function of effective
+% temperature, \teff, because it is easier to divide stars into bins of roughly
+% equal numbers in \teff-space than in color-space.
 

### 2020-02-21 paper/conclusion.tex
Elisabeth's comments and some of Gus's
Gus's comments

@@
 increase with decreasing mass.
-Although {\it vertical} velocity, \vz, is the quantity that most strongly
-traces time-dependent orbital heating in the disc of the Galaxy, most stars
-with measured rotation periods do not yet have radial velocities, so we used
-velocity in the direction of Galactic latitude, \vb\, as a proxy for \vz.
+Although {\it vertical} velocity, \vz, is a quantity that has been
+demonstrated to trace time-dependent orbital heating in the disc of the
+Galaxy, most stars with measured rotation periods do not yet have radial
+velocities, so we used velocity in the direction of Galactic latitude, \vb\,
+as a proxy for \vz.
 Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
@@
 describe the period-age-\teff\ relation for old stars.
-At young ages, rotation period is anti-correlated with \teff: cooler stars
-spin more slowly than hotter stars of the same age.
+Instead we found that, at young ages, rotation period is anti-correlated with
+\teff: cooler stars spin more slowly than hotter stars of the same age.
 However, at intermediate ages the relation flattens out and K dwarfs rotate at
@@
 gap, indicating that the gap does not separate two distinct populations.
-Finally, we used kinematics to reveal a population of synchronized binaries
-with rotation periods less than around 10 days.
+Finally, we used kinematics to indicate that there is a population of
+synchronized binaries with rotation periods less than around 10 days.
 

### 2020-02-21 paper/results.tex
Gus's comments

@@
 The results shown in figure \ref{fig:vplot} indicate that stars of spectral
-type ranging from late G to late K follow a braking law that changes over
-time.
+type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
+that changes over time.
 In particular, the relationship between rotation period and effective
@@
     Galactic latitude, $b$) are shown as a function of rotation period.
+    The black line indicates the velocity standard deviation as a function of
+    period.
 }
@@
 
-Figure \ref{fig:gap} shows the velocity dispersions of stars in the \mct\
-sample, with stars subdivided into three groups: those that rotate more
-quickly than the major rotation period distribution (green triangles), those
-with rotation periods shorter than the gap (orange squares), and those with
-rotation periods longer than the gap (blue circles).
+The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
+stars in the \mct\ sample, with stars subdivided into three groups: those that
+rotate more quickly than the major rotation period distribution (green
+triangles), those with rotation periods shorter than the gap (orange squares),
+and those with rotation periods longer than the gap (blue circles).
 Stars were separated into these three groups using \citet{angus2019}

### 2020-02-25 paper/results.tex
Rocio's comments

@@
     velocity dispersion.
-    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
+    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
     and bottom panels.
@@
 increases with age as expected.
+This result is insensitive to the choice of bin position and size.
 Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
@@
 periods over time.
-These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
+These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
 At the youngest ages, these gyrochrones describe the data well: the palest
@@
 However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
-velocity dispersion/age among the field stars, by 1.5 Gyrs the gyrochrones
+velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
 start to {\it cross} different velocity dispersion regimes.
@@
 gyrochronology model shown as white lines.
-These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
+These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
 Although these models do not trace lines of constant velocity dispersion at
@@
 sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
-whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
+whose MS lifetimes exceed around 11 Gyr and are therefore representative of
 the initial mass function.
@@
 We subtracted this offset and examined the residuals of the \vz\ -- \vb\
-relationship to investigate the variance as a function of Galactic latitude.
+relationship to investigate the variance as a function of Galactic latitude
+(shown in figure \ref{fig:vb_vz}).
 We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
@@
 % the \vz-\vb\ transformation.
-The RVs of most of these stars will become available in \Gaia\ DR3, allowing
-calculations of \vz, which can be used to calculate more reliable ages via an
-AVR and suitable modeling of the selection function.
+% The RVs of most of these stars will become available in \Gaia\ DR3, allowing
+% calculations of \vz, which can be used to calculate more reliable ages via an
+% AVR and suitable modeling of the selection function.
 

### 2020-02-27 paper/abstract.tex
affiliations

@@
 
-The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
+The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
 billion years due to a lack of precise ages and rotation periods for old

### 2020-02-27 paper/method.tex
comments from Jason

@@
 \citep{green2018}, and {\tt astropy} \citep{astropy2013}.
-We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
-using an 8th-order polynomial relation calibrated using .... stars
-\racomment{ask Jason for details}.
+
+For this work, we used the precise \textit{Gaia} DR2 photometric color,
+$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
+Curtis \etal\ (2020, in prep) combined \teff s for nearby, unreddened field
+stars from benchmark samples, including FGK stars characterized with
+high-resolution optical spectroscopy \citep{brewer2016}, M dwarfs
+characterized with low-resolution optical and near-infrared spectroscopy
+\citep{mann2015}, and K and M dwarfs characterized with interferometry and
+bolometric flux analyses \citep{boyajian2012}.
+This empirical color--temperature relation is valid over the color range $0.55
+< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $6470 < T_{\rm eff} <
+3070$~K.
+The dispersion about the relation implies a high precision of 50~K. Using
+these benchmark data enable us to accurately estimate \teff\ for cool dwarfs
+\citep[\eg][]{rabus2019}, and allows to correct for interstellar reddening at
+all temperatures\footnote{The color--temperature relation is described in
+detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
+\etal\ (2020, in prep).}.
+The equation we used to calculate photometric temperatures is,
 \begin{equation}
-    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
-    301C^6 - 26C^7,
+    \mathrm{T_{eff}} = -416.585 + 39780.0C -84190.5C^2 + 85203.9C^3
+    -48225.9C^4 + 15598.5C^5 -2694.76C^6 + 192.865C^7,
+% Jason's updated parameters:
+% −416.585
+% 39780.0
+% −84190.5
+% 85203.9
+% −48225.9
+% 15598.5
+% −2694.76
+% 192.865
+    % \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
+    % 301C^6 - 26C^7,
 % 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
@@
 
-Visual binaries and subgiants were removed from the sample by applying cuts to
-the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
+Photometric binaries and subgiants were removed from the sample by applying
+cuts to the color-magnitude diagram (CMD), shown in figure
+\ref{fig:age_gradient}.
 A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
-approximate the division between single stars and visual binaries.
+approximate the division between single stars and photometrci binaries.
 All stars above this line were removed from the sample.
@@
     on a \gaia\ CMD.
-    We removed visual binaries and subgiants from the sample by excluding
+    We removed photometric binaries and subgiants from the sample by excluding
     stars above the dashed lines.

### 2020-02-27 paper/results.tex
comments from Jason

@@
 This result is in agreement with a recent study which found that a large
-fraction of visual binaries were rapid rotators, and the probability of a star
-being a synchronized binary system substantially increased below rotation
+fraction of photometric binaries were rapid rotators, and the probability of a
+star being a synchronized binary system substantially increased below rotation
 periods of around 7 days \citep{simonian2019}.
@@
 4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
-We also eliminated visual binaries by removing stars above a 6th order
+We also eliminated photometric binaries by removing stars above a 6th order
 polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in

### 2020-02-27 code/data.py
Jason's new coefficients

@@ -188,4 +188,4 @@ gaia_mc["v_dec"] = vdec.value
 gaia_mc["v_b"] = v_b
 
 print("Saving file")
-gaia_mc.to_csv("gaia_mc3.csv")
+gaia_mc.to_csv("gaia_mc4.csv")

### 2020-02-27 code/photometric_teff.py

@@ -16,7 +16,12 @@ def bprp_to_teff(bprp):
 
     """
 
-    coeffs = [8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
-            -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
-            301.13205924630165, -25.923997443169355]
+    # coeffs = [8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
+    #         -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
+    #         301.13205924630165, -25.923997443169355]
+
+    # Jason's updated parameters:
+    coeffs = [-416.585, 39780.0, -84190.5, 85203.9, -48225.9, 15598.5,
+              -2694.76, 192.865]
+
     return np.polyval(coeffs[::-1], bprp)

### 2020-03-04 paper/abstract.tex
moving notebooks to figure_codez and a couple of Jason's comments

@@
 The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
-billion years due to a lack of precise ages and rotation periods for old
-main-sequence stars.
+Gyr due to a lack of precise ages and rotation periods for old main-sequence
+stars.
 % However, recent evidence suggests that the relationship between rotation
@@
 % to the Praesepe cluster, accurately predicts the relative ages of stars.
-We find that the period-\teff\ relationship of the Praesepe cluster does not
-apply to stars older than around 1 Gyr.
+We find that a gyrochronology model, calibrated to fit the period-\teff\
+relationship of the Praesepe cluster, does not apply to stars older than
+around 1 Gyr.
 % : a polynomial gyrochronology model, fit

### 2020-03-04 paper/intro.tex
Elisabeth's comments and some of Gus's
moving notebooks to figure_codez and a couple of Jason's comments

@@
 generate stronger magnetic fields and more efficient magnetic braking.
-However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
-same} rate as early-K dwarfs \citep{curtis2019}.
+However, in the 1.1 Gyr NGC 6811 cluster \citep{janes2011}, late-K dwarfs
+rotate at the {\it same} rate as early-K dwarfs \citep{curtis2019}.
 In other words, convection zone depth cannot be the only variable that affects
@@
 velocities.
+In section \ref{sec:results} we use kinematics to investigate the
+relationship between rotation period, age and color/\teff\ in the field and
+interpret our results.
 We also establish that \vb\ velocity dispersion, \sigmavb, can be used as an
@@
 selection function is observed to strongly affect our sample.
-In section \ref{sec:results} we use kinematics to investigate the
-relationship between rotation period, age and color/\teff\ in the field and
-interpret our results.
+Finally, we examine the rotation period gap and the kinematics of synchronized
+binaries in section \ref{sec:results}.
 % We show that the period-color relation of Praesepe is not applicable to

### 2020-03-04 paper/method.tex
moving notebooks to figure_codez and a couple of Jason's comments

@@
 rotation periods at old ages, the latter scenario seems more likely.
+An alternative explanation could be that the rotation periods of the oldest
+early M dwarfs are so long that they are not measurable with Kepler data.
+Ground-based rotation period measurements of mid and late M dwarfs
+indicate that there is an upper limit to the rotation periods of M
+dwarfs of around 140 days \citep{newton2017}, which is much longer than the
+longest rotation periods measured in the \mct\ sample (around 70 days).
+For this reason, it is likely that slowly rotating M dwarfs are missing from
+the \mct\ sample and this could be another reason for the lack of old M
+dwarfs.
+
 The \citet{angus2019} gyrochronology relation is a simple polynomial model,

### 2020-03-09 code/data.py
jason's comments and exploring point-by-point

@@ -36,6 +36,7 @@ with fits.open("../kepler_dr2_1arcsec.fits") as data:
     gaia = pd.DataFrame(data[1].data, dtype="float64")
 
 gaia_mc = pd.merge(mc, gaia, on="kepid", how="left")
+print(len(gaia_mc), "stars")
 
 # S/N cuts
 sn = gaia_mc.parallax.values/gaia_mc.parallax_error.values
@@ -43,6 +44,17 @@ sn = gaia_mc.parallax.values/gaia_mc.parallax_error.values
 m = (sn > 10)
 m &= (gaia_mc.parallax.values > 0) * np.isfinite(gaia_mc.parallax.values)
 m &= gaia_mc.astrometric_excess_noise.values < 5
+print(len(gaia_mc.iloc[m]), "stars after S/N cuts")
+
+# Jason's wide binary cuts
+# m &= gaia_mc.astrometric_excess_noise.values > 0
+# m &= gaia_mc.astrometric_excess_noise_sig.values > 6
+
+# Jason's short-period binary cuts
+# m &= radial_velocity_error < 4
+# print(len(gaia_mc.iloc[m]), "stars after Jason's binary cuts")
+# assert 0
+
 gaia_mc = gaia_mc.iloc[m]
 
 print("Loading Dustmaps")
@@ -57,6 +69,7 @@ ebv, flags = bayestar(coords, mode='percentile', pct=[16., 50., 84.],
 
 # Calculate Av
 Av_bayestar = 2.742 * ebv
+print(np.shape(Av_bayestar), "shape")
 Av = Av_bayestar[:, 1]
 Av_errm = Av - Av_bayestar[:, 0]
 Av_errp = Av_bayestar[:, 2] - Av
@@ -111,6 +124,12 @@ m &= (y < np.polyval(w, x) - extra) + (subcut > y)
 flag = np.zeros(len(gaia_mc))
 flag[~m] = np.ones(len(flag[~m]))
 gaia_mc["flag"] = flag
+
+test = gaia_mc.iloc[gaia_mc.flag.values == 1]
+plt.plot(gaia_mc.bp_dered - gaia_mc.rp_dered, gaia_mc.abs_G, ".", alpha=.1)
+plt.plot(test.bp_dered - test.rp_dered, test.abs_G, ".", alpha=.1)
+plt.ylim(10, 1)
+plt.savefig("test")
 
[…]

### 2020-03-09 code/dispersion.py

@@ -1,6 +1,7 @@
 import numpy as np
 from tqdm import trange
 import scipy.stats as sps
+import astropy.stats as aps
 import matplotlib.pyplot as plt
 
 
@@ -288,3 +289,44 @@ def err_on_sample_std_dev(std_dev_of_distribution, n):
 
     sig = std_dev_of_distribution
     return 1./(2*sig) * np.sqrt(2*sig**4/(n-1))
+
+# Calculate tangential distance.
+def tan_dist(x1, y1, x2, y2):
+    return np.sqrt((x1 - x2)**2 + (y1 - y2)**2)
+
+
+# Find N nearest points
+def n_nearest_points(x1, y1, x2, y2, z2, N):
+    td = tan_dist(x1, y1, x2, y2)
+    inds = np.argsort(td)
+    return x2[inds[:N]], y2[inds[:N]], z2[inds[:N]]
+
+
+def make_bin(x1, y1, x2, y2, z2, xrange, yrange):
+    dx, dy = xrange/2., yrange/2.
+    xlower, xupper = x1 - dx, x1 + dx
+    ylower, yupper = y1 - dy, y1 + dy
+
+    m = (xlower < x2) * (x2 < xupper) * (ylower < y2) * (y2 < yupper)
+    return x2[m], y2[m], z2[m]
+
+
+# Run on each star
+def calc_dispersion_nearest(x, y, z, N):
+
+    dispersions = np.zeros(len(x))
+    for i in trange(len(x)):
+        nx, ny, nz = n_nearest_points(x[i], y[i], x, y, z, N)
+        dispersions[i] = 1.5*aps.median_absolute_deviation(nz, ignore_nan=True)
+
+    return dispersions
+
+
+def calc_dispersion_bins(x, y, z, xrange, yrange):
+
+    dispersions = np.zeros(len(x))
+    for i in trange(len(x)):
+        nx, ny, nz = make_bin(x[i], y[i], x, y, z, xrange, yrange)
+        dispersions[i] = 1.5*aps.median_absolute_deviation(nz, ignore_nan=True)
+
+    return dispersions

### 2020-03-09 code/test_dispersion.py

@@ -113,6 +113,39 @@ def test_err_to_log10_err():
     assert np.isclose(10**(np.log10(value) + err_to_log10_err(value, err)),
                       value + err, atol=.01*value)
 
+
+def test_tan_dist():
+    x1, y1 = 1, 1
+    x2, y2 = 2, 2
+    assert tan_dist(x1, y1, x2, y2) == np.sqrt(2)
+
+
+def test_n_nearest_points():
+    x1, y1 = 10, 12
+    np.random.seed(42)
+    x2, y2 = [np.random.randn(1000) + 10 for i in range(2)]
+    z2 = np.random.randn(1000)*y2
+    nx, ny, nz = n_nearest_points(x1, y1, x2, y2, z2, 50)
+
+
+def test_make_bin():
+    np.random.seed(42)
+    x, y, z = [np.random.randn(1000) + 10 for i in range(3)]
+    bx, by, bz = make_bin(10, 10, x, y, z, 1, 1)
+    plt.plot(x, y, ".")
+    plt.plot(bx, by, ".")
+
+
+def test_calc_dispersion():
+    x2, y2 = [np.random.randn(1000) + 10 for i in range(2)]
+    z2 = np.random.randn(1000)*y2
+
+    dispersions_nearest = calc_dispersion_nearest(x2, y2, z2, 100);
+    dispersions_bins = calc_dispersion_bins(x2, y2, z2, .5, .5);
+
+    return dispersions_nearest, dispersions_bins
+
+
 if __name__ == "__main__":
     # test_dispersion()
     # test_MC()
@@ -120,3 +153,9 @@ if __name__ == "__main__":
     # test_select_stars()
     # test_fit_line()
     test_err_to_log10_err()
+    test_tan_dist()
+    test_n_nearest_points()
+    test_make_bin()
+    test_calc_dispersion()
+
+

### 2020-03-09 kinematics_and_rotation/kinematics.py
adding extra bits

@@ -1,0 +1,158 @@
+"""
+Calculate velocities and vertical actions
+"""
+
+
+import astropy.units as u
+from astropy.coordinates import ICRS
+from astropy.coordinates import Galactic
+import astropy.coordinates as coord
+import numpy as np
+
+
+def calc_vb(mean_list):
+    """
+    Calculate latitudinal velocity.
+
+    Args:
+        mean_list (list): list of arrays of,
+            ra: Right Ascension in degrees.
+            dec: Declination in degrees.
+            parallax: parallax in milliarcseconds.
+            pmra: RA proper motion in milliarcseconds per year.
+            pmdec: Dec proper motion in milliarcseconds per year.
+
+    Returns:
+        The array of latitudinal velocities.
+
+    """
+    ra, dec, parallax, pmra, pmdec = mean_list
+
+#     icrs = ICRS(ra=ra*u.degree,
+#                 dec=dec*u.degree,
+#                 distance=distance*u.pc,
+#                 pm_ra_cosdec=pmra*u.mas/u.yr,
+#                 pm_dec=pmdec*u.mas/u.yr)
+#     vels = icrs.transform_to(Galactic)
+
+    d = coord.Distance(parallax=parallax*u.mas)
+    vra = (pmra*u.mas/u.yr * d).to(u.km/u.s, u.dimensionless_angles())
+    vdec = (pmdec*u.mas/u.yr * d).to(u.km/u.s, u.dimensionless_angles())
+
+    c = coord.SkyCoord(ra=ra*u.deg, dec=dec*u.deg, distance=d,
+                       pm_ra_cosdec=pmra*u.mas/u.yr,
+                       pm_dec=pmdec*u.mas/u.yr)
+    gal = c.galactic
+    v_b = (gal.pm_b * gal.distance).to(u.km/u.s, u.dimensionless_angles())
+
+    return v_b
+
+
+def vb_with_err(mean_list, cov_list, Nsamples):
+    """
+    Calculate latitudinal velocities with uncertainties.
+
+    Args:
+        mean_list (list): A list of arrays of astrometric data.
+            [ra, dec, plx, pmra, pmdec]
+        cov_list (array): A list of all the uncertainties and covariances:
[…]

### 2020-03-09 kinematics_and_rotation/photometric_teff.py

@@ -1,0 +1,22 @@
+# Calculate photometric Teff from Gaia color
+# This should be dereddened colour!
+
+import numpy as np
+
+
+def bprp_to_teff(bprp):
+    """
+    Calculate photometric Teff from Gaia color (use dereddened color!)
+
+    Args:
+        bprp (array): Gaia G_BP colour minus Gaia G_RP colour.
+
+    Returns:
+        teffs (array): Photometric effective temperatures.
+
+    """
+
+    coeffs = [8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
+            -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
+            301.13205924630165, -25.923997443169355]
+    return np.polyval(coeffs[::-1], bprp)

### 2020-03-09 kinematics_and_rotation/teff_bv.py

@@ -1,0 +1,16 @@
+# Convert Teffs to B-V colours, or vice-versa
+# using the sekiguchi and fukugita conversion.
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+def teff2bv(teff, logg, feh):
+    # best fit parameters
+    t = [-813.3175, 684.4585, -189.923, 17.40875]
+    f = [1.2136, 0.0209]
+    d1 = -0.294
+    g1 = -1.166
+    e1 = 0.3125
+    return t[0] + t[1]*np.log10(teff) + t[2]*(np.log10(teff))**2 + \
+            t[3]*(np.log10(teff))**3 + f[0]*feh + f[1]*feh**2 \
+            + d1*feh*np.log10(teff) + g1*logg + e1*logg*np.log10(teff)

### 2020-03-09 kinematics_and_rotation/test_kinematics.py

@@ -1,0 +1,67 @@
+import kinematics as k
+import numpy as np
+
+
+def test_construct_and_sample_from_cov():
+    N = 2
+    ra_err, dec_err, plx_err = [np.ones(N)*.1 for i in range(3)]
+    pmra_err, pmdec_err = [np.ones(N)*.1 for i in range(2)]
+    ra_dec_corr, ra_plx_corr = np.ones(N)*.2, np.ones(N)*.3
+    ra_pmra_corr, ra_pmdec_corr = np.ones(N)*.4, np.ones(N)*.5
+    dec_plx_corr, dec_pmra_corr = np.ones(N)*.6, np.ones(N)*.7
+    dec_pmdec_corr = np.ones(N)*.8
+    plx_pmra_corr, plx_pmdec_corr = np.ones(N)*.9, np.ones(N)*.15
+    pmra_pmdec_corr = np.ones(N)*.25
+
+    cov_list = [ra_err, dec_err, plx_err, pmra_err, pmdec_err, ra_dec_corr,
+                ra_plx_corr, ra_pmra_corr, ra_pmdec_corr, dec_plx_corr,
+                dec_pmra_corr, dec_pmdec_corr, plx_pmra_corr, plx_pmdec_corr,
+                pmra_pmdec_corr]
+    cov1 = k.construct_cov(cov_list, 5)
+    assert np.shape(cov1) == (5, 5, N)
+
+    # cov = np.zeros((5, 5, len(ra_err)))
+    # cov[0, 0, :] = ra_err**2
+    # cov[0, 1, :] = ra_dec_corr
+    # cov[0, 2, :] = ra_plx_corr
+    # cov[0, 3, :] = ra_pmra_corr
+    # cov[0, 4, :] = ra_pmdec_corr
+
+    # cov[1, 0, :] = ra_dec_corr
+    # cov[1, 1, :] = dec_err**2
+    # cov[1, 2, :] = dec_plx_corr
+    # cov[1, 3, :] = dec_pmra_corr
+    # cov[1, 4, :] = dec_pmdec_corr
+
+    # cov[2, 0, :] = ra_plx_corr
+    # cov[2, 1, :] = dec_plx_corr
+    # cov[2, 2, :] = plx_err**2
+    # cov[2, 3, :] = plx_pmra_corr
+    # cov[2, 4, :] = plx_pmdec_corr
+
+    # cov[3, 0, :] = ra_pmra_corr
+    # cov[3, 1, :] = dec_pmra_corr
[…]

### 2020-03-09 paper/results.tex
comments about mass plot

@@
     Bottom: Same as top panel with rotation period vs {\it mass}
-    \citep[from the Kepler Input Catalog][]{brown2011}.
+    \citep[from][]{berger2020}.
     White lines show gyrochrones from a model that includes mass and
@@
 The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
-function of rotation period and {\it mass}, obtained from the Kepler Input
-Catalog \citep{brown2011}, with gyrochrones from the \citep{spada2019}
-gyrochronology model shown as white lines.
+function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
+gyrochrones from the \citep{spada2019} gyrochronology model shown as white
+lines.
+Each point is assigned the same color as in the top panel, so bin outlines
+appear smeared-out because temperature does not convert directly to mass.
 These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
-Although these models do not trace lines of constant velocity dispersion at
-0.5 and 1 Gyr, they do appear to reproduce trends in the data at ages of 1.5
-Gyr and older.
-The \citet{spada2019} models appear to qualitatively agree with the data:
-their rotation period-\teff\ relation flattens out over time and eventually
-inverts.
+The \citet{spada2019} models trace lines of constant velocity dispersion, and
+reproduce the trends in the data at all ages.
+These models qualitatively agree with the data: the rotation period-\teff/mass
+relation flattens out over time and eventually inverts.
 

### 2020-03-10 paper/intro.tex
some small edits

@@
 pinsonneault1989}.
-Although stars are born with random rotation periods, from 1 to 10 days,
+Although stars are born with random rotation periods, between 1 and 10 days,
 observations of young open clusters reveal that their rotation periods
@@
 For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
-spin more slowly than the G dwarfs.
-In theory this is because lower-mass stars have deeper convections zones which
-generate stronger magnetic fields and more efficient magnetic braking.
-However, in the 1.1 Gyr NGC 6811 cluster \citep{janes2011}, late-K dwarfs
-rotate at the {\it same} rate as early-K dwarfs \citep{curtis2019}.
+spin more slowly than the G dwarfs; in theory because lower-mass stars have
+deeper convections zones which generate stronger magnetic fields and more
+efficient magnetic braking.
+However, in the NGC 6811 cluster which is around 1.1 Gyr \citep{janes2011},
+late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
+\citep{curtis2019}.
 In other words, convection zone depth cannot be the only variable that affects
@@
 evolution in the MW]{sellwood2014}.
-Although the velocity of any individual star will only provide a weak age
-constraint because velocity is dependent on the position of a star in its
-orbit, the velocity {\it dispersion} of a {\it group} of stars can, in
-principle, average over this effect and indicate whether that group is old or
-young relative to other groups.
+The velocity of any individual star will only provide a weak age constraint
+because its velocity depends on its position in orbit, however the velocity
+{\it dispersion} of a {\it group} of stars will indicate whether that group is
+old or young relative to other groups.
 Vertical velocity dispersion is thought to increase monotonically with age.

### 2020-03-10 paper/method.tex
some small edits

@@
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-Curtis \etal\ (2020, in prep) combined \teff s for nearby, unreddened field
-stars from benchmark samples, including FGK stars characterized with
-high-resolution optical spectroscopy \citep{brewer2016}, M dwarfs
-characterized with low-resolution optical and near-infrared spectroscopy
-\citep{mann2015}, and K and M dwarfs characterized with interferometry and
-bolometric flux analyses \citep{boyajian2012}.
+Curtis \etal\ (2020, in prep) combined effective temperature measurements for
+nearby, unreddened field stars in benchmark samples, including FGK stars
+characterized with high-resolution optical spectroscopy \citep{brewer2016}, M
+dwarfs characterized with low-resolution optical and near-infrared
+spectroscopy \citep{mann2015}, and K and M dwarfs characterized with
+interferometry and bolometric flux analyses \citep{boyajian2012}.
 This empirical color--temperature relation is valid over the color range $0.55
@@
 3070$~K.
-The dispersion about the relation implies a high precision of 50~K. Using
-these benchmark data enable us to accurately estimate \teff\ for cool dwarfs
-\citep[\eg][]{rabus2019}, and allows to correct for interstellar reddening at
-all temperatures\footnote{The color--temperature relation is described in
+The dispersion about the relation implies a high precision of 50~K.
+These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
+\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
+at all temperatures\footnote{The color--temperature relation is described in
 detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
 \etal\ (2020, in prep).}.
-The equation we used to calculate photometric temperatures is,
-\begin{equation}
-    \mathrm{T_{eff}} = -416.585 + 39780.0C -84190.5C^2 + 85203.9C^3
-    -48225.9C^4 + 15598.5C^5 -2694.76C^6 + 192.865C^7,
-% Jason's updated parameters:
-% −416.585
-% 39780.0
-% −84190.5
-% 85203.9
-% −48225.9
-% 15598.5
-% −2694.76
-% 192.865
-    % \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
-    % 301C^6 - 26C^7,
-% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
-%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
-%           301.13205924630165, -25.923997443169355]
-\label{eq:curtis}
-\end{equation}
-where C is \gaia\ \gcolor.
+The equation we used to calculate photometric temperatures from Gaia \gcolor\
+color is a seventh-order polynomial with coefficients given in table
+\ref{tab:coeffs}.
+% \ref{tab:coeffs}.
+% \begin{equation}
+%     \mathrm{T_{eff}} = a + bC cC^2 + dC^3
+%     eC^4 + 15598.5C^5 -2694.76C^6 + 192.865C^7,
+% % Jason's updated parameters:
+% % −416.585
+% % 39780.0
+% % −84190.5
+% % 85203.9
+% % −48225.9
+% % 15598.5
+% % −2694.76
+% % 192.865
+%     % \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
+%     % 301C^6 - 26C^7,
+% % 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
+% %           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
+% %           301.13205924630165, -25.923997443169355]
+% \label{eq:curtis}
+% \end{equation}
+% where C is \gaia\ \gcolor and the coefficients are provided in table
+% \ref{coeffs}.
+\begin{table}[h!]
+  \begin{center}
+      \caption{
+          Coefficient values for a 7th-order polynomial used to estimate
+      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
+      prep).}
+    \label{tab:coeffs}
+    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
+      Exponent & Coefficient  \\
+      \hline
+      $0$ & -416.585 \\
+      $1$ & 39780.0  \\
+      $2$ & -84190.5 \\
+      $3$ & 85203.9  \\
+      $4$ & -48225.9 \\
+      $5$ & 15598.5  \\
+      $6$ & -2694.76 \\
+      $7$ & 192.865  \\
+    \end{tabular}
+  \end{center}
+\end{table}
 
@@
 All stars above this line were removed from the sample.
-Subgiants were also removed by eliminating stars brighter than 6th magnitude
+Subgiants were also removed by eliminating stars brighter than 4th magnitude
 in \gaia\ G-band.
@@
 early M dwarfs are so long that they are not measurable with Kepler data.
-Ground-based rotation period measurements of mid and late M dwarfs
-indicate that there is an upper limit to the rotation periods of M
-dwarfs of around 140 days \citep{newton2017}, which is much longer than the
+Ground-based rotation period measurements of mid and late M dwarfs indicate
+that there is an upper limit to the rotation periods of M dwarfs of around 140
+days \citep{newton2016, newton2017, newton2018}, which is much longer than the
 longest rotation periods measured in the \mct\ sample (around 70 days).

### 2020-03-10 paper/results.tex
some small edits

@@
 same without sigma-clipping.
-We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
-3500 K to avoid biases caused by the selection function at the faint, cool
-end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
+% We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
+% 3500 K to avoid biases caused by the selection function at the faint, cool
+% end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
 Finally, we removed stars with rotation periods shorter than the bulk of
@@
     These gyrochrones do not appear to reflect the evolution of field stars at
-    long rotation periods/old ages: they do not trace lines of constant
+    long rotation periods/old ages because they do not trace lines of constant
     velocity dispersion.
@@
 rotation period across all temperatures, implying that rotation period
-increases with age as expected.
+increases with age, as expected.
 This result is insensitive to the choice of bin position and size.
@@
 The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
-If these were true isochrones however, they should follow lines of constant
-velocity dispersion.
+If these were true isochrones they would follow lines of constant velocity
+dispersion.
 At ages older than around 1 Gyr, it appears that gyrochrones should have a more
@@
 function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
-gyrochrones from the \citep{spada2019} gyrochronology model shown as white
-lines.
-Each point is assigned the same color as in the top panel, so bin outlines
-appear smeared-out because temperature does not convert directly to mass.
+gyrochrones from the \citep{spada2019} model shown in white.
 These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
-The \citet{spada2019} models trace lines of constant velocity dispersion, and
-reproduce the trends in the data at all ages.
-These models qualitatively agree with the data: the rotation period-\teff/mass
-relation flattens out over time and eventually inverts.
+Each point plotted in the top panel also appears in the bottom panel with the
+same color.
+Because velocity dispersion was calculated in bins of \teff, not mass, bin
+outlines are clearly visible in the top panel but appear smeared-out in the
+bottom panel.
+In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
+{\it do} trace lines of constant velocity dispersion, and reproduce the trends
+in the data at all ages.
+These models qualitatively agree with the data and reproduces the apparent
+flattening and inversion in the rotation period-\teff/mass relations.
 

### 2020-03-10 paper/intro.tex
more small edits

@@
 evolution in the MW]{sellwood2014}.
-The velocity of any individual star will only provide a weak age constraint
-because its velocity depends on its position in orbit, however the velocity
-{\it dispersion} of a {\it group} of stars will indicate whether that group is
-old or young relative to other groups.
-Vertical velocity dispersion is thought to increase monotonically with age.
+The velocity of any individual star will only provide a weak age constraint,
+because its velocity depends on its current position in orbit, however the
+velocity {\it dispersion} of a {\it group} of stars will indicate whether that
+group is old or young relative to other groups.
+% Vertical velocity dispersion is thought to increase monotonically with age.
 In this work we compare the velocity dispersions of groups of field stars in
-the Galactic thin disk to ascertain which groups are older and which younger
-and draw conclusions based on the implied relative ages.
+the Galactic thin disk, to ascertain which groups are older and which younger,
+and draw conclusions about the rotational evolution of stars based on their
+implied relative ages.
 % The age-velocity dispersion reations (AVRs) are still actively being
@@
 indicator \citep{beane2018, ting2019}.
-Unfortunately most field stars with measured rotation periods do not have
+Unfortunately, most field stars with measured rotation periods do not have
 radial velocity (RV) measurements because they are relatively faint \kepler\
@@
 \ref{sec:method}) approximation to \vz.
-Because we use \vb\ rather than \vz\, we cannot calculate absolute kinematic
-ages using a published age-velocity dispersion relation (AVR), but this could
-be addressed in follow-up work that also takes into account the Kepler
-selection function.
-However, regardless of direction, velocity dispersion is expected to
+Because we use \vb\ rather than \vz\, we do not calculate absolute kinematic
+ages using a published age-velocity dispersion relation (AVR), calibrated with
+vertical velocity, in this work.
+In future it may be possible to account for the differences between \vb\ and
+\vz, or marginalize over missing RV measurements and the \kepler\ selection
+function, in order to infer absolute ages.
+Regardless of direction however, velocity dispersion is expected to
 monotonically increase over time \citep[\eg][]{holmberg2009}, and can
@@
 velocities.
-In section \ref{sec:results} we use kinematics to investigate the
-relationship between rotation period, age and color/\teff\ in the field and
-interpret our results.
+In section \ref{sec:results} we use kinematics to investigate the relationship
+between stellar rotation period, age and color/\teff\ and interpret the
+results.
 We also establish that \vb\ velocity dispersion, \sigmavb, can be used as an
 age proxy by demonstrating that neither mass-dependent heating nor the
-selection function is observed to strongly affect our sample.
+\kepler/\gaia\ selection function is observed to strongly affect our sample.
 Finally, we examine the rotation period gap and the kinematics of synchronized

### 2020-03-10 paper/method.tex
more small edits

@@
 A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
-approximate the division between single stars and photometrci binaries.
+approximate the division between single stars and photometric binaries (shown
+as the curved dashed line in figure \ref{fig:age_gradient}).
 All stars above this line were removed from the sample.

### 2020-03-11 paper/method.tex
minor edits to middle sections

@@
       \caption{
-          Coefficient values for a 7th-order polynomial used to estimate
+          Coefficient values for the 7th-order polynomial used to estimate
       \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
@@
     \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
-      Exponent & Coefficient  \\
+        (\gcolor\ ) exponent & Coefficient  \\
       \hline
@@
 
-Photometric binaries and subgiants were removed from the sample by applying
-cuts to the color-magnitude diagram (CMD), shown in figure
+Photometric binaries and subgiants were removed from the \mct\ sample by
+applying cuts to the color-magnitude diagram (CMD), shown in figure
 \ref{fig:age_gradient}.
@@
 
-The dwarf stars in the \mct\ sample are shown on a \gaia\ color-magnitude
-diagram (CMD) in figure \ref{fig:age_gradient}.
+The rotation periods of the dwarf stars in the \mct\ sample are shown on a
+\gaia\ color-magnitude diagram (CMD) in the top panel of figure
+\ref{fig:age_gradient}.
 In the bottom panel, the stars are colored by their gyrochronal age,
@@
 Given that lower-mass stars stay active for longer than higher-mass stars
-\citep[\eg][]{west2011}, and are therefore more likely to have measurable
-rotation periods at old ages, the latter scenario seems more likely.
-An alternative explanation could be that the rotation periods of the oldest
-early M dwarfs are so long that they are not measurable with Kepler data.
+\citep[\eg][]{west2011, kiman2019}, and are therefore more likely to have
+measurable rotation periods at old ages, the latter scenario seems likely.
+However, it is also possible that the rotation periods of the oldest early M
+dwarfs are so long that they are not measurable with Kepler data.
 Ground-based rotation period measurements of mid and late M dwarfs indicate
-that there is an upper limit to the rotation periods of M dwarfs of around 140
-days \citep{newton2016, newton2017, newton2018}, which is much longer than the
-longest rotation periods measured in the \mct\ sample (around 70 days).
-For this reason, it is likely that slowly rotating M dwarfs are missing from
-the \mct\ sample and this could be another reason for the lack of old M
-dwarfs.
-
+that there is an upper limit to the rotation periods of {\it late} M dwarfs of
+around 140 days \citep{newton2016, newton2017, newton2018}, which is much
+longer than the longest rotation periods measured in the \mct\ sample (around
+70 days).
+The apparent lack of old ages for M dwarfs in figure \ref{fig:age_gradient}
+may be caused by a combination of ages being underestimated by a poorly
+calibrated model, and rotation period detection bias.
+% For this reason, it is likely that slowly rotating M dwarfs are missing from
+% the \mct\ sample and this could be another reason for the lack of old M
+% dwarfs.
 The \citet{angus2019} gyrochronology relation is a simple polynomial model,
 fit to the period-color relation of Praesepe.
-Since this relation predicts that the oldest stars in the \mct\ sample are
-late-G and early-K dwarfs, it is probably under-predicting the ages of late-K
-and early-M dwarfs.
-This is a typical feature of empirically calibrated gyrochronology models:
-since there are no (or at least very few) old M dwarfs with rotation periods,
-the models are poorly calibrated for these stars.
+% Since this relation predicts that the oldest stars in the \mct\ sample are
+% late-G and early-K dwarfs, it is probably under-predicting the ages of late-K
+% and early-M dwarfs.
+Inaccuracies are a typical feature of empirically calibrated gyrochronology
+models: since there are no (or at least very few) old M dwarfs with rotation
+periods, the models are poorly calibrated for these stars.
 \begin{figure}
@@
 The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
-astropy2018} {\it Python} packages were used to calculate stellar velocities.
+astropy2018} {\it Python} packages were used to calculate velocities for the
+\mct\ sample.
 {\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
 covariance matrix via Monte Carlo sampling, thereby accounting for the
-ovariances between \gaia\ positions, parallaxes and proper motions.
+covariances between \gaia\ positions, parallaxes and proper motions.
 Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,

### 2020-03-11 paper/results.tex
minor edits to middle sections

@@
 
-To explore the relationship between rotation period, \teff\ and velocity
-dispersion, we first removed high and low velocity outliers from the \mct\
-sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
+To explore the relationship between rotation period, effective temperature
+(\teff ) and velocity dispersion, we first removed high and low velocity
+outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
+\vb\ velocities.
 Without sigma-clipping, we found that a small number of high velocity
 outliers at the low-temperature end of our sample substantially raised the
-velocity dispersion for cooler stars, however the overall trends remain the
-same without sigma-clipping.
+velocity dispersion for cooler stars, although this did not change the overall
+trends in the data.
 % We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
@@
 % end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
-Finally, we removed stars with rotation periods shorter than the bulk of
-periods, since this area of the period-\teff\ diagram is sparsely populated.
-We removed rapid rotators by cutting out stars with gyrochronal ages less than
-0.5 Gyr, because a 0.5 Gyr gyrochrone\footnote{A gyrochrone is a
-gyrochronological isochrone, or a line of constant age in period-\teff, or
-period-color space.} traces the bottom edge of the main population of rotation
-periods.
-After these cuts, 6879 stars were included in the sample.
+Finally, we removed stars with rotation periods shorter than the main
+population of periods, since this area of the period-\teff\ diagram is
+sparsely populated.
+We removed these rapid rotators by cutting out stars with gyrochronal ages
+less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
+Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
+line of constant age in period-\teff, or period-color space.} traces the
+bottom edge of the main population of rotation periods.
+After these cuts, around 13,000 stars out of the original $\sim$34,000 were
+included in the sample.
 
@@
 similar color in figure \ref{fig:vplot} are similar ages.
+We discuss this assumption further in section
+\ref{sec:mass-dependent-heating}.
 \begin{figure}
@@
 dispersion.
-At ages older than around 1 Gyr, it appears that gyrochrones should have a more
-flattened, or even inverted relationship between rotation period and effective
-temperature than these Praesepe-based models.
+At ages older than around 1 Gyr, it appears that gyrochrones should have a
+more flattened, or even inverted, shape in rotation period-\teff\ space than
+these Praesepe-based models.
 % For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
@@
 in the data at all ages.
-These models qualitatively agree with the data and reproduces the apparent
+These models qualitatively agree with the data and reproduce the apparent
 flattening and inversion in the rotation period-\teff/mass relations.
@@
 the main sequence and have transitioned from the `I' sequence to the `C'
-sequence ][]{barnes2003}, stellar rotation period decreases with increasing
-mass.
+sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
+increasing} mass.
 This is likely because lower-mass stars with deeper convection zones have
 stronger magnetic fields, larger Alfv\'en radii and therefore experience
-greater angular momentum loss rate.
+greater angular momentum loss rate \citep[\eg][]{schatzman1962, parker1970,
+kawaler1988, charbonneau2010}.
 According to the \citet{spada2019} model, there is minimal transportation of
@@
 so the surface slows down but the core keeps spinning rapidly.
-At intermediate ages, rotation period is constant with mass, and at late ages
-rotation period {\it increases} with mass for K-dwarfs.
+This aligns with the current assumptions about stellar spin-down, dynamo
+theory, and the gyrochronology paradigm that has been in place for decades
+\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
+According to the data presented in figure \ref{fig:vplot} however, at
+intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
+mass, and at late ages rotation period {\it increases} with {\it increaasing}
+mass.
 The interpretation of this, according to the \citet{spada2019} model, is that

### 2020-03-11 paper/abstract.tex
some small edits
hyphen to en-dash

@@
 % to the Praesepe cluster, accurately predicts the relative ages of stars.
-We find that a gyrochronology model, calibrated to fit the period-\teff\
+We find that a gyrochronology model, calibrated to fit the period--\teff\
 relationship of the Praesepe cluster, does not apply to stars older than
@@
 surfaces and stay magnetically active longer than more massive stars.
-Finally, based on their kinematics, we find that many rapidly rotating
+Finally, based on their kinematics, we confirm that many rapidly rotating
 K-dwarfs are likely to be synchronized binaries.

### 2020-03-11 paper/intro.tex
intro edits

@@
 spin more slowly than the G dwarfs; in theory because lower-mass stars have
-deeper convections zones which generate stronger magnetic fields and more
+deeper convective zones which generate stronger magnetic fields and more
 efficient magnetic braking.
@@
 redistribution in the interiors of stars are able to reproduce this flattened
-period-color relation \citep{spada2019}.
+period--color relation \citep{spada2019}.
 These models suggest that mass and age-dependent angular momentum transport
@@
 \citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
-On average, the vertical velocities of stars increase over time
+On average, the vertical velocities of older stars is observed to be larger
 \citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
 casagrande2011}.
-Although the cause of vertical dynamical heating is not well understood,
-interactions with giant molecular clouds, spiral arms and the galactic bar are
-thought to play an important role \citep[see][for a review of secular
-evolution in the MW]{sellwood2014}.
-The velocity of any individual star will only provide a weak age constraint,
-because its velocity depends on its current position in orbit, however the
-velocity {\it dispersion} of a {\it group} of stars will indicate whether that
+This is likely either a signature of dynamical heating, such as from
+interactions with giant molecular clouds, spiral arms and the galactic bar
+\citep[see][for a review of secular evolution in the MW]{sellwood2014}, or an
+indication that stars formed dynamically ``hotter'' in the past
+\citep[e.g.,][]{bird2013}.
+In either case, the vertical velocity distribution is observed to depend
+significantly on stellar age.
+While the velocity of any individual star only provides a weak age constraint,
+because its velocity depends on its current position in its orbit, the
+velocity {\it dispersion} of a {\it group} of stars indicates whether that
 group is old or young relative to other groups.
 % Vertical velocity dispersion is thought to increase monotonically with age.
-In this work we compare the velocity dispersions of groups of field stars in
-the Galactic thin disk, to ascertain which groups are older and which younger,
+In this work, we compare the velocity dispersions of groups of field stars in
+the Galactic thin disk to ascertain which groups are older and which younger,
 and draw conclusions about the rotational evolution of stars based on their
@@
 % vertical
-Although {\it vertical} velocity, \vz, is a well-established age proxy, it can
+Although {\it vertical} velocity, \vz, is an established age proxy, it can
 only be calculated with full 6-dimensional position and velocity information.
 In fact, with full 6D phase space and an assumed Galactic potential, it is
-possible to calculate vertical {\it action}, which may be an even better age
-indicator \citep{beane2018, ting2019}.
+possible to calculate the dynamically-invariant vertical {\it action}, which may
+be an even better age indicator \citep{beane2018, ting2019}.
 Unfortunately, most field stars with measured rotation periods do not have
@@
 For this reason, we used velocity in the direction of galactic latitude, \vb,
-to approximate \vz.
+as a proxy for \vz.
 The \kepler\ field is positioned at low galactic latitude
@@
 Because we use \vb\ rather than \vz\, we do not calculate absolute kinematic
-ages using a published age-velocity dispersion relation (AVR), calibrated with
-vertical velocity, in this work.
+ages using a published age--velocity dispersion relation (AVR), calibrated with
+vertical velocity.
 In future it may be possible to account for the differences between \vb\ and

### 2020-03-11 paper/conclusion.tex
rocio;s comments
hyphen to en-dash

@@
 Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
-\vz\ introduces an additional velocity dispersion, which increases with
+\vz\ introduces an additional pseudo-velocity dispersion, which increases with
 increasing Galactic latitude.
@@
 We found that the \citet{angus2019} relation, which is based on the
-period-color relation of the 650 Myr Praesepe cluster, does not correctly
-describe the period-age-\teff\ relation for old stars.
+period--color relation of the 650 Myr Praesepe cluster, does not correctly
+describe the period--age--\teff\ relation for old stars.
 Instead we found that, at young ages, rotation period is anti-correlated with
@@
 dwarfs of the same age.
-We showed that the period-\teff\ relations change shape over time in a way
+We showed that the period--\teff\ relations change shape over time in a way
 that qualitatively agrees with theoretical models which include a
@@
 
-% We outlined a number of scenarios which could provide alternative
 % explanations for these observations, including incorrect dust corrections for

### 2020-03-11 paper/results.tex
making some edits to the gap/binaries section

@@
 
-\subsection{The period gap and synchronized binaries}
-\label{sec:gap}
-
 There is a sharp gap in the population of rotation periods, which lies just
@@
 produces the gap.
+
+\subsection{Synchronized binaries and the \kepler\ period gap}
+\label{sec:gap}
 
@@
 
+In this section, we explored the kinematic properties of the \mct\ sample in
+more detail, investigating the velocity dispersions of stars either side of the
+\kepler\ period gap, and identifying rapidly rotating stars that may be
+synchronized binaries.
+
 The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
 stars in the \mct\ sample, with stars subdivided into three groups: those that
-rotate more quickly than the major rotation period distribution (green
+rotate more quickly than the major rotation period population (green
 triangles), those with rotation periods shorter than the gap (orange squares),
@@
 Stars were separated into these three groups using \citet{angus2019}
-gyrochronology models, according to the scheme shown in the legend.
+gyrochronology model, according to the scheme shown in the legend.
 Only stars cooler than 5000 K are included in the bottom panel in order to
@@
 rotation periods below and above the gap (orange squares to blue circles),
-suggesting that these groups are part of the same galactic population.
+suggesting that these groups are part of the same Galactic population.
 Previously, only the overall velocity dispersions of all stars above and below
@@
 to two distinct populations \citep{mcquillan2014}.
-On the other hand, the velocity dispersions of stars with rotation periods
-shorter than the lower edge of the rotation period distribution (green
-triangles) are not significantly smaller than the, presumed older,
-orange-colored stars.
+
+Synchronized binaries are pairs of stars whose rotation periods are equal to
+their orbital period.
+Since synchronization appears to happen at rotation periods of 7 days or
+shorter \citep{simonian2019}, and most isolated stars have rotation periods
+longer than 7 days, the rotation periods of synchronized binaries are likely
+to be {\it shorter} than they would be if they were isolated stars.
+In other words, their rotation periods do not reflect their ages.
+Synchronized binaries are therefore a source of contamination for
+gyrochronology and should be removed from samples before performing a
+gyrochronal age analysis.
+The gyrochronal age of a synchronized binary is likely to be much younger than
+the true age of the system.
+For this reason, stars that rotate rapidly but appear kinematically old are
+likely to be synchronized binaries.
+In general, the velocity dispersions of stars with rotation periods shorter
+than the lower edge of the rotation period distribution (green triangles in
+figure \ref{fig:gap}) are
+not significantly smaller than the, presumed older, orange-colored stars.
+In addition, some of the most rapidly rotating stars sample have relatively
+large velocities.
+The large velocity dispersions of the most rapidly rotating stars in the \mct\
+sample indicates that at least some fraction of these stars are likely to be
+synchronized binaries.
 % This is an indication that many of these are synchronized binaries.
@@
 % short rotation periods.
-The large velocity dispersions of the most rapidly rotating stars indicates
-that some fraction of these stars are old but rotating rapidly, and therefore
-likely to be synchronized binaries.
 % Given the strong correlation between binary fraction and primary mass
@@
 % rotating) sequence, it is likely that
-Figure \ref{fig:gap} indicates that there is an increased probability of stars
-with rotation periods less than $\sim$10 days being synchronized binaries.
+It general, there appears to be an increased probability of stars with
+rotation periods less than $\sim$10 days being synchronized binaries.
 This result is in agreement with a recent study which found that a large

### 2020-03-11 paper/conclusion.tex
incorporating Adrian's suggestions

@@
-\section{Conclusion}
+\section{Conclusions}
 
-In this paper, we demonstrated that the dispersion of velocities in the
-direction of galactic latitude, \vb, can be used as an age proxy, by showing
-that there is no strong evidence for mass-dependent heating in low-mass
-\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
-main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
-increase with decreasing mass.
+In this paper, we used the \vb\ velocity dispersions of stars in the \mct\
+catalog to explore the evolution of stellar rotation period as a function of
+effective temperature and age.
+Our conclusions are as follows:
+\begin{itemize}
+\item{{\bf Magnetic braking efficiency doesn't always increase with decreasing
+mass for K dwarfs.}
+% We found that the \citet{angus2019} relation, which is based on the
+% period--color relation of the 650 Myr Praesepe cluster, does not correctly
+% describe the period--age--\teff\ relation for old stars.
+Although at young ages, rotation period is anti-correlated with \teff\ (as
+seen in many young open clusters, include Praesepe), at intermediate ages the
+relation flattens out and K dwarfs of different masses rotate at the same
+rate.
+At old ages, cooler K dwarfs spin more rapidly than hotter K dwarfs of the
+same age.}
+
+\item{{\bf Variable core-envelope coupling may be the cause.} We showed that
+the period--\teff\ relations change shape over time in a way that
+qualitatively agrees with theoretical models which include mass and
+time-dependent core-envelope angular momentum transport \citep{spada2019}.}
+
+% The period-color/\teff\ relation seems to start flattening out after $\sim$1
+% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
+% older than the period gap which is located at a gyrochronal age of around
+% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
+\item{{\bf Low-mass stars stay active longer.}
+We found that the oldest stars in the \mct\ catalog are cooler than 4500
+K, which suggests that lower-mass stars remain active for longer, allowing
+their rotation periods to be measured at older ages.}
+
+\item{{\bf The \kepler\ period gap may be related to core-envelope coupling.}
+We speculated that the rotation period gap \citep{mcquillan2014} may separate
+a young regime where stellar rotation periods decrease with increasing mass
+from an old regime where periods increase with increasing mass, however more
+data are needed to provide a conclusive result.
+The velocity dispersions of stars increase smoothly across the rotation period
+gap, indicating that the gap does not separate two distinct stellar
+populations.}
+
+\item{{\bf Rapidly rotating stars with large absolute velocities may be
+synchronized binaries.}
+We used kinematics to indicate that there is a population of
+synchronized binaries with rotation periods less than around 10 days.}
+
+\item{{\bf \sigmavb\ can be used as an age proxy for \kepler\ stars.} In the
+    appendix of this paper we demonstrate that the dispersion of velocities in
+        the direction of Galactic latitude, \vb, can be used as an age proxy,
+        by showing that there is no strong evidence for mass-dependent heating
+        in low-mass \kepler\ dwarfs: the velocity dispersions of K and M
+        dwarfs, whose main-sequence lifetimes are longer than around 11 Gyrs,
+        do not appear to increase with decreasing mass.
 Although {\it vertical} velocity, \vz, is a quantity that has been
@@
 as a proxy for \vz.
-Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
-\vz\ introduces an additional pseudo-velocity dispersion, which increases with
-increasing Galactic latitude.
-For this reason we did not attempt to convert \vb\ dispersions into ages using
-an age-velocity dispersion relation.
+        Using stars in the GUMS simulation \citep{robin2012}, we showed that
+        using \vb\ as a proxy for \vz\ introduces an additional
+        pseudo-velocity dispersion, which increases with increasing Galactic
+        latitude.
+% For this reason we did not attempt to convert \vb\ dispersions into ages using
+% an age-velocity dispersion relation.
 However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
@@
 mass-dependent velocity dispersion bias into the sample.
-We therefore assumed that \vb\ velocity dispersion can be used to accurately
-rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
-on average, older than a group of stars with a small velocity dispersion.
+We therefore confirmed that \vb\ velocity dispersion can be used to accurately
+rank stars by age.}
+% , \ie\ a group of stars with a large velocity dispersion is,
+% on average, older than a group of stars with a small velocity dispersion.}
 
-% We found that old groups of cool dwarfs, selected to be coeval using the
-% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
-% dispersion across all temperatures.
-We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
-the evolution of stellar rotation period as a function of effective
-temperature and age.
-We found that the \citet{angus2019} relation, which is based on the
-period--color relation of the 650 Myr Praesepe cluster, does not correctly
-describe the period--age--\teff\ relation for old stars.
-Instead we found that, at young ages, rotation period is anti-correlated with
-\teff: cooler stars spin more slowly than hotter stars of the same age.
-However, at intermediate ages the relation flattens out and K dwarfs rotate at
-the same rate, regardless of mass.
-At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
-dwarfs of the same age.
-We showed that the period--\teff\ relations change shape over time in a way
-that qualitatively agrees with theoretical models which include a
-mass-dependent core-envelope angular momentum transport \citep{spada2019}.
-
-% The period-color/\teff\ relation seems to start flattening out after $\sim$1
-% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
-% older than the period gap which is located at a gyrochronal age of around
-% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
-We also found that the oldest stars in the \mct\ catalog are cooler than 4500
-K, which suggests that lower-mass stars remain active for longer, allowing
-their rotation periods to be measured at older ages.
-We speculate that the rotation period gap \citep{mcquillan2014} may separate
-a young regime where stellar rotation periods decrease with increasing mass
-from an old regime where periods increase with increasing mass, however more
-data are needed to provide a conclusive result.
-The velocity dispersions of stars increase smoothly across the rotation period
-gap, indicating that the gap does not separate two distinct populations.
-Finally, we used kinematics to indicate that there is a population of
-synchronized binaries with rotation periods less than around 10 days.
+\end{itemize}
 

### 2020-03-11 paper/intro.tex
incorporating Adrian's suggestions

@@
 The \kepler\ field is positioned at low galactic latitude
-(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see section
-\ref{sec:method}) approximation to \vz.
+(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see
+appendix) approximation to \vz.
 Because we use \vb\ rather than \vz\, we do not calculate absolute kinematic
@@
 results.
-We also establish that \vb\ velocity dispersion, \sigmavb, can be used as an
-age proxy by demonstrating that neither mass-dependent heating nor the
-\kepler/\gaia\ selection function is observed to strongly affect our sample.
-Finally, we examine the rotation period gap and the kinematics of synchronized
-binaries in section \ref{sec:results}.
+We also examine the rotation period gap and the kinematics of synchronized
+binaries.
+In the appendix, we establish that \vb\ velocity dispersion, \sigmavb, can be
+used as an age proxy by demonstrating that neither mass-dependent heating nor
+the \kepler/\gaia\ selection function is observed to strongly affect our
+sample.
 % We show that the period-color relation of Praesepe is not applicable to

### 2020-03-11 paper/method.tex
incorporating Adrian's suggestions

@@
 Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
-\citep{green2018}, and {\tt astropy} \citep{astropy2013}.
+\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.
 
@@
 15\degrees\ (justification provided below) were removed from the sample.
+Finally, we removed stars with rotation periods shorter than the main
+population of periods, since this area of the period-\teff\ diagram is
+sparsely populated.
+We removed these rapid rotators by cutting out stars with gyrochronal ages
+less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
+Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
+line of constant age in period-\teff, or period-color space.} traces the
+bottom edge of the main population of rotation periods.
+After these cuts, around 13,000 stars out of the original $\sim$34,000 were
+included in the sample.
+
 % Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,

### 2020-03-11 paper/results.tex
incorporating Adrian's suggestions

@@
 
+% To explore the relationship between rotation period, effective temperature
+% (\teff ) and velocity dispersion,
+% % , we first removed high and low velocity
+% % outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
+% % \vb\ velocities.
+% % Without sigma-clipping, we found that a small number of high velocity
+% % outliers at the low-temperature end of our sample substantially raised the
+% % velocity dispersion for cooler stars, although this did not change the overall
+% % trends in the data.
+% % We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
+% % 3500 K to avoid biases caused by the selection function at the faint, cool
+% % end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
+% we removed stars with rotation periods shorter than the main population of
+% periods, since this area of the period-\teff\ diagram is sparsely populated.
+% We removed these rapid rotators by cutting out stars with gyrochronal ages
+% less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
+% Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
+% line of constant age in period-\teff, or period-color space.} traces the
+% bottom edge of the main population of rotation periods.
+% After these cuts, around 13,000 stars out of the original $\sim$34,000 were
+% included in the sample.
+
 To explore the relationship between rotation period, effective temperature
-(\teff ) and velocity dispersion, we first removed high and low velocity
-outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
-\vb\ velocities.
-Without sigma-clipping, we found that a small number of high velocity
-outliers at the low-temperature end of our sample substantially raised the
-velocity dispersion for cooler stars, although this did not change the overall
-trends in the data.
-% We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
-% 3500 K to avoid biases caused by the selection function at the faint, cool
-% end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
-Finally, we removed stars with rotation periods shorter than the main
-population of periods, since this area of the period-\teff\ diagram is
-sparsely populated.
-We removed these rapid rotators by cutting out stars with gyrochronal ages
-less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
-Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
-line of constant age in period-\teff, or period-color space.} traces the
-bottom edge of the main population of rotation periods.
-After these cuts, around 13,000 stars out of the original $\sim$34,000 were
-included in the sample.
-
+(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb
+was calculated as 1.5$\times$ the median absolute deviation, to mitigate
+sensitivity to outliers.} for groups of stars with similar rotation periods
+and temperatures, and presumed similar age.
 The top panel of figure \ref{fig:vplot} shows rotation period versus effective
-temperature for the \mct\ sample, coloured by the standard deviation of their
-(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
-grid in $\log_{10}$(period) and temperature.
+temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
+calculated for groups of stars over a grid in $\log_{10}$(period) and
+temperature.
 If we assume that mass dependent heating does not strongly affect this sample
@@
 similar color in figure \ref{fig:vplot} are similar ages.
-We discuss this assumption further in section
-\ref{sec:mass-dependent-heating}.
+We discuss this assumption further in the appendix.
 \begin{figure}
@@
 \end{figure}
+
 Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
@@
 
-There is a sharp gap in the population of rotation periods, which lies just
-above the 1 Gyr gyrochrone in the upper panel of figure \ref{fig:vplot}, whose
-origin is unknown and is the subject of much speculation \citep{mcquillan2014,
-davenport2018, reinhold2019}.
-The Praesepe-based model appears to be valid below the gap but not above.
-Although this may be coincidental (and more data would be needed to confirm a
-connection) the gap may indeed separate a young regime where stellar cores are
-decoupled from their envelopes from an old regime where these layers are more
-tightly coupled.
-If so, this could indicate that the phenomenon responsible for changing the
-shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
-produces the gap.
-
 \subsection{Synchronized binaries and the \kepler\ period gap}
@@
   \caption{
-      Top: rotation period vs. effective temperate for stars in the \mct\
+      Top: rotation period vs. effective temperature for stars in the \mct\
     sample, separated into three groups. Blue circles
@@
 
+There is a sharp gap in the population of rotation periods (often called the
+\kepler\ period gap), which lies just above the 1 Gyr gyrochrone in the upper
+panel of figure \ref{fig:vplot}, whose origin is unknown and is the subject of
+much speculation \citep{mcquillan2014, davenport2018, reinhold2019}.
+This gap was first identified by \mct, and roughly follows a line of constant
+gyrochronal age of around 1.1 Gyr \citep[according to the][gyrochronology
+relation]{angus2019}.
+Several explanations for the gap's origin have been proposed, including a
+discontinuous star formation history \citep{mcquillan2013, davenport2017,
+davenport2018} and a change in magnetic field structure causing a brief period
+where rotational variability is reduced and rotation periods cannot be
+measured \citep{reinhold2019}.
+
+The top panel of figure \ref{fig:vplot} suggests that the \citet{angus2019},
+Praesepe-based gyrochronology model is valid below the gap but not above.
+Gyrochrones follow lines of constant velocity dispersion below the gap, but
+{\it cross} lines of constant velocity dispersion above the gap.
+Although we do not provide an in-depth analysis here (and more data may be
+needed to confirm a connection) these data suggest that the gap may indeed
+separate a young regime where stellar cores are decoupled from their envelopes
+from an old regime where these layers are more tightly coupled.
+If so, this could indicate that the phenomenon responsible for changing the
+shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
+produces the gap.
+
+An alternate explanation for the gap is that the \mct\ sample contains two
+distinct stellar populations: one young and one old.
+If so, the kinematic properties of stars above and below the gap are likely to
+be distinctly different.
 The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
 stars in the \mct\ sample, with stars subdivided into three groups: those that
-rotate more quickly than the major rotation period population (green
+rotate more quickly than the main rotation period population (green
 triangles), those with rotation periods shorter than the gap (orange squares),
@@
 to two distinct populations \citep{mcquillan2014}.
+The smooth increase in velocity dispersion across the gap shown here supports
+the hypothesis that the gap is caused by a phenomenon of magnetic or
+structural stellar evolution, not a discontinuity in the local star formation
+history.
 
@@
 to be {\it shorter} than they would be if they were isolated stars.
-In other words, their rotation periods do not reflect their ages.
+For this reason, their rotation periods do not reflect their ages and the
+gyrochronal age of a synchronized binary is likely to be much younger than the
+true age of the system.
 Synchronized binaries are therefore a source of contamination for
@@
 gyrochronal age analysis.
-The gyrochronal age of a synchronized binary is likely to be much younger than
-the true age of the system.
-For this reason, stars that rotate rapidly but appear kinematically old are
-likely to be synchronized binaries.
-In general, the velocity dispersions of stars with rotation periods shorter
-than the lower edge of the rotation period distribution (green triangles in
-figure \ref{fig:gap}) are
-not significantly smaller than the, presumed older, orange-colored stars.
-In addition, some of the most rapidly rotating stars sample have relatively
-large velocities.
-The large velocity dispersions of the most rapidly rotating stars in the \mct\
-sample indicates that at least some fraction of these stars are likely to be
-synchronized binaries.
+Figure \ref{fig:gap} shows that some of the most rapidly rotating stars in the
+\mct\ sample have relatively large absolute velocities, indicating that they
+are likely synchronized binaries.
+% The large velocity dispersions of the most rapidly rotating stars in the \mct\
+% sample indicates that at least some fraction of these stars are likely to be
+% synchronized binaries.
+For this reason, the velocity dispersions of stars with rotation periods
+shorter than the lower edge of the rotation period distribution (green
+triangles in figure \ref{fig:gap}) are not significantly smaller than the,
+presumed older, orange-colored stars.
 % This is an indication that many of these are synchronized binaries.
@@
 % rotating) sequence, it is likely that
-It general, there appears to be an increased probability of stars with
-rotation periods less than $\sim$10 days being synchronized binaries.
+In general, stars with rotation periods less than $\sim$10 days have an
+increased chance of being synchronized binaries.
 This result is in agreement with a recent study which found that a large
@@
 periods of around 7 days \citep{simonian2019}.
+We caution users of rotation period catalogs that rapid rotators with large
+absolute velocities should be flagged as potential synchronized binaries
+before applying any gyrochronal analysis.
 
-\subsection{Validating \vb\ dispersion as an age proxy}
-\label{sec:mass-dependent-heating}
+% \subsection{Validating \vb\ dispersion as an age proxy}
+% \label{sec:mass-dependent-heating}
 
-There are two main reasons why \vb\ velocity dispersion may not be a good age
-proxy.
-Firstly, mass-dependent heating may act on the sample, meaning that velocity
-dispersion depends on both age and mass, so cannot be interpreted as a simple
-age proxy.
-Secondly, since stars in the \kepler\ field have a range of galactic
-latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
-stars, and introduce a velocity bias for high latitude stars (which are more
-likely to be cooler and older).
-In this section we demonstrate that neither of these problems seem to be a
-significant issue for our data.
+% The conclusions drawn in this paper depend on the assumption that velocity
+% dispersion in the direction of Galactic latitude (\sigmavb) can be used as an
+% age proxy.
+% There are two main reasons however, why \vb\ velocity dispersion may {\it not}
+% be a good age proxy.
+% Firstly, mass-dependent heating may act on the sample, meaning that velocity
+% dispersion depends on both age and mass, so cannot be interpreted as a simple
+% age proxy.
+% Secondly, since stars in the \kepler\ field have a range of Galactic
+% latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
+% stars, and introduce a velocity bias for high latitude stars (which are more
+% likely to be cooler and older).
+% In this section we demonstrate that neither of these problems seem to be a
+% significant issue for our data.
 
-In order to establish whether \sigmavb\ can be used as an age proxy, we
-searched for signs of mass-dependent heating within the \kepler\ field.
-Mass-dependent dynamical heating may result from lower-mass stars experiencing
-greater velocity changes when gravitationally perturbed than more massive
-stars.
-It has not been unambiguously observed in the galactic disk because of the
-strong anti-correlation between stellar mass and stellar age.
-Less massive stars do indeed have larger velocity dispersions, however they
-are also older on average.
-This mass-age degeneracy is highly reduced in M dwarfs because their
-main-sequence lifetimes are longer than the age of the Universe, and no
-evidence for mass-dependent heating has previously been found in M dwarfs
-\citep[\eg][]{faherty2009, newton2016}.
+% In order to establish whether \sigmavb\ can be used as an age proxy, we
+% searched for signs of mass-dependent heating within the \kepler\ field.
+% Mass-dependent dynamical heating may result from lower-mass stars experiencing
+% greater velocity changes when gravitationally perturbed than more massive
+% stars.
+% It has not been unambiguously observed in the galactic disk because of the
+% strong anti-correlation between stellar mass and stellar age.
+% Less massive stars do indeed have larger velocity dispersions, however they
+% are also older on average.
+% This mass-age degeneracy is highly reduced in M dwarfs because their
+% main-sequence lifetimes are longer than the age of the Universe, and no
+% evidence for mass-dependent heating has previously been found in M dwarfs
+% \citep[\eg][]{faherty2009, newton2016}.
 
-To investigate whether mass-dependent heating could be acting on the \kepler\
-sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
-whose MS lifetimes exceed around 11 Gyr and are therefore representative of
-the initial mass function.
-We could not perform this analysis on the \mct\ sample, because it only
-includes stars with {\it detectable} rotation periods, and since lower-mass
-stars stay active for longer it is likely that it contains a strong mass-age
-correlation.
-We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
-greater than 1.2 (corresponding to an effective temperature $\lesssim$
-4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
-We also eliminated photometric binaries by removing stars above a 6th order
-polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
-figure \ref{fig:age_gradient}).
-We then applied the quality cuts described above in section
-\ref{sec:the_data}.
-To search for evidence of mass-dependent heating we calculated the (\vb)
-velocity dispersion of stars in effective temperature bins.
-Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
-outliers before calculating the standard deviation of stars in each bin.
-These extreme velocity outliers may be very old late K and M dwarfs, or they
-result from using \vb\ instead of \vz, which introduces additional velocity
-scatter.
+% To investigate whether mass-dependent heating could be acting on the \kepler\
+% sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
+% whose MS lifetimes exceed around 11 Gyr and are therefore representative of
+% the initial mass function.
+% We could not perform this analysis on the \mct\ sample, because it only
+% includes stars with {\it detectable} rotation periods, and since lower-mass
+% stars stay active for longer it is likely that it contains a strong mass-age
+% correlation.
+% We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
+% greater than 1.2 (corresponding to an effective temperature $\lesssim$
+% 4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+% We also eliminated photometric binaries by removing stars above a 6th order
+% polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
+% figure \ref{fig:age_gradient}).
+% We then applied the quality cuts described above in section
+% \ref{sec:the_data}.
+% To search for evidence of mass-dependent heating we calculated the (\vb)
+% velocity dispersion of stars in effective temperature bins.
+% Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
+% outliers before calculating the standard deviation of stars in each bin.
+% These extreme velocity outliers may be very old late K and M dwarfs, or they
+% result from using \vb\ instead of \vz, which introduces additional velocity
+% scatter.
 
-Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
-function of effective temperature \footnote{calculated by transforming
-dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
-\kepler\ dwarf sample.
-Velocity dispersion very slightly {\it decreases} with decreasing temperature,
-the opposite of the trend expected for mass-dependent heating, however the
-slope is only inconsistent with zero at 1.3 $\sigma$.
-This trend may be due to a selection bias: cooler stars are fainter and
-therefore typically closer, with smaller heights above the galactic plane and
-smaller velocities.
-The essential point however, is that we do not see evidence for mass-dependent
-heating acting on stars in the \kepler\ field, indicating that velocity
-dispersion {\it can} be used as an age proxy (with the caveat that there is
-still a chance, albeit a small one, that the opposing effects of the selection
-function and mass-dependent heating are working to cancel each other out).
-This analysis was performed using \vb\ but we also examined the {\it vertical}
-velocities of the 537 stars in this sample with RV measurements.
-Again, no evidence was found for mass-dependent heating: the slope of the
-velocity dispersion-temperature relation was consistent with zero.
-\begin{figure}
-  \caption{
-      Top: Stellar velocity (\vb) as a function of \teff\ for
-      \kepler\ K and M dwarfs.
-Vertical lines indicate different \teff-groupings used to calculate velocity
-    dispersion.
-Pink stars were not included in velocity dispersion calculations as they were
-    either removed as outliers during a sigma clipping process, or they lie at
-    the sparcely populated, extremely cool end of the temperature range.
-    Velocity dispersion and \teff\ are slightly positively correlated, likely
-    due to a brightness-related selection bias, indicating that mass-dependent
-    heating does not significantly affect low-mass stars in the \kepler\
-    field.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vb_vs_teff}
-\label{fig:vb_vs_teff}
-\end{figure}
-
-Having found no strong evidence for mass-dependent heating, we next tested
-the validity of \vb\ as a proxy for \vz\ in more detail.
-At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
-of $b$, this equivalence becomes an approximation that grows noisier with $b$.
-To test the validity of the \vb$\sim$\vz\ approximation over a range of
-latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
-(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
-We downloaded stars from four pointings in the \kepler\ field with galactic
-latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
-a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
-The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
-\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
-Galaxy.
-We subtracted this offset and examined the residuals of the \vz\ -- \vb\
-relationship to investigate the variance as a function of Galactic latitude
-(shown in figure \ref{fig:vb_vz}).
-We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
-with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
-The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
-$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
-20$^\circ$.
-
-% Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
-% over different Galactic latitudes.
-% Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
-% fits are shown as dashed blue lines.
-% The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
-% the variance increases with increasing Galactic latitude.
-\begin{figure}
-  \caption{
-This figure demonstrates the variance in the relationship between \vb\ and
-    \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
-The panels show a kernel density estimator (KDE) (black solid line) for
-    the \vz -- \vb\ residuals of stars in the GUMS simulation at four
-    different Galactic latitudes.
-Blue dashed lines show Gaussian fits to these KDEs.
-The distributions are close to Gaussian, with slightly heavy tails.
-The standard deviations of the Gaussian fits increase with Galactic latitude.
-This figure illustrates how using \vb\ instead of \vz\ artificially
-    increases velocity dispersion, especially at high latitudes.
-}
-  \centering
-    \includegraphics[width=1\textwidth]{vb_vz}
-\label{fig:vb_vz}
-\end{figure}
-
-Since we are concerned with velocity {\it dispersions}, rather than velocities
-themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
-function of temperature for
-stars downloaded from the GUMS simulation.
-For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
-consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
-latitudes the two quantities became significantly different.
-For this reason we proceeded by only including stars with galactic latitudes
-less than 15\degrees\ in our analysis.
-Although we find that the transformation between \vz\ and \vb\ does not {\it
-strongly} affect our results, we cannot rule out the possibility that it
-introduces systematic biases into the velocity dispersions we present here.
-In \gaia\ DR3, RVs will be available for most stars in this sample, providing
-an opportunity to validate (or correct) the results presented here, and to
-work in action-space, rather than velocity-space.
-
-% Since we use \vb, not \vz\ in our analysis, it is possible that the sample
-% selection function could influence our results.
-% For example, higher-mass stars tend to be younger and reside closer to the
-% galactic disk mid-plane (low galactic latitudes).
-% Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
-% lower-mass stars typically have greater velocity dispersions due to .
-% \teff\ and galactic latitude, $b$,
-% could result in
-% Stars at higher latitudes have additional velocity components in the ${\bf x}$
-% and ${\bf y}$ directions, which could increase \vb\ but not \vz.
-% Again however, since the relationship between \sigmavb\ and \teff\ is
-% positively, not negatively correlated for cool stars in the \kepler\ field,
-% this effect is probably too small to influence our results.
-
-% Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
-% strongly affect the \mct\ sample of rotating \kepler\ stars.
-% For this reason, we assume that age difference is the major cause of velocity
-% dispersion differences between groups.
-% In other words, (\vb) velocity dispersion is a reliable age proxy for the
-% \mct\ sample.
-
-Because of the noisy relationship between \vb\ and \vz\, in this paper we do
-not attempt to convert velocity dispersion (\sigmavb) into an age via an
-age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
-Although we find that \sigmavb\ can be used to rank groups of stars by age, a
-more careful analysis that includes formal modeling of the \vb\ -- \vz\
-relationship will be needed to calculate absolute ages.
-% For example, the velocity distributions could be modeled as a mixture of
-% Gaussians in order to account for the additional velocity dispersion caused by
-% the \vz-\vb\ transformation.
-% The RVs of most of these stars will become available in \Gaia\ DR3, allowing
-% calculations of \vz, which can be used to calculate more reliable ages via an
-% AVR and suitable modeling of the selection function.
-
-% We then selected groups of stars within different gyrochronal age ranges and
-% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
-% of effective temperature for each age group.
-% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-% throughout this paper we show rotation periods as a function of effective
-% temperature, \teff, because it is easier to divide stars into bins of roughly
-% equal numbers in \teff-space than in color-space.
-
-% We calculated the overall velocity dispersions of stars as a function of
-% rotation period and gyrochronal age.
-
-% Each star is assigned the same velocity dispersion as in figure
-% \ref{fig:vplot}, and because stars in the same temperature group are not
-% necessarily in the same mass group, which is why stars of the same mass and period do not nece
-
-% The shape of the period-\teff\ relations at old ages appears to follow the
-% shape of the upper detection edge.
-% The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
-% period catalog characterized by a dearth of slowly rotating M dwarfs between
-% $\sim$3750 and 4250 K, is reflected in the lines of constant velocity
-% dispersion (and presumed age) in the top panel of figure \ref{fig:vplot}.
-% If the shape of the upper edge of rotation period measurements is created by a
-% detection limit, it could indicate the rotation periods at which stars of
-% different temperatures become relatively inactive (and their rotation periods
-% therefore become undetectable).
-% % Once inactive, stars would have little photometric variability induced by
-% % star spots and their rotation periods would be difficult to measure, so may
-% % not feature in the \mct\ catalog.
-% If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
-% $\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.
-
-% The velocity dispersions of the coolest stars may be affected by a selection
-% bias -- these extremely faint stars are more difficult to detect at larger
-% distances, larger heights above the plane, and therefore larger velocities.
-% It is possible that some high velocity stars of temperatures cooler than
-% $\sim$4000 K are missing from this sample, and the velocity dispersions may
-% therefore appear lower than they truly are.
-% This could be why the velocity dispersion appears to decrease towards the
-% right of figure \ref{fig:dispersion_period_teff}.
-
+% Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
+% function of effective temperature for the K and M \kepler\ dwarf sample.
+% Velocity dispersion very slightly {\it decreases} with decreasing temperature,
+% the opposite of the trend expected for mass-dependent heating, however the
+% slope is only inconsistent with zero at 1.3 $\sigma$.
+% This trend may be due to a selection bias: cooler stars are fainter and
+% therefore typically closer, with smaller heights above the galactic plane and
+% smaller velocities.
+% The essential point however, is that we do not see evidence for mass-dependent
+% heating acting on stars in the \kepler\ field, indicating that velocity
+% dispersion {\it can} be used as an age proxy (with the caveat that there is
+% still a chance, albeit a small one, that the opposing effects of the selection
+% function and mass-dependent heating are working to cancel each other out).
+% This analysis was performed using \vb\ but we also examined the {\it vertical}
+% velocities of the 537 stars in this sample with RV measurements.
+% Again, no evidence was found for mass-dependent heating: the slope of the
+% velocity dispersion-temperature relation was consistent with zero.
 % \begin{figure}
 %   \caption{
-%     Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
-%     x-axis.
-%     Masses are from the \kepler\ input catalog.
-%     The white lines show the \citet{spada2019} rotational evolution models at
-%     0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
-%     period.
-%     These models include age and mass-dependent coupling between the stellar
-%     core and envelope.
-%     The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
-%     an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
-%     decreases with decreasing mass.
-%     A similar phenomenon is visible in the velocity dispersions of field stars
-%     shown as colored points in the background of this figure.
+%       Top: Stellar velocity (\vb) as a function of \teff\ for
+%       \kepler\ K and M dwarfs.
+% Vertical lines indicate different \teff-groupings used to calculate velocity
+%     dispersion.
+% Pink stars were not included in velocity dispersion calculations as they were
+%     either removed as outliers during a sigma clipping process, or they lie at
+%     the sparcely populated, extremely cool end of the temperature range.
+%     Velocity dispersion and \teff\ are slightly positively correlated, likely
+%     due to a brightness-related selection bias, indicating that mass-dependent
+%     heating does not significantly affect low-mass stars in the \kepler\
+%     field.
 % }
 %   \centering
-%     \includegraphics[width=1\textwidth]{masses_with_models}
-% \label{fig:masses_with_models}
+%     \includegraphics[width=1\textwidth]{vb_vs_teff}
+% \label{fig:vb_vs_teff}
 % \end{figure}
 
-% \subsection{The period gap}
-% \label{sec:period_gap}
+% Having found no strong evidence for mass-dependent heating, we next tested
+% the validity of \vb\ as a proxy for \vz\ in more detail.
+% At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
+% of $b$, this equivalence becomes an approximation that grows noisier with $b$.
+% To test the validity of the \vb$\sim$\vz\ approximation over a range of
+% latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
+% (GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
+% We downloaded stars from four pointings in the \kepler\ field with galactic
+% latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
+% a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
+% The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
+% \vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
+% Galaxy.
+% We subtracted this offset and examined the residuals of the \vz\ -- \vb\
+% relationship to investigate the variance as a function of Galactic latitude
+% (shown in figure \ref{fig:vb_vz}).
+% We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
+% with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
+% The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
+% $, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
+% 20$^\circ$.
+% This demonstrates that using \vb\ instead of \vz\ for stars in the \kepler\
+% field will introduce an additional velocity scatter, inflating \sigmavb\
+% relative to \sigmavz.
+% This additional velocity scatter will be greatest for stars at the highest
+% Galactic latitudes.
 
-% The origin of the rotation period gap, first identified
-% by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
-% \ref{fig:dispersion_period_teff} still remains a mystery.
-% This gap can be seen as an under-density of points between the 0.7-1.0 and
-% 1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
-% of constant gyrochronal age of around 1.1 Gyr \citep[according to the
-% gyrochronology relation of][]{angus2019}, as shown in figure
-% \ref{fig:dispersion_period_teff}.
-% Several explanations for the gap's origin have been proposed, including a
-% discontinuous star formation history \citep{mcquillan2013, davenport2017,
-% davenport2018}, a rapid change in magnetic field structure
-% \citep{reinhold2019}, and erroneous rotation period measurements that are
-% incorrect by a factor of two \citep{koen2018}.
-% The latter explanation can be ruled out because stars below the gap have
-% smaller velocity dispersions than the stars above the gap, indicating that
-% they are kinematically younger \citep{mcquillan2013, davenport2018}, as
-% evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
-% For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
-% \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
-% temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
-% velocity dispersion increases with \teff.
-% The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
-% dispersion as the hottest stars in the age range above which indicates that
-% the period-\teff\ relations are {\it flat} at these rotation periods.
-% This is also visible in figure \ref{fig:dispersion_period_teff}.
-% Below the gap, velocity dispersion within a given period range appears to {\it
-% decrease} with decreasing temperature.
-% The opposite appears to be true above the gap.
-% It could be that the gap is positioned at a significant Rossby number/age at
-% which stellar magnetic dynamos go through a transition.
-% Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
-% magnetic braking is more efficient for stars with deeper convection zones.
-% Once stars reach this critical age or Rossby number their magnetic fields
-% undergo some radical transition, which produces the gap in the rotation
-% period-\teff\ plane.
-% After this transition, magnetic braking efficiency no longer increases with
-% decreasing mass.
-% Of course, it may be a coincidence that the gyrochronology relations seem to
-% only flatten off above the period gap and we lack a sufficient quantity of
-% data to do more than speculate here.
-% New rotation periods from the \ktwo\ and \tess\ missions may be able to
-% validate or rule out this hypothesis in the future.
+% % Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
+% % over different Galactic latitudes.
+% % Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
+% % fits are shown as dashed blue lines.
+% % The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
+% % the variance increases with increasing Galactic latitude.
+% \begin{figure}
+%   \caption{
+% This figure demonstrates the variance in the relationship between \vb\ and
+%     \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
+% The panels show a kernel density estimator (KDE) (black solid line) for
+%     the \vz -- \vb\ residuals of stars in the GUMS simulation at four
+%     different Galactic latitudes.
+% Blue dashed lines show Gaussian fits to these KDEs.
+% The distributions are close to Gaussian, with slightly heavy tails.
+% The standard deviations of the Gaussian fits increase with Galactic latitude.
+% This figure illustrates how using \vb\ instead of \vz\ artificially
+%     increases velocity dispersion, especially at high latitudes.
+% }
+%   \centering
+%     \includegraphics[width=1\textwidth]{vb_vz}
+% \label{fig:vb_vz}
+% \end{figure}
 
-% In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
-% are faster than expected; their rotational evolution appears to have stalled,
-% and they have similar rotation periods to the 650 Myr Praesepe cluster
-% \citep{curtis2019}.
-% The rotation periods of the K dwarfs in this cluster are plotted in figure
-% \ref{fig:dispersion_period_teff}.
-% Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
-% dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
-% NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
-% K dwarfs lie below it.
-% This cluster may be the `missing link' that connects two epoch of stellar
-% spin-down.
-% % , however without more observations of middle-aged open clusters it
-% % : an early stage where the period-\teff\ relation for cool dwarfs has
-% % a negative slope and a late stage where it has a positive slope.
-% % In this case the period gap may delineate the transition between these two
-% % regimes and is the point at which stellar magnetic dynamos likely undergo a
-% \subsection{Velocity dispersion of coeval groups}
-% \label{sec:age_cut}
-% % dramatic structural shift at an age of $\sim$ 1.1 Gyr.
+% Since we are concerned with velocity {\it dispersions}, rather than velocities
+% themselves, we also compared \sigmavb\ and \sigmavz\ as a function of
+% temperature for stars downloaded from the GUMS simulation.
+% For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
+% consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
+% latitudes the two quantities became significantly different.
+% For this reason we proceeded by only including stars with galactic latitudes
+% less than 15\degrees\ in our analysis.
+% Although we find that the transformation between \vz\ and \vb\ does not {\it
+% strongly} affect our results, we cannot rule out the possibility that it
+% introduces systematic biases into the velocity dispersions we present here.
+% In \gaia\ DR3, RVs will be available for most stars in this sample, providing
+% an opportunity to validate (or correct) the results presented here, and to
+% work in action-space, rather than velocity-space.
+
+% % Since we use \vb, not \vz\ in our analysis, it is possible that the sample
+% % selection function could influence our results.
+% % For example, higher-mass stars tend to be younger and reside closer to the
+% % galactic disk mid-plane (low galactic latitudes).
+% % Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
+% % lower-mass stars typically have greater velocity dispersions due to .
+% % \teff\ and galactic latitude, $b$,
+% % could result in
+% % Stars at higher latitudes have additional velocity components in the ${\bf x}$
+% % and ${\bf y}$ directions, which could increase \vb\ but not \vz.
+% % Again however, since the relationship between \sigmavb\ and \teff\ is
+% % positively, not negatively correlated for cool stars in the \kepler\ field,
+% % this effect is probably too small to influence our results.
+
+% % Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
+% % strongly affect the \mct\ sample of rotating \kepler\ stars.
+% % For this reason, we assume that age difference is the major cause of velocity
+% % dispersion differences between groups.
+% % In other words, (\vb) velocity dispersion is a reliable age proxy for the
+% % \mct\ sample.
+
+% Because of the noisy relationship between \vb\ and \vz\, in this paper we do
+% not attempt to convert velocity dispersion (\sigmavb) into an age via an
+% age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
+% Although we find that \sigmavb\ can be used to rank groups of stars by age, a
+% more careful analysis that includes formal modeling of the \vb\ -- \vz\
+% relationship will be needed to calculate absolute ages.
+% % For example, the velocity distributions could be modeled as a mixture of
+% % Gaussians in order to account for the additional velocity dispersion caused by
+% % the \vz-\vb\ transformation.
+% % The RVs of most of these stars will become available in \Gaia\ DR3, allowing
+% % calculations of \vz, which can be used to calculate more reliable ages via an
+% % AVR and suitable modeling of the selection function.
+
+% % We then selected groups of stars within different gyrochronal age ranges and
+% % calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
+% % of effective temperature for each age group.
+% % Ages were calculated using dereddened \gaia\ \gcolor\ color, however
+% % throughout this paper we show rotation periods as a function of effective
+% % temperature, \teff, because it is easier to divide stars into bins of roughly
+% % equal numbers in \teff-space than in color-space.
+
+% % We calculated the overall velocity dispersions of stars as a function of
+% % rotation period and gyrochronal age.
+
+% % Each star is assigned the same velocity dispersion as in figure
+% % \ref{fig:vplot}, and because stars in the same temperature group are not
+% % necessarily in the same mass group, which is why stars of the same mass and period do not nece
+
+% % The shape of the period-\teff\ relations at old ages appears to follow the
+% % shape of the upper detection edge.
+% % The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
+% % period catalog characterized by a dearth of slowly rotating M dwarfs between
+% % $\sim$3750 and 4250 K, is reflected in the lines of constant velocity
+% % dispersion (and presumed age) in the top panel of figure \ref{fig:vplot}.
+% % If the shape of the upper edge of rotation period measurements is created by a
+% % detection limit, it could indicate the rotation periods at which stars of
+% % different temperatures become relatively inactive (and their rotation periods
+% % therefore become undetectable).
+% % % Once inactive, stars would have little photometric variability induced by
+% % % star spots and their rotation periods would be difficult to measure, so may
+% % % not feature in the \mct\ catalog.
+% % If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
+% % $\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.
+
+% % The velocity dispersions of the coolest stars may be affected by a selection
+% % bias -- these extremely faint stars are more difficult to detect at larger
+% % distances, larger heights above the plane, and therefore larger velocities.
+% % It is possible that some high velocity stars of temperatures cooler than
+% % $\sim$4000 K are missing from this sample, and the velocity dispersions may
+% % therefore appear lower than they truly are.
+% % This could be why the velocity dispersion appears to decrease towards the
+% % right of figure \ref{fig:dispersion_period_teff}.
+
+% % \begin{figure}
+% %   \caption{
+% %     Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
+% %     x-axis.
+% %     Masses are from the \kepler\ input catalog.
+% %     The white lines show the \citet{spada2019} rotational evolution models at
+% %     0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
+% %     period.
+% %     These models include age and mass-dependent coupling between the stellar
+% %     core and envelope.
+% %     The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
+% %     an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
+% %     decreases with decreasing mass.
+% %     A similar phenomenon is visible in the velocity dispersions of field stars
+% %     shown as colored points in the background of this figure.
+% % }
+% %   \centering
+% %     \includegraphics[width=1\textwidth]{masses_with_models}
+% % \label{fig:masses_with_models}
+% % \end{figure}
+
+% % \subsection{The period gap}
+% % \label{sec:period_gap}
+
+% % The origin of the rotation period gap, first identified
+% % by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
+% % \ref{fig:dispersion_period_teff} still remains a mystery.
+% % This gap can be seen as an under-density of points between the 0.7-1.0 and
+% % 1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
+% % of constant gyrochronal age of around 1.1 Gyr \citep[according to the
+% % gyrochronology relation of][]{angus2019}, as shown in figure
+% % \ref{fig:dispersion_period_teff}.
+% % Several explanations for the gap's origin have been proposed, including a
+% % discontinuous star formation history \citep{mcquillan2013, davenport2017,
+% % davenport2018}, a rapid change in magnetic field structure
+% % \citep{reinhold2019}, and erroneous rotation period measurements that are
+% % incorrect by a factor of two \citep{koen2018}.
+% % The latter explanation can be ruled out because stars below the gap have
+% % smaller velocity dispersions than the stars above the gap, indicating that
+% % they are kinematically younger \citep{mcquillan2013, davenport2018}, as
+% % evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
+% % For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
+% % \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
+% % temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
+% % velocity dispersion increases with \teff.
+% % The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
+% % dispersion as the hottest stars in the age range above which indicates that
+% % the period-\teff\ relations are {\it flat} at these rotation periods.
+% % This is also visible in figure \ref{fig:dispersion_period_teff}.
+% % Below the gap, velocity dispersion within a given period range appears to {\it
+% % decrease} with decreasing temperature.
+% % The opposite appears to be true above the gap.
+% % It could be that the gap is positioned at a significant Rossby number/age at
+% % which stellar magnetic dynamos go through a transition.
+% % Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
+% % magnetic braking is more efficient for stars with deeper convection zones.
+% % Once stars reach this critical age or Rossby number their magnetic fields
+% % undergo some radical transition, which produces the gap in the rotation
+% % period-\teff\ plane.
+% % After this transition, magnetic braking efficiency no longer increases with
+% % decreasing mass.
+% % Of course, it may be a coincidence that the gyrochronology relations seem to
+% % only flatten off above the period gap and we lack a sufficient quantity of
+% % data to do more than speculate here.
+% % New rotation periods from the \ktwo\ and \tess\ missions may be able to
+% % validate or rule out this hypothesis in the future.
+
+% % In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
+% % are faster than expected; their rotational evolution appears to have stalled,
+% % and they have similar rotation periods to the 650 Myr Praesepe cluster
+% % \citep{curtis2019}.
+% % The rotation periods of the K dwarfs in this cluster are plotted in figure
+% % \ref{fig:dispersion_period_teff}.
+% % Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
+% % dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
+% % NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
+% % K dwarfs lie below it.
+% % This cluster may be the `missing link' that connects two epoch of stellar
+% % spin-down.
+% % % , however without more observations of middle-aged open clusters it
+% % % : an early stage where the period-\teff\ relation for cool dwarfs has
+% % % a negative slope and a late stage where it has a positive slope.
+% % % In this case the period gap may delineate the transition between these two
+% % % regimes and is the point at which stellar magnetic dynamos likely undergo a
+% % \subsection{Velocity dispersion of coeval groups}
+% % \label{sec:age_cut}
+% % % dramatic structural shift at an age of $\sim$ 1.1 Gyr.

### 2020-03-11 kinematics_and_rotation/dispersion.py

@@ -70,7 +70,7 @@ def dispersion(x, y, bins, method):
         if method == "std":
             d[i] = np.std(y[m])
         if method == "mad":
-            d[i] = 1.5*np.median(np.abs(y[m] - np.median(y[m])))
+            d[i] = 1.5*aps.median_absolute_deviation(y[m])
 
         # Calculate kurtosis
         k[i] = sps.kurtosis(y[m])

### 2020-03-12 paper/conclusion.tex
small typos

@@
 Although at young ages, rotation period is anti-correlated with \teff\ (as
-seen in many young open clusters, include Praesepe), at intermediate ages the
+seen in many young open clusters, including Praesepe), at intermediate ages the
 relation flattens out and K dwarfs of different masses rotate at the same
@@
         using \vb\ as a proxy for \vz\ introduces an additional
-        pseudo-velocity dispersion, which increases with increasing Galactic
+        velocity scatter, which increases with increasing Galactic
         latitude.

### 2020-03-12 paper/method.tex
small typos

@@
 greater than 1 \kms\, and stars with galactic latitudes greater than
-15\degrees\ (justification provided below) were removed from the sample.
+15\degrees\ (justification provided in the appendix) were removed from the
+sample.
 Finally, we removed stars with rotation periods shorter than the main

### 2020-03-12 paper/conclusion.tex
removing commented blocks of text

@@
 mass for K dwarfs.}
-% We found that the \citet{angus2019} relation, which is based on the
-% period--color relation of the 650 Myr Praesepe cluster, does not correctly
-% describe the period--age--\teff\ relation for old stars.
 Although at young ages, rotation period is anti-correlated with \teff\ (as
@@
 
-% The period-color/\teff\ relation seems to start flattening out after $\sim$1
-% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
-% older than the period gap which is located at a gyrochronal age of around
-% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
 \item{{\bf Low-mass stars stay active longer.}
@@
         latitude.
-% For this reason we did not attempt to convert \vb\ dispersions into ages using
-% an age-velocity dispersion relation.
 However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
@@
 rank stars by age.}
-% , \ie\ a group of stars with a large velocity dispersion is,
-% on average, older than a group of stars with a small velocity dispersion.}
 
 \end{itemize}
-
-% explanations for these observations, including incorrect dust corrections for
-% the lowest-mass stars and an excess of companions increasing the velocity
-% dispersion for these stars.
-
-% If the period-color/\teff\ relation does invert at old ages as our results
-% suggest, this would be a paradigm shift for gyrochronology.
-% Stellar spin-down rate is thought to be directly tied to magnetic field
-% strength, and the deeper convection zones of cooler stars generate stronger
-% magnetic fields which {\it should} lead to more efficient angular momentum
-% loss.
-% However, the micro- and macro-physics of stellar structure and evolution and
-% magnetic dynamo models are extremely complicated and a lot is still unknown
-% about the magnetic behavior of stars.
-% Observations like these can provide useful constraints for physical models,
-% and may help to reveal new physical processes at work in stars like our own
-% Sun and other planet hosts.

### 2020-03-13 paper/conclusion.tex
comments from Gus

@@
 
-\item{{\bf \sigmavb\ can be used as an age proxy for \kepler\ stars.} In the
-    appendix of this paper we demonstrate that the dispersion of velocities in
-        the direction of Galactic latitude, \vb, can be used as an age proxy,
-        by showing that there is no strong evidence for mass-dependent heating
-        in low-mass \kepler\ dwarfs: the velocity dispersions of K and M
-        dwarfs, whose main-sequence lifetimes are longer than around 11 Gyrs,
-        do not appear to increase with decreasing mass.
+\item{{\bf \sigmavb\ can be used as an age proxy for (low Galactic latitude)
+    \kepler\ stars.} In the appendix of this paper we demonstrate that the
+        dispersion of velocities in the direction of Galactic latitude, \vb,
+        can be used as an age proxy, by showing that there is no strong
+        evidence for mass-dependent heating in low-mass \kepler\ dwarfs: the
+        velocity dispersions of K and M dwarfs, whose main-sequence lifetimes
+        are longer than around 11 Gyrs, do not appear to increase with
+        decreasing mass.
 Although {\it vertical} velocity, \vz, is a quantity that has been

### 2020-03-13 paper/intro.tex
removing commented blocks of text
comments from Gus

@@
-% Word limit: 500
 \section{Introduction}
@@
 barnes2010, meibom2011, meibom2015}.
-% It is well established that magnetized stellar winds cause the rotation
-% periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
-% weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
-% calibrated, the relationship between period, mass and age can be used to
-% age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
-% meibom2011, meibom2015}.
 However, new photometric rotation periods made available by the \kepler\
@@
 reliable ages for stars spanning a range of ages and masses.
-% however relying on open clusters (which are typically young) and asteroseismic
-% stars (which are typically G-type stars or more massive) to calibrate
-% gyrochronology relations limits the mass and age coverage of the calibration
-% sample.
 In this paper, we use the velocity dispersions of field stars to qualitatively
@@
 implied relative ages.
-% The age-velocity dispersion reations (AVRs) are still actively being
 
-% Vertical action is thought to be a better age indicator than vertical
-% velocity \citep{ting2019}, although still only a weak age indicator for
-% individual stars \citep{beane2018}, however both vertical action ($J_z$) and
-% vertical
 Although {\it vertical} velocity, \vz, is an established age proxy, it can
@@
 vertical velocity.
-In future it may be possible to account for the differences between \vb\ and
-\vz, or marginalize over missing RV measurements and the \kepler\ selection
-function, in order to infer absolute ages.
+In the future it may be possible to account for the differences between \vb\
+and \vz, or marginalize over missing RV measurements and the \kepler\
+selection function, in order to infer the absolute ages of groups of stars.
 Regardless of direction however, velocity dispersion is expected to
@@
 sample.
-% We show that the period-color relation of Praesepe is not applicable to
-% old stars in section \ref{sec:age_cut}, and reveal the true shape of the
-% period-color relations in section \ref{sec:the_reveal}.
-% We discuss a possible connection with the rotation period gap in section
-% \ref{sec:period_gap}.
-
-% We also tested the \citet{angus2019} gyrochronology relation, a separable
-% relation in color and age, calibrated using the period-color relation of
-% Praesepe (in \gaia\ \gcolor\ color) and the period-age relation of Praesepe
-% and the Sun.  The large number of Praesepe members with precise rotation
-% periods from the
-% \ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
-% through early M, makes it a good cluster for calibrating the period-color
-% relation of stars at 650 Myrs.
-% However, although this relation accurately describes the rotation periods of F
-% and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
-% with available rotation periods), it over-predicts the rotation periods of K
-% dwarfs in the 1.1 Gyr NGC 6811 cluster.
-% Very few reliable age estimates exist for K dwarfs with rotation periods older
-% than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
-% evolution of middle-aged K dwarfs is largely unknown.
-% The oldest M dwarfs with reliable ages and rotation periods are members of
-% Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
-% evolution.
-
-% We calculated gyrochronal ages of cool field dwarfs in the \kepler\ field
-% using the \citep{angus2019} gyrochronology model, with de-reddened \Gaia\
-% \gcolor\ color and rotation periods reported by \mct\ (we explain our data
-% selection process in section \ref{sec:the_data}).
-% These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
-% figure \ref{fig:age_gradient}.
-% The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
-% along the upper edge of the MS, where stellar evolution models predict old
-% stars to be, however the majority of these `old' stars are bluer than \gcolor\
-% $\sim$ 1.5 dex.
-% This suggests that the \citet{angus2019} gyrochronology relation
-% under-predicts the ages of low-mass stars.
-% There is no reason to expect the oldest stars in this sample to be the bluer
-% ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
-% active for longer, so should therefore have measurable rotation periods at
-% older ages.
-% Since the \citet{angus2019} gyrochronology model, which is based on the
-% period-color relation of Praesepe, predicts the oldest stars in this sample to
-% be K dwarfs, it is probably either under-predicting M dwarf ages or
-% over-predicting K dwarf ages.
-% % In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
-% % slowly than the model predicts.
-% In what follows, we use a population-based stellar age indicator, velocity
-% dispersion, to investigate the gyrochronology relations at old ages in
-% the field.

### 2020-03-14 paper/results.tex
removing commented blocks of text
incorporating jen's comments

@@
-% word limit: 500
 \section{Results and Discussion}
@@
 
-% \subsection{Velocity dispersion of coeval groups}
-% \label{sec:age_cut}
-
-% To explore the relationship between rotation period, \teff\ and age for field
-% K dwarfs, we calculated gyrochronal ages using dereddened \Gaia\ photometry
-% ($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
-% We used the \citet{angus2019} gyrochronology relation which is a simple,
-% separable relation in \gcolor\ color and age, calibrated using the
-% period-color relation of Praesepe and the period-age relation of Praesepe and
-% the Sun.
-% The large number of Praesepe members with precise rotation periods from the
-% \ktwo\ mission \citep{howell2014, douglas2017, rebull2017}, spanning spectral
-% types F through early M, makes it a good cluster for calibrating the
-% period-color relation of stars at young ages.
-% However, this is an extremely simple model and although it accurately
-% describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
-% of NGC 6819 -- the oldest cluster with available rotation periods), it
-% over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
-% cluster.
-
-% We then selected groups of stars within different gyrochronal age ranges and
-% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
-% of effective temperature for each age group.
-% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-% throughout this paper we show rotation periods as a function of effective
-% temperature, \teff, because it is easier to divide stars into bins of roughly
-% equal numbers in \teff-space than in color-space.
-
-% \begin{figure}
-%   \caption{
-% Top: rotation period vs effective temperature for stars in the \mct\
-%     catalog.
-%     The full catalog, with subgiants and visual binaries removed is shown in
-%     grey, and stars selected to be in different age groups (between \tmin\ and
-%     \tmax\ K) are overlayed in color.
-% These age groups were selected using the \citet{angus2019} gyrochronology
-%     relation.
-% The legend in the center of the figure lists the age range, in Gyr, of each
-%     group.
-% Bottom: velocity dispersion vs effective temperature for each age
-%     group.
-% The color of the line corresponds to the color of the group shown in the top
-%     panel.
-% If the gyrochronal model were correct at all ages, and the stars in each group
-%     were the same age across temperatures, the velocity dispersion would be
-%     constant as a function of \teff.
-% However, the velocity dispersions of the oldest age groups increase with
-%     \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
-%     the the ages of late-K dwarfs relative to the ages of late G and early K
-%     dwarfs at old ages.
-% % An alternative explanation could be that the gyrochronology relation is
-% %     correct and {\it mass-dependent heating} is responsible for the greater
-% %     velocity dispersions of cooler stars.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{age_cut}
-% \label{fig:age_cut}
-% \end{figure}
-% The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
-% (excluding visual binaries and subgiants) in grey, with coeval groups
-% \citep[according to the][relation]{angus2019} shown in color.
-% The color of the points corresponds to the age ranges specified in the legend
-% (in Gyr), which also apply to the lines in the lower panel.
-% The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
-% as a function of effective temperature.
-% % We only included stars within a temperature range of 5000 - 3500 K in our
-% % analysis, as hotter stars are more likely to have stopped magnetic braking
-% % \citep{vansaders2016}, which could bias the results.
-% % Late M dwarfs were not included in our analysis because such faint stars
-% % cannot be observed at large heights above the plane (because of the low
-% % galactic latitude of the \kepler\ field, stars at high-Z are more distant),
-% % which introduces a mass-dependent velocity bias: cooler populations of stars
-% % are skewed towards lower velocity dispersions.
-% % The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
-% % have low velocity dispersions, indicating that this effect may already become
-% % important at temperatures lower than $\sim$ 4000 K.
-% Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
-% with gyrochronal age across all temperatures, implying that both velocity
-% dispersion and rotation period increases with age as expected.
-% The flat velocity dispersion of young stars as a function of temperature
-% shows that the Praesepe-calibrated gyrochronology relation accurately predicts
-% the relative ages of {\it young} field stars.
-% % To our knowledge, no gyrochronology relation has previously been demonstrated to
-% % correctly predict ages (either relative or absolute) for such cool or such
-% % young field stars.
-% % This is not particularly remarkable however, since these young stars are a
-% % similar age to the Praesepe cluster, which was used to calibrate the
-% % period-color relation.
-% % However, if the stars in each selected age group had the {\it same} age across
-% % the temperature range, their velocity dispersion would be a constant function
-% % of \teff.
-% If \citet{angus2019} gyrochronology relation worked at all ages and
-% temperatures, the bottom panel of figure \ref{fig:age_cut} would show a flat
-% relationship between velocity dispersion and \teff at all ages.
-% However, velocity dispersion {\it increases} as a function of temperature for
-% old stars, meaning the \citet{angus2019} gyrochronology relation either
-% under-predicts the ages of old, late-K dwarfs, or over-predicts the ages of
-% old early-K and late-G dwarfs\footnote{Since \vb\ velocity dispersion only
-% provides relative and not absolute ages, it is difficult to tell whether the
-% ages of cool stars are being under-predicted, the ages of hot stars being are
-% over-predicted, or both.}.
-% This suggests that the relationship between rotation period and photometric
-% color or \teff\ flattens out over time, and possibly even inverts.
-% % The lines of constant age (isochrones) sweeping diagonally upwards in the top
-% % panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.
-
 \subsection{The period-\teff\ relations, revealed}
 \label{sec:the_reveal}
-
-% To explore the relationship between rotation period, effective temperature
-% (\teff ) and velocity dispersion,
-% % , we first removed high and low velocity
-% % outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
-% % \vb\ velocities.
-% % Without sigma-clipping, we found that a small number of high velocity
-% % outliers at the low-temperature end of our sample substantially raised the
-% % velocity dispersion for cooler stars, although this did not change the overall
-% % trends in the data.
-% % We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
-% % 3500 K to avoid biases caused by the selection function at the faint, cool
-% % end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
-% we removed stars with rotation periods shorter than the main population of
-% periods, since this area of the period-\teff\ diagram is sparsely populated.
-% We removed these rapid rotators by cutting out stars with gyrochronal ages
-% less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
-% Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
-% line of constant age in period-\teff, or period-color space.} traces the
-% bottom edge of the main population of rotation periods.
-% After these cuts, around 13,000 stars out of the original $\sim$34,000 were
-% included in the sample.
 
@@
 the 0.5 Gyr gyrochrone.
-% Lines of constant age (isochrones) appear to follow the shape of the
-% Praesepe-based gyrochronology model (black dotted line) at young ages.
-% However, at older ages it appears that the relation between rotation period
-% and \teff\ flattens out, until eventually rotation period {\it decreases} with
-% decreasing effective temperature at a given age.
 However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
@@
 these Praesepe-based models.
-% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
-% rotate more rapidly than higher-mass stars.
-% This is the opposite of the trend that is observed in young open clusters such
-% as Praesepe and the Hyades.
 
@@
 massive stars.
+To investigate this idea further, we compared the velocity dispersions of
+stars with measured rotation periods, to the velocity dispersions of the
+overall \kepler\ sample.
+We calculated the velocity dispersions for all stars in the \kepler\ field,
+after removing visual binaries, subgiants, stars fainter than 16th magnitude,
+and high Galactic latitude stars, following the method described in section
+\ref{sec:method}.
+We then compared these velocity dispersions to the velocity dispersions of
+stars in the \mct\ sample.
+If the rotation periods of G stars are only detectable when they are young,
+G stars with measured periods should have smaller velocity dispersions than
+the overall \kepler\ field.
+We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
+\sigmavb\ for stars with rotation periods published in \mct, as a function of
+\teff\ (see figure \ref{fig:compare}).
+A larger ratio means the rotating star sample is {\it younger}, on average,
+than the overall \Kepler\ population, and a ratio of 1 means that the rotating
+stars have the {\it same} age distribution as the overall Kepler sample.
+Figure \ref{fig:compare} shows that this ratio is largest for G stars and
+approaches unity for K and early M dwarfs.
+This indicates that the G stars with detectable rotation periods are, on
+average, {\it younger} than the total population of G stars in the Kepler
+field.
+On the other hand, the late K and early M dwarfs with detectable rotation
+periods have a similar age distribution to the overall Kepler population which
+suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
+This result bolsters the evidence that M dwarf rotation periods are measurable
+at older ages than G dwarf rotation periods.
+G stars become magnetically inactive and have fewer active surface regions
+{\it at a younger age than M dwarfs}.
+
+\begin{figure}
+  \caption{
+    Velocity dispersions for the entire \kepler\ field divided by the velocity
+    dispersion of stars with measured rotation periods in \mct,
+    as a function of effective temperature.
+    A larger ratio means that the overall \kepler\ field is older, on average,
+    than stars in the \mct\ catalog.
+    As this ratio approaches unity the two populations have similar kinematic
+    ages.
+    The large ratio for the hottest stars indicates that G dwarfs become
+    inactive at young ages.
+    This ratio approaches unity at low temperatures, showing that K and early
+    M dwarf rotation periods are measurable over a large range of ages.
+}
+  \centering
+    \includegraphics[width=1\textwidth]{field_comparison}
+\label{fig:compare}
+\end{figure}
 
@@
 {\it cross} lines of constant velocity dispersion above the gap.
+This phenomenon is robust to the choice of bin size and position.
 Although we do not provide an in-depth analysis here (and more data may be
@@
 to two distinct populations \citep{mcquillan2014}.
-The smooth increase in velocity dispersion across the gap shown here supports
-the hypothesis that the gap is caused by a phenomenon of magnetic or
-structural stellar evolution, not a discontinuity in the local star formation
-history.
+The smooth increase in velocity dispersion across the gap shown here suggests
+that stars above and below are part of the same stellar population.
+However, it does not rule out the possibility that a brief cessation of star
+formation in the Solar neighborhood, around one Gyr ago, may have caused this
+gap.
 
+In the final part of our analysis, we investigated the potential for using
+kinematics to identify synchronized binaries in the \mct\ sample.
 Synchronized binaries are pairs of stars whose rotation periods are equal to
@@
 are likely synchronized binaries.
-% The large velocity dispersions of the most rapidly rotating stars in the \mct\
-% sample indicates that at least some fraction of these stars are likely to be
-% synchronized binaries.
 For this reason, the velocity dispersions of stars with rotation periods
@@
 presumed older, orange-colored stars.
-% This is an indication that many of these are synchronized binaries.
-% The rotation periods of stars in synchronized binaries are tidally locked to
-% the orbital period of the binary system and since synchronized binaries have
-% short orbital periods this can produce old, seemingly isolated stars, with
-% short rotation periods.
-% Given the strong correlation between binary fraction and primary mass
-% \citep[\eg][]{price-whelan2020}, and the fact that M dwarfs take longer to
-% transition from the `I` (rapidly rotating) sequence to the `C' (slowly
-% rotating) sequence, it is likely that
 In general, stars with rotation periods less than $\sim$10 days have an
@@
 before applying any gyrochronal analysis.
-
-% \subsection{Validating \vb\ dispersion as an age proxy}
-% \label{sec:mass-dependent-heating}
-
-% The conclusions drawn in this paper depend on the assumption that velocity
-% dispersion in the direction of Galactic latitude (\sigmavb) can be used as an
-% age proxy.
-% There are two main reasons however, why \vb\ velocity dispersion may {\it not}
-% be a good age proxy.
-% Firstly, mass-dependent heating may act on the sample, meaning that velocity
-% dispersion depends on both age and mass, so cannot be interpreted as a simple
-% age proxy.
-% Secondly, since stars in the \kepler\ field have a range of Galactic
-% latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
-% stars, and introduce a velocity bias for high latitude stars (which are more
-% likely to be cooler and older).
-% In this section we demonstrate that neither of these problems seem to be a
-% significant issue for our data.
-
-% In order to establish whether \sigmavb\ can be used as an age proxy, we
-% searched for signs of mass-dependent heating within the \kepler\ field.
-% Mass-dependent dynamical heating may result from lower-mass stars experiencing
-% greater velocity changes when gravitationally perturbed than more massive
-% stars.
-% It has not been unambiguously observed in the galactic disk because of the
-% strong anti-correlation between stellar mass and stellar age.
-% Less massive stars do indeed have larger velocity dispersions, however they
-% are also older on average.
-% This mass-age degeneracy is highly reduced in M dwarfs because their
-% main-sequence lifetimes are longer than the age of the Universe, and no
-% evidence for mass-dependent heating has previously been found in M dwarfs
-% \citep[\eg][]{faherty2009, newton2016}.
-
-% To investigate whether mass-dependent heating could be acting on the \kepler\
-% sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
-% whose MS lifetimes exceed around 11 Gyr and are therefore representative of
-% the initial mass function.
-% We could not perform this analysis on the \mct\ sample, because it only
-% includes stars with {\it detectable} rotation periods, and since lower-mass
-% stars stay active for longer it is likely that it contains a strong mass-age
-% correlation.
-% We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
-% greater than 1.2 (corresponding to an effective temperature $\lesssim$
-% 4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
-% We also eliminated photometric binaries by removing stars above a 6th order
-% polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
-% figure \ref{fig:age_gradient}).
-% We then applied the quality cuts described above in section
-% \ref{sec:the_data}.
-% To search for evidence of mass-dependent heating we calculated the (\vb)
-% velocity dispersion of stars in effective temperature bins.
-% Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
-% outliers before calculating the standard deviation of stars in each bin.
-% These extreme velocity outliers may be very old late K and M dwarfs, or they
-% result from using \vb\ instead of \vz, which introduces additional velocity
-% scatter.
-
-% Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
-% function of effective temperature for the K and M \kepler\ dwarf sample.
-% Velocity dispersion very slightly {\it decreases} with decreasing temperature,
-% the opposite of the trend expected for mass-dependent heating, however the
-% slope is only inconsistent with zero at 1.3 $\sigma$.
-% This trend may be due to a selection bias: cooler stars are fainter and
-% therefore typically closer, with smaller heights above the galactic plane and
-% smaller velocities.
-% The essential point however, is that we do not see evidence for mass-dependent
-% heating acting on stars in the \kepler\ field, indicating that velocity
-% dispersion {\it can} be used as an age proxy (with the caveat that there is
-% still a chance, albeit a small one, that the opposing effects of the selection
-% function and mass-dependent heating are working to cancel each other out).
-% This analysis was performed using \vb\ but we also examined the {\it vertical}
-% velocities of the 537 stars in this sample with RV measurements.
-% Again, no evidence was found for mass-dependent heating: the slope of the
-% velocity dispersion-temperature relation was consistent with zero.
-% \begin{figure}
-%   \caption{
-%       Top: Stellar velocity (\vb) as a function of \teff\ for
-%       \kepler\ K and M dwarfs.
-% Vertical lines indicate different \teff-groupings used to calculate velocity
-%     dispersion.
-% Pink stars were not included in velocity dispersion calculations as they were
-%     either removed as outliers during a sigma clipping process, or they lie at
-%     the sparcely populated, extremely cool end of the temperature range.
-%     Velocity dispersion and \teff\ are slightly positively correlated, likely
-%     due to a brightness-related selection bias, indicating that mass-dependent
-%     heating does not significantly affect low-mass stars in the \kepler\
-%     field.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{vb_vs_teff}
-% \label{fig:vb_vs_teff}
-% \end{figure}
-
-% Having found no strong evidence for mass-dependent heating, we next tested
-% the validity of \vb\ as a proxy for \vz\ in more detail.
-% At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
-% of $b$, this equivalence becomes an approximation that grows noisier with $b$.
-% To test the validity of the \vb$\sim$\vz\ approximation over a range of
-% latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
-% (GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
-% We downloaded stars from four pointings in the \kepler\ field with galactic
-% latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
-% a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
-% The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
-% \vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
-% Galaxy.
-% We subtracted this offset and examined the residuals of the \vz\ -- \vb\
-% relationship to investigate the variance as a function of Galactic latitude
-% (shown in figure \ref{fig:vb_vz}).
-% We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
-% with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
-% The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
-% $, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
-% 20$^\circ$.
-% This demonstrates that using \vb\ instead of \vz\ for stars in the \kepler\
-% field will introduce an additional velocity scatter, inflating \sigmavb\
-% relative to \sigmavz.
-% This additional velocity scatter will be greatest for stars at the highest
-% Galactic latitudes.
-
-% % Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
-% % over different Galactic latitudes.
-% % Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
-% % fits are shown as dashed blue lines.
-% % The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
-% % the variance increases with increasing Galactic latitude.
-% \begin{figure}
-%   \caption{
-% This figure demonstrates the variance in the relationship between \vb\ and
-%     \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
-% The panels show a kernel density estimator (KDE) (black solid line) for
-%     the \vz -- \vb\ residuals of stars in the GUMS simulation at four
-%     different Galactic latitudes.
-% Blue dashed lines show Gaussian fits to these KDEs.
-% The distributions are close to Gaussian, with slightly heavy tails.
-% The standard deviations of the Gaussian fits increase with Galactic latitude.
-% This figure illustrates how using \vb\ instead of \vz\ artificially
-%     increases velocity dispersion, especially at high latitudes.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{vb_vz}
-% \label{fig:vb_vz}
-% \end{figure}
-
-% Since we are concerned with velocity {\it dispersions}, rather than velocities
-% themselves, we also compared \sigmavb\ and \sigmavz\ as a function of
-% temperature for stars downloaded from the GUMS simulation.
-% For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
-% consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
-% latitudes the two quantities became significantly different.
-% For this reason we proceeded by only including stars with galactic latitudes
-% less than 15\degrees\ in our analysis.
-% Although we find that the transformation between \vz\ and \vb\ does not {\it
-% strongly} affect our results, we cannot rule out the possibility that it
-% introduces systematic biases into the velocity dispersions we present here.
-% In \gaia\ DR3, RVs will be available for most stars in this sample, providing
-% an opportunity to validate (or correct) the results presented here, and to
-% work in action-space, rather than velocity-space.
-
-% % Since we use \vb, not \vz\ in our analysis, it is possible that the sample
-% % selection function could influence our results.
-% % For example, higher-mass stars tend to be younger and reside closer to the
-% % galactic disk mid-plane (low galactic latitudes).
-% % Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
-% % lower-mass stars typically have greater velocity dispersions due to .
-% % \teff\ and galactic latitude, $b$,
-% % could result in
-% % Stars at higher latitudes have additional velocity components in the ${\bf x}$
-% % and ${\bf y}$ directions, which could increase \vb\ but not \vz.
-% % Again however, since the relationship between \sigmavb\ and \teff\ is
-% % positively, not negatively correlated for cool stars in the \kepler\ field,
-% % this effect is probably too small to influence our results.
-
-% % Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
-% % strongly affect the \mct\ sample of rotating \kepler\ stars.
-% % For this reason, we assume that age difference is the major cause of velocity
-% % dispersion differences between groups.
-% % In other words, (\vb) velocity dispersion is a reliable age proxy for the
-% % \mct\ sample.
-
-% Because of the noisy relationship between \vb\ and \vz\, in this paper we do
-% not attempt to convert velocity dispersion (\sigmavb) into an age via an
-% age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
-% Although we find that \sigmavb\ can be used to rank groups of stars by age, a
-% more careful analysis that includes formal modeling of the \vb\ -- \vz\
-% relationship will be needed to calculate absolute ages.
-% % For example, the velocity distributions could be modeled as a mixture of
-% % Gaussians in order to account for the additional velocity dispersion caused by
-% % the \vz-\vb\ transformation.
-% % The RVs of most of these stars will become available in \Gaia\ DR3, allowing
-% % calculations of \vz, which can be used to calculate more reliable ages via an
-% % AVR and suitable modeling of the selection function.
-
-% % We then selected groups of stars within different gyrochronal age ranges and
-% % calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
-% % of effective temperature for each age group.
-% % Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-% % throughout this paper we show rotation periods as a function of effective
-% % temperature, \teff, because it is easier to divide stars into bins of roughly
-% % equal numbers in \teff-space than in color-space.
-
-% % We calculated the overall velocity dispersions of stars as a function of
-% % rotation period and gyrochronal age.
-
-% % Each star is assigned the same velocity dispersion as in figure
-% % \ref{fig:vplot}, and because stars in the same temperature group are not
-% % necessarily in the same mass group, which is why stars of the same mass and period do not nece
-
-% % The shape of the period-\teff\ relations at old ages appears to follow the
-% % shape of the upper detection edge.
-% % The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
-% % period catalog characterized by a dearth of slowly rotating M dwarfs between
-% % $\sim$3750 and 4250 K, is reflected in the lines of constant velocity
-% % dispersion (and presumed age) in the top panel of figure \ref{fig:vplot}.
-% % If the shape of the upper edge of rotation period measurements is created by a
-% % detection limit, it could indicate the rotation periods at which stars of
-% % different temperatures become relatively inactive (and their rotation periods
-% % therefore become undetectable).
-% % % Once inactive, stars would have little photometric variability induced by
-% % % star spots and their rotation periods would be difficult to measure, so may
-% % % not feature in the \mct\ catalog.
-% % If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
-% % $\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.
-
-% % The velocity dispersions of the coolest stars may be affected by a selection
-% % bias -- these extremely faint stars are more difficult to detect at larger
-% % distances, larger heights above the plane, and therefore larger velocities.
-% % It is possible that some high velocity stars of temperatures cooler than
-% % $\sim$4000 K are missing from this sample, and the velocity dispersions may
-% % therefore appear lower than they truly are.
-% % This could be why the velocity dispersion appears to decrease towards the
-% % right of figure \ref{fig:dispersion_period_teff}.
-
-% % \begin{figure}
-% %   \caption{
-% %     Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
-% %     x-axis.
-% %     Masses are from the \kepler\ input catalog.
-% %     The white lines show the \citet{spada2019} rotational evolution models at
-% %     0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
-% %     period.
-% %     These models include age and mass-dependent coupling between the stellar
-% %     core and envelope.
-% %     The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
-% %     an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
-% %     decreases with decreasing mass.
-% %     A similar phenomenon is visible in the velocity dispersions of field stars
-% %     shown as colored points in the background of this figure.
-% % }
-% %   \centering
-% %     \includegraphics[width=1\textwidth]{masses_with_models}
-% % \label{fig:masses_with_models}
-% % \end{figure}
-
-% % \subsection{The period gap}
-% % \label{sec:period_gap}
-
-% % The origin of the rotation period gap, first identified
-% % by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
-% % \ref{fig:dispersion_period_teff} still remains a mystery.
-% % This gap can be seen as an under-density of points between the 0.7-1.0 and
-% % 1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
-% % of constant gyrochronal age of around 1.1 Gyr \citep[according to the
-% % gyrochronology relation of][]{angus2019}, as shown in figure
-% % \ref{fig:dispersion_period_teff}.
-% % Several explanations for the gap's origin have been proposed, including a
-% % discontinuous star formation history \citep{mcquillan2013, davenport2017,
-% % davenport2018}, a rapid change in magnetic field structure
-% % \citep{reinhold2019}, and erroneous rotation period measurements that are
-% % incorrect by a factor of two \citep{koen2018}.
-% % The latter explanation can be ruled out because stars below the gap have
-% % smaller velocity dispersions than the stars above the gap, indicating that
-% % they are kinematically younger \citep{mcquillan2013, davenport2018}, as
-% % evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
-% % For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
-% % \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
-% % temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
-% % velocity dispersion increases with \teff.
-% % The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
-% % dispersion as the hottest stars in the age range above which indicates that
-% % the period-\teff\ relations are {\it flat} at these rotation periods.
-% % This is also visible in figure \ref{fig:dispersion_period_teff}.
-% % Below the gap, velocity dispersion within a given period range appears to {\it
-% % decrease} with decreasing temperature.
-% % The opposite appears to be true above the gap.
-% % It could be that the gap is positioned at a significant Rossby number/age at
-% % which stellar magnetic dynamos go through a transition.
-% % Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
-% % magnetic braking is more efficient for stars with deeper convection zones.
-% % Once stars reach this critical age or Rossby number their magnetic fields
-% % undergo some radical transition, which produces the gap in the rotation
-% % period-\teff\ plane.
-% % After this transition, magnetic braking efficiency no longer increases with
-% % decreasing mass.
-% % Of course, it may be a coincidence that the gyrochronology relations seem to
-% % only flatten off above the period gap and we lack a sufficient quantity of
-% % data to do more than speculate here.
-% % New rotation periods from the \ktwo\ and \tess\ missions may be able to
-% % validate or rule out this hypothesis in the future.
-
-% % In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
-% % are faster than expected; their rotational evolution appears to have stalled,
-% % and they have similar rotation periods to the 650 Myr Praesepe cluster
-% % \citep{curtis2019}.
-% % The rotation periods of the K dwarfs in this cluster are plotted in figure
-% % \ref{fig:dispersion_period_teff}.
-% % Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
-% % dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
-% % NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
-% % K dwarfs lie below it.
-% % This cluster may be the `missing link' that connects two epoch of stellar
-% % spin-down.
-% % % , however without more observations of middle-aged open clusters it
-% % % : an early stage where the period-\teff\ relation for cool dwarfs has
-% % % a negative slope and a late stage where it has a positive slope.
-% % % In this case the period gap may delineate the transition between these two
-% % % regimes and is the point at which stellar magnetic dynamos likely undergo a
-% % \subsection{Velocity dispersion of coeval groups}
-% % \label{sec:age_cut}
-% % % dramatic structural shift at an age of $\sim$ 1.1 Gyr.

### 2020-03-15 paper/abstract.tex
removing commented blocks of text
more changes

@@
-% Word limit: 250
-
 The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
@@
 stars.
-% However, recent evidence suggests that the relationship between rotation
-% period and color/temperature/mass does not remain constant over large
-% timescales, contradicting a long-held assumption that underlies many
-% gyrochronology relations.
 In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
@@
 kinematics could be a useful tool for calibrating gyrochronology.
-% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
-% to the Praesepe cluster, accurately predicts the relative ages of stars.
-We find that a gyrochronology model, calibrated to fit the period--\teff\
-relationship of the Praesepe cluster, does not apply to stars older than
-around 1 Gyr.
-% : a polynomial gyrochronology model, fit
-% to Praesepe stars, under-predicts the ages of late K dwarfs in the field.
+We find that a linear gyrochronology model, calibrated to fit the
+period--\teff\ relationship of the Praesepe cluster, does not apply to stars
+older than around 1 Gyr.
 Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
@@
 the {\it same rate} or faster than late-K dwarfs of the same age.
-% This is probably because their stronger magnetic fields lead to more efficient
-% angular momentum loss and reflects the behavior of young open clusters such as
-% the Pleaides, Praesepe, and the Hyades.
-% However, at older ages we find that late G and early K dwarfs rotate at the
-% {\it same rate} or faster than late K dwarfs of the same age.
-% These oThis may be a feature of angular momentum redistribution in the stellar
-% interior.
-% This implies that low-mass stars lose angular momentum more slowly than
-% higher-mass stars after a certain age or Rossby number.
 This result agrees qualitatively with semi-empirical models that vary the rate
@@
 K-dwarfs are likely to be synchronized binaries.
-% Firstly, we
-% find no evidence for mass-dependent heating in a sample of K and M dwarfs in
-% the \kepler\ field.
-
-% We find that groups of stars, selected to be the same age using a
-% gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
-% the same velocity dispersion at all temperatures; cooler stars have greater
-% velocity dispersions than hotter stars.
-% We interpret this to mean that these cooler stars are older than the hotter
-% ones, which implies that the period-color relation of Praesepe is not
-% appropriate for old stars.
-% Futhermore, our results indicate that the slope of the period-color relation
-% actually {\it inverts}: at young ages early K stars rotate more rapidly than
-% late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
-% These observations could be reproduced by mass-dependent dynamical heating,
-% where lower-mass stars experience a greater rate of dynamical heating than
-% higher-mass stars, however we find no evidence for mass-dependent heating
-% within K and M dwarfs in the \kepler\ field.

### 2020-03-15 paper/conclusion.tex
more changes

@@
 \begin{itemize}
-\item{{\bf Magnetic braking efficiency doesn't always increase with decreasing
-mass for K dwarfs.}
+\item{{\bf Spin-down rate doesn't always increase with decreasing mass for K
+    dwarfs.}
 Although at young ages, rotation period is anti-correlated with \teff\ (as
@@
 We found that the oldest stars in the \mct\ catalog are cooler than 4500
-K, which suggests that lower-mass stars remain active for longer, allowing
-their rotation periods to be measured at older ages.}
+K, which agrees with previous results which show that lower-mass stars remain
+        active for longer, allowing their rotation periods to be measured at
+        older ages.}
 

### 2020-03-15 paper/method.tex
removing commented blocks of text
more changes

@@
-% 270 words
-% Limit: 500 words
-
 \section{Method}
@@
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-Curtis \etal\ (2020, in prep) combined effective temperature measurements for
-nearby, unreddened field stars in benchmark samples, including FGK stars
-characterized with high-resolution optical spectroscopy \citep{brewer2016}, M
-dwarfs characterized with low-resolution optical and near-infrared
-spectroscopy \citep{mann2015}, and K and M dwarfs characterized with
-interferometry and bolometric flux analyses \citep{boyajian2012}.
+To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
+temperature measurements for nearby, unreddened field stars in benchmark
+samples, including FGK stars characterized with high-resolution optical
+spectroscopy \citep{brewer2016}, M dwarfs characterized with low-resolution
+optical and near-infrared spectroscopy \citep{mann2015}, and K and M dwarfs
+characterized with interferometry and bolometric flux analyses
+\citep{boyajian2012}.
 This empirical color--temperature relation is valid over the color range $0.55
@@
 \ref{tab:coeffs}.
-% \ref{tab:coeffs}.
-% \begin{equation}
-%     \mathrm{T_{eff}} = a + bC cC^2 + dC^3
-%     eC^4 + 15598.5C^5 -2694.76C^6 + 192.865C^7,
-% % Jason's updated parameters:
-% % −416.585
-% % 39780.0
-% % −84190.5
-% % 85203.9
-% % −48225.9
-% % 15598.5
-% % −2694.76
-% % 192.865
-%     % \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
-%     % 301C^6 - 26C^7,
-% % 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
-% %           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
-% %           301.13205924630165, -25.923997443169355]
-% \label{eq:curtis}
-% \end{equation}
-% where C is \gaia\ \gcolor and the coefficients are provided in table
-% \ref{coeffs}.
 \begin{table}[h!]
@@
 All stars above this line were removed from the sample.
-Subgiants were also removed by eliminating stars brighter than 4th magnitude
-in \gaia\ G-band.
+Subgiants were also removed by eliminating stars brighter than 4th absolute
+magnitude in \gaia\ G-band.
 
@@
 Given that lower-mass stars stay active for longer than higher-mass stars
-\citep[\eg][]{west2011, kiman2019}, and are therefore more likely to have
-measurable rotation periods at old ages, the latter scenario seems likely.
+\citep[\eg][]{west2008, newton2017, kiman2019}, and are therefore more likely
+to have measurable rotation periods at old ages, the latter scenario seems
+likely.
 However, it is also possible that the rotation periods of the oldest early M
@@
 that there is an upper limit to the rotation periods of {\it late} M dwarfs of
-around 140 days \citep{newton2016, newton2017, newton2018}, which is much
+around 140 days \citep{newton2016, newton2018}, which is much
 longer than the longest rotation periods measured in the \mct\ sample (around
 70 days).
-The apparent lack of old ages for M dwarfs in figure \ref{fig:age_gradient}
-may be caused by a combination of ages being underestimated by a poorly
-calibrated model, and rotation period detection bias.
-% For this reason, it is likely that slowly rotating M dwarfs are missing from
-% the \mct\ sample and this could be another reason for the lack of old M
-% dwarfs.
+The apparent lack of old gyro-ages for M dwarfs in figure
+\ref{fig:age_gradient} may be caused by a combination of ages being
+underestimated by a poorly calibrated model, and rotation period detection
+bias.
 The \citet{angus2019} gyrochronology relation is a simple polynomial model,
 fit to the period-color relation of Praesepe.
-% Since this relation predicts that the oldest stars in the \mct\ sample are
-% late-G and early-K dwarfs, it is probably under-predicting the ages of late-K
-% and early-M dwarfs.
-Inaccuracies are a typical feature of empirically calibrated gyrochronology
-models: since there are no (or at least very few) old M dwarfs with rotation
-periods, the models are poorly calibrated for these stars.
+Inaccuracies at low masses are a typical feature of empirically calibrated
+gyrochronology models: since there are no (or at least very few) old M dwarfs
+with rotation periods, the models are poorly calibrated for these stars.
 \begin{figure}
@@
   \centering
-    % \includegraphics[width=1\textwidth]{age_gradient_combo}
     \includegraphics[width=1\textwidth]{CMD_cuts_double}
@@
 included in the sample.
-
-% Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
-% and because latitude is correlated with stellar mass (lower mass stars are
-% older and tend to be preferentially located at high $b$, although this trends
-% is greatly reduced for stars cooler than 5000 K), we found that excluding
-% stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
-% stellar mass.
-% 15\degrees\ was found to be an adequate compromise between maintaining a close
-% relationship between \vb\ and \vz\ and retaining a sample large enough to
-% provide meaningful results.
-% Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
-% trends in the data, and intermediate cuts did not significantly change the
-% results.
-% We also found that the anti-correlation between stellar mass and $b$ is
-% negligable for stars cooler than 5000 K.
-% For this reason, and because the rotational evolution of old, G stars does not
-% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
-% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
-% dwarfs.
-
-% \subsection{Validating \vb\ dispersion as an age proxy}
-% \label{sec:mass-dependent-heating}
-
-% There are two main reasons why \vb\ velocity dispersion may not be a good age
-% proxy.
-% Firstly, mass-dependent heating may act on the sample, meaning that velocity
-% dispersion depends on both age and mass, so cannot be interpreted as a simple
-% age proxy.
-% Secondly, since stars in the \kepler\ field have a range of galactic
-% latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
-% stars, and introduce a velocity bias for high latitude stars (which are more
-% likely to be cooler and older).
-% In this section we demonstrate that neither of these problems seem to be a
-% significant issue for our data.
-
-% In order to establish whether \sigmavb\ can be used as an age proxy, we
-% searched for signs of mass-dependent heating within the \kepler\ field.
-% Mass-dependent dynamical heating may result from lower-mass stars experiencing
-% greater velocity changes when gravitationally perturbed than more massive
-% stars.
-% It has not been unambiguously observed in the galactic disk because of the
-% strong anti-correlation between stellar mass and stellar age.
-% Less massive stars do indeed have larger velocity dispersions, however they
-% are also older on average.
-% This mass-age degeneracy is highly reduced in M dwarfs because their
-% main-sequence lifetimes are longer than the age of the Universe, and no
-% evidence for mass-dependent heating has previously been found in M dwarfs
-% \citep[\eg][]{faherty2009, newton2016}.
-
-% To investigate whether mass-dependent heating could be acting on the \kepler\
-% sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
-% whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
-% the initial mass function.
-% We could not perform this analysis on the \mct\ sample, because it only
-% includes stars with {\it detectable} rotation periods, and since lower-mass
-% stars stay active for longer it is likely that it contains a strong mass-age
-% correlation.
-% We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
-% greater than 1.2 (corresponding to an effective temperature $\lesssim$
-% 4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
-% We also eliminated visual binaries by removing stars above a 6th order
-% polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
-% figure \ref{fig:age_gradient}).
-% We then applied the quality cuts described above in section
-% \ref{sec:the_data}.
-% To search for evidence of mass-dependent heating we calculated the (\vb)
-% velocity dispersion of stars in effective temperature bins.
-% Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
-% outliers before calculating the standard deviation of stars in each bin.
-% These high velocity outliers may be very old late K and M dwarfs, or they
-% result from using \vb\ instead of \vz, which introduces additional velocity
-% scatter.
-
-% Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
-% function of effective temperature \footnote{calculated by transforming
-% dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
-% \kepler\ dwarf sample.
-% Velocity dispersion very slightly {\it decreases} with decreasing temperature,
-% the opposite of the trend expected for mass-dependent heating, however the
-% slope is only inconsistent with zero at the 1.3 $\sigma$ level.
-% This trend may be due to a selection bias: cooler stars are fainter and
-% therefore typically closer, with smaller heights above the galactic plane and
-% smaller velocities.
-% The essential point however, is that we do not see evidence for mass-dependent
-% heating acting on stars in the \kepler\ field, indicating that velocity
-% dispersion {\it can} be used as an age proxy.
-% This analysis was performed using \vb\ but we also examined the {\it vertical}
-% velocities of the 537 stars in this sample with RV measurements.
-% Again, no evidence was found for mass-dependent heating: the slope of the
-% velocity dispersion-temperature relation was consistent with zero.
-% \begin{figure}
-%   \caption{
-%       Top: Stellar velocity (\vb) as a function of \teff\ for
-%       \kepler\ K and M dwarfs.
-% Vertical lines indicate different \teff-groupings used to calculate velocity
-%     dispersion.
-% Pink stars were not included in velocity dispersion calculations as they were
-%     either removed as outliers during a sigma clipping process, or they lie at
-%     the sparcely populated, extremely cool end of the temperature range.
-%     Velocity dispersion and \teff\ are slightly positively correlated, likely
-%     due to a brightness-related selection bias, indicating that mass-dependent
-%     heating does not significantly affect low-mass stars in the \kepler\
-%     field.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{vb_vs_teff}
-% \label{fig:vb_vs_teff}
-% \end{figure}
-
-% Having found no strong evidence for mass-dependent heating, we next tested
-% the validity of \vb\ as a proxy for \vz\ in more detail.
-% At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
-% of $b$, this equivalence becomes an approximation that grows noisier with $b$.
-% To test the validity of the \vb$\sim$\vz\ approximation over a range of
-% latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
-% (GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
-% We downloaded stars from four pointings in the \kepler\ field with galactic
-% latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
-% a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
-% The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
-% \vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
-% Galaxy.
-% We subtracted this offset and examined the residuals of the \vz\ -- \vb\
-% relationship to investigate the variance as a function of Galactic latitude.
-% We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
-% with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
-% The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
-% $, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
-% 20$^\circ$.
-
-% % Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
-% % over different Galactic latitudes.
-% % Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
-% % fits are shown as dashed blue lines.
-% % The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
-% % the variance increases with increasing Galactic latitude.
-% \begin{figure}
-%   \caption{
-% This figure demonstrates the variance in the relationship between \vb\ and
-%     \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
-% The panels show a kernel density estimator (KDE) (black solid line) for
-%     the \vz -- \vb\ residuals of stars in the GUMS simulation at four
-%     different Galactic latitudes.
-% Blue dashed lines show Gaussian fits to these KDEs.
-% The distributions are close to Gaussian, with slightly heavy tails.
-% The standard deviations of the Gaussian fits increase with Galactic latitude.
-% This figure illustrates how using \vb\ instead of \vz\ artificially
-%     increases velocity dispersion, especially at high latitudes.
-% }
-%   \centering
-%     \includegraphics[width=1\textwidth]{vb_vz}
-% \label{fig:vb_vz}
-% \end{figure}
-
-% Since we are concerned with velocity {\it dispersions}, rather than velocities
-% themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
-% function of temperature for
-% stars downloaded from the GUMS simulation.
-% For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
-% consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
-% latitudes the two quantities became significantly different.
-% For this reason we proceeded by only including stars with galactic latitudes
-% less than 15\degrees\ in our analysis.
-% Although it seems that the transformation between \vz\ and \vb\ does not {\it
-% strongly} affect our results, we cannot rule out the possibility that it
-% introduces systematic biases into the velocity dispersions we present here.
-% In \gaia\ DR3, RVs will be available for most stars in this sample, providing
-% an opportunity to validate (or correct) the results presented here.
-
-% % Since we use \vb, not \vz\ in our analysis, it is possible that the sample
-% % selection function could influence our results.
-% % For example, higher-mass stars tend to be younger and reside closer to the
-% % galactic disk mid-plane (low galactic latitudes).
-% % Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
-% % lower-mass stars typically have greater velocity dispersions due to .
-% % \teff\ and galactic latitude, $b$,
-% % could result in
-% % Stars at higher latitudes have additional velocity components in the ${\bf x}$
-% % and ${\bf y}$ directions, which could increase \vb\ but not \vz.
-% % Again however, since the relationship between \sigmavb\ and \teff\ is
-% % positively, not negatively correlated for cool stars in the \kepler\ field,
-% % this effect is probably too small to influence our results.
-
-% % Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
-% % strongly affect the \mct\ sample of rotating \kepler\ stars.
-% % For this reason, we assume that age difference is the major cause of velocity
-% % dispersion differences between groups.
-% % In other words, (\vb) velocity dispersion is a reliable age proxy for the
-% % \mct\ sample.
-
-% Because of the noisy relationship between \vb\ and \vz\, in this paper we do
-% not attempt to convert velocity dispersion (\sigmavb) into an age via an
-% age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
-% Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
-% careful analysis that includes formal modeling of the \vb\ -- \vz\
-% relationship will be needed to calculate absolute ages.
-% % For example, the velocity distributions could be modeled as a mixture of
-% % Gaussians in order to account for the additional velocity dispersion caused by
-% % the \vz-\vb\ transformation.
-% The RVs of most of these stars will become available in \Gaia\ DR3, allowing
-% calculations of \vz, which can be used to calculate more reliable ages via an
-% AVR.
-
-% % We then selected groups of stars within different gyrochronal age ranges and
-% % calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
-% % of effective temperature for each age group.
-% % Ages were calculated using dereddened \gaia\ \gcolor\ color, however
-% % throughout this paper we show rotation periods as a function of effective
-% % temperature, \teff, because it is easier to divide stars into bins of roughly
-% % equal numbers in \teff-space than in color-space.

### 2020-03-15 paper/results.tex
more changes

@@
 similar color in figure \ref{fig:vplot} are similar ages.
-We discuss this assumption further in the appendix.
+In the appendix of this paper, we show that this assumption appears valid for
+stars with Galactic latitude $<$ 15\degrees.
 \begin{figure}
@@
     velocity dispersion.
-    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
-    and bottom panels.
+    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr (Solar age)
+    in both top and bottom panels.
     Bottom: Same as top panel with rotation period vs {\it mass}
@@
 periods over time.
-These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
+These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 (Solar age)
+Gyr.
 At the youngest ages, these gyrochrones describe the data well: the palest
@@
 function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
-gyrochrones from the \citep{spada2019} model shown in white.
+gyrochrones from the \citet{spada2019} model shown in white.
 These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
@@
 intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
-mass, and at late ages rotation period {\it increases} with {\it increaasing}
+mass, and at late ages rotation period {\it increases} with {\it increasing}
 mass.
@@
     gyrochronology models, with the scheme shown in the legend.
-    Only stars cooler than 5000 K are plotted in
-    the bottom panel in order to isolate populations above and below the
-    period gap, which only extends up to temperatures of $\sim$4600 K.
     Bottom: the velocities of these groups of stars (in the direction of
     Galactic latitude, $b$) are shown as a function of rotation period.
+    Only stars cooler than 5000 K are plotted in the bottom panel in order to
+    isolate populations above and below the period gap, which only extends up
+    to temperatures of $\sim$4600 K.
     The black line indicates the velocity standard deviation as a function of
@@
 isolate populations above and below the period gap, which only extends up to a
-temperature of $\sim$4600 K in our sample \citep[Although][found that the gap
-extends to temperatures as hot as 6000 K]{davenport2017}.
+temperature of $\sim$4600 K in our sample, although \citet{davenport2017}
+found that the gap extends to temperatures as hot as 6000 K.
 In general, velocity dispersion increases with rotation period because both

### 2020-03-16 paper/conclusion.tex
adding jen's small comments and a core-envelope coupling section

@@
 
-\item{{\bf \sigmavb\ can be used as an age proxy for (low Galactic latitude)
-    \kepler\ stars.} In the appendix of this paper we demonstrate that the
-        dispersion of velocities in the direction of Galactic latitude, \vb,
-        can be used as an age proxy, by showing that there is no strong
-        evidence for mass-dependent heating in low-mass \kepler\ dwarfs: the
-        velocity dispersions of K and M dwarfs, whose main-sequence lifetimes
-        are longer than around 11 Gyrs, do not appear to increase with
-        decreasing mass.
-Although {\it vertical} velocity, \vz, is a quantity that has been
-demonstrated to trace time-dependent orbital heating in the disc of the
-Galaxy, most stars with measured rotation periods do not yet have radial
-velocities, so we used velocity in the direction of Galactic latitude, \vb\,
-as a proxy for \vz.
-        Using stars in the GUMS simulation \citep{robin2012}, we showed that
-        using \vb\ as a proxy for \vz\ introduces an additional
-        velocity scatter, which increases with increasing Galactic
-        latitude.
-However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
-confirmed that using \vb\ instead of \vz\ does not introduce any
-mass-dependent velocity dispersion bias into the sample.
-We therefore confirmed that \vb\ velocity dispersion can be used to accurately
-rank stars by age.}
+% \item{{\bf \sigmavb\ can be used as an age proxy for (low Galactic latitude)
+%     \kepler\ stars.} In the appendix of this paper we demonstrate that the
+%         dispersion of velocities in the direction of Galactic latitude, \vb,
+%         can be used as an age proxy, by showing that there is no strong
+%         evidence for mass-dependent heating in low-mass \kepler\ dwarfs: the
+%         velocity dispersions of K and M dwarfs, whose main-sequence lifetimes
+%         are longer than around 11 Gyrs, do not appear to increase with
+%         decreasing mass.
+% Although {\it vertical} velocity, \vz, is a quantity that has been
+% demonstrated to trace time-dependent orbital heating in the disc of the
+% Galaxy, most stars with measured rotation periods do not yet have radial
+% velocities, so we used velocity in the direction of Galactic latitude, \vb\,
+% as a proxy for \vz.
+%         Using stars in the GUMS simulation \citep{robin2012}, we showed that
+%         using \vb\ as a proxy for \vz\ introduces an additional
+%         velocity scatter, which increases with increasing Galactic
+%         latitude.
+% However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
+% confirmed that using \vb\ instead of \vz\ does not introduce any
+% mass-dependent velocity dispersion bias into the sample.
+% We therefore confirmed that \vb\ velocity dispersion can be used to accurately
+% rank stars by age.}
 

### 2020-03-16 paper/intro.tex
more changes
adding jen's small comments and a core-envelope coupling section

@@
 strong magnetic fields and slowly lose angular momentum via magnetic braking
-\citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
+\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
 pinsonneault1989}.
@@
 efficient magnetic braking.
-However, in the NGC 6811 cluster which is around 1.1 Gyr \citep{janes2011},
+However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
+sandquist2016},
 late-K dwarfs rotate at the {\it same rate} as early-K dwarfs

### 2020-03-16 paper/method.tex
adding jen's small comments and a core-envelope coupling section

@@
 All stars above this line were removed from the sample.
-Subgiants were also removed by eliminating stars brighter than 4th absolute
-magnitude in \gaia\ G-band.
+Potential subgiants were also removed by eliminating stars brighter than 4th
+absolute magnitude in \gaia\ G-band.
+This cut also removed a number of main sequence F stars from our sample,
+however these hot stars are not the focus of our gyrochronology study since
+their small convective zones inhibit the generation of a strong magnetic
+field.
+The removal of photometric binaries and evolved/hot stars reduced the total
+sample of around 34,000 stars by almost 10,000.
 
@@
 covariances between \gaia\ positions, parallaxes and proper motions.
-Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
-stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
-greater than 1 \kms\, and stars with galactic latitudes greater than
-15\degrees\ (justification provided in the appendix) were removed from the
-sample.
-Finally, we removed stars with rotation periods shorter than the main
-population of periods, since this area of the period-\teff\ diagram is
+Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
+(around 3,000 stars), stars fainter than 16th magnitude (200 stars), stars
+with absolute \vb\ uncertainties greater than 1 \kms\ (1000 stars), and stars
+with galactic latitudes greater than 15\degrees\ (5500 stars, justification
+provided in the appendix) were removed from the sample.
+Finally, we removed almost 2000 stars with rotation periods shorter than the
+main population of periods, since this area of the period-\teff\ diagram is
 sparsely populated.

### 2020-03-16 paper/results.tex
adding jen's small comments and a core-envelope coupling section

@@
     sample, colored by the velocity dispersions of stars calculated over a
-    grid in $\log_{10}$(period) and \teff.
+    grid in $\log_{10}$(period) and \teff\ (this grid causes the quantized
+    appearance).
     Black lines show gyrochrones from a gyrochronology model that projects the
@@
     stars at long rotation periods/old ages: they trace lines of constant
-    velocity dispersion by reproducing periods of `stalled' rotational
+    velocity dispersion by reproducing periods of `stalled' surface rotational
     evolution for K-dwarfs.
@@
 temperature appears to flatten out and eventually invert.
-These results provide further evidence for `stalled' rotational evolution of K
-dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
-by models that vary angular momentum transport between stellar core and
-envelope with time and mass \citep{spada2019}.
+These results provide further evidence for `stalled' surface rotational
+evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
+and reproduced by models that vary angular momentum transport between stellar
+core and envelope with time and mass \citep{spada2019}.
 The velocity dispersions of stars in the \mct\ sample provide the following
@@
 
-It has been demonstrated that lower-mass stars remain magnetically active
-longer than more massive stars, \citep[\eg][]{west2008, kiman2019}.
+A period of core-envelope decoupling in the evolution of cool dwarfs has been
+explored in theoretical models for decades \citep[\eg][]{endal1981,
+macgregor1991, denissenkov2010, gallet2013}.
+In such models, the angular momenta of the radiative core and convective
+envelope are permitted to evolve separately once a radiative core develops on
+the pre-main sequence.
+A core-envelope decoupling phase in the early angular momentum history of
+stars is required in order to reproduce observations of young clusters and
+star forming regions \citep[\eg][]{irwin2007, bouvier2008, denissenkov2010,
+spada2011, reiners2012} and has become an established element of theoretical
+gyrochronology models.
+The \citet{lanzafame2015} gyrochronology model quantified the mass-dependence
+of the core-envelope coupling timescale and this model was fit to open-cluster
+observations in \citet{spada2019}, including new rotation period measurements
+for stars down to M$\sim$0.6M$_\odot$ in the NGC 6811 cluster
+\citep{curtis2019}.
+A similar mass-dependent core-envelope coupling timescale was also found to
+explain the observed lithium depletion trends in open clusters by an
+independent study \citep{somers2016}.
+Although variable angular momentum transport between the surfaces and cores of
+stars has been an essential ingredient of stellar evolution models for
+decades, the transport mechanism is still unknown.
+Among the proposed mechanisms are magneto-hydrodynamical waves resulting from
+various magnetic field geometries, and gravity waves \citep[see,
+\eg][]{charbonneau1993, ruediger1996, spruit2002, talon2003, spada2010,
+brun2011, oglethorpe2013}.
+
+Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
+K and M dwarfs remain magnetically active for longer than G dwarfs.
+The mass dependence of magnetic activity lifetimes has been demonstrated
+previously \citep[\eg][]{west2008, newton2017, kiman2019}.
 If the detectability of a rotation period is considered to be a magnetic

### 2020-03-16 paper/abstract.tex
small updates

@@
 It also aligns with recent observations of stars in the NGC 6811 cluster,
-which indicate that K dwarfs experience an epoch of stalled spin-down.
+which indicate that the surface rotation rates of K dwarfs go through an epoch
+of inhibited evolution.
 We find that the oldest \kepler\ stars with measured rotation periods are

### 2020-03-16 paper/intro.tex
small updates

@@
 \citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
-meibom2015, curtis2019} have revealed that rotational evolution is more
-complicated than previously thought.
+meibom2015, curtis2019} confirm that rotational evolution is a highly complex
+process.
 For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster

### 2020-03-16 paper/results.tex
small updates

@@
 stronger magnetic fields, larger Alfv\'en radii and therefore experience
-greater angular momentum loss rate \citep[\eg][]{schatzman1962, parker1970,
-kawaler1988, charbonneau2010}.
-According to the \citet{spada2019} model, there is minimal transportation of
-angular momentum from the surface to the core of the star at these young ages,
-so the surface slows down but the core keeps spinning rapidly.
+greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
+parker1970, kawaler1988, charbonneau2010}.
 This aligns with the current assumptions about stellar spin-down, dynamo
@@
 \citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
-According to the data presented in figure \ref{fig:vplot} however, at
-intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
-mass, and at late ages rotation period {\it increases} with {\it increasing}
-mass.
+According to the \citet{spada2019} model, the radiative cores and convective
+envelopes of stars are decoupled at these young ages, \ie\ transportation of
+angular momentum from the surface to the core of the star is reduced, so the
+surface slows down due to wind-braking but the core keeps spinning rapidly.
+According to the data presented in figure \ref{fig:vplot}, at intermediate
+ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
+late ages rotation period {\it increases} with {\it increasing} mass.
 The interpretation of this, according to the \citet{spada2019} model, is that

### 2020-03-16 paper/abstract.tex
small edits from a big re-read

@@
 stars.
-In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
-an age proxy, to reveal their rotational evolution and demonstrate that
-kinematics could be a useful tool for calibrating gyrochronology.
+In this work, we use the velocity dispersions of groups of low-mass \kepler\
+dwarfs as an age proxy, to reveal the temperature-dependent rotational
+evolution of these stars, and demonstrate that kinematic ages could be a
+useful tool for calibrating gyrochronology in future.
 We find that a linear gyrochronology model, calibrated to fit the

### 2020-03-16 paper/intro.tex
small edits from a big re-read

@@
 \citep{curtis2019}.
-In other words, convection zone depth cannot be the only variable that affects
-stellar spin-down rate.
+% In other words, convection zone depth cannot be the only variable that affects
+% stellar spin-down rate.
 New semi-empirical models that vary the rate of angular momentum
@@
 surface rotation rates.
+
 Another example of unexpected rotational evolution is seen in old field stars
@@
 
-\subsection{Using kinematics as an age proxy}
+\subsection{Using velocity dispersion as an age proxy}
 
@@
 significantly on stellar age.
-While the velocity of any individual star only provides a weak age constraint,
+While the velocity of any individual star only provides a weak age constrain
 because its velocity depends on its current position in its orbit, the

### 2020-03-16 paper/method.tex
small edits from a big re-read

@@
 Inaccuracies at low masses are a typical feature of empirically calibrated
-gyrochronology models: since there are no (or at least very few) old M dwarfs
-with rotation periods, the models are poorly calibrated for these stars.
+gyrochronology models since there are no (or at least very few) old M dwarfs
+with rotation periods and the models are poorly calibrated for these stars.
 \begin{figure}

### 2020-03-16 paper/results.tex
small edits from a big re-read

@@
 To explore the relationship between rotation period, effective temperature
-(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb
-was calculated as 1.5$\times$ the median absolute deviation, to mitigate
-sensitivity to outliers.} for groups of stars with similar rotation periods
-and temperatures, and presumed similar age.
+(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb\
+was calculated as 1.5$\times$ the median absolute deviation of velocities, to
+mitigate sensitivity to outliers.} for groups of stars with similar rotation
+periods and temperatures, and presumed similar age.
 The top panel of figure \ref{fig:vplot} shows rotation period versus effective
@@
 core and envelope with time and mass \citep{spada2019}.
-The velocity dispersions of stars in the \mct\ sample provide the following
-picture of rotational evolution.
+
+The velocity dispersions of stars in the \mct\ sample, shown in figure
+\ref{fig:vplot}, provide the following picture of rotational evolution.
 At young ages \citep[younger than around 1 Gyr but still old enough to be on
@@
 the pre-main sequence.
-A core-envelope decoupling phase in the early angular momentum history of
-stars is required in order to reproduce observations of young clusters and
-star forming regions \citep[\eg][]{irwin2007, bouvier2008, denissenkov2010,
-spada2011, reiners2012} and has become an established element of theoretical
-gyrochronology models.
-The \citet{lanzafame2015} gyrochronology model quantified the mass-dependence
-of the core-envelope coupling timescale and this model was fit to open-cluster
-observations in \citet{spada2019}, including new rotation period measurements
-for stars down to M$\sim$0.6M$_\odot$ in the NGC 6811 cluster
-\citep{curtis2019}.
+A decoupled core and envelope is required to reproduce observations of young
+clusters and star forming regions \citep[\eg][]{irwin2007, bouvier2008,
+denissenkov2010, spada2011, reiners2012} and has become an established element
+of theoretical gyrochronology models.
+During this phase, angular momentum transport between radiative core and
+convective envelope is reduced.
+Over time, models increase the efficiency of angular momentum transport
+between the core and envelope in order to reproduce the close-to solid body
+rotation observed for the Sun \citep[\eg][]{thompson1996}.
+The core-envelope coupling timescale affects the predicted surface rotation
+periods of young and intermediate-age stars and is usually constrained using
+observations of open clusters.
+The \citet{lanzafame2015} gyrochronology model uses a mass-dependent
+core-envelope coupling timescale, and \citet{spada2019} fit this model to
+open-cluster observations, including new rotation period measurements for
+K dwarfs in the NGC 6811 cluster \citep{curtis2019}.
 A similar mass-dependent core-envelope coupling timescale was also found to
@@
 brun2011, oglethorpe2013}.
+% These different processes may make subtly different predictions for the
+% rotational evolution of stars, and could potentially be tested with the large
+% number of field stars with measured ages and rotation periods which could be
+% provided by kinematic ages for the \mct\ sample.
 
@@
 The mass dependence of magnetic activity lifetimes has been demonstrated
-previously \citep[\eg][]{west2008, newton2017, kiman2019}.
-If the detectability of a rotation period is considered to be a magnetic
-activity proxy, then our results provide further evidence for a mass-dependent
-activity lifetime.
+previously \citep[\eg][]{west2008, newton2017, kiman2019}, and if the
+detectability of a rotation period is considered to be a magnetic activity
+proxy, then our results provide further evidence for a mass-dependent activity
+lifetime.
 Figure \ref{fig:vplot} shows that the groups of stars with the largest
@@
 stars in the \mct\ sample.
-If the rotation periods of G stars are only detectable when they are young,
-G stars with measured periods should have smaller velocity dispersions than
-the overall \kepler\ field.
+If the rotation periods of G stars are only detectable when they are young, G
+stars with measured periods should have smaller velocity dispersions than G
+stars in the overall \kepler\ field.
 We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
@@
 at older ages than G dwarf rotation periods.
-G stars become magnetically inactive and have fewer active surface regions
-{\it at a younger age than M dwarfs}.
+In other words, G stars become magnetically inactive and have fewer active
+surface regions {\it at a younger age than M dwarfs}.
 
@@
 panel of figure \ref{fig:vplot}, whose origin is unknown and is the subject of
-much speculation \citep{mcquillan2014, davenport2018, reinhold2019}.
+much speculation \citep{mcquillan2014, davenport2017, davenport2018,
+reinhold2019, reinhold2020}.
 This gap was first identified by \mct, and roughly follows a line of constant
@@
 where rotational variability is reduced and rotation periods cannot be
-measured \citep{reinhold2019}.
+measured \citep{reinhold2019, reinhold2020}.
 
@@
 quantities increase with age.
-There is a smooth transition in velocity dispersion between stars with
-rotation periods below and above the gap (orange squares to blue circles),
-suggesting that these groups are part of the same Galactic population.
 Previously, only the overall velocity dispersions of all stars above and below
@@
 to two distinct populations \citep{mcquillan2014}.
-The smooth increase in velocity dispersion across the gap shown here suggests
-that stars above and below are part of the same stellar population.
-However, it does not rule out the possibility that a brief cessation of star
-formation in the Solar neighborhood, around one Gyr ago, may have caused this
-gap.
+However, figure \ref{fig:gap} shows a smooth increase in velocity dispersion
+with rotation period across the gap (from orange squares to blue circles),
+suggesting that these groups are part of the same Galactic population.
+% The smooth increase in velocity dispersion across the gap shown here suggests
+% that stars above and below are part of the same stellar population.
+This observation does not rule out the possibility that a brief cessation of
+star formation in the Solar neighborhood, around one Gyr ago, may have caused
+this gap, however.
 

### 2020-03-16 paper/results.tex
typo

@@
     Velocity dispersions for the entire \kepler\ field divided by the velocity
-    dispersion of stars with measured rotation periods in \mct,
+    dispersions of stars with measured rotation periods in \mct,
     as a function of effective temperature.
-    A larger ratio means that the overall \kepler\ field is older, on average,
-    than stars in the \mct\ catalog.
+    A larger ratio indicates that the overall \kepler\ field is older, on
+    average, than stars in the \mct\ catalog.
     As this ratio approaches unity the two populations have similar kinematic

### 2020-03-18 paper/one_big_file.tex
adding one_big_file.tex

% zip all_in_one.zip one_big_file.tex CMD_cuts_double.pdf main_figure.pdf gap.pdf vb_vs_teff.pdf vb_vz.pdf field_comparison.pdf bib.bib ms.pdf

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}

\AtBeginDocument{\let\textlabel\label}

\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\etal}{{\it et al.}}

\newcommand{\kepler}{{\it Kepler}}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\tess}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\lsst}{{\it LSST}}
\newcommand{\Wfirst}{{\it WFIRST}}
\newcommand{\wfirst}{{\it WFIRST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\plato}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\panstarrs}{{\it PanSTARRS}}
\newcommand{\LAMOST}{{\it LAMOST}}

\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\prot}{$P_{\mathrm{rot}}$}
\newcommand{\pmega}{$\bar{\omega}$}
\newcommand{\mj}{$m_j$}
\newcommand{\mh}{$m_h$}
\newcommand{\mk}{$m_k$}
\newcommand{\mx}{$m_x$}
\newcommand{\logg}{log(g)}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\degrees}{$^\circ$}
\newcommand{\vz}{$v_{\bf z}$}
\newcommand{\vb}{$v_{\bf b}$}
\newcommand{\kms}{kms$^{-1}$}
\newcommand{\sigmavb}{$\sigma_{v{\bf b}}$}
\newcommand{\sigmavz}{$\sigma_{v{\bf z}}$}

\newcommand{\tmin}{3500}
\newcommand{\tmax}{5500}

% \newcommand{\nsim_stars}{841}

\newcommand{\amnh}{1}
\newcommand{\cca}{2}
\newcommand{\columbia}{3}
\newcommand{\cfa}{4}
\newcommand{\dartmouth}{5}
\newcommand{\hawaii}{6}
\newcommand{\cuny}{7}
\newcommand{\carnegie}{8}

\newcommand{\sd}{{\tt stardate}}
\newcommand{\gcolor}{$G_{BP} - G_{RP}$}
\newcommand{\mcp}{\citep{mcquillan2014}}
\newcommand{\mct}{\citet{mcquillan2014}}
\newcommand{\bvector}{${\bf b}$}
\newcommand{\python}{{\it Python}}

\newcommand{\racomment}[1]{{\color{blue}#1}}

\begin{document}

\title{Exploring the evolution of stellar rotation using Galactic kinematics}

\author{%
    Ruth Angus$^{1, 2, 3}$,  % signed off
    Angus Beane$^4$,  % signed off
    Adrian M. Price-Whelan$^2$,  % signed off
    Jason Curtis$^1$,  % signed off
    Travis Berger$^6$,  % signed off
    Elisabeth Newton$^5$,  % signed off
    Jennifer van Saders$^6$,  % signed off
    Rocio Kiman$^{7, 1}$,  % signed off
    Dan Foreman-Mackey$^2$,  % signed off
    Yuxi Lu$^{3, 1}$,  % signed off
    Lauren Anderson$^8$}  % signed off

\altaffiltext{\amnh}{Department of Astrophysics, American Museum of Natural
History, 200 Central Park West, Manhattan, NY, USA}
\altaffiltext{\cca}{Center for Computational Astrophysics, Flatiron Institute,
162 5th Avenue, Manhattan, NY, USA}
\altaffiltext{\columbia}{Department of Astronomy, Columbia
University, Manhattan, NY, USA}
\altaffiltext{\cfa}{Center for Astrophysics {\normalfont |} Harvard \&
Smithsonian, 60 Garden Street, Cambridge, MA 02138, USA}
\altaffiltext{\dartmouth}{Department of Physics and Astronomy, Dartmouth
College, Hanover, NH, USA}
\altaffiltext{\hawaii}{Institute for Astronomy, University of Hawai`i, 2680
Woodlawn Drive, Honolulu, HI 96822, USA}
\altaffiltext{\cuny}{Department of physics, CUNY Graduate Center, City
University of New York, Manhattan, NY, USA}
\altaffiltext{\carnegie}{Observatories of the Carnegie Institution for
Science, 813 Santa Barbara Street, Pasadena, CA 91101, USA}

\begin{abstract}
The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of groups of low-mass \kepler\
dwarfs as an age proxy, to reveal the temperature-dependent rotational
evolution of these stars, and demonstrate that kinematic ages could be a
useful tool for calibrating gyrochronology in future.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
they are young, at old ages we find that late-G and early-K dwarfs rotate at
the {\it same rate} or faster than late-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
GKM dwarfs are likely to be synchronized binaries.
\end{abstract}

\section{Introduction}

\subsection{Gyrochronology}

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} confirm that rotational evolution is a highly complex
process.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
% In other words, convection zone depth cannot be the only variable that affects
% stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.

Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

The rotational evolution of stars is clearly a complicated process and, to
fully calibrate the gyrochronology relations we need a large sample of
reliable ages for stars spanning a range of ages and masses.
In this paper, we use the velocity dispersions of field stars to qualitatively
explore the rotational evolution of GKM dwarfs, and show that kinematics could
provide a gyrochronology calibration sample.

\subsection{Using velocity dispersion as an age proxy}

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the Galaxy with a low out-of-plane, or vertical, velocity (\vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of older stars is observed to be larger
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
This is likely either a signature of dynamical heating, such as from
interactions with giant molecular clouds, spiral arms and the galactic bar
\citep[see][for a review of secular evolution in the MW]{sellwood2014}, or an
indication that stars formed dynamically ``hotter'' in the past
\citep[e.g.,][]{bird2013}.
In either case, the vertical velocity distribution is observed to depend
significantly on stellar age.
While the velocity of any individual star only provides a weak age constrain
because its velocity depends on its current position in its orbit, the
velocity {\it dispersion} of a {\it group} of stars indicates whether that
group is old or young relative to other groups.
% Vertical velocity dispersion is thought to increase monotonically with age.
In this work, we compare the velocity dispersions of groups of field stars in
the Galactic thin disk to ascertain which groups are older and which younger,
and draw conclusions about the rotational evolution of stars based on their
implied relative ages.

Although {\it vertical} velocity, \vz, is an established age proxy, it can
only be calculated with full 6-dimensional position and velocity information.
In fact, with full 6D phase space and an assumed Galactic potential, it is
possible to calculate the dynamically-invariant vertical {\it action}, which may
be an even better age indicator \citep{beane2018, ting2019}.
Unfortunately, most field stars with measured rotation periods do not have
radial velocity (RV) measurements because they are relatively faint \kepler\
targets ($\sim$12th-16th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb,
as a proxy for \vz.
The \kepler\ field is positioned at low galactic latitude
(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see
appendix) approximation to \vz.
Because we use \vb\ rather than \vz\, we do not calculate absolute kinematic
ages using a published age--velocity dispersion relation (AVR), calibrated with
vertical velocity.
In the future it may be possible to account for the differences between \vb\
and \vz, or marginalize over missing RV measurements and the \kepler\
selection function, in order to infer the absolute ages of groups of stars.
Regardless of direction however, velocity dispersion is expected to
monotonically increase over time \citep[\eg][]{holmberg2009}, and can
therefore be used to {\it rank} groups of stars by age.

This paper is laid out as follows: in section \ref{sec:method} we describe our
sample selection process and the methods used to calculate stellar
velocities.
In section \ref{sec:results} we use kinematics to investigate the relationship
between stellar rotation period, age and color/\teff\ and interpret the
results.
We also examine the rotation period gap and the kinematics of synchronized
binaries.
In the appendix, we establish that \vb\ velocity dispersion, \sigmavb, can be
used as an age proxy by demonstrating that neither mass-dependent heating nor
the \kepler/\gaia\ selection function is observed to strongly affect our
sample.

\section{Method}
\label{sec:method}

\subsection{The data}
\label{sec:the_data}

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog of parallaxes, proper motions and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.

For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
temperature measurements for nearby, unreddened field stars in benchmark
samples, including FGK stars characterized with high-resolution optical
spectroscopy \citep{brewer2016}, M dwarfs characterized with low-resolution
optical and near-infrared spectroscopy \citep{mann2015}, and K and M dwarfs
characterized with interferometry and bolometric flux analyses
\citep{boyajian2012}.
This empirical color--temperature relation is valid over the color range $0.55
< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $6470 < T_{\rm eff} <
3070$~K.
The dispersion about the relation implies a high precision of 50~K.
These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
at all temperatures\footnote{The color--temperature relation is described in
detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
\etal\ (2020, in prep).}.
The equation we used to calculate photometric temperatures from Gaia \gcolor\
color is a seventh-order polynomial with coefficients given in table
\ref{tab:coeffs}.
\begin{table}[h!]
  \begin{center}
      \caption{
          Coefficient values for the 7th-order polynomial used to estimate
      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
      prep).}
    \label{tab:coeffs}
    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
        (\gcolor\ ) exponent & Coefficient  \\
      \hline
      $0$ & -416.585 \\
      $1$ & 39780.0  \\
      $2$ & -84190.5 \\
      $3$ & 85203.9  \\
      $4$ & -48225.9 \\
      $5$ & 15598.5  \\
      $6$ & -2694.76 \\
      $7$ & 192.865  \\
    \end{tabular}
  \end{center}
\end{table}

Photometric binaries and subgiants were removed from the \mct\ sample by
applying cuts to the color-magnitude diagram (CMD), shown in figure
\ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and photometric binaries (shown
as the curved dashed line in figure \ref{fig:age_gradient}).
All stars above this line were removed from the sample.
Potential subgiants were also removed by eliminating stars brighter than 4th
absolute magnitude in \gaia\ G-band.
This cut also removed a number of main sequence F stars from our sample,
however these hot stars are not the focus of our gyrochronology study since
their small convective zones inhibit the generation of a strong magnetic
field.
The removal of photometric binaries and evolved/hot stars reduced the total
sample of around 34,000 stars by almost 10,000.

The rotation periods of the dwarf stars in the \mct\ sample are shown on a
\gaia\ color-magnitude diagram (CMD) in the top panel of figure
\ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2008, newton2017, kiman2019}, and are therefore more likely
to have measurable rotation periods at old ages, the latter scenario seems
likely.
However, it is also possible that the rotation periods of the oldest early M
dwarfs are so long that they are not measurable with Kepler data.
Ground-based rotation period measurements of mid and late M dwarfs indicate
that there is an upper limit to the rotation periods of {\it late} M dwarfs of
around 140 days \citep{newton2016, newton2018}, which is much
longer than the longest rotation periods measured in the \mct\ sample (around
70 days).
The apparent lack of old gyro-ages for M dwarfs in figure
\ref{fig:age_gradient} may be caused by a combination of ages being
underestimated by a poorly calibrated model, and rotation period detection
bias.
The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
Inaccuracies at low masses are a typical feature of empirically calibrated
gyrochronology models since there are no (or at least very few) old M dwarfs
with rotation periods and the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
(around 3,000 stars), stars fainter than 16th magnitude (200 stars), stars
with absolute \vb\ uncertainties greater than 1 \kms\ (1000 stars), and stars
with galactic latitudes greater than 15\degrees\ (5500 stars, justification
provided in the appendix) were removed from the sample.
Finally, we removed almost 2000 stars with rotation periods shorter than the
main population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.

\section{Results and Discussion}
\label{sec:results}

\subsection{The period-\teff\ relations, revealed}
\label{sec:the_reveal}

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb\
was calculated as 1.5$\times$ the median absolute deviation of velocities, to
mitigate sensitivity to outliers.} for groups of stars with similar rotation
periods and temperatures, and presumed similar age.
The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
In the appendix of this paper, we show that this assumption appears valid for
stars with Galactic latitude $<$ 15\degrees.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff\ (this grid causes the quantized
    appearance).
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr (Solar age)
    in both top and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' surface rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}

Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age, as expected.
This result is insensitive to the choice of bin position and size.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 (Solar age)
Gyr.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones they would follow lines of constant velocity
dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a
more flattened, or even inverted, shape in rotation period-\teff\ space than
these Praesepe-based models.

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citet{spada2019} model shown in white.
These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduce the apparent
flattening and inversion in the rotation period-\teff/mass relations.

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' surface rotational
evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
and reproduced by models that vary angular momentum transport between stellar
core and envelope with time and mass \citep{spada2019}.

The velocity dispersions of stars in the \mct\ sample, shown in figure
\ref{fig:vplot}, provide the following picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010}.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

A period of core-envelope decoupling in the evolution of cool dwarfs has been
explored in theoretical models for decades \citep[\eg][]{endal1981,
macgregor1991, denissenkov2010, gallet2013}.
In such models, the angular momenta of the radiative core and convective
envelope are permitted to evolve separately once a radiative core develops on
the pre-main sequence.
A decoupled core and envelope is required to reproduce observations of young
clusters and star forming regions \citep[\eg][]{irwin2007, bouvier2008,
denissenkov2010, spada2011, reiners2012} and has become an established element
of theoretical gyrochronology models.
During this phase, angular momentum transport between radiative core and
convective envelope is reduced.
Over time, models increase the efficiency of angular momentum transport
between the core and envelope in order to reproduce the close-to solid body
rotation observed for the Sun \citep[\eg][]{thompson1996}.
The core-envelope coupling timescale affects the predicted surface rotation
periods of young and intermediate-age stars and is usually constrained using
observations of open clusters.
The \citet{lanzafame2015} gyrochronology model uses a mass-dependent
core-envelope coupling timescale, and \citet{spada2019} fit this model to
open-cluster observations, including new rotation period measurements for
K dwarfs in the NGC 6811 cluster \citep{curtis2019}.
A similar mass-dependent core-envelope coupling timescale was also found to
explain the observed lithium depletion trends in open clusters by an
independent study \citep{somers2016}.
Although variable angular momentum transport between the surfaces and cores of
stars has been an essential ingredient of stellar evolution models for
decades, the transport mechanism is still unknown.
Among the proposed mechanisms are magneto-hydrodynamical waves resulting from
various magnetic field geometries, and gravity waves \citep[see,
\eg][]{charbonneau1993, ruediger1996, spruit2002, talon2003, spada2010,
brun2011, oglethorpe2013}.
% These different processes may make subtly different predictions for the
% rotational evolution of stars, and could potentially be tested with the large
% number of field stars with measured ages and rotation periods which could be
% provided by kinematic ages for the \mct\ sample.

Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
K and M dwarfs remain magnetically active for longer than G dwarfs.
The mass dependence of magnetic activity lifetimes has been demonstrated
previously \citep[\eg][]{west2008, newton2017, kiman2019}, and if the
detectability of a rotation period is considered to be a magnetic activity
proxy, then our results provide further evidence for a mass-dependent activity
lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
To investigate this idea further, we compared the velocity dispersions of
stars with measured rotation periods, to the velocity dispersions of the
overall \kepler\ sample.
We calculated the velocity dispersions for all stars in the \kepler\ field,
after removing visual binaries, subgiants, stars fainter than 16th magnitude,
and high Galactic latitude stars, following the method described in section
\ref{sec:method}.
We then compared these velocity dispersions to the velocity dispersions of
stars in the \mct\ sample.
If the rotation periods of G stars are only detectable when they are young, G
stars with measured periods should have smaller velocity dispersions than G
stars in the overall \kepler\ field.
We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
\sigmavb\ for stars with rotation periods published in \mct, as a function of
\teff\ (see figure \ref{fig:compare}).
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods.
In other words, G stars become magnetically inactive and have fewer active
surface regions {\it at a younger age than M dwarfs}.

\begin{figure}
  \caption{
    Velocity dispersions for the entire \kepler\ field divided by the velocity
    dispersions of stars with measured rotation periods in \mct,
    as a function of effective temperature.
    A larger ratio indicates that the overall \kepler\ field is older, on
    average, than stars in the \mct\ catalog.
    As this ratio approaches unity the two populations have similar kinematic
    ages.
    The large ratio for the hottest stars indicates that G dwarfs become
    inactive at young ages.
    This ratio approaches unity at low temperatures, showing that K and early
    M dwarf rotation periods are measurable over a large range of ages.
}
  \centering
    \includegraphics[width=1\textwidth]{field_comparison}
\label{fig:compare}
\end{figure}

\subsection{Synchronized binaries and the \kepler\ period gap}
\label{sec:gap}

\begin{figure}
  \caption{
      Top: rotation period vs. effective temperature for stars in the \mct\
    sample, separated into three groups. Blue circles
      show stars with rotation periods longer than the
    period gap, orange squares show stars with rotation periods shorter than
    the gap, but longer than the lower edge of the main rotation period
    distribution, and green triangles show stars with rotation periods shorter
    than this lower edge.
    Stars were separated into these three groups using \citet{angus2019}
    gyrochronology models, with the scheme shown in the legend.
    Bottom: the velocities of these groups of stars (in the direction of
    Galactic latitude, $b$) are shown as a function of rotation period.
    Only stars cooler than 5000 K are plotted in the bottom panel in order to
    isolate populations above and below the period gap, which only extends up
    to temperatures of $\sim$4600 K.
    The black line indicates the velocity standard deviation as a function of
    period.
}
  \centering
    \includegraphics[width=1\textwidth]{gap}
\label{fig:gap}
\end{figure}

In this section, we explored the kinematic properties of the \mct\ sample in
more detail, investigating the velocity dispersions of stars either side of the
\kepler\ period gap, and identifying rapidly rotating stars that may be
synchronized binaries.

There is a sharp gap in the population of rotation periods (often called the
\kepler\ period gap), which lies just above the 1 Gyr gyrochrone in the upper
panel of figure \ref{fig:vplot}, whose origin is unknown and is the subject of
much speculation \citep{mcquillan2014, davenport2017, davenport2018,
reinhold2019, reinhold2020}.
This gap was first identified by \mct, and roughly follows a line of constant
gyrochronal age of around 1.1 Gyr \citep[according to the][gyrochronology
relation]{angus2019}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018} and a change in magnetic field structure causing a brief period
where rotational variability is reduced and rotation periods cannot be
measured \citep{reinhold2019, reinhold2020}.

The top panel of figure \ref{fig:vplot} suggests that the \citet{angus2019},
Praesepe-based gyrochronology model is valid below the gap but not above.
Gyrochrones follow lines of constant velocity dispersion below the gap, but
{\it cross} lines of constant velocity dispersion above the gap.
This phenomenon is robust to the choice of bin size and position.
Although we do not provide an in-depth analysis here (and more data may be
needed to confirm a connection) these data suggest that the gap may indeed
separate a young regime where stellar cores are decoupled from their envelopes
from an old regime where these layers are more tightly coupled.
If so, this could indicate that the phenomenon responsible for changing the
shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
produces the gap.

An alternate explanation for the gap is that the \mct\ sample contains two
distinct stellar populations: one young and one old.
If so, the kinematic properties of stars above and below the gap are likely to
be distinctly different.
The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
stars in the \mct\ sample, with stars subdivided into three groups: those that
rotate more quickly than the main rotation period population (green
triangles), those with rotation periods shorter than the gap (orange squares),
and those with rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology model, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample, although \citet{davenport2017}
found that the gap extends to temperatures as hot as 6000 K.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
However, figure \ref{fig:gap} shows a smooth increase in velocity dispersion
with rotation period across the gap (from orange squares to blue circles),
suggesting that these groups are part of the same Galactic population.
% The smooth increase in velocity dispersion across the gap shown here suggests
% that stars above and below are part of the same stellar population.
This observation does not rule out the possibility that a brief cessation of
star formation in the Solar neighborhood, around one Gyr ago, may have caused
this gap, however.

In the final part of our analysis, we investigated the potential for using
kinematics to identify synchronized binaries in the \mct\ sample.
Synchronized binaries are pairs of stars whose rotation periods are equal to
their orbital period.
Since synchronization appears to happen at rotation periods of 7 days or
shorter \citep{simonian2019}, and most isolated stars have rotation periods
longer than 7 days, the rotation periods of synchronized binaries are likely
to be {\it shorter} than they would be if they were isolated stars.
For this reason, their rotation periods do not reflect their ages and the
gyrochronal age of a synchronized binary is likely to be much younger than the
true age of the system.
Synchronized binaries are therefore a source of contamination for
gyrochronology and should be removed from samples before performing a
gyrochronal age analysis.
Figure \ref{fig:gap} shows that some of the most rapidly rotating stars in the
\mct\ sample have relatively large absolute velocities, indicating that they
are likely synchronized binaries.
For this reason, the velocity dispersions of stars with rotation periods
shorter than the lower edge of the rotation period distribution (green
triangles in figure \ref{fig:gap}) are not significantly smaller than the,
presumed older, orange-colored stars.
In general, stars with rotation periods less than $\sim$10 days have an
increased chance of being synchronized binaries.
This result is in agreement with a recent study which found that a large
fraction of photometric binaries were rapid rotators, and the probability of a
star being a synchronized binary system substantially increased below rotation
periods of around 7 days \citep{simonian2019}.
We caution users of rotation period catalogs that rapid rotators with large
absolute velocities should be flagged as potential synchronized binaries
before applying any gyrochronal analysis.

\section{Conclusions}

In this paper, we used the \vb\ velocity dispersions of stars in the \mct\
catalog to explore the evolution of stellar rotation period as a function of
effective temperature and age.
Our conclusions are as follows:
\begin{itemize}
\item{{\bf Spin-down rate doesn't always increase with decreasing mass for K
    dwarfs.}
Although at young ages, rotation period is anti-correlated with \teff\ (as
seen in many young open clusters, including Praesepe), at intermediate ages the
relation flattens out and K dwarfs of different masses rotate at the same
rate.
At old ages, cooler K dwarfs spin more rapidly than hotter K dwarfs of the
same age.}

\item{{\bf Variable core-envelope coupling may be the cause.} We showed that
the period--\teff\ relations change shape over time in a way that
qualitatively agrees with theoretical models which include mass and
time-dependent core-envelope angular momentum transport \citep{spada2019}.}

\item{{\bf Low-mass stars stay active longer.}
We found that the oldest stars in the \mct\ catalog are cooler than 4500
K, which agrees with previous results which show that lower-mass stars remain
        active for longer, allowing their rotation periods to be measured at
        older ages.}

\item{{\bf The \kepler\ period gap may be related to core-envelope coupling.}
We speculated that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
The velocity dispersions of stars increase smoothly across the rotation period
gap, indicating that the gap does not separate two distinct stellar
populations.}

\item{{\bf Rapidly rotating stars with large absolute velocities may be
synchronized binaries.}
We used kinematics to indicate that there is a population of
synchronized binaries with rotation periods less than around 10 days.}

% \item{{\bf \sigmavb\ can be used as an age proxy for (low Galactic latitude)
%     \kepler\ stars.} In the appendix of this paper we demonstrate that the
%         dispersion of velocities in the direction of Galactic latitude, \vb,
%         can be used as an age proxy, by showing that there is no strong
%         evidence for mass-dependent heating in low-mass \kepler\ dwarfs: the
%         velocity dispersions of K and M dwarfs, whose main-sequence lifetimes
%         are longer than around 11 Gyrs, do not appear to increase with
%         decreasing mass.
% Although {\it vertical} velocity, \vz, is a quantity that has been
% demonstrated to trace time-dependent orbital heating in the disc of the
% Galaxy, most stars with measured rotation periods do not yet have radial
% velocities, so we used velocity in the direction of Galactic latitude, \vb\,
% as a proxy for \vz.
%         Using stars in the GUMS simulation \citep{robin2012}, we showed that
%         using \vb\ as a proxy for \vz\ introduces an additional
%         velocity scatter, which increases with increasing Galactic
%         latitude.
% However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
% confirmed that using \vb\ instead of \vz\ does not introduce any
% mass-dependent velocity dispersion bias into the sample.
% We therefore confirmed that \vb\ velocity dispersion can be used to accurately
% rank stars by age.}

\end{itemize}

We would like to thank Suzanne Aigrain for providing thoughtful insight that
improved this paper.

This work was partly developed at the 2019 KITP conference `Better stars,
better planets'.
This research was supported in part by the National Science Foundation under
Grant No. NSF PHY-1748958.

Parts of this project are based on ideas explored at the Gaia sprints at the
Flatiron Institute in New York City, 2016 and MPIA, Heidelberg, 2017.

This work made use of the gaia-kepler.fun crossmatch database created by Megan
Bedell.

Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
\Kepler\ mission is provided by the NASA Science Mission directorate.

This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it
Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
Funding for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral Agreement.

\section{Appendix: Validating \vb\ dispersion as an age proxy}
\label{sec:mass-dependent-heating}

The conclusions drawn in this paper depend on the assumption that velocity
dispersion in the direction of Galactic latitude (\sigmavb) can be used as an
age proxy.
There are two main reasons however, why \vb\ velocity dispersion may {\it not}
be a good age proxy.
Firstly, mass-dependent heating may act on the sample, meaning that velocity
dispersion depends on both age and mass.
Secondly, since stars in the \kepler\ field have a range of Galactic
latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
stars, and introduce a velocity bias for high latitude stars (which are more
likely to be cooler and older).
In this section we demonstrate that neither of these problems seem to be a
significant issue for our data.

In order to establish whether \sigmavb\ can be used as an age proxy, we
searched for signs of mass-dependent heating within the \kepler\ field.
Mass-dependent dynamical heating may result from lower-mass stars experiencing
greater velocity changes when gravitationally perturbed than more massive
stars.
It has not been unambiguously observed in the galactic disk because of the
strong anti-correlation between stellar mass and stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, and no
evidence for mass-dependent heating has previously been found in M dwarfs
\citep[\eg][]{faherty2009, newton2016}.

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyr and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because it only
includes stars with {\it detectable} rotation periods, and since lower-mass
stars stay active for longer it is likely that it contains a strong mass-age
correlation.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated photometric binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
outliers before calculating the standard deviation of stars in each bin.
These extreme velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
function of effective temperature for the K and M \kepler\ dwarf sample.
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
the opposite of the trend expected for mass-dependent heating, however the
slope is only inconsistent with zero at 1.3 $\sigma$.
The sharp uptick in velocity dispersion in the coolest bin is probably noise
caused by the small number of stars in that bin.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point however, is that we do not see evidence for mass-dependent
heating acting on stars in the \kepler\ field, indicating that velocity
dispersion {\it can} be used as an age proxy (with the caveat that there is
still a chance, albeit a small one, that the opposing effects of the selection
function and mass-dependent heating are working to cancel each other out).
This analysis was performed using \vb\ but we also examined the {\it vertical}
velocities of the 537 stars in this sample with RV measurements.
Again, no evidence was found for mass-dependent heating: the slope of the
velocity dispersion-temperature relation was consistent with zero.
\begin{figure}
  \caption{
      Top: Stellar velocity (\vb) as a function of \teff\ for
      \kepler\ K and M dwarfs.
Vertical lines indicate different \teff-groupings used to calculate velocity
    dispersion.
Pink stars were not included in velocity dispersion calculations as they were
    either removed as outliers during a sigma clipping process, or they lie at
    the sparcely populated, extremely cool end of the temperature range.
    Velocity dispersion and \teff\ are slightly positively correlated, likely
    due to a brightness-related selection bias, indicating that mass-dependent
    heating does not significantly affect low-mass stars in the \kepler\
    field.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vs_teff}
\label{fig:vb_vs_teff}
\end{figure}

Having found no strong evidence for mass-dependent heating, we next tested
the validity of \vb\ as a proxy for \vz\ in more detail.
At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
of $b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
Galaxy.
We subtracted this offset and examined the residuals of the \vz\ -- \vb\
relationship to investigate the variance as a function of Galactic latitude
(shown in figure \ref{fig:vb_vz}).
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
20$^\circ$.
This demonstrates that using \vb\ instead of \vz\ for stars in the \kepler\
field will introduce an additional velocity scatter, inflating \sigmavb\
relative to \sigmavz.
This additional velocity scatter will be greatest for stars at the highest
Galactic latitudes.

\begin{figure}
  \caption{
This figure demonstrates the variance in the relationship between \vb\ and
    \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
The panels show a kernel density estimator (KDE) (black solid line) for
    the \vz -- \vb\ residuals of stars in the GUMS simulation at four
    different Galactic latitudes.
Blue dashed lines show Gaussian fits to these KDEs.
The distributions are close to Gaussian, with slightly heavy tails.
The standard deviations of the Gaussian fits increase with Galactic latitude.
This figure illustrates how using \vb\ instead of \vz\ artificially
    increases velocity dispersion, especially at high latitudes.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vz}
\label{fig:vb_vz}
\end{figure}

Since we are concerned with velocity {\it dispersions}, rather than velocities
themselves, we also compared \sigmavb\ and \sigmavz\ as a function of
temperature for stars downloaded from the GUMS simulation.
For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
latitudes the two quantities became significantly different.
For this reason we proceeded by only including stars with galactic latitudes
less than 15\degrees\ in our analysis.
Although we find that the transformation between \vz\ and \vb\ does not {\it
strongly} affect our results, we cannot rule out the possibility that it
introduces systematic biases into the velocity dispersions we present here.
In \gaia\ DR3, RVs will be available for most stars in this sample, providing
an opportunity to validate (or correct) the results presented here, and to
work in action-space, rather than velocity-space.

Because of the noisy relationship between \vb\ and \vz\, in this paper we do
not attempt to convert velocity dispersion (\sigmavb) into an age via an
age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
Although we find that \sigmavb\ can be used to rank groups of stars by age, a
more careful analysis that includes formal modeling of the \vb\ -- \vz\
relationship will be needed to calculate absolute ages.

\bibliographystyle{plainnat}
\bibliography{bib.bib}
\end{document}

### 2020-03-18 paper/acknowledgements.tex
Elisabeth's comments and some of Gus's
adding jen's small comments and a core-envelope coupling section
changing bib and adding acknowledgement

@@
-This work was partly developed at the 2019 KITP conference 'Better stars,
+We would like to thank Suzanne Aigrain for providing thoughtful insight that
+improved this paper.
+
+This work was partly developed at the 2019 KITP conference `Better stars,
 better planets'.
+This research was supported in part by the National Science Foundation under
+Grant No. NSF PHY-1748958.
+
 Parts of this project are based on ideas explored at the Gaia sprints at the

### 2020-03-20 paper/ms63.tex
updating to 6.3

--- 2020-03-18 paper/one_big_file.tex
@@
-% zip all_in_one.zip one_big_file.tex CMD_cuts_double.pdf main_figure.pdf gap.pdf vb_vs_teff.pdf vb_vz.pdf field_comparison.pdf bib.bib ms.pdf
-
-\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
-\usepackage{cite, natbib}
-\usepackage{float}
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
-\usepackage{hyperref}
-\usepackage{epsfig}
-\usepackage{cases}
-\usepackage[section]{placeins}
-\usepackage{graphicx, subfigure}
-\usepackage{color}
-\usepackage{bm}
-
-\AtBeginDocument{\let\textlabel\label}
+% zip files.zip ms63.tex CMD_cuts_double.pdf main_figure.pdf gap.pdf vb_vs_teff.pdf vb_vz.pdf field_comparison.pdf bib.bib ms.pdf
+
+\documentclass{aastex63}
 
@@
 \newcommand{\pmega}{$\bar{\omega}$}
-\newcommand{\mj}{$m_j$}
-\newcommand{\mh}{$m_h$}
-\newcommand{\mk}{$m_k$}
-\newcommand{\mx}{$m_x$}
 \newcommand{\logg}{log(g)}
@@
 
-\newcommand{\tmin}{3500}
-\newcommand{\tmax}{5500}
-
-% \newcommand{\nsim_stars}{841}
-
 \newcommand{\amnh}{1}
@@
 
+% \received{June 1, 2019}
+% \revised{January 10, 2019}
+% \accepted{\today}
+% \submitjournal{AJ}
+
+%% The following command can be used to set the latex table counters.  It
+%% is needed in this document because it uses a mix of latex tabular and
+%% AASTeX deluxetables.  In general it should not be needed.
+%\setcounter{table}{1}
+
+% \shorttitle{Sample article}
+% \shortauthors{Schwarz et al.}
+
 \begin{document}
@@
 
-\author{%
-    Ruth Angus$^{1, 2, 3}$,  % signed off
-    Angus Beane$^4$,  % signed off
-    Adrian M. Price-Whelan$^2$,  % signed off
-    Jason Curtis$^1$,  % signed off
-    Travis Berger$^6$,  % signed off
-    Elisabeth Newton$^5$,  % signed off
-    Jennifer van Saders$^6$,  % signed off
-    Rocio Kiman$^{7, 1}$,  % signed off
-    Dan Foreman-Mackey$^2$,  % signed off
-    Yuxi Lu$^{3, 1}$,  % signed off
-    Lauren Anderson$^8$}  % signed off
-
-\altaffiltext{\amnh}{Department of Astrophysics, American Museum of Natural
-History, 200 Central Park West, Manhattan, NY, USA}
-\altaffiltext{\cca}{Center for Computational Astrophysics, Flatiron Institute,
+%% The \author command is the same as before except it now takes an optional
+%% argument which is the 16 digit ORCID. The syntax is:
+%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
+%%
+%% This will hyperlink the author name to the author's ORCID page. Note that
+%% during compilation, LaTeX will do some limited checking of the format of
+%% the ID to make sure it is valid. If the "orcid-ID.png" image file is
+%% present or in the LaTeX pathway, the OrcID icon will appear next to
+%% the authors name.
+%%
+%% Use \affiliation for affiliation information. The old \affil is now aliased
+%% to \affiliation. AASTeX v6.3 will automatically index these in the header.
+%% When a duplicate is found its index will be the same as its previous entry.
+%%
+%% Note that \altaffilmark and \altaffiltext have been removed and thus
+%% can not be used to document secondary affiliations. If they are used latex
+%% will issue a specific error message and quit. Please use multiple
+%% \affiliation calls for to document more than one affiliation.
+%%
+%% Use \email to set provide email addresses. Each \email will appear on its
+%% own line so you can put multiple email address in one \email call. A new
+%% \correspondingauthor command is available in V6.3 to identify the
+%% corresponding author of the manuscript. It is the author's responsibility
+%% to make sure this name is also in the author list.
+
+\correspondingauthor{Ruth Angus}
+\email{rangus@amnh.org}
+
+\author{Ruth Angus}
+\affiliation{Department of Astrophysics, American Museum of Natural History,
+200 Central Park West, Manhattan, NY, USA}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute,
 162 5th Avenue, Manhattan, NY, USA}
-\altaffiltext{\columbia}{Department of Astronomy, Columbia
-University, Manhattan, NY, USA}
-\altaffiltext{\cfa}{Center for Astrophysics {\normalfont |} Harvard \&
+\affiliation{Department of Astronomy, Columbia University, Manhattan, NY, USA}
+
+\author{Angus Beane}
+\affiliation{Center for Astrophysics {\normalfont |} Harvard \&
 Smithsonian, 60 Garden Street, Cambridge, MA 02138, USA}
-\altaffiltext{\dartmouth}{Department of Physics and Astronomy, Dartmouth
+
+\author{Adrian M. Price-Whelan}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute,
+162 5th Avenue, Manhattan, NY, USA}
+
+\author{Jason Curtis}
+\affiliation{Department of Astrophysics, American Museum of Natural History,
+200 Central Park West, Manhattan, NY, USA}
+
+\author{Travis Berger}
+\affiliation{Institute for Astronomy, University of Hawai`i, 2680
+Woodlawn Drive, Honolulu, HI 96822, USA}
+
+\author{Elisabeth Newton}
+\affiliation{Department of Physics and Astronomy, Dartmouth
 College, Hanover, NH, USA}
-\altaffiltext{\hawaii}{Institute for Astronomy, University of Hawai`i, 2680
+
+\author{Jennifer van Saders}
+\affiliation{Institute for Astronomy, University of Hawai`i, 2680
 Woodlawn Drive, Honolulu, HI 96822, USA}
-\altaffiltext{\cuny}{Department of physics, CUNY Graduate Center, City
+
+\author{Rocio Kiman}
+\affiliation{Department of physics, CUNY Graduate Center, City
 University of New York, Manhattan, NY, USA}
-\altaffiltext{\carnegie}{Observatories of the Carnegie Institution for
+
+\author{Dan Foreman-Mackey}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute,
+162 5th Avenue, Manhattan, NY, USA}
+
+\author{Yuxi (Lucy) Lu}
+\affiliation{Department of Astronomy, Columbia University, Manhattan, NY, USA}
+\affiliation{Department of Astrophysics, American Museum of Natural History,
+200 Central Park West, Manhattan, NY, USA}
+
+\author{Lauren Anderson}
+\affiliation{Observatories of the Carnegie Institution for
 Science, 813 Santa Barbara Street, Pasadena, CA 91101, USA}
@@
 \end{abstract}
+
+%% Keywords should appear after the \end{abstract} command.
+%% See the online documentation for the full list of available subject
+%% keywords and the rules for their use.
+% \keywords{editorials, notices ---
+% miscellaneous --- catalogs --- surveys}
 
@@
 improved this paper.
-
 This work was partly developed at the 2019 KITP conference `Better stars,
@@
 Grant No. NSF PHY-1748958.
-
 Parts of this project are based on ideas explored at the Gaia sprints at the
 Flatiron Institute in New York City, 2016 and MPIA, Heidelberg, 2017.
-
 This work made use of the gaia-kepler.fun crossmatch database created by Megan
@@
 
-\bibliographystyle{plainnat}
-\bibliography{bib.bib}
+%% For this sample we use BibTeX plus aasjournals.bst to generate the
+%% the bibliography. The sample63.bib file was populated from ADS. To
+%% get the citations to show in the compiled file do the following:
+%%
+%% pdflatex sample63.tex
+%% bibtext sample63
+%% pdflatex sample63.tex
+%% pdflatex sample63.tex
+
+\bibliography{bib}{}
+\bibliographystyle{aasjournal}
+
+%% This command is needed to show the entire author+affiliation list when
+%% the collaboration and author truncation commands are used.  It has to
+%% go at the end of the manuscript.
+%\allauthors
+
+%% Include this line if you are using the \added, \replaced, \deleted
+%% commands to see a summary list of all changes at the end of the article.
+%\listofchanges
+
 \end{document}

### 2020-04-16 kinematics_and_rotation/barnes2010.py
adding data tables

@@ -1,0 +1,102 @@
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from scipy import interpolate
+
+
+def teff_to_mass(teff):
+    df = pd.read_csv("mamajek_table.csv", usecols=["Teff", "Msun"])
+    f = interpolate.interp1d(df.Teff, df.Msun)
+    return f(teff)
+
+
+def convective_overturn_time(M):
+    """Estimate the convective overturn time.
+
+    Estimate the convective overturn time using equation 11 in Wright et al.
+    (2011): https://arxiv.org/abs/1109.4634
+    log tau = 1.16 - 1.49log(M/M⊙) - 0.54log^2(M/M⊙)
+
+    Args:
+        M (float): Mass in Solar units
+
+    Returns:
+        The convective overturn time in days.
+
+    """
+    log_tau = 1.16 - 1.49*np.log10(M) - .54*(np.log10(M))**2
+    return 10**log_tau
+
+
+def dpdt(prot, tau):
+    ki = 740
+    return tau / (ki * prot)
+
+
+def barnes_2010_model(p_init, age, dt, tau):
+
+    p_current = p_init
+    for i in np.arange(0, age, dt):
+        dp_dt = dpdt(p_current, tau)
+        p_current += (dp_dt * dt)
+
+    return p_current
+
+
+def test_teff_to_mass():
+    assert np.isclose(teff_to_mass(5777), 1., atol=.1)
+
+
+def test_barnes_2010_model():
+    assert np.isclose(barnes_2010_model(1., 4560, 1000, mass=1.), 26, atol=3)
+    assert np.isclose(barnes_2010_model(1., 4560, 1000, teff=5777), 26, atol=3)
+
+
+def mass_to_tau_barnes(mass):
+    df = pd.read_csv("../data/barnes_and_kim_table.csv")
+    f = interpolate.interp1d(df.Mass, df.Global_tauc)
+    return f(mass)
+
+
+if __name__ == "__main__":
+    print(mass_to_tau_barnes(1.))
+    test_mass = np.linspace(1.2, .4, 100)
+    test_tau = mass_to_tau_barnes(test_mass)
+
+    ages = np.linspace(.1, 12, 6)*1e3
+    ages[2] = 4560
+    for a in ages:
+        print(a)
+        p = barnes_2010_model(15., a, 100, test_tau)
+        plt.plot(test_mass, p)
+
[…]

### 2020-04-16 kinematics_and_rotation/__init__.py
adding barnes model to __init__

@@ -5,3 +5,4 @@ from .dispersion import *
 from .sigma_clip import *
 from .photometric_teff import *
 from .teff_bv import *
+from .barnes2010 import *

### 2020-04-16 paper/ms63.tex
moving the table to the appendix
adding some changes from the referee report

@@
 older than around 1 Gyr.
-Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
-they are young, at old ages we find that late-G and early-K dwarfs rotate at
-the {\it same rate} or faster than late-K dwarfs of the same age.
+Although late-K dwarfs spin more slowly than early-K dwarfs when they are
+    young, at old ages we find that late-K dwarfs rotate at the {\it same
+    rate} or faster than early-K dwarfs of the same age.
 This result agrees qualitatively with semi-empirical models that vary the rate
@@
 \citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.
-
 For this work, we used the precise \textit{Gaia} DR2 photometric color,
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
-temperature measurements for nearby, unreddened field stars in benchmark
-samples, including FGK stars characterized with high-resolution optical
-spectroscopy \citep{brewer2016}, M dwarfs characterized with low-resolution
-optical and near-infrared spectroscopy \citep{mann2015}, and K and M dwarfs
-characterized with interferometry and bolometric flux analyses
-\citep{boyajian2012}.
-This empirical color--temperature relation is valid over the color range $0.55
-< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $6470 < T_{\rm eff} <
-3070$~K.
-The dispersion about the relation implies a high precision of 50~K.
-These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
-\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
-at all temperatures\footnote{The color--temperature relation is described in
-detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
-\etal\ (2020, in prep).}.
-The equation we used to calculate photometric temperatures from Gaia \gcolor\
-color is a seventh-order polynomial with coefficients given in table
-\ref{tab:coeffs}.
-\begin{table}[h!]
-  \begin{center}
-      \caption{
-          Coefficient values for the 7th-order polynomial used to estimate
-      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
-      prep).}
-    \label{tab:coeffs}
-    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
-        (\gcolor\ ) exponent & Coefficient  \\
-      \hline
-      $0$ & -416.585 \\
-      $1$ & 39780.0  \\
-      $2$ & -84190.5 \\
-      $3$ & 85203.9  \\
-      $4$ & -48225.9 \\
-      $5$ & 15598.5  \\
-      $6$ & -2694.76 \\
-      $7$ & 192.865  \\
-    \end{tabular}
-  \end{center}
-\end{table}
+\racomment{The calibration of this relation is described Curtis \etal\ (2020,
+in prep) and briefly summarized in the Appendix of this paper.}
 
@@
 gyrochrones from the \citet{spada2019} model shown in white.
-These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
+These gyrochrones are \racomment{plotted for the same ages as above.}
 Each point plotted in the top panel also appears in the bottom panel with the
@@
 
+\racomment{
+In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
+gyrochronology model is over-plotted for comparison.
+We used this model as an example, however all empirical gyrochronology models
+are similar in essence to the \citet{angus2019} relation and none is able to
+reproduce the observed velocity dispersions, or capture the evolving shape of
+the period-\teff\ relations.
+The \citet{angus2019} relation is a new fully-empirical gyrochronology
+relation, calibrated using recently measured rotation periods for members of
+the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
+douglas2017}.
+These are the oldest cluster M dwarfs with measured photometric rotation
+periods and the \citet{angus2019} model therefore encapsulates the behavior of
+these cool stars at Praesepe age.
+
+Alternatively, we could have chosen to compare our data with one of the many
+other available empirical gyrochronology models.
+For example, the \citet{barnes2003, barnes2007, meibom2011, angus2015}
+relations all have a functional form that was first introduced by
+\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
+are free parameters, and B and V are photometric magnitudes.
+These relations have been widely used in the literature, and are similar to
+the \citet{angus2019} relation in that they consist of a color-dependent term
+multiplied by an age-dependent term, \ie\ they are separable in color and age.
+None of these relations follow lines of constant velocity dispersion in figure
+\ref{fig:vplot} because they have a universal power-law index, $n$ (around
+0.5), which applies to stars of all masses and ages.
+They do not have the flexibility to capture the evolving period-color
+relationship seen in the data.
+The \citet{barnes2010} model has a different functional form: it uses Rossby
+number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
+time) to encode the mass-dependent evolution of rotation periods:
+$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
+This simple relation neatly reproduces the rotation periods of young cluster
+stars, however, like the other empirical gyrochronology models, it does not
+reproduce the observed trends seen in the velocity dispersion data.
+It has a similar period-\teff\ relation to the Praesepe-based
+\citet{angus2019} model, because the period-\teff\ relation of Praesepe
+roughly follows a line of constant Rossby number.
+Consequently, like the \citet{angus2019} model, the \citet{barnes2010}
+gyrochrones do not reproduce the data explored here and do not have the
+flexibility to capture mass and time-variable rotational evolution.
+Although it seems that the period-\teff\ relations follow lines of constant
+}
+
 The velocity dispersions of stars in the \mct\ sample, shown in figure
@@
 At young ages \citep[younger than around 1 Gyr but still old enough to be on
-the main sequence and have transitioned from the `I' sequence to the `C'
+the main sequence and have transitioned from the `C' sequence to the `I'
 sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
@@
 
-\section{Appendix: Validating \vb\ dispersion as an age proxy}
+\section{Appendix A: Validating \vb\ dispersion as an age proxy}
 \label{sec:mass-dependent-heating}
@@
 greater than 1.2 (corresponding to an effective temperature $\lesssim$
-4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
+4800 K) and absolute \gaia\ $G$-band magnitudes $>$ 4.
 We also eliminated photometric binaries by removing stars above a 6th order
@@
 
+\racomment{
+\section{Appendix B: Photometric temperatures}
+
+For this work, we used the precise \textit{Gaia} DR2 photometric color,
+$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
+To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
+temperature measurements for nearby, unreddened field stars in benchmark
+samples, including FGK stars characterized with high-resolution optical
+spectroscopy \citep{brewer2016}, M dwarfs characterized with low-resolution
+optical and near-infrared spectroscopy \citep{mann2015}, and K and M dwarfs
+characterized with interferometry and bolometric flux analyses
+\citep{boyajian2012}.
+This empirical color--temperature relation is valid over the color range $0.55
+< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $3070 > T_{\rm eff} >
+6470$~K.
+The dispersion about the relation implies a high precision of 50~K.
+These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
+\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
+at all temperatures\footnote{The color--temperature relation is described in
+detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
+\etal\ (2020, in prep).}.
+The equation we used to calculate photometric temperatures from Gaia \gcolor\
+color is a seventh-order polynomial with coefficients given in table
+\ref{tab:coeffs}.
+\begin{table}[h!]
+  \begin{center}
+      \caption{
+          Coefficient values for the 7th-order polynomial used to estimate
+      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
+      prep).}
+    \label{tab:coeffs}
+    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
+        (\gcolor\ ) exponent & Coefficient  \\
+      \hline
+      $0$ & -416.585 \\
+      $1$ & 39780.0  \\
+      $2$ & -84190.5 \\
+      $3$ & 85203.9  \\
+      $4$ & -48225.9 \\
+      $5$ & 15598.5  \\
+      $6$ & -2694.76 \\
+      $7$ & 192.865  \\
+    \end{tabular}
+  \end{center}
+\end{table}
+}
+
 %% For this sample we use BibTeX plus aasjournals.bst to generate the

### 2020-04-16 paper/abstract.tex
abstract typo
clarifying sigma comparison but

--- 2020-03-16 paper/abstract.tex
@@
 stars.
-In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
-an age proxy, to reveal their rotational evolution and demonstrate that
-kinematics could be a useful tool for calibrating gyrochronology.
+In this work, we use velocity dispersion as an age proxy to reveal the
+temperature-dependent rotational evolution of groups of low-mass \kepler\
+dwarfs, and demonstrate that kinematic ages could be a useful tool for
+calibrating gyrochronology in future.
 We find that a linear gyrochronology model, calibrated to fit the
@@
 Finally, based on their kinematics, we confirm that many rapidly rotating
-K-dwarfs are likely to be synchronized binaries.
+GKM dwarfs are likely to be synchronized binaries.

### 2020-04-16 paper/ms63.tex
clarifying sigma comparison but

@@
 stars.
-In this work, we use the velocity dispersions of groups of low-mass \kepler\
-dwarfs as an age proxy, to reveal the temperature-dependent rotational
-evolution of these stars, and demonstrate that kinematic ages could be a
-useful tool for calibrating gyrochronology in future.
+\racomment{
+In this work, we use velocity dispersion as an age proxy, to reveal the
+temperature-dependent rotational evolution of low-mass \kepler\ dwarfs and
+demonstrate that kinematic ages could be a useful tool for calibrating
+gyrochronology in future.
+}
+% In this work, we use the velocity dispersions of groups of low-mass \kepler\
+% dwarfs as an age proxy, to reveal the temperature-dependent rotational
+% evolution of these stars, and demonstrate that kinematic ages could be a
+% useful tool for calibrating gyrochronology in future.
 We find that a linear gyrochronology model, calibrated to fit the
@@
 flexibility to capture mass and time-variable rotational evolution.
-Although it seems that the period-\teff\ relations follow lines of constant
+
+The lack of flexibility that is baked-in to these empirical gyrochronology
+models should not be seen as a failing of gyrochronologists: with a paucity of
+calibration data for old stars all one can do is fit simple relations to young
+stars and extrapolate to old ages.
+Naturally however, in this era of precise space-based photometry, new data are
+revealing flaws in our old models (which no one ever thought were perfect),
+and are providing opportunities to improve them.
 }
@@
 theory, and the gyrochronology paradigm that has been in place for decades
-\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
+\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, barnes2010,
+angus2019}.
 According to the \citet{spada2019} model, the radiative cores and convective
@@
 velocity dispersions are cooler than 4500 K.
-This implies that the oldest stars with detectable rotation periods, are
+This implies that the oldest stars with detectable rotation periods are
 cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
 massive stars.
+
+\racomment{
     To investigate this idea further, we compared the velocity dispersions of
-stars with measured rotation periods, to the velocity dispersions of the
-overall \kepler\ sample.
-We calculated the velocity dispersions for all stars in the \kepler\ field,
+stars with {\it measured rotation periods} to the velocity dispersions of the
+entire \kepler\ sample.
+If, for example, the rotation periods of G stars are only detectable when they
+are young, G stars with {\it measured periods} should have smaller velocity
+dispersions than {\it all} G stars, because they are younger on average.
+We calculated the velocity dispersions for all stars in the \kepler\ field
 after removing visual binaries, subgiants, stars fainter than 16th magnitude,
@@
 \ref{sec:method}.
-We then compared these velocity dispersions to the velocity dispersions of
-stars in the \mct\ sample.
-If the rotation periods of G stars are only detectable when they are young, G
-stars with measured periods should have smaller velocity dispersions than G
-stars in the overall \kepler\ field.
-We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
-\sigmavb\ for stars with rotation periods published in \mct, as a function of
-\teff\ (see figure \ref{fig:compare}).
+We then compared these velocity dispersions, as a function of
+\teff, to the velocity dispersions of stars with {\it measured} rotation
+periods \citep[\ie\ stars that appear in table 1 of][]{mcquillan2014}.
+We show the ratio of total \sigmavb\ to McQuillan-\sigmavb\ as a
+function of \teff\ in figure \ref{fig:compare}.}
 A larger ratio means the rotating star sample is {\it younger}, on average,

### 2020-04-17 paper/ms63.tex
changes from jackie's comments

@@
 demonstrate that kinematic ages could be a useful tool for calibrating
-gyrochronology in future.
+gyrochronology in the future.
 }
@@
 significantly on stellar age.
-While the velocity of any individual star only provides a weak age constrain
-because its velocity depends on its current position in its orbit, the
-velocity {\it dispersion} of a {\it group} of stars indicates whether that
-group is old or young relative to other groups.
+While the velocity of any individual star only provides a weak age constraint
+\racomment{(if any at all)} because its velocity depends on its current
+position in its orbit, the velocity {\it dispersion} of a {\it population} of
+stars indicates whether that population is old or young relative to other
+populations.
 % Vertical velocity dispersion is thought to increase monotonically with age.
-In this work, we compare the velocity dispersions of groups of field stars in
-the Galactic thin disk to ascertain which groups are older and which younger,
-and draw conclusions about the rotational evolution of stars based on their
-implied relative ages.
+In this work, we compare the velocity dispersions of populations of field
+stars in the Galactic thin disk to ascertain which populations are older and
+which younger, and draw conclusions about the rotational evolution of stars
+based on their implied relative ages.
+
+\racomment{ Kinematic ages have been used to explore the evolution of cool
+dwarfs many times in the past, \citep[\eg][]{reid1995, gizis2000, west2004,
+west2006, schmidt2007, faherty2009, kiman2019}.
+For example, \citet{west2004, west2006} used the vertical distances of stars
+from the Galactic mid-plane as an age proxy, and found that the fraction of
+magnetically active M dwarfs decreases over time.
+\citet{faherty2009} used tangential velocities to infer the ages of M, L and T
+dwarfs, and showed that dwarfs with lower surface gravities tended to be
+kinematically younger, and \citet{kiman2019} used velocity dispersion as an
+age proxy to explore the evolution of H$\alpha$ equivalent width (a magnetic
+activity indicator), in M dwarfs.
+}
 
@@
 and \vz, or marginalize over missing RV measurements and the \kepler\
-selection function, in order to infer the absolute ages of groups of stars.
+selection function, in order to infer the absolute ages of populations of
+stars.
 Regardless of direction however, velocity dispersion is expected to
 monotonically increase over time \citep[\eg][]{holmberg2009}, and can
-therefore be used to {\it rank} groups of stars by age.
+therefore be used to {\it rank} populations of stars by age.
 
@@
 was calculated as 1.5$\times$ the median absolute deviation of velocities, to
-mitigate sensitivity to outliers.} for groups of stars with similar rotation
-periods and temperatures, and presumed similar age.
+mitigate sensitivity to outliers.} for populations of stars with similar
+rotation periods and temperatures, and presumed similar age.
 The top panel of figure \ref{fig:vplot} shows rotation period versus effective
@@
 other available empirical gyrochronology models.
-For example, the \citet{barnes2003, barnes2007, meibom2011, angus2015}
-relations all have a functional form that was first introduced by
+For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
+angus2015} relations all have a functional form that was first introduced by
 \citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
@@
 greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
-parker1970, kawaler1988, charbonneau2010}.
-This aligns with the current assumptions about stellar spin-down, dynamo
-theory, and the gyrochronology paradigm that has been in place for decades
-\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, barnes2010,
-angus2019}.
+parker1970, kawaler1988, charbonneau2010, matt2012, matt2015}.
+% This aligns with the current assumptions about stellar spin-down, dynamo
+% theory, and the gyrochronology paradigm that has been in place for decades.
+% \citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, barnes2010}.
 According to the \citet{spada2019} model, the radiative cores and convective
@@
 angular momentum transport between the two layers.
-Angular momentum resurfaces and prevents the stellar envelopes from
-spinning-down rapidly, and this effect is strongest for late K-dwarfs with
-effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.
+Angular momentum resurfaces from the core and prevents the stellar envelopes
+from spinning-down rapidly, and this effect is strongest for late K-dwarfs
+with effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7
+M$_\odot$.
 
@@
 lifetime.
-Figure \ref{fig:vplot} shows that the groups of stars with the largest
+Figure \ref{fig:vplot} shows that the populations of stars with the largest
 velocity dispersions are cooler than 4500 K.
@@
 massive stars.
+\racomment{ Although later-type stars are known to remain active for longer,
+this result is not necessarily altogether expected.
+For example, \citet{matt2015}, found that the upper-envelope of the \kepler\
+rotation period distribution corresponds to a single-age population of around
+4 Gyr.
+Their gyrochronology model coincided with the upper-envelope at 4 Gyrs, and
+the natural conclusion was that stars with rotation periods near the
+upper-edge of the \mct\ rotation period distribution are all the same age.
+However, the velocity dispersion data presented in figure \ref{fig:vplot}
+indicate that the upper-envelope is {\it not} a single-age population, and
+that later-type stars are in fact older.
+}
 
@@
 To investigate whether mass-dependent heating could be acting on the \kepler\
-sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
-whose MS lifetimes exceed around 11 Gyr and are therefore representative of
-the initial mass function.
+sample, we selected late K and early M dwarfs observed by both \kepler\ and
+\gaia, whose MS lifetimes exceed around 11 Gyr and are therefore
+representative of the initial mass function.
 We could not perform this analysis on the \mct\ sample, because it only
@@
 outliers before calculating the standard deviation of stars in each bin.
-These extreme velocity outliers may be very old late K and M dwarfs, or they
-result from using \vb\ instead of \vz, which introduces additional velocity
-scatter.
+These extreme velocity outliers may be very old late K and early M dwarfs, or
+they result from using \vb\ instead of \vz, which introduces additional
+velocity scatter.
 
@@
 age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
-Although we find that \sigmavb\ can be used to rank groups of stars by age, a
-more careful analysis that includes formal modeling of the \vb\ -- \vz\
+Although we find that \sigmavb\ can be used to rank populations of stars by
+age, a more careful analysis that includes formal modeling of the \vb\ -- \vz\
 relationship will be needed to calculate absolute ages.

### 2020-04-17 paper/ms63.tex
some more comments from Jackie

@@
 Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
-(around 3,000 stars), stars fainter than 16th magnitude (200 stars), stars
-with absolute \vb\ uncertainties greater than 1 \kms\ (1000 stars), and stars
-with galactic latitudes greater than 15\degrees\ (5500 stars, justification
-provided in the appendix) were removed from the sample.
+(around 3,000 stars), stars fainter than 16th magnitude \racomment{in \gaia\
+G-band} (200 stars), stars with absolute \vb\ uncertainties greater than 1
+\kms\ (1000 stars), and stars with galactic latitudes greater than 15\degrees\
+(5500 stars, justification provided in the appendix) were removed from the
+sample.
 Finally, we removed almost 2000 stars with rotation periods shorter than the
@@
 bottom edge of the main population of rotation periods.
+\racomment{These rapidly rotating stars are likely to be either very young
+stars, or synchronized binaries (see section \ref{sec:gap}).}
 After these cuts, around 13,000 stars out of the original $\sim$34,000 were
@@
 lower-mass stars are still braking more efficiently at these intermediate and
-old ages but their cores are more tightly coupled to their envelopes, allowing
-angular momentum transport between the two layers.
-Angular momentum resurfaces from the core and prevents the stellar envelopes
-from spinning-down rapidly, and this effect is strongest for late K-dwarfs
-with effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7
-M$_\odot$.
+old ages \racomment{compared to higher-mass stars}, but their cores are more
+tightly coupled to their envelopes, allowing angular momentum transport
+between the two layers.
+Angular momentum resurfaces \racomment{from the core} and prevents the stellar
+envelopes from spinning-down rapidly, and this effect is strongest for late
+K-dwarfs with effective temperatures of $\sim$4000-4500K and masses
+$\sim$0.5-0.7 M$_\odot$.
 
@@
 of theoretical gyrochronology models.
-During this phase, angular momentum transport between radiative core and
-convective envelope is reduced.
+During this phase, angular momentum transport between the radiative core and
+the convective envelope is reduced.
 Over time, models increase the efficiency of angular momentum transport
@@
 
-Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
-K and M dwarfs remain magnetically active for longer than G dwarfs.
+% Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
+% K and M dwarfs remain magnetically active for longer than G dwarfs.
+
+\racomment{
+In order to measure a rotation period at all, there must be some magnetically
+active regions (either bright plages or dark spots), of a reasonable size on
+the surface of a star.
+    The relatively sharp upper-edge of the rotation period distribution seen
+in figure \ref{fig:vplot} may therefore be caused by a finite active lifetime
+of stars.
 The mass dependence of magnetic activity lifetimes has been demonstrated
-previously \citep[\eg][]{west2008, newton2017, kiman2019}, and if the
-detectability of a rotation period is considered to be a magnetic activity
-proxy, then our results provide further evidence for a mass-dependent activity
-lifetime.
+previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
+kiman2019}, and if the detectability of a rotation period is considered to be
+a magnetic activity proxy, then our results provide further evidence for a
+{\it mass-dependent} activity lifetime.
 Figure \ref{fig:vplot} shows that the populations of stars with the largest
@@
 massive stars.
-\racomment{ Although later-type stars are known to remain active for longer,
-this result is not necessarily altogether expected.
-For example, \citet{matt2015}, found that the upper-envelope of the \kepler\
-rotation period distribution corresponds to a single-age population of around
-4 Gyr.
-Their gyrochronology model coincided with the upper-envelope at 4 Gyrs, and
-the natural conclusion was that stars with rotation periods near the
-upper-edge of the \mct\ rotation period distribution are all the same age.
-However, the velocity dispersion data presented in figure \ref{fig:vplot}
-indicate that the upper-envelope is {\it not} a single-age population, and
-that later-type stars are in fact older.
 }
+% a decrease in magnetic activity after stars exceed a certain age.
+% If so, it is possible that the ages of stars near this upper-edge represent
+% the activity lifetimes of stars.}
+% % We therefore assume that stars {\it with} measured rotation periods are,
+% on % average, more magnetically active than stars {\it without} measured
+% rotation % periods.
+% \citet{matt2015} found that a 4 Gyr gyrochrone, calculated with their model,
+% coincided with the upper-envelope at 4 Gyrs and concluded that the
+% upper-envelope of the \kepler\ rotation period distribution corresponds to a
+% single-age population of around 4 Gyr.
+% However, the velocity dispersion data presented in figure \ref{fig:vplot}
+% suggest that the activity lifetime of stars is mass-dependent.
+% In this case, stars near the upper-dge of the period distribution are not all
+% the same age that later-type stars are in fact older than earlier types.
+% }
 
@@
 entire \kepler\ sample.
-If, for example, the rotation periods of G stars are only detectable when they
-are young, G stars with {\it measured periods} should have smaller velocity
-dispersions than {\it all} G stars, because they are younger on average.
-We calculated the velocity dispersions for all stars in the \kepler\ field
-after removing visual binaries, subgiants, stars fainter than 16th magnitude,
-and high Galactic latitude stars, following the method described in section
-\ref{sec:method}.
+% We assumed that stars with measured rotation periods are more magnetically
+% active than stars without measured rotation periods.
+If stars with measured rotation periods are more magnetically active and
+younger than stars without rotation period measurements, they should also have
+smaller velocity dispersions.
+% This means that stars with measured rotation periods should be kinematically
+% younger on average than the entire \kepler\ sample.
+% they are young, because G stars are only magnetically active when they are
+% young, G stars with {\it measured periods} should have smaller velocity
+% dispersions than {\it all} G stars.
+To test this theory, we calculated the velocity dispersions for all stars in
+the \kepler\ field after removing visual binaries, subgiants, stars fainter
+than 16th magnitude, and high Galactic latitude stars, following the method
+described in section \ref{sec:method}.
 We then compared these velocity dispersions, as a function of
@@
 In this section, we explored the kinematic properties of the \mct\ sample in
-more detail, investigating the velocity dispersions of stars either side of the
-\kepler\ period gap, and identifying rapidly rotating stars that may be
+more detail, investigating the velocity dispersions of stars on either side of
+the \kepler\ period gap, and identifying rapidly rotating stars that may be
 synchronized binaries.
@@
 from an old regime where these layers are more tightly coupled.
-If so, this could indicate that the phenomenon responsible for changing the
-shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
-produces the gap.
+If so, this could indicate that \racomment{these phenomena are related, \ie\
+the process that is responsible for changing the shape of gyrochrones in
+rotation-\teff\ space is related to the process that produces the gap.
+% Of course this may just be a coincidence.
+% Alternatively, the fact that the gap is located near a changing stellar
+% evolution regime may be a coincidence.
+}
 

### 2020-04-17 paper/ms63.tex
adding jackie as coathor
commas in the abstract

@@
 
+\author{Jacqueline K. Faherty}
+\affiliation{Department of Astrophysics, American Museum of Natural History,
+200 Central Park West, Manhattan, NY, USA}
+
 \begin{abstract}
@@
 \racomment{
-In this work, we use velocity dispersion as an age proxy, to reveal the
-temperature-dependent rotational evolution of low-mass \kepler\ dwarfs and
+In this work we use velocity dispersion as an age proxy to reveal the
+temperature-dependent rotational evolution of low-mass \kepler\ dwarfs, and
 demonstrate that kinematic ages could be a useful tool for calibrating

### 2020-04-17 paper/ms63.tex
wording better

@@
 
-\racomment{ Kinematic ages have been used to explore the evolution of cool
-dwarfs many times in the past, \citep[\eg][]{reid1995, gizis2000, west2004,
+\racomment{ There is a long history of using kinematic ages to explore the
+evolution of cool dwarfs \citep[\eg][]{reid1995, gizis2000, west2004,
 west2006, schmidt2007, faherty2009, kiman2019}.
@@
 In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
-gyrochronology model is over-plotted for comparison.
-We used this model as an example, however all empirical gyrochronology models
-are similar in essence to the \citet{angus2019} relation and none is able to
+gyrochronology model is plotted for comparison with the data.
+This model was chosen as an example but there are many other empirical
+gyrochronology models we could have used.
+All the available empirical gyrochronology models are similar in essence to
+the \citet{angus2019} relation and, like that relation, none is able to
 reproduce the observed velocity dispersions, or capture the evolving shape of

### 2020-04-17 paper/ms63.tex
a few more edits

@@
 the period-\teff\ relations.
+
 The \citet{angus2019} relation is a new fully-empirical gyrochronology
@@
 these cool stars at Praesepe age.
-
-Alternatively, we could have chosen to compare our data with one of the many
-other available empirical gyrochronology models.
+There are many other empirical gyrochronology models we could have chosen to
+compare with our data.
 For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
@@
 None of these relations follow lines of constant velocity dispersion in figure
-\ref{fig:vplot} because they have a universal power-law index, $n$ (around
+\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
 0.5), which applies to stars of all masses and ages.
@@
 stars, however, like the other empirical gyrochronology models, it does not
-reproduce the observed trends seen in the velocity dispersion data.
-It has a similar period-\teff\ relation to the Praesepe-based
-\citet{angus2019} model, because the period-\teff\ relation of Praesepe
-roughly follows a line of constant Rossby number.
-Consequently, like the \citet{angus2019} model, the \citet{barnes2010}
-gyrochrones do not reproduce the data explored here and do not have the
-flexibility to capture mass and time-variable rotational evolution.
-
-The lack of flexibility that is baked-in to these empirical gyrochronology
-models should not be seen as a failing of gyrochronologists: with a paucity of
-calibration data for old stars all one can do is fit simple relations to young
-stars and extrapolate to old ages.
+reproduce the observed trends seen in the velocity dispersion data here.
+The \citet{barnes2010} model has a similar period-\teff\ relation to the
+Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
+Praesepe roughly follows a line of constant Rossby number.
+Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
+in its current form does not have the flexibility to capture the mass and
+time-variable rotational evolution revealed by the data explored here.
+
+The lack of flexibility in these empirical gyrochronology models should not be
+seen as a failing of gyrochronologists.
+With a lack of old calibration stars, the only possible approach has been to
+fit simple relations to young stars and extrapolate to old ages.
 Naturally however, in this era of precise space-based photometry, new data are
@@
 the surface of a star.
-    The relatively sharp upper-edge of the rotation period distribution seen
-in figure \ref{fig:vplot} may therefore be caused by a finite active lifetime
-of stars.
+In other words, there is presumably some minimum magnetic activity level that
+is required in order to produce coherent light curve variability from which a
+rotation period can be confidently measured.
+The relatively sharp upper-edge of the rotation period distribution seen in
+figure \ref{fig:vplot} may be caused by a finite active lifetime of stars,
+\ie\ stars older than a critical age are no longer active enough to produce
+periodic, high-amplitude variability.
+The data presented in figure \ref{fig:vplot} indicate that this active
+lifetime is a function of stellar mass.
 The mass dependence of magnetic activity lifetimes has been demonstrated
@@
 velocity dispersions are cooler than 4500 K.
-This implies that the oldest stars with detectable rotation periods are
-cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
-massive stars.
+This implies that \racomment{most of} the oldest stars with detectable
+rotation periods are cooler than 4500 K, \ie\ these-low mass stars stay active
+longer than more massive stars.
 }

### 2020-04-20 paper/ms63.tex
editing blue bits

@@
 While the velocity of any individual star only provides a weak age constraint
-\racomment{(if any at all)} because its velocity depends on its current
+\racomment{(if any at all)}, because its velocity depends on its current
 position in its orbit, the velocity {\it dispersion} of a {\it population} of
@@
 
-Although {\it vertical} velocity, \vz, is an established age proxy, it can
-only be calculated with full 6-dimensional position and velocity information.
+Although {\it vertical} velocity, \vz, is \racomment{a well-established} age
+proxy, it can only be calculated with full 6-dimensional position and velocity
+information.
 In fact, with full 6D phase space and an assumed Galactic potential, it is
@@
 \racomment{The calibration of this relation is described Curtis \etal\ (2020,
-in prep) and briefly summarized in the Appendix of this paper.}
+in prep) and summarized in the Appendix of this paper.}
 
@@
 bottom edge of the main population of rotation periods.
-\racomment{These rapidly rotating stars are likely to be either very young
-stars, or synchronized binaries (see section \ref{sec:gap}).}
-After these cuts, around 13,000 stars out of the original $\sim$34,000 were
-included in the sample.
+\racomment{These rapid rotators are probably either very young stars, or
+synchronized binaries (see section \ref{sec:gap}).} After these cuts, around
+13,000 stars out of the original $\sim$34,000 were included in the sample.
 
@@
 gyrochrones from the \citet{spada2019} model shown in white.
-These gyrochrones are \racomment{plotted for the same ages as above.}
+These gyrochrones are \racomment{plotted for the same ages described above.}
 Each point plotted in the top panel also appears in the bottom panel with the
@@
 
-The lack of flexibility in these empirical gyrochronology models should not be
-seen as a failing of gyrochronologists.
-With a lack of old calibration stars, the only possible approach has been to
-fit simple relations to young stars and extrapolate to old ages.
+The lack of flexibility in these empirical models should not be seen as a
+failing of gyrochronology.
+With a paucity of old calibration stars, the only possible approach has been
+to fit simple relations to young stars and extrapolate to old ages.
 Naturally however, in this era of precise space-based photometry, new data are
@@
 between the two layers.
-Angular momentum resurfaces \racomment{from the core} and prevents the stellar
+Angular momentum surfaces \racomment{from the core} and prevents the stellar
 envelopes from spinning-down rapidly, and this effect is strongest for late
@@
 the surface of a star.
-In other words, there is presumably some minimum magnetic activity level that
-is required in order to produce coherent light curve variability from which a
-rotation period can be confidently measured.
-The relatively sharp upper-edge of the rotation period distribution seen in
-figure \ref{fig:vplot} may be caused by a finite active lifetime of stars,
-\ie\ stars older than a critical age are no longer active enough to produce
-periodic, high-amplitude variability.
+In other words, some minimum magnetic activity level is presumably required to
+produce coherent light curve-variability, from which a rotation period can be
+confidently measured.
+In this case, the relatively sharp upper-edge of the rotation period
+distribution seen in figure \ref{fig:vplot} may be caused by a finite active
+lifetime of stars, \ie\ stars older than a critical age are no longer active
+enough to produce periodic, high-amplitude variability.
 The data presented in figure \ref{fig:vplot} indicate that this active
-lifetime is a function of stellar mass.
+lifetime is a function of stellar mass, because the velocity dispersions of
+stars along the upper edge of the rotation period distribution increase with
+decreasing mass.
+
 The mass dependence of magnetic activity lifetimes has been demonstrated
 previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
-kiman2019}, and if the detectability of a rotation period is considered to be
-a magnetic activity proxy, then our results provide further evidence for a
-{\it mass-dependent} activity lifetime.
+kiman2019}.
+If the detectability of a rotation period is considered to be a magnetic
+activity proxy, then our results provide further evidence for a {\it
+mass-dependent} activity lifetime.
 Figure \ref{fig:vplot} shows that the populations of stars with the largest
@@
 % active than stars without measured rotation periods.
-If stars with measured rotation periods are more magnetically active and
-younger than stars without rotation period measurements, they should also have
-smaller velocity dispersions.
+If stars with measured rotation periods are more magnetically active, and
+younger, than stars without rotation period measurements, they should also
+have smaller velocity dispersions.
 % This means that stars with measured rotation periods should be kinematically
@@
 To test this theory, we calculated the velocity dispersions for all stars in
-the \kepler\ field after removing visual binaries, subgiants, stars fainter
+the \kepler\ field (after removing visual binaries, subgiants, stars fainter
 than 16th magnitude, and high Galactic latitude stars, following the method
-described in section \ref{sec:method}.
+described in section \ref{sec:method}).
 We then compared these velocity dispersions, as a function of

### 2020-04-20 paper/ms63.tex
edits from more thorough read through and saving plot with yaxis tick labes

@@
 first order, by its color and age alone.
-This is the principle behind gyrochronology, the method of inferring a
-star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
-barnes2010, meibom2011, meibom2015}.
+This is the principle behind gyrochronology, the method of inferring a star’s
+age from its rotation period \citep[\eg][]{skumanich1972, barnes2003,
+barnes2007, mamajek2008, barnes2010, meibom2011, meibom2015}.
 However, new photometric rotation periods made available by the \kepler\
@@
 dwarfs are so long that they are not measurable with Kepler data.
-Ground-based rotation period measurements of mid and late M dwarfs indicate
-that there is an upper limit to the rotation periods of {\it late} M dwarfs of
-around 140 days \citep{newton2016, newton2018}, which is much
-longer than the longest rotation periods measured in the \mct\ sample (around
-70 days).
+\racomment{Ground-based rotation period measurements of early M dwarfs
+(spectral types earlier than $\sim$ M2.5, \teff $\gtrsim$3500) indicate that
+there is an upper limit to their rotation periods of around 80 days
+\citep{newton2016, newton2018}, which is longer than the longest rotation
+periods measured for the early M dwarfs in the \mct\ sample (around 50 days).}
 The apparent lack of old gyro-ages for M dwarfs in figure
@@
 
-\racomment{
-In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
-gyrochronology model is plotted for comparison with the data.
-This model was chosen as an example but there are many other empirical
-gyrochronology models we could have used.
-All the available empirical gyrochronology models are similar in essence to
-the \citet{angus2019} relation and, like that relation, none is able to
-reproduce the observed velocity dispersions, or capture the evolving shape of
-the period-\teff\ relations.
-
-The \citet{angus2019} relation is a new fully-empirical gyrochronology
-relation, calibrated using recently measured rotation periods for members of
-the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
-douglas2017}.
-These are the oldest cluster M dwarfs with measured photometric rotation
-periods and the \citet{angus2019} model therefore encapsulates the behavior of
-these cool stars at Praesepe age.
-There are many other empirical gyrochronology models we could have chosen to
-compare with our data.
-For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
-angus2015} relations all have a functional form that was first introduced by
-\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
-are free parameters, and B and V are photometric magnitudes.
-These relations have been widely used in the literature, and are similar to
-the \citet{angus2019} relation in that they consist of a color-dependent term
-multiplied by an age-dependent term, \ie\ they are separable in color and age.
-None of these relations follow lines of constant velocity dispersion in figure
-\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
-0.5), which applies to stars of all masses and ages.
-They do not have the flexibility to capture the evolving period-color
-relationship seen in the data.
-The \citet{barnes2010} model has a different functional form: it uses Rossby
-number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
-time) to encode the mass-dependent evolution of rotation periods:
-$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
-This simple relation neatly reproduces the rotation periods of young cluster
-stars, however, like the other empirical gyrochronology models, it does not
-reproduce the observed trends seen in the velocity dispersion data here.
-The \citet{barnes2010} model has a similar period-\teff\ relation to the
-Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
-Praesepe roughly follows a line of constant Rossby number.
-Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
-in its current form does not have the flexibility to capture the mass and
-time-variable rotational evolution revealed by the data explored here.
-
-The lack of flexibility in these empirical models should not be seen as a
-failing of gyrochronology.
-With a paucity of old calibration stars, the only possible approach has been
-to fit simple relations to young stars and extrapolate to old ages.
-Naturally however, in this era of precise space-based photometry, new data are
-revealing flaws in our old models (which no one ever thought were perfect),
-and are providing opportunities to improve them.
-}
-
 The velocity dispersions of stars in the \mct\ sample, shown in figure
@@
 % K and M dwarfs remain magnetically active for longer than G dwarfs.
+
+\racomment{
+In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
+gyrochronology model is plotted for comparison with the data.
+This model was chosen as an example but there are many other empirical
+gyrochronology models we could have used.
+All the available empirical gyrochronology models are similar in essence to
+the \citet{angus2019} relation and, like that relation, none is able to
+reproduce the observed velocity dispersions, or capture the evolving shape of
+the period-\teff\ relations.
+
+The \citet{angus2019} relation is a new fully-empirical gyrochronology
+relation, calibrated using recently measured rotation periods for members of
+the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
+douglas2017}.
+These are the oldest cluster M dwarfs with measured photometric rotation
+periods and the \citet{angus2019} model therefore encapsulates the behavior of
+these cool stars at Praesepe age.
+There are many other empirical gyrochronology models we could have chosen to
+compare with our data.
+For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
+angus2015} relations all have a functional form that was first introduced by
+\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
+are free parameters, and B and V are photometric magnitudes.
+These relations have been widely used in the literature, and are similar to
+the \citet{angus2019} relation in that they consist of a color-dependent term
+multiplied by an age-dependent term, \ie\ they are separable in color and age.
+None of these relations follow lines of constant velocity dispersion in figure
+\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
+0.5), which applies to stars of all masses and ages.
+They do not have the flexibility to capture the evolving period-color
+relationship seen in the data.
+The \citet{barnes2010} model has a different functional form: it uses Rossby
+number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
+time) to encode the mass-dependent evolution of rotation periods:
+$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
+This simple relation neatly reproduces the rotation periods of young cluster
+stars, however, like the other empirical gyrochronology models, it does not
+reproduce the observed trends seen in the velocity dispersion data here.
+The \citet{barnes2010} model has a similar period-\teff\ relation to the
+Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
+Praesepe roughly follows a line of constant Rossby number.
+Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
+in its current form does not have the flexibility to capture the mass and
+time-variable rotational evolution revealed by the data explored here.
+
+The lack of flexibility in these empirical models should not be seen as a
+failing of gyrochronology.
+With a paucity of old calibration stars, the only possible approach has been
+to fit simple relations to young stars and extrapolate to old ages.
+Naturally however, in this era of precise space-based photometry, new data are
+revealing flaws in our old models (which no one ever thought were perfect),
+and are providing opportunities to improve them.
+}
 

### 2020-04-20 paper/ms63.tex
adding a section for activity lifetime

@@
 \racomment{
+
+\subsection{The mass-dependence of magnetic activity lifetimes}
+\label{sec:activity}
+
+The mass dependence of magnetic activity lifetimes has been demonstrated
+previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
+kiman2019} and, if the detectability of a rotation period is considered to be
+a magnetic activity proxy, then our results provide further evidence for a
+mass-dependent activity lifetime.
+
 In order to measure a rotation period at all, there must be some magnetically
@@
 confidently measured.
-In this case, the relatively sharp upper-edge of the rotation period
-distribution seen in figure \ref{fig:vplot} may be caused by a finite active
-lifetime of stars, \ie\ stars older than a critical age are no longer active
-enough to produce periodic, high-amplitude variability.
-The data presented in figure \ref{fig:vplot} indicate that this active
-lifetime is a function of stellar mass, because the velocity dispersions of
-stars along the upper edge of the rotation period distribution increase with
-decreasing mass.
-
-The mass dependence of magnetic activity lifetimes has been demonstrated
-previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
-kiman2019}.
-If the detectability of a rotation period is considered to be a magnetic
-activity proxy, then our results provide further evidence for a {\it
-mass-dependent} activity lifetime.
+The relatively sharp upper-edge of the rotation period distribution seen in
+figure \ref{fig:vplot} may therefore be caused by a finite active lifetime of
+stars, \ie\ stars older than a critical age are no longer active enough to
+produce periodic, high-amplitude variability.
+% The data presented in figure \ref{fig:vplot} indicate that this active
+% lifetime is a function of stellar mass, because the velocity dispersions of
+% stars along the upper edge of the rotation period distribution increase with
+% decreasing mass.
+In figure \ref{fig:vplot}, the velocity dispersions of stars along the upper
+edge of the rotation period distribution increase with decreasing mass,
+indicating that magnetic activity lifetimes are mass-dependent.
 Figure \ref{fig:vplot} shows that the populations of stars with the largest
 velocity dispersions are cooler than 4500 K.
-This implies that \racomment{most of} the oldest stars with detectable
-rotation periods are cooler than 4500 K, \ie\ these-low mass stars stay active
-longer than more massive stars.
+This implies that most of the oldest stars with detectable rotation periods
+are cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
+massive stars.
 }
@@
 periods have a similar age distribution to the overall Kepler population which
-suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
+suggests that \racomment{most of} the oldest K and M dwarfs are represented in
+the \mct\ sample.
 This result bolsters the evidence that M dwarf rotation periods are measurable
-at older ages than G dwarf rotation periods.
-In other words, G stars become magnetically inactive and have fewer active
-surface regions {\it at a younger age than M dwarfs}.
+at older ages than G dwarf rotation periods and that G stars become
+magnetically inactive and have fewer active surface regions {\it at a younger
+age than M dwarfs}.
 

### 2020-04-21 paper/ms63.tex
I'm not procrastinating... YOU'RE procrastinating!
update apw orcid
Minor text fixes, ORCID

@@
 
-\author{Angus Beane}
-\affiliation{Center for Astrophysics {\normalfont |} Harvard \&
+\author[0000-0002-8658-1453]{Angus~Beane}
+\affiliation{Center for Astrophysics \textpipe{} Harvard \&
 Smithsonian, 60 Garden Street, Cambridge, MA 02138, USA}
 
-\author{Adrian M. Price-Whelan}
+\author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
 \affiliation{Center for Computational Astrophysics, Flatiron Institute,
@@
 
-\author{Dan Foreman-Mackey}
+\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
 \affiliation{Center for Computational Astrophysics, Flatiron Institute,
@@
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-\racomment{The calibration of this relation is described Curtis \etal\ (2020,
+\racomment{The calibration of this relation is described in Curtis \etal\ (2020,
 in prep) and summarized in the Appendix of this paper.}
@@
 \racomment{Ground-based rotation period measurements of early M dwarfs
-(spectral types earlier than $\sim$ M2.5, \teff $\gtrsim$3500) indicate that
+(spectral types earlier than $\sim$ M2.5, \teff $\gtrsim$3500 K) indicate that
 there is an upper limit to their rotation periods of around 80 days

### 2020-04-22 paper/ms63.tex
adding orcids and acknowledgements

@@
 
-\author{Ruth Angus}
+\author[0000-0003-4540-5661]{Ruth Angus}
 \affiliation{Department of Astrophysics, American Museum of Natural History,
@@
 \author[0000-0002-8658-1453]{Angus~Beane}
-\affiliation{Center for Astrophysics \textpipe{} Harvard \&
+\affiliation{Center for Astrophysics $\vert$ Harvard \&
 Smithsonian, 60 Garden Street, Cambridge, MA 02138, USA}
@@
 
+\author{Elisabeth Newton}
+\affiliation{Department of Physics and Astronomy, Dartmouth
+College, Hanover, NH, USA}
+
 \author{Jason Curtis}
@@
 
-\author{Travis Berger}
+\author[0000-0002-2580-3614]{Travis Berger}
 \affiliation{Institute for Astronomy, University of Hawai`i, 2680
 Woodlawn Drive, Honolulu, HI 96822, USA}
-
-\author{Elisabeth Newton}
-\affiliation{Department of Physics and Astronomy, Dartmouth
-College, Hanover, NH, USA}
 
@@
 
-\author{Rocio Kiman}
+\author[0000-0003-2102-3159]{Rocio Kiman}
 \affiliation{Department of physics, CUNY Graduate Center, City
 University of New York, Manhattan, NY, USA}
+\affiliation{Department of Astrophysics, American Museum of Natural History,
+200 Central Park West, Manhattan, NY, USA}
 
@@
 
+We thank the anonymous referee for their comments which greatly improved the
+manuscript.
 We would like to thank Suzanne Aigrain for providing thoughtful insight that
@@
 Bedell.
+T.A.B. acknowledges support by a NASA FINESST award (80NSSC19K1424).
 

### 2020-04-22 paper/ms63.tex
adding some new citations and orcid
another orcid
adding modified text from referee report

@@
 
-\author{Jason Curtis}
+\author[0000-0002-2792-134X]{Jason L. Curtis}
 \affiliation{Department of Astrophysics, American Museum of Natural History,
@@
 
-\author{Jacqueline K. Faherty}
+\author[0000-0001-6251-0573]{Jacqueline K. Faherty}
 \affiliation{Department of Astrophysics, American Museum of Natural History,
@@
 these cool stars at Praesepe age.
+Most semi-physical spin-down models predict that, once the rotation periods of
+stars converge to a tight sequence, they approximately spin down with a common
+braking index \citep[e.g., Fig. 5 in][]{gallet2015}.
+While some semi-physical models do not predict that stars have fully converged
+by the age of Praesepe, the observations suggest that they have
+\citep{douglas2019}, in which case, projecting the Praesepe sequence forward
+with a single braking index that is common to all stars and constant in time
+is appropriate.
+However, observations of low-mass stars in older clusters \citep[\eg, NGC
+6811, NGC 752, and Ruprecht 147][Curtis \etal, submitted]{curtis2019,
+agueros2018} demonstrate that the assumptions adopted in those models must be
+invalid, or inaccurate in some quantitative way (for example, the
+core-envelope coupling timescale is inaccurately calibrated).
+
 There are many other empirical gyrochronology models we could have chosen to
@@
 
-The lack of flexibility in these empirical models should not be seen as a
-failing of gyrochronology.
-With a paucity of old calibration stars, the only possible approach has been
-to fit simple relations to young stars and extrapolate to old ages.
-Naturally however, in this era of precise space-based photometry, new data are
-revealing flaws in our old models (which no one ever thought were perfect),
-and are providing opportunities to improve them.
+% The lack of flexibility in these empirical models should not be seen as a
+% failing of gyrochronology.
+% With a paucity of old calibration stars, the only possible approach has been
+% to fit simple relations to young stars and extrapolate to old ages.
+% Naturally however, in this era of precise space-based photometry, new data are
+% revealing flaws in our old models (which no one ever thought were perfect),
+% and are providing opportunities to improve them.
+
+It is not surprising that the \citet{angus2019} model, or other gyrochronology
+models do not reproduce the data, since tensions with
+\citet{skumanich1972}-type spin-down have been revealed whenever new rotation
+periods of benchmark stars have become available.
+With the influx of new stellar rotation periods provided by \kepler/\ktwo\ and
+\tess, our understanding of spin-down is changing rapidly.
+New data are revealing flaws in not just the calibration of but also the
+functional forms of old models.
+We have shown that velocity dispersion data provides support for more
+sophisticated models incorporating additional physical effects \citep[\eg,
+core-envelope coupling][]{spada2019} over simple empirical models.
+Alternatively, more flexible empirical frameworks, like a Gaussian process
+model, could capture everything we now know about stellar spin-down.
 }

### 2020-04-23 paper/ms63.tex
jen's comments

@@
 
-\author{Jennifer van Saders}
+\author[0000-0002-4284-8638]{Jennifer van Saders}
 \affiliation{Institute for Astronomy, University of Hawai`i, 2680
@@
 these cool stars at Praesepe age.
-Most semi-physical spin-down models predict that, once the rotation periods of
+Most semi-empirical spin-down models predict that, once the rotation periods of
 stars converge to a tight sequence, they approximately spin down with a common
 braking index \citep[e.g., Fig. 5 in][]{gallet2015}.
-While some semi-physical models do not predict that stars have fully converged
+While some semi-empirical models do not predict that stars have fully converged
 by the age of Praesepe, the observations suggest that they have

### 2020-04-24 paper/ms63.tex
adding citation to Ruprecht paper
small edits

@@
 
-\author{Elisabeth Newton}
+\author[0000-0003-4150-841X]{Elisabeth Newton}
 \affiliation{Department of Physics and Astronomy, Dartmouth
@@
 
-\author{Yuxi (Lucy) Lu}
+\author[0000-0003-4769-3273]{Yuxi (Lucy) Lu}
 \affiliation{Department of Astronomy, Columbia University, Manhattan, NY, USA}
@@
 
-\author{Lauren Anderson}
+\author[0000-0001-5725-9329]{Lauren Anderson}
 \affiliation{Observatories of the Carnegie Institution for
@@
 \citep{curtis2019}.
-% In other words, convection zone depth cannot be the only variable that affects
-% stellar spin-down rate.
+\racomment{
+    In addition, new rotation period measurements for low-mass stars in
+the 2.7-Gyr-old cluster Ruprecht 147 show that its slow-rotator sequence is
+flat compared to younger clusters like Praesepe (Curtis \etal\ 2020, submitted
+to AAS Journals), and departures from classical Skumanich-like spin-down were
+also noted in observations of the $\sim$1.3 Gyr NGC 752 open cluster
+\citep{agueros2018}.
+}
+% In other words, convection zone depth cannot be the only variable that
+% affects stellar spin-down rate.
 New semi-empirical models that vary the rate of angular momentum
-redistribution in the interiors of stars are able to reproduce this flattened
-period--color relation \citep{spada2019}.
+redistribution in the interiors of stars are able to reproduce the flattened
+period--color relation
+\\racomment{
+    seen in these clusters \citep{spada2019}.
+}
 These models suggest that mass and age-dependent angular momentum transport
@@
 While the velocity of any individual star only provides a weak age constraint
-\racomment{(if any at all)}, because its velocity depends on its current
+\racomment{
+    (if any at all)
+}, because its velocity depends on its current
 position in its orbit, the velocity {\it dispersion} of a {\it population} of
@@
 
-\racomment{ There is a long history of using kinematic ages to explore the
+\racomment{
+    There is a long history of using kinematic ages to explore the
 evolution of cool dwarfs \citep[\eg][]{reid1995, gizis2000, west2004,
@@
 
-Although {\it vertical} velocity, \vz, is \racomment{a well-established} age
-proxy, it can only be calculated with full 6-dimensional position and velocity
-information.
+Although {\it vertical} velocity, \vz, is \racomment{
+    a well-established
+}
+age proxy, it can only be calculated with full 6-dimensional position and
+velocity information.
 In fact, with full 6D phase space and an assumed Galactic potential, it is
@@
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-\racomment{The calibration of this relation is described in Curtis \etal\ (2020,
-in prep) and summarized in the Appendix of this paper.}
+\racomment{
+    The calibration of this relation is described in Curtis \etal\ (2020,
+in prep) and summarized in the Appendix of this paper.
+}
 
@@
 dwarfs are so long that they are not measurable with Kepler data.
-\racomment{Ground-based rotation period measurements of early M dwarfs
-(spectral types earlier than $\sim$ M2.5, \teff $\gtrsim$3500 K) indicate that
-there is an upper limit to their rotation periods of around 80 days
+\racomment{
+    Ground-based rotation period measurements of early M dwarfs
+(spectral types earlier than $\sim$ M2.5, \teff\ $\gtrsim$ 3500 K) indicate
+there is a $\sim$ 80 day upper limit to their rotation periods
 \citep{newton2016, newton2018}, which is longer than the longest rotation
-periods measured for the early M dwarfs in the \mct\ sample (around 50 days).}
+periods measured for the early M dwarfs in the \mct\ sample (around 50 days).
+}
+
 The apparent lack of old gyro-ages for M dwarfs in figure
@@
 Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
-(around 3,000 stars), stars fainter than 16th magnitude \racomment{in \gaia\
-G-band} (200 stars), stars with absolute \vb\ uncertainties greater than 1
+(around 3,000 stars), stars fainter than 16th magnitude \racomment{
+    in \gaia\ G-band
+}
+(200 stars), stars with absolute \vb\ uncertainties greater than 1
 \kms\ (1000 stars), and stars with galactic latitudes greater than 15\degrees\
@@
 bottom edge of the main population of rotation periods.
-\racomment{These rapid rotators are probably either very young stars, or
-synchronized binaries (see section \ref{sec:gap}).} After these cuts, around
+\racomment{
+    These rapid rotators are probably either very young stars, or
+synchronized binaries (see section \ref{sec:gap}).
+}
+After these cuts, around
 13,000 stars out of the original $\sim$34,000 were included in the sample.
@@
 gyrochrones from the \citet{spada2019} model shown in white.
-These gyrochrones are \racomment{plotted for the same ages described above.}
+These gyrochrones are \racomment{
+    plotted for the same ages described above.
+}
 Each point plotted in the top panel also appears in the bottom panel with the
@@
 lower-mass stars are still braking more efficiently at these intermediate and
-old ages \racomment{compared to higher-mass stars}, but their cores are more
+old ages \racomment{
+    compared to higher-mass stars,
+}
+but their cores are more
 tightly coupled to their envelopes, allowing angular momentum transport
 between the two layers.
-Angular momentum surfaces \racomment{from the core} and prevents the stellar
+Angular momentum surfaces \racomment{
+    from the core
+}
+and prevents the stellar
 envelopes from spinning-down rapidly, and this effect is strongest for late
@@
 braking index \citep[e.g., Fig. 5 in][]{gallet2015}.
-While some semi-empirical models do not predict that stars have fully converged
-by the age of Praesepe, the observations suggest that they have
-\citep{douglas2019}, in which case, projecting the Praesepe sequence forward
-with a single braking index that is common to all stars and constant in time
-is appropriate.
+So, if stellar rotation periods have fully converged by the age of Praesepe,
+as the observations suggest they have \citep{douglas2019}, it is appropriate
+to project the Praesepe sequence forward in time with a single braking index
+that is common to all stars and constant in time.
+% While some semi-empirical models do not predict that stars have
+% by the age of Praesepe, the observations suggest that they have
+% \citep{douglas2019}, in which case, projecting the Praesepe sequence forward
+% with a single braking index that is common to all stars and constant in time
+% is appropriate.
 However, observations of low-mass stars in older clusters \citep[\eg, NGC
-6811, NGC 752, and Ruprecht 147][Curtis \etal, submitted]{curtis2019,
-agueros2018} demonstrate that the assumptions adopted in those models must be
-invalid, or inaccurate in some quantitative way (for example, the
-core-envelope coupling timescale is inaccurately calibrated).
+6811, NGC 752, and Ruprecht 147][Curtis \etal\ 2020, submitted]{curtis2019,
+agueros2018} now demonstrate that the assumptions adopted in the
+semi-empirical models which support a common braking index must be invalid, or
+inaccurate in some quantitative way (for example, the core-envelope coupling
+timescale is inaccurately calibrated).
 
@@
 None of these relations follow lines of constant velocity dispersion in figure
-\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
-0.5), which applies to stars of all masses and ages.
+\ref{fig:vplot} because they have a universal power-law index, $n$
+\citep[$\approx$ 0.5,][]{skumanich1972}, which applies to stars of all masses
+and ages.
 They do not have the flexibility to capture the evolving period-color
@@
 It is not surprising that the \citet{angus2019} model, or other gyrochronology
-models do not reproduce the data, since tensions with
-\citet{skumanich1972}-type spin-down have been revealed whenever new rotation
-periods of benchmark stars have become available.
-With the influx of new stellar rotation periods provided by \kepler/\ktwo\ and
-\tess, our understanding of spin-down is changing rapidly.
+models do not reproduce the data, since tensions with Skumanich-like spin-down
+have been revealed whenever new rotation periods of benchmark stars have
+become available \citep{angus2015, vansaders2016, metcalfe2019, agueros2018,
+curtis2019}.
+With the influx of new stellar rotation periods provided by \kepler/\ktwo,
+\tess, and ground-based facilities, our understanding of spin-down is changing
+rapidly.
 New data are revealing flaws in not just the calibration of but also the
@@
 We have shown that velocity dispersion data provides support for more
-sophisticated models incorporating additional physical effects \citep[\eg,
-core-envelope coupling][]{spada2019} over simple empirical models.
+sophisticated models that incorporate additional physical effects \citep[\eg,
+core-envelope coupling][]{spada2019} over the simple empirical models used
+today.
 Alternatively, more flexible empirical frameworks, like a Gaussian process
@@
 periods have a similar age distribution to the overall Kepler population which
-suggests that \racomment{most of} the oldest K and M dwarfs are represented in
+suggests that \racomment{
+    most of
+}
+the oldest K and M dwarfs are represented in
 the \mct\ sample.
@@
 from an old regime where these layers are more tightly coupled.
-If so, this could indicate that \racomment{these phenomena are related, \ie\
+If so, this could indicate that
+\racomment{
+    these phenomena are related, \ie\
 the process that is responsible for changing the shape of gyrochrones in
@@
 
-We thank the anonymous referee for their comments which greatly improved the
+We thank the anonymous referee for their comments which greatly improved this
 manuscript.
-We would like to thank Suzanne Aigrain for providing thoughtful insight that
-improved this paper.
+We also would like to thank Suzanne Aigrain for providing thoughtful insight
+that improved the paper.
 This work was partly developed at the 2019 KITP conference `Better stars,
@@
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
+To calibrate this relation, Curtis \etal\ (2020, submitted) combined effective
 temperature measurements for nearby, unreddened field stars in benchmark
@@
 detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
-\etal\ (2020, in prep).}.
+\etal\ (2020, submitted).}.
 The equation we used to calculate photometric temperatures from Gaia \gcolor\

### 2020-04-24 paper/ms63.tex
v. small changes

@@
 \ref{fig:vplot}, provide the following picture of rotational evolution.
+\racomment{
     At young ages \citep[younger than around 1 Gyr but still old enough to be on
@@
 increasing} mass.
+}
 This is likely because lower-mass stars with deeper convection zones have

### 2020-04-29 paper/ms63.tex
adding citations

@@
 The mass dependence of magnetic activity lifetimes has been demonstrated
-previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
-kiman2019} and, if the detectability of a rotation period is considered to be
-a magnetic activity proxy, then our results provide further evidence for a
-mass-dependent activity lifetime.
+previously \racomment{for M dwarfs} \citep[\eg][]{west2008, reiners2009,
+west2009, newton2017, kiman2019} and, if the detectability of a rotation
+period is considered to be a magnetic activity proxy, then our results provide
+further evidence for a mass-dependent activity lifetime.
 

### 2020-04-30 paper/ms63.tex
adding an additional concluding statement

@@
 $\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
-This simple relation neatly reproduces the rotation periods of young cluster
-stars, however, like the other empirical gyrochronology models, it does not
-reproduce the observed trends seen in the velocity dispersion data here.
+This simple relation neatly reproduces the rotation periods of stars in the
+slow-rotation sequences of young ($\sim$100-700 Gyr) clusters, however, like
+the other empirical gyrochronology models, it does not reproduce the observed
+trends seen in the velocity dispersion data here.
 The \citet{barnes2010} model has a similar period-\teff\ relation to the
@@
 At old ages, cooler K dwarfs spin more rapidly than hotter K dwarfs of the
-same age.}
+same age.
+        \racomment{ While most empirical gyrochronology models calibrated to
+        date can broadly reproduce the rotation periods of young
+        ($\sim$100-700 Gyr) cluster stars, they do not match the rotational
+        behavior of intermediate-age and old ($\gtrsim$1 Gyr) K dwarfs.
+In addition, their simple functional forms, with a universal age-rotation
+        power-law index, are not flexible enough to capture rotational
+        evolution at all ages and masses.
+This is because they have been calibrated with young open clusters: stellar
+        spin down is relatively straightforward at young ages, and can often
+        be represented with a color and age-separable power-law relation.
+Until recently, more complex rotational evolution models were not needed.
+We now know however, from new observations of intermediate-age clusters, and
+        the kinematic data presented here, that these empirical models are not
+        flexible enough to reproduce the mass and time-dependent spin-down
+        rate of cool dwarfs at old ages.
+We advocate for the adoption of more flexible models, like semi-parametric
+        Gaussian process models, in the future.
+        }
+        }
 

### 2020-05-01 paper/ms63.tex
submitting

@@
 \newcommand{\racomment}[1]{{\color{blue}#1}}
+% \newcommand{\racomment}{}
 
@@
 stars.
-\racomment{
-In this work we use velocity dispersion as an age proxy to reveal the
+\racomment{In this work we use velocity dispersion as an age proxy to reveal the
 temperature-dependent rotational evolution of low-mass \kepler\ dwarfs, and
 demonstrate that kinematic ages could be a useful tool for calibrating
-gyrochronology in the future.
-}
+gyrochronology in the future.}
 % In this work, we use the velocity dispersions of groups of low-mass \kepler\
@@
 \citep{curtis2019}.
-\racomment{
-    In addition, new rotation period measurements for low-mass stars in
+\racomment{In addition, new rotation period measurements for low-mass stars in
 the 2.7-Gyr-old cluster Ruprecht 147 show that its slow-rotator sequence is
@@
 also noted in observations of the $\sim$1.3 Gyr NGC 752 open cluster
-\citep{agueros2018}.
-}
+\citep{agueros2018}.}
 % In other words, convection zone depth cannot be the only variable that
@@
 redistribution in the interiors of stars are able to reproduce the flattened
-period--color relation
-\\racomment{
-    seen in these clusters \citep{spada2019}.
-}
+period--color relation \racomment{seen in these clusters \citep{spada2019}.}
 These models suggest that mass and age-dependent angular momentum transport
@@
 While the velocity of any individual star only provides a weak age constraint
-\racomment{
-    (if any at all)
-}, because its velocity depends on its current
+\racomment{(if any at all)}, because its velocity depends on its current
 position in its orbit, the velocity {\it dispersion} of a {\it population} of
@@
 
-\racomment{
-    There is a long history of using kinematic ages to explore the
+\racomment{There is a long history of using kinematic ages to explore the
 evolution of cool dwarfs \citep[\eg][]{reid1995, gizis2000, west2004,
@@
 
-Although {\it vertical} velocity, \vz, is \racomment{
-    a well-established
-}
-age proxy, it can only be calculated with full 6-dimensional position and
-velocity information.
+Although {\it vertical} velocity, \vz, is \racomment{a well-established} age
+proxy, it can only be calculated with full 6-dimensional position and velocity
+information.
 In fact, with full 6D phase space and an assumed Galactic potential, it is
@@
 $G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
-\racomment{
-    The calibration of this relation is described in Curtis \etal\ (2020,
-in prep) and summarized in the Appendix of this paper.
-}
+\racomment{The calibration of this relation is described in Curtis \etal\
+(2020, in prep) and summarized in the Appendix of this paper.}
 
@@
 dwarfs are so long that they are not measurable with Kepler data.
-\racomment{
-    Ground-based rotation period measurements of early M dwarfs
+\racomment{Ground-based rotation period measurements of early M dwarfs
 (spectral types earlier than $\sim$ M2.5, \teff\ $\gtrsim$ 3500 K) indicate
@@
 Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
-(around 3,000 stars), stars fainter than 16th magnitude \racomment{
-    in \gaia\ G-band
-}
-(200 stars), stars with absolute \vb\ uncertainties greater than 1
+(around 3,000 stars), stars fainter than 16th magnitude \racomment{in \gaia\
+G-band} (200 stars), stars with absolute \vb\ uncertainties greater than 1
 \kms\ (1000 stars), and stars with galactic latitudes greater than 15\degrees\
@@
 bottom edge of the main population of rotation periods.
-\racomment{
-    These rapid rotators are probably either very young stars, or
-synchronized binaries (see section \ref{sec:gap}).
-}
+\racomment{These rapid rotators are probably either very young stars, or
+synchronized binaries (see section \ref{sec:gap}).}
 After these cuts, around
@@
 gyrochrones from the \citet{spada2019} model shown in white.
-These gyrochrones are \racomment{
-    plotted for the same ages described above.
-}
+These gyrochrones are \racomment{plotted for the same ages described above.}
 Each point plotted in the top panel also appears in the bottom panel with the
@@
 \ref{fig:vplot}, provide the following picture of rotational evolution.
-\racomment{
-    At young ages \citep[younger than around 1 Gyr but still old enough to be on
-the main sequence and have transitioned from the `C' sequence to the `I'
-sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
-increasing} mass.
-}
-This is likely because lower-mass stars with deeper convection zones have
-stronger magnetic fields, larger Alfv\'en radii and therefore experience
-greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
-parker1970, kawaler1988, charbonneau2010, matt2012, matt2015}.
+\racomment{At young ages \citep[younger than around 1 Gyr but still old enough
+to be on the main sequence and have transitioned from the `C' sequence to the
+`I' sequence ][]{barnes2003}, stellar rotation period {\it decreases} with
+{\it increasing} mass.}
+This is likely because lower-mass stars with deeper
+convection zones have stronger magnetic fields, larger Alfv\'en radii and
+therefore experience greater angular momentum loss rate
+\citep[\eg][]{schatzman1962, kraft1967, parker1970, kawaler1988,
+charbonneau2010, matt2012, matt2015}.
 % This aligns with the current assumptions about stellar spin-down, dynamo
@@
 lower-mass stars are still braking more efficiently at these intermediate and
-old ages \racomment{
-    compared to higher-mass stars,
-}
-but their cores are more
+old ages \racomment{compared to higher-mass stars,} but their cores are more
 tightly coupled to their envelopes, allowing angular momentum transport
 between the two layers.
-Angular momentum surfaces \racomment{
-    from the core
-}
-and prevents the stellar
+Angular momentum surfaces \racomment{from the core} and prevents the stellar
 envelopes from spinning-down rapidly, and this effect is strongest for late
@@
 
-\racomment{
-In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
+\racomment{In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
 gyrochronology model is plotted for comparison with the data.
@@
 periods have a similar age distribution to the overall Kepler population which
-suggests that \racomment{
-    most of
-}
+suggests that \racomment{most of}
 the oldest K and M dwarfs are represented in
@@
 If so, this could indicate that
-\racomment{
-    these phenomena are related, \ie\
+\racomment{these phenomena are related, \ie\
 the process that is responsible for changing the shape of gyrochrones in
-rotation-\teff\ space is related to the process that produces the gap.
+rotation-\teff\ space is related to the process that produces the gap.}
 % Of course this may just be a coincidence.
@@
 % evolution regime may be a coincidence.
-}
 

### 2020-05-01 paper/ms63.tex
typo in citation

@@
 The mass dependence of magnetic activity lifetimes has been demonstrated
-previously \racomment{for M dwarfs} \citep[\eg][]{west2008, reiners2009,
+previously \racomment{for M dwarfs} \citep[\eg][]{west2008, reiners2008,
 west2009, newton2017, kiman2019} and, if the detectability of a rotation

### 2020-05-06 paper/ms63.tex
adding a table and fixing appendix typo

@@
 
-\newcommand{\racomment}[1]{{\color{blue}#1}}
-% \newcommand{\racomment}{}
+% \newcommand{\racomment}[1]{{\color{blue}#1}}
+\newcommand{\racomment}{}
 
@@
 % miscellaneous --- catalogs --- surveys}
+\keywords{Stellar magnetic fields --- Milky Way dynamics ---
+Milky Way disk --- Stellar activity --- Solar analogs ---
+Stellar phenomena --- Late-type stars --- Low mass stars ---
+Stellar evolution}
 
@@
 sample.
+The data used in this project is described in table \ref{tab:data}, at the end
+of this paper, and is available online.
 
@@
 
-The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
+The {\tt Pyia} \citep{price-whelan2018} and {\tt astropy} \citep{astropy2013,
 astropy2018} {\it Python} packages were used to calculate velocities for the
@@
     evolution for K-dwarfs.
+    The data used to create this figure is available in table \ref{tab:data}.
 }
@@
 This empirical color--temperature relation is valid over the color range $0.55
-< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $3070 > T_{\rm eff} >
+< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $3070 < T_{\rm eff} <
 6470$~K.
@@
 
+\ref{tab:data}.
+\begin{table}[h!]
+  \begin{center}
+      \caption{
+The data used for this project (the full table is available online).
+      Masses were obtained from \citet{berger2020}, rotation periods from
+      \mct, effective temperatures were calculated from dereddened \gaia\
+      photometry, as described in the appendix and Curtis \etal\ (Submitted).
+Velocities were calculated using {\tt pyia} and {\tt astropy}
+      \citep{astropy2013, astropy2018, price-whelan2018}.
+Uncertainties on velocity dispersions were calculated as the standard
+      error of the sample standard deviation \citep{rao1973}.
+      }
+    \label{tab:data}
+\begin{tabular}{cccccc}
+KIC ID & Mass [M$_\odot$] & T$_\mathrm{eff}$ [K] & P$_\mathrm{rot}$ [days] & v$_b$ [kms$^{-1}$] & $\sigma_\mathrm{vb}$ [kms$^{-1}$] \\
+\hline
+1164102 & 0.63 & 4053 & 31.5 $\pm$ 0.5 & 13.6 $\pm$ 0.2 & 21.2 $\pm$ 0.4 \\
+1292688 & 0.52 & 3752 & 42.7 $\pm$ 2.1 & -14.2 $\pm$ 0.2 & 21.9 $\pm$ 0.7 \\
+1297303 & 0.67 & 4318 & 27.3 $\pm$ 0.2 & 16.7 $\pm$ 0.3 & 16.7 $\pm$ 0.2 \\
+1429921 & 0.65 & 4258 & 23.1 $\pm$ 0.1 & -7.0 $\pm$ 0.2 & 14.8 $\pm$ 0.2 \\
+1430349 & 0.65 & 4368 & 34.7 $\pm$ 0.8 & 12.2 $\pm$ 0.2 & 19.7 $\pm$ 0.4 \\
+1430893 & 0.61 & 3985 & 17.0 $\pm$ 0.0 & 0.4 $\pm$ 0.1 & 9.2 $\pm$ 0.1 \\
+1431116 & 0.72 & 4415 & 38.8 $\pm$ 1.0 & -16.5 $\pm$ 0.1 & 19.7 $\pm$ 0.4 \\
+1432745 & 0.7 & 4413 & 22.2 $\pm$ 0.2 & -2.1 $\pm$ 0.0 & 13.4 $\pm$ 0.2 \\
+1435229 & 0.57 & 4080 & 23.5 $\pm$ 0.0 & -8.8 $\pm$ 1.0 & 13.2 $\pm$ 0.2 \\
+1569682 & 0.57 & 3909 & 17.4 $\pm$ 0.2 & 4.4 $\pm$ 0.1 & 9.2 $\pm$ 0.1 \\
+\end{tabular}
+\end{center}
+\end{table}
+
 \bibliography{bib}{}

### 2020-05-06 kinematics_and_rotation/dispersion.py

@@ -274,6 +274,7 @@ def err_on_sample_std_dev(std_dev_of_distribution, n):
     Which takes the derivation from
     Rao (1973) Linear Statistical Inference and its Applications 2nd Ed, John
     Wiley & Sons, NY
+    https://onlinelibrary.wiley.com/doi/book/10.1002/9780470316436
 
     Derivation for standard error on the variance is here:
     https://math.stackexchange.com/questions/72975/variance-of-sample-variance

### 2020-05-06 setup.py

@@ -1,12 +1,8 @@
 from setuptools import setup
 
-setup(name='kinematics_and_rotation',
+setup(name='cds',
       version='0.1rc0',
-      description='Tools for kinematics',
-      url='http://github.com/RuthAngus/kinematics-and-rotation',
-      author='Ruth Angus',
-      author_email='ruthangus@gmail.com',
-      license='MIT',
-      packages=['kinematics_and_rotation'],
-      install_requires=['numpy', 'pandas', 'tqdm', 'astropy', 'matplotlib', 'scipy'],
+      description='Make machine readable table for Visier',
+      packages=['cds'],
+      install_requires=['numpy', "astropy"],
       zip_safe=False)
